### 2025-04-02 documentation/hierarchicalmodels_seqruns.qmd
add functions

---
title: "Hierarchical models and seq model fits"
author: "Leontine"
format: html
editor: visual
---

# Overview

In this document we explain

1.  how we set up a hierarchical model for a parameter;

2.  how we fix higher level parameters for that set up when doing sequential runs.

The current goal is to document the set up and include a simplified example of use, to help with code checks and to make it easier to do code updates. Longer term, we can see if we can pull out this related set of functions into a separate R package, to make it easier to use in other modeling exercises.

The code set up was introduced in <https://github.com/AlkemaLab/BayesTransitionModels>

# Set up

Goal: create a simplified example that uses our data set and helper functions from the package (functions in R for processing, functions and code blocks in stan to set up the stan model).

Load relevant packages, including our own

```{r}
#| include: false
library(here) 
library(tidyverse) 
library(ggplot2) 
library(haven) 
library(stringr)
library(cmdstanr) 
library(bayesplot) # diagnostic plots 
devtools::load_all(here::here())
```

Read in data and preprocess as usual

```{r}
#| include: false
data_folder <- "data_raw" 
dat <- read_csv(here::here(data_folder, "coverage_data.csv"))
```

Let's use columns `cluster` and `subcluster` and `iso`:

```{r}
dat %>% select(iso, cluster, subcluster)
```

Summary of clusters and subclusters, after selecting one row per country

```{r}
dat %>%
  group_by(iso) %>%
  filter(year == max(year)) %>%
  select(country, iso, subcluster, cluster) %>%
  group_by(cluster) %>%
  summarize(nsubclusters = length(unique(subcluster)), ncountries = n())
```

# Model fitting

## Model set up

Let's consider the following model (we won't bother with logit-transforms):

-   parameter to estimate = $\mu_c$ = country-specific coverage (during entire period)

-   data model: $y_i \sim N(\mu_{c[i]}, \sigma_y^2)$

-   hierarchical model for $\mu_c$:

    -   $\mu_c \sim N(\mu_{s[c]}^{(subregion)}, \sigma_{\mu, across-country}^2)$ ,

    -   $\mu_s^{(subregion)} \sim N(\mu_{r[s]}^{(region)}, \sigma_{\mu, across-subregions}^2)$,

    -   $\mu_r^{(region)} \sim N(\mu^{(global)}, \sigma_{\mu, across-regions}^2)$

Note that an equivalent specification of the hierarchical model (the one we'll use in the stan model) is as follows:

-   $\mu_c =  \mu^{(global)} + \eta_{r[s[c]]]}^{(region)} + \eta_{s[c]}^{(subregion)} + \eta_c$ with

-   $\eta_c \sim N(0, \sigma_{\mu, across-country}^2)$ ,

-   $\eta_s^{(subregion)} \sim N(0, \sigma_{\mu, across-subregions}^2)$,

-   $\eta_r^{(region)} \sim N(0, \sigma_{\mu, across-regions}^2)$

In addition, in the stan model, we will use that we can write $\eta = \sigma\cdot \gamma$ where $\gamma \sim N(0,1)$. In addition, with the same set up, if $\mu^{(global)}$ has prior

$$\mu^{(global)} \sim N(0, s_{\mu, global}^2),$$then we can write $\mu^{(global)} = s_{\mu, global}\cdot \gamma^{(global)}$, with $s$ being the fixed prior standard deviation (as opposed to a SD that's estimated). All that combined, we get

-   $\mu_c =  s_{global}\cdot \gamma_{global} + \sigma_{\mu, region}\cdot \gamma_{r[s[c]]]}^{(region)} + \sigma_{\mu, subregion}\cdot\gamma_{s[c]}^{(subregion)} + \sigma_{\mu, country}\cdot\gamma_{\mu, c}$ with

-   $\gamma \sim N(0,1)$ for each gamma (one global, $R$ region, $S$ subregions, $C$ countries.

The set up and notation in stan/R, in terms of parameters, using parameter Omega as an example, is currently as follows:

-   Omega_raw refers to the vector with ALL $\gamma$s, including the global one.The ordering is global-allregional-allsubregional-country.

-   Omega_sigma referred to the vector of stacked sigmas, excluding $s_{\mu, global}$ (in the original package, it was called Omega_sigma_raw).

-   Omega_star refers to the vector with (mu global and ALL $\eta$s). we get Omega_stars from Omega_raw and the vector ($s_{\mu, global}$ , Omega_sigma), using the stan `scale_blocks` function.

-   Omega refers to the final $\mu$s at the country level. Matrix multiplication is used to obtain the Omega country vector from the Omega_stars. The full matrix `Omega_model_matrix` is a sparse matrix, with 1s and 0s (columns refer to $\mu_{global}$, the regional etas, the subregional etas, the country etas). The matrix multiplication is implemented using the stan function csr_matrix_times_vector, using a decomposition of `Omega_model_matrix` into different parts U, V, W.

-   Because of the sequential fitting, for Omega_raw and Omega_sigma, we also had to distinguish between parameters that were fixed vs those that were estimated. We do so with the \_fixed and \_estimate additions, i.e. the vector of Omega_raws is produced from Omega_raw_fixed and Omega_raw_estimate.

## Model fitting

global model fit

```{r}
fit1a <- fit_model_simplified(runstep = "step1a",
        hierarchical_level     = c("intercept",  "subcluster", "iso"),

                  survey_df = dat,
                  chains = 4)

```

create summary object

```{r}
fit1a$post_summ <- get_posterior_summaries_simplified(fit1a) %>%
  dplyr::rename(median = mean)
fit1a$post_summ

```

do local fit... here we fix things but do get results for all countries

```{r}
#fit1a$samples <- NULL
fit_local <- fit_model_simplified(runstep = "local_national",
                  global_fit = fit1a,
                  survey_df = dat,
                  chains = 4)
```

```{r}
#fit1a$samples$summary()$variable
```

#### Produce and check outputs

in the simpler model used, various Omegas are defined in the transformed parameters block so we can check those directly. (Note that this not (yet) true for the transition model because the creation of parameters takes place within the transition function, see notes in last section on this).

here just checking the sigmas and first Omega:

```{r}

fit <- fit1a
 parname_list <- list( c("Omega_raw_estimate[1]",
                          "Omega_sigma_estimate"))
#    plot_dirandname <- file.path(fit$output_dir, paste0(runname, "diagnostics.pdf"))
#    pdf(plot_dirandname, width = 11, height = 11)
    for (i in 1:length(parname_list)){
      parname <- parname_list[[i]]
      samp <- fit$samples$draws(parname )
      #mcmc_trace(samp)
      p <- mcmc_dens_overlay(samp)
      # for more than 1 parameter
      #mcmc_pairs(samp)
      print(p)
    }
#    dev.off()
```

Re pulling out information at non-country levels (ie estimates for regions and subregions): The orginal code set up in <https://github.com/AlkemaLab/BayesTransitionModels> includes helper functions to pull out various outputs, for example, $\mu$s at the regional or subnational level. We moved some over and use them in the package here. However, they are not yet very user friendly given limited documentation, hardcoding of prior SDs for global means, and reliance on overall fit object. It is work in progress to clean this up further. What I did here for the simplified example (and will move over to the general package later):

-   To deal with the set up where things are fixed vs estimated, we let the stan model spit out the Omega_stars, that's what the original helper functions use as inputs.

-   We clean up helper functions, so that they take in the Omega_stars, and specific other stan_data, and are better documented. See function extract_parameter_subhierarchical_simplified in fit_model_simplified R script, and its usage below.

```{r}
fit <- fit1a
# some notes, to understand updates between this code base and original one
# and where to find info 
#fit1a$country_index is now fit$geo_unit
# see fit1a$hierarchical_level for info
# hierarchical_data(fit$geo_unit, fit$hierarchical_level)
# note that subhierarchies are called column in hierarchical data
Omega <- list()
for(subhierarchy in fit$hierarchical_level) {
  Omega[[subhierarchy]] <-
    extract_parameter_subhierarchical_simplified (
      hierarchical_data = hierarchical_data(fit$geo_unit, fit$hierarchical_level),
      subhierarchy = subhierarchy,
      parname = "Omega",
      fit_samples = fit$samples)
}
# Omega[[]] has posterior samples for the subhierarchy level
# to get the summaries
quantile_df <- function(x, probs = c(0.025, 0.5, 0.975)) {
  tibble(
    val = quantile(x, probs, na.rm = TRUE),
    quant = probs
  )
}
res_global <- map(Omega, function(tibble_samples) 
  tibble_samples %>% select(name, value)  %>% reframe(quantile_df(value), .by = name))
res_global

```

```{r}
fit <- fit_local
Omega <- list()
for(subhierarchy in fit$hierarchical_level) {
  Omega[[subhierarchy]] <-
    extract_parameter_subhierarchical_simplified (
      hierarchical_data = hierarchical_data(fit$geo_unit, fit$hierarchical_level),
      subhierarchy = subhierarchy,
      parname = "Omega",
      fit_samples = fit$samples)
}
res_local <- map(Omega, function(tibble_samples) 
  tibble_samples %>% select(name, value)  %>% reframe(quantile_df(value), .by = name))
res_local
```

checks that results from local model approximate those from global model

at subcluster level

```{r}
bind_rows(res_local$subcluster %>% mutate(model = "local"),
          res_global$subcluster %>% mutate(model = "global")) %>%
  group_by(name, model) %>% 
  reframe(y = val[quant == 0.5], ymin = val[quant == 0.025], ymax = val[quant == 0.975]) %>%
  ggplot(aes(y = y, x = name, color = model)) +
  geom_errorbar(aes(ymin = ymin, ymax = ymax)) +
  geom_point()

```

at country level

```{r}
bind_rows(res_local$iso %>% mutate(model = "local"),
          res_global$iso %>% mutate(model = "global")) %>%
  group_by(name, model) %>% 
  reframe(y = val[quant == 0.5], ymin = val[quant == 0.025], ymax = val[quant == 0.975]) %>%
  ggplot(aes(y = y, x = name, color = model)) +
  geom_errorbar(aes(ymin = ymin, ymax = ymax)) +
  geom_point()   
```

# Related notes and to dos to come back to

-   To do next: clean code to fix parameters

-   How to get the Omegas in full model set up: in the simpler model used here, various Omegas are defined in the transformed parameters block so we can check those directly. Note that this not (yet) true for the full transition model because the creation of parameters takes place within the process model function. Options to obtain the Omegas in the full model are to

    1.  add them to generated quantities: pull out omegas with an `if {add_output_omegas}` type statement, re-using functions/statements introduced in the simpler file here.

    2.  rewrite transformed parameter block and pull Omegas out of the process model function (not sure if that slows things down when not used).

    3.  create them in R (would need more work to get functions going, could consider exposing stan functions)

    Options 1-2 seem easiest to LA.

-   This example does not yet get into estimation of parameter vectors and associated code for that (as we have for spline coefficients). Simplest way to set that up is to estimate the same parameter $\mu$ twice.

-   More general, on how to organize stan functions: Stan functions lived in a separate `include` folder in the original code base. In the fpet2 package we ended up including stan functions in the stan model directly. That inclusion was needed to ship a precompiled model with the package for use in the webtool (using the `instantiate` package). We can move functions back to an `include` subfolder in this documentation/package.

-   Note on approach used to get this example together (can be followed when extending further): Start subfolder for simplified model and code, added current stan_model and fit_model function and simplify to produce just the toy example.

### 2025-04-02 R/check_nas.R

@@ -1,0 +1,8 @@
+check_nas <- function(data, column) {
+  if(!(column %in% names(data))) {
+    stop(glue::glue("Could not find column {column}."))
+  }
+  if(sum(is.na(data[[column]]))) {
+    stop(glue::glue("There are NAs in column {column}."))
+  }
+}

### 2025-04-02 R/extract_parameter_subhierarchical.R

@@ -1,0 +1,51 @@
+
+extract_parameter_subhierarchical_simplified <- function(
+    hierarchical_data, # hierarchical_data, obtained from hierarchical_data(fit$geo_unit, fit$hierarchical_level)
+    subhierarchy, # selected hierarchical_level (example: "intercept" or "region")
+    parname, # selected parameter name (example: "Omega")
+    fit_samples #fit$samples
+) {
+
+
+  start <- hierarchical_data$model_matrix$index %>%
+    dplyr::filter(column == subhierarchy) %>%
+    dplyr::pull(i) %>%
+    min()
+
+  end <- hierarchical_data$model_matrix$index %>%
+    dplyr::filter(column == subhierarchy) %>%
+    dplyr::pull(i) %>%
+    max()
+
+  pars <- c(glue::glue("{parname}_star"))
+
+  star <- fit_samples$draws(pars) %>%
+    tidybayes::spread_draws((!!sym(pars[1]))[i]) %>%
+    dplyr::group_by(.chain, .iteration, .draw) %>%
+    tidyr::nest() %>%
+    dplyr::mutate(star = map(data, `[[`, glue::glue("{parname}_star"))) %>%
+    dplyr::select(-data)
+
+  uniq <- unique(hierarchical_data$model_matrix$mat[, 1:end, drop = FALSE])
+
+  titles <- c()
+  for(i in 1:nrow(uniq)) {
+    index <- rep(0, hierarchical_data$n_terms)
+    index[1:end] <- uniq[i, 1:end]
+    title <- hierarchical_data$model_matrix$index %>%
+      dplyr::filter(i == last(which(index == 1))) %>%
+      dplyr::pull(level)
+    titles <- c(titles, title)
+
+    star[[title]] = map_dbl(star[["star"]], function(star) {
+      index %*% star
+    })
+  }
+
+  star <- star %>%
+    tidyr::pivot_longer(cols = all_of(titles)) %>%
+    dplyr::ungroup() %>%
+    dplyr::select(-star)
+
+  return(star)
+}

### 2025-04-02 R/fit_model_simplified.R

@@ -233,20 +233,21 @@ fit_model_simplified <- function(
   )) %>%
     setdiff("intercept")
 
-  # Make sure there are no NAs in any of the columns
-  for(column in hierarchical_column_names) {
-    if(column == "intercept") next
-    if (column == area) {
-      check_nas_or_pops(data, column, year, population_data)
-    } else {
-      check_nas(data, column)
-    }
-  }
+  print("to do: check what NA check is needed")
+  # # Make sure there are no NAs in any of the columns
+  # for(column in hierarchical_column_names) {
+  #   if(column == "intercept") next
+  #   if (column == area) {
+  #     check_nas_or_pops(data, column, year, population_data)
+  #   } else {
+  #     check_nas(data, column)
+  #   }
+  # }
 
 
   ##### Load model #####
   if (compile_model){
-    stan_file_path <- file.path(here::here("documentation/code_for_hierarchicalmodels_seqrun", "hierfunctions_seq.stan"))
+    stan_file_path <- file.path(here::here("inst/stan/", "hierfunctions_seq.stan"))
     stan_model <- cmdstanr::cmdstan_model(
       stan_file = stan_file_path,
       force_recompile = force_recompile
@@ -384,120 +385,7 @@ fit_model_simplified <- function(
 }
 
 
-get_posterior_summaries_simplified <- function(
-    fit,
-    process_indicator_prefixes = c(""),
-    dm_indicator_prefixes = c(""),
-    process_params = c("Omega_raw", "Omega_sigma"),
-    dm_params = c("nonse")
-    ) {
-  params_to_collect <- dplyr::bind_rows(
-    expand.grid(
-      prefix = process_indicator_prefixes,
-      param = process_params),
-    expand.grid(
-      prefix = dm_indicator_prefixes,
-      param = dm_params)
-  ) |>
-    dplyr::mutate(prefixed_param_name = paste0(prefix, param))
 
-  # split parameters into those that are "fixable" in local fits and those that
-  # are "unfixable"
-  #  unfixable_param_names <- c("local_shrinkage_dm")
-  # unfixable_param_names <- c("BLA")
-  fixable_params_to_collect <- params_to_collect |>
-    #   dplyr::filter(!param %in% unfixable_param_names) |>
-    dplyr::pull(prefixed_param_name)
-
-  # unfixable_params_to_collect <- params_to_collect |>
[…]

### 2025-04-02 R/get_posterior_summaries.R

@@ -1,0 +1,61 @@
+get_posterior_summaries_simplified <- function(
+    fit,
+    process_indicator_prefixes = c(""),
+    dm_indicator_prefixes = c(""),
+    process_params = c("Omega_raw", "Omega_sigma"),
+    dm_params = c("nonse")
+) {
+  params_to_collect <- dplyr::bind_rows(
+    expand.grid(
+      prefix = process_indicator_prefixes,
+      param = process_params),
+    expand.grid(
+      prefix = dm_indicator_prefixes,
+      param = dm_params)
+  ) |>
+    dplyr::mutate(prefixed_param_name = paste0(prefix, param))
+
+  # split parameters into those that are "fixable" in local fits and those that
+  # are "unfixable"
+  #  unfixable_param_names <- c("local_shrinkage_dm")
+  # unfixable_param_names <- c("BLA")
+  fixable_params_to_collect <- params_to_collect |>
+    #   dplyr::filter(!param %in% unfixable_param_names) |>
+    dplyr::pull(prefixed_param_name)
+
+  # unfixable_params_to_collect <- params_to_collect |>
+  #   dplyr::filter(param %in% unfixable_param_names) |>
+  #   dplyr::pull(prefixed_param_name)
+
+  # get parameter summaries for each group
+  # get parameter summaries for each group
+  print(
+    "Warning: in obtaining posterior summary in get_posterior_summary_one_fixable_param, we have hardcoded num_inds based on names of parameters."
+  )
+  fixable_combined_summary <- purrr::map(
+    fixable_params_to_collect,
+    get_posterior_summary_one_fixable_param,
+    fit = fit
+  ) |>
+    purrr::list_rbind()%>%
+    # LA added column wtith parname w/o []
+    dplyr::mutate(
+      variable_no_index = stringr::str_split_i(string = variable, pattern = fixed("["), i = 1)
+    )
+
+  # unfixable_combined_summary <- fit$samples$summary(unfixable_params_to_collect,
+  #                                                   "mean"#,
+  #                                                   #"median"
+  #                                                   )
+
+  return(fixable_combined_summary)
+  # return(dplyr::bind_rows(
+  #   fixable_combined_summary,
+  #   unfixable_combined_summary
+  #     ) %>%
+  #     # LA added column wtith parname w/o []
+  #   dplyr::mutate(
+  #     variable_no_index = stringr::str_split_i(string = variable, pattern = fixed("["), i = 1)
+  #   )
+  # )
+}

### 2025-04-02 R/get_posterior_summary_one_fixable_param.R

@@ -1,0 +1,107 @@
+
+get_posterior_summary_one_fixable_param <- function(
+    fit,
+    prefixed_param_name) {
+  # extract parameter summaries for estimated parameters
+  estimate_param_name <- paste0(prefixed_param_name, "_estimate")
+  #print(estimate_param_name)
+  if (variable_exists_in_draws(fit, estimate_param_name)) {
+    param_estimate_summary <- fit$samples$summary(
+      estimate_param_name,
+      "mean"
+      #,
+      #"median"
+    )
+  } else {
+    # empty result
+    param_estimate_summary <- tibble(variable = character(0),
+                                     #                                   median = numeric(0))
+                                     mean = numeric(0))
+  }
+
+  #extract fixed values used for estimation
+  param_fixed <- fit$stan_data[[paste0(prefixed_param_name, "_fixed")]]
+
+  # if there are no fixed param values, nothing to do other than update the
+  # variable name, dropping "_estimate"
+  if (length(param_fixed) == 0) {
+    return(
+      param_estimate_summary |>
+        dplyr::mutate(variable = gsub(pattern = "_estimate", replacement = "",
+                                      x = variable))
+    )
+  }
+
+  # what is the dimension of the array we're expecting?
+  # param_inf <- fit$stan_model$variables()[["parameters"]][[estimate_param_name]]
+  # num_inds <- param_inf[["dimensions"]]
+  # update la: when reading in a run, this `fit$stan_model$variables()` does not exist
+  # it's used here just to get the dimension, 1 or 2
+  # easiest to check in main function, so passing here as an argument
+  # hardcoding here
+  #warning
+  print("Warning: in obtaining posterior summary in get_posterior_summary_one_fixable_param, we have hardcoded num_inds based on names of parameters.")
+  num_inds = ifelse(prefixed_param_name %in% c("a_raw", "a_sigma", "d_a_raw", "d_a_sigma",
+                                               "z_a_raw", "z_a_sigma"), 2, 1)
+
+
+
+  # process variable names in param_estimate_summary, splitting
+  # into columns var_name, i, and possibly j, where i and j
+  # are indices
+  param_estimate_summary <- split_var_name_and_index(param_estimate_summary,
+                                                     num_inds)
+
+  # shape of param_fixed
+  # if one-dimensional, length (no columns)
+  # if two-dimensional, number of rows and columns
[…]

### 2025-04-02 R/hier_data_helpers.R

@@ -1,0 +1,138 @@
+
+
+is_there_a_comma <- function(variable){
+  grepl(
+    pattern = ",",
+    x = variable,
+    fixed = TRUE)
+}
+# is_there_a_comma("a[1,1]")
+
+# get_dimension <- function(variable){
+#   comma_matches <- gregexpr(
+#     pattern = ",",
+#     text = variable,
+#     fixed = TRUE)[[1]]
+#   if (length(comma_matches) == 1 && comma_matches[[1]] == -1) {
+#     num_inds <- 1
+#   } else {
+#     num_inds <- length(comma_matches) + 1
+#   }
+#   return(num_inds)
+# }
+#
+# #get_dimension(c("a[1,1]", "a[2,1]"))
+# #get_dimension(c("a[1]", "a[2]"))
+# #get_dimension(c("a[1,1]", "a[1,2]"))
+
+
+# more complicated than need to be...
+create_a_matrix <- function(post_summ){
+  res <- post_summ %>%
+    mutate(
+    row = gsub(pattern = ".*\\[(.*),.*", replacement = "\\1", x = variable),
+    col = gsub(pattern = ".*\\[.*,(.*)\\]", replacement = "\\1", x = variable)
+    ) %>%
+    select(-variable) %>%
+    # make integers
+    tidyr::pivot_wider(names_from = col, values_from = median) %>%
+    tidyr::pivot_longer(cols = -row, names_to = "col", values_to = "median") %>%
+    mutate(row = as.integer(row),
+           col = as.integer(col))
+  m <- matrix(NA, max(res$row), max(res$col))
+  for (rowi in unique(res$row)){
+    for (coli in unique(res$col)){
+      m[rowi, coli] <- res %>%
+        filter(row == rowi, col == coli) %>%
+        pull(median)
+    }
+  }
+  m
+}
+#create_a_matrix( tibble(variable = c("a[1,1]", "a[2,1]", "a[1,2]", "a[2,2]"), median = seq(1,4)) )
+#create_a_matrix( tibble(variable = c("a[1,1]", "a[2,1]"), median = seq(1,2)) )
+#robust to ordering of the variables
+#create_a_matrix( tibble(variable = c("a[2,1]", "a[1,1]"), median = seq(1,2)) )
+#create_a_matrix( tibble(variable = c("a[1,1]",  "a[1,2]", "a[2,1]","a[2,2]"), median = seq(1,4)) )
+
+#' Determine whether or not a variable exists in draws from a fit
[…]

### 2025-04-02 R/hierarchical_data.R

@@ -1,0 +1,17 @@
+#' @import purrr
+hierarchical_data <- function(data, hierarchy) {
+  model_matrix <- hierarchical_model_matrix(hierarchy, data)
+  n_terms <- ncol(model_matrix$mat)
+  re <- unique(model_matrix$assign)
+  n_re <- length(re)
+  re_start <- purrr::map_int(re, function(x) min(which(model_matrix$assign == x)))
+  re_end   <- purrr::map_int(re, function(x) max(which(model_matrix$assign == x)))
+
+  list(
+    model_matrix = model_matrix,
+    n_terms = n_terms,
+    n_re = n_re,
+    re_start = re_start,
+    re_end = re_end
+  )
+}

### 2025-04-02 R/hierarchical_model_matrix.R

@@ -1,0 +1,35 @@
+
+#' @import tibble
+hierarchical_model_matrix <- function(columns, data) {
+  mat <- matrix(NA, nrow = nrow(data), ncol = 0)
+  assign <- c()
+
+  index <- tibble::tibble(i = numeric(0), column = character(0), level = character(0))
+
+  i <- 1
+  if("intercept" %in% columns) {
+    mat <- cbind(mat, rep(1, nrow(mat)))
+    assign <- c(assign, 0)
+
+    index[i, ] <- tibble(i = i, column = "intercept", level = "intercept")
+
+    i <- i + 1
+  }
+
+  for(column in columns) {
+    for(l in levels(factor(data[[column]]))) {
+      mat <- cbind(mat, as.numeric(data[[column]] == l))
+      assign <- c(assign, which(column == columns))
+      index[i, ] <- tibble::tibble(i = i, column = column, level = l)
+
+      i <- i + 1
+    }
+  }
+
+  list(
+    assign = assign,
+    matrix = mat,
+    index = index
+  )
+}
+

### 2025-04-02 R/hierarchical_param_stan_data.R

@@ -1,0 +1,110 @@
+
+#' Set up data for Stan related to hierarchical parameters
+#'
+#' @param global_fit an optional "global" fit that will be used to extract
+#'   parameter estimates for any specified hierarchical units to fix
+#' @param param_name The name of the parameter we are working with:
+#'   "P_tilde", "Omega", or "a"
+#' @param param_data Data structures as constructed by hierarchical_data
+#' @param hierarchical_terms_fixed character vector specifying hierarchical
+#'   levels for which the terms should be fixed
+#' @param hierarchical_terms_fixed character vector specifying hierarchical
+#'   levels for which the terms should be fixed
+#'
+#' @return named list with Stan data relevant to the hierarchical set up for
+#'   this parameter, e.g. if the `param_name` is `"Omega"`, these will be
+#'   `Omega_n_terms, Omega_n_sigma, Omega_re_start, Omega_re_end,
+#'   Omega_model_matrix, Omega_n_terms_fixed, Omega_n_terms_estimate,
+#'   Omega_raw_fixed, Omega_n_sigma_fixed, Omega_n_sigma_estimate,
+#'   Omega_sigma_fixed`
+hierarchical_param_stan_data <- function(param_name, param_data,
+                                         hierarchical_terms_fixed,
+                                         hierarchical_sigmas_fixed,
+                                         global_fit) {
+  result <- list()
+  result[[paste0(param_name, "_n_terms")]] <- param_data$n_terms
+  result[[paste0(param_name, "_n_sigma")]] <- param_data$n_re
+  result[[paste0(param_name, "_re_start")]] <- array(param_data$re_start)
+  result[[paste0(param_name, "_re_end")]] <- array(param_data$re_end)
+  result[[paste0(param_name, "_model_matrix")]] <- param_data$model_matrix$mat
+
+
+  # get nfixed etc, no estimates yet
+  if (is.null(global_fit)) {
+    result[[paste0(param_name, "_n_terms_fixed")]] <- 0
+    result[[paste0(param_name, "_n_terms_estimate")]] <- param_data$n_terms
+    result[[paste0(param_name, "_raw_fixed")]] <- numeric(0)
+
+    result[[paste0(param_name, "_n_sigma_fixed")]] <- 0
+    result[[paste0(param_name, "_n_sigma_estimate")]] <- param_data$n_re - 1
+    result[[paste0(param_name, "_sigma_fixed")]] <- numeric(0)
+    return(result)
+  } else {
+
+    # extract relevant data
[…]

### 2025-04-02 documentation/tmp_script.R

@@ -1,0 +1,10 @@
+devtools::load_all(here::here())
+fit1a <- fit_model_simplified(runstep = "step1a",
+                              hierarchical_level     = c("intercept",  "subcluster", "iso"),
+
+                              survey_df = dat,
+                              chains = 4)
+
+# functions to clean out
+## check_nas
+## hier data helpers

### 2025-04-03 documentation/hierarchicalmodels_seqruns.qmd
document hierarchical data functions

@@
 
-The set up and notation in stan/R, in terms of parameters, using parameter Omega as an example, is currently as follows:
+The set up and notation in stan/R, in terms of parameters, using parameter mu as an example, is currently as follows:
 
--   Omega_raw refers to the vector with ALL $\gamma$s, including the global one.The ordering is global-allregional-allsubregional-country.
+-   mu_raw refers to the vector with ALL $\gamma$s, including the global one.The ordering is global-allregional-allsubregional-country.
 
--   Omega_sigma referred to the vector of stacked sigmas, excluding $s_{\mu, global}$ (in the original package, it was called Omega_sigma_raw).
+-   mu_sigma referred to the vector of stacked sigmas, excluding $s_{\mu, global}$ (in the original package, it was called mu_sigma_raw).
 
--   Omega_star refers to the vector with (mu global and ALL $\eta$s). we get Omega_stars from Omega_raw and the vector ($s_{\mu, global}$ , Omega_sigma), using the stan `scale_blocks` function.
+-   mu_star refers to the vector with (mu global and ALL $\eta$s). we get mu_stars from mu_raw and the vector ($s_{\mu, global}$ , mu_sigma), using the stan `scale_blocks` function.
 
--   Omega refers to the final $\mu$s at the country level.
-Matrix multiplication is used to obtain the Omega country vector from the Omega_stars.
-The full matrix `Omega_model_matrix` is a sparse matrix, with 1s and 0s (columns refer to $\mu_{global}$, the regional etas, the subregional etas, the country etas).
-The matrix multiplication is implemented using the stan function csr_matrix_times_vector, using a decomposition of `Omega_model_matrix` into different parts U, V, W.
+-   mu refers to the final $\mu$s at the country level.
+Matrix multiplication is used to obtain the mu country vector from the mu_stars.
+The full matrix `mu_model_matrix` is a sparse matrix, with 1s and 0s (columns refer to $\mu_{global}$, the regional etas, the subregional etas, the country etas).
+The matrix multiplication is implemented using the stan function csr_matrix_times_vector, using a decomposition of `mu_model_matrix` into different parts U, V, W.
 
--   Because of the sequential fitting, for Omega_raw and Omega_sigma, we also had to distinguish between parameters that were fixed vs those that were estimated.
+-   Because of the sequential fitting, for mu_raw and mu_sigma, we also had to distinguish between parameters that were fixed vs those that were estimated.
 We do so with the \_fixed and \_estimate additions, i.e.
-the vector of Omega_raws is produced from Omega_raw_fixed and Omega_raw_estimate.
+the vector of mu_raws is produced from mu_raw_fixed and mu_raw_estimate.
 
@@
 
-in the simpler model used, various Omegas are defined in the transformed parameters block so we can check those directly.
+in the simpler model used, various mus are defined in the transformed parameters block so we can check those directly.
 (Note that this not (yet) true for the transition model because the creation of parameters takes place within the transition function, see notes in last section on this).
 
-here just checking the sigmas and first Omega:
+here just checking the sigmas and first mu:
 
@@
 fit <- fit1a
- parname_list <- list( c("Omega_raw_estimate[1]",
-                          "Omega_sigma_estimate"))
+ parname_list <- list( c("mu_raw_estimate[1]",
+                          "mu_sigma_estimate"))
 #    plot_dirandname <- file.path(fit$output_dir, paste0(runname, "diagnostics.pdf"))
@@
 
--   To deal with the set up where things are fixed vs estimated, we let the stan model spit out the Omega_stars, that's what the original helper functions use as inputs.
+-   To deal with the set up where things are fixed vs estimated, we let the stan model spit out the mu_stars, that's what the original helper functions use as inputs.
 
--   We clean up helper functions, so that they take in the Omega_stars, and specific other stan_data, and are better documented.
+-   We clean up helper functions, so that they take in the mu_stars, and specific other stan_data, and are better documented.
 See function extract_parameter_subhierarchical_simplified in fit_model_simplified R script, and its usage below.
@@
 # note that subhierarchies are called column in hierarchical data
-Omega <- list()
+mu <- list()
 for(subhierarchy in fit$hierarchical_level) {
-  Omega[[subhierarchy]] <-
+  mu[[subhierarchy]] <-
     extract_parameter_subhierarchical_simplified (
@@
       subhierarchy = subhierarchy,
-      parname = "Omega",
+      parname = "mu",
       fit_samples = fit$samples)
 }
-# Omega[[]] has posterior samples for the subhierarchy level
+# mu[[]] has posterior samples for the subhierarchy level
 # to get the summaries
@@
 }
-res_global <- map(Omega, function(tibble_samples) 
+res_global <- map(mu, function(tibble_samples) 
   tibble_samples %>% select(name, value)  %>% reframe(quantile_df(value), .by = name))
@@
 fit <- fit_local
-Omega <- list()
+mu <- list()
 for(subhierarchy in fit$hierarchical_level) {
-  Omega[[subhierarchy]] <-
+  mu[[subhierarchy]] <-
     extract_parameter_subhierarchical_simplified (
@@
       subhierarchy = subhierarchy,
-      parname = "Omega",
+      parname = "mu",
       fit_samples = fit$samples)
 }
-res_local <- map(Omega, function(tibble_samples) 
+res_local <- map(mu, function(tibble_samples) 
   tibble_samples %>% select(name, value)  %>% reframe(quantile_df(value), .by = name))
@@
 
--   How to get the Omegas in full model set up: in the simpler model used here, various Omegas are defined in the transformed parameters block so we can check those directly.
+-   How to get the mus in full model set up: in the simpler model used here, various mus are defined in the transformed parameters block so we can check those directly.
 Note that this not (yet) true for the full transition model because the creation of parameters takes place within the process model function.
-Options to obtain the Omegas in the full model are to
+Options to obtain the mus in the full model are to
 
-    1.  add them to generated quantities: pull out omegas with an `if {add_output_omegas}` type statement, re-using functions/statements introduced in the simpler file here.
+    1.  add them to generated quantities: pull out mus with an `if {add_output_mus}` type statement, re-using functions/statements introduced in the simpler file here.
 
-    2.  rewrite transformed parameter block and pull Omegas out of the process model function (not sure if that slows things down when not used).
+    2.  rewrite transformed parameter block and pull mus out of the process model function (not sure if that slows things down when not used).
 

### 2025-04-03 R/extract_parameter_subhierarchical.R

@@ -1,8 +1,19 @@
 
+#' Title
+#'
+#' @param hierarchical_data
+#' @param subhierarchy
+#' @param parname
+#' @param fit_samples
+#'
+#' @returns
+#' @export
+#'
+#' @examples
 extract_parameter_subhierarchical_simplified <- function(
     hierarchical_data, # hierarchical_data, obtained from hierarchical_data(fit$geo_unit, fit$hierarchical_level)
     subhierarchy, # selected hierarchical_level (example: "intercept" or "region")
-    parname, # selected parameter name (example: "Omega")
+    parname, # selected parameter name (example: "mu")
     fit_samples #fit$samples
 ) {
 

### 2025-04-03 R/fit_model_simplified.R

@@ -227,13 +227,15 @@ fit_model_simplified <- function(
 
 
 
-  # Create district index for matching district and district index
-  hierarchical_column_names <- unique(c(
-    hierarchical_level
-  )) %>%
-    setdiff("intercept")
+
 
   print("to do: check what NA check is needed")
+  # no longer needed
+  # # Create district index for matching district and district index
+  # hierarchical_column_names <- unique(c(
+  #   hierarchical_level
+  # )) %>%
+  #   setdiff("intercept")
   # # Make sure there are no NAs in any of the columns
   # for(column in hierarchical_column_names) {
   #   if(column == "intercept") next
@@ -256,10 +258,12 @@ fit_model_simplified <- function(
 
   ##### Setup data for Stan #####
 
-  geo_unit_index <- data[!is.na(data[[area]]), ] %>%
-    dplyr::distinct(!!! syms(hierarchical_column_names)) %>%
-    dplyr::mutate(c = 1:n())
-
+  geo_unit_index <- get_geo_unit_index_data(data,
+                                            hierarchical_levels = c(hierarchical_level),
+                                            area = area)
+  # need it here after all
+  hierarchical_column_names <- unique(hierarchical_level) %>%
+    setdiff("intercept")
   data <- data %>%
     dplyr::left_join(geo_unit_index, by = hierarchical_column_names)
 
@@ -301,7 +305,7 @@ fit_model_simplified <- function(
 
   ##### Set up hierarchical structures ######
   hier_data <- hier_stan_data  <- list()
-  hier_data[["Omega_data"]] <- hierarchical_data(geo_unit_index, hierarchical_level)
+  hier_data[["mu_data"]] <- hierarchical_data(geo_unit_index_data = geo_unit_index, hierarchical_level = hierarchical_level)
   #hier_data[["a_data"]] <- hierarchical_data(geo_unit_index, hierarchical_splines)
   # hier_stan_data[["a"]] <- hierarchical_param_stan_data(
   #   global_fit = global_fit,
@@ -309,13 +313,14 @@ fit_model_simplified <- function(
   #   param_data = hier_data[["a_data"]],
   #   hierarchical_terms_fixed = hierarchical_splines_terms_fixed,
   #   hierarchical_sigmas_fixed = hierarchical_splines_sigmas_fixed)
-  hier_stan_data[["Omega"]] <- hierarchical_param_stan_data(
[…]

### 2025-04-03 R/get_geo_unit_index_data.R

@@ -1,0 +1,29 @@
+
+
+#' get_geo_unit_index_data
+#'
+#' @param data survey data of interest that contains the hierarchical_column_names
+#' @param hierarchical_levels vector that contains one of more specs of hierarchical_level (incl intercept)
+#' @param area unit of analysis (eg country or subnational region)
+#'
+#' @returns tibble with unique hierarchical levels (one column for each level) and index "c" assigned based on area
+#' @export
+#'
+#' @examples
+#' survey_dat <- tibble(subcluster = c("A", "A", "B", "B"), iso = c("iso1", "iso2", "iso3", "iso3"))
+#' get_geo_unit_index_data(survey_dat,
+#'         hierarchical_levels = c("intercept", "subcluster", "iso"),
+#'         area = "iso")
+get_geo_unit_index_data <- function(data,
+                                    hierarchical_levels,
+                                    area){
+  hierarchical_column_names <- unique(hierarchical_levels) %>%
+    setdiff("intercept")
+  if (!all(hierarchical_column_names %in% names(data))) {
+    stop("data does not contain all hierarchical levels")
+  }
+  data[!is.na(data[[area]]), ] %>%
+    dplyr::distinct(!!! syms(hierarchical_column_names)) %>%
+    dplyr::mutate(c = 1:n())
+}
+

### 2025-04-03 R/get_posterior_summaries.R

@@ -2,7 +2,7 @@ get_posterior_summaries_simplified <- function(
     fit,
     process_indicator_prefixes = c(""),
     dm_indicator_prefixes = c(""),
-    process_params = c("Omega_raw", "Omega_sigma"),
+    process_params = c("mu_raw", "mu_sigma"),
     dm_params = c("nonse")
 ) {
   params_to_collect <- dplyr::bind_rows(

### 2025-04-03 R/hierarchical_data.R

@@ -1,6 +1,16 @@
-#' @import purrr
-hierarchical_data <- function(data, hierarchy) {
-  model_matrix <- hierarchical_model_matrix(hierarchy, data)
+
+#' Hierarchical_data
+#'
+#' @param geo_unit_index_data tibble with unique hierarchical levels (one column for each level)
+#' @param hierarchical_level vector with hierarchical spec
+#'
+#' @returns list with model matrix and info on start and end indices associated with different levels
+#' @export
+#'
+#' @examples
+hierarchical_data <- function(geo_unit_index_data, hierarchical_level) {
+  model_matrix <- hierarchical_model_matrix(geo_unit_index_data = geo_unit_index_data,
+                                            hierarchical_level = hierarchical_level)
   n_terms <- ncol(model_matrix$mat)
   re <- unique(model_matrix$assign)
   n_re <- length(re)

### 2025-04-03 R/hierarchical_model_matrix.R

@@ -1,13 +1,25 @@
 
-#' @import tibble
-hierarchical_model_matrix <- function(columns, data) {
-  mat <- matrix(NA, nrow = nrow(data), ncol = 0)
+
+#' hierarchical_model_matrix
+#'
+#' @param geo_unit_index_data tibble with unique hierarchical levels (one column for each level)
+#' @param hierarchical_level vector with hierarchical spec from highest to lowest levels
+#'
+#' @returns list (modelmatrix, assign, index) associated with the geo_unit_data and hierarchical_levels
+#' in modelmatrix each column refers to one eta (and nrows is the number of lowest level units).
+#' assign is length of etas, with 0 for intercept, then an hierarchical level index (starting at 2) for each eta level
+#' index is a tibble with n_eta rows, with info for each eta in and columns i [index of eta], column [hierarchical level], and level [name of hierarchical level]
+#' @export
+#'
+#' @examples
+hierarchical_model_matrix <- function(geo_unit_index_data, hierarchical_level) {
+  mat <- matrix(NA, nrow = nrow(geo_unit_index_data), ncol = 0)
   assign <- c()
 
   index <- tibble::tibble(i = numeric(0), column = character(0), level = character(0))
 
   i <- 1
-  if("intercept" %in% columns) {
+  if("intercept" %in% hierarchical_level) {
     mat <- cbind(mat, rep(1, nrow(mat)))
     assign <- c(assign, 0)
 
@@ -15,11 +27,11 @@ hierarchical_model_matrix <- function(columns, data) {
 
     i <- i + 1
   }
-
-  for(column in columns) {
-    for(l in levels(factor(data[[column]]))) {
-      mat <- cbind(mat, as.numeric(data[[column]] == l))
-      assign <- c(assign, which(column == columns))
+  # columns in model matrix are ordered by hierarchical level
+  for(column in hierarchical_level) {
+    for(l in levels(factor(geo_unit_index_data[[column]]))) {
+      mat <- cbind(mat, as.numeric(geo_unit_index_data[[column]] == l))
+      assign <- c(assign, which(column == hierarchical_level))
       index[i, ] <- tibble::tibble(i = i, column = column, level = l)
 
       i <- i + 1

### 2025-04-03 R/hierarchical_param_stan_data.R

@@ -1,33 +1,34 @@
 
 #' Set up data for Stan related to hierarchical parameters
 #'
+#' @param param_name The name of the parameter we are working with, e.g. "mu"
+#' @param param_data Data structures as constructed by hierarchical_data for that parameter
 #' @param global_fit an optional "global" fit that will be used to extract
 #'   parameter estimates for any specified hierarchical units to fix
-#' @param param_name The name of the parameter we are working with:
-#'   "P_tilde", "Omega", or "a"
-#' @param param_data Data structures as constructed by hierarchical_data
 #' @param hierarchical_terms_fixed character vector specifying hierarchical
-#'   levels for which the terms should be fixed
+#'   levels for which the terms should be fixed (subset of `hierarchical_levels`)
 #' @param hierarchical_terms_fixed character vector specifying hierarchical
-#'   levels for which the terms should be fixed
+#'   levels for which the terms should be fixed (subset of `hierarchical_levels`)
 #'
 #' @return named list with Stan data relevant to the hierarchical set up for
-#'   this parameter, e.g. if the `param_name` is `"Omega"`, these will be
-#'   `Omega_n_terms, Omega_n_sigma, Omega_re_start, Omega_re_end,
-#'   Omega_model_matrix, Omega_n_terms_fixed, Omega_n_terms_estimate,
-#'   Omega_raw_fixed, Omega_n_sigma_fixed, Omega_n_sigma_estimate,
-#'   Omega_sigma_fixed`
+#'   this parameter, e.g. if the `param_name` is `"mu"`, these will be
+#'   `mu_n_terms, mu_n_sigma, mu_re_start, mu_re_end,
+#'   mu_model_matrix, mu_n_terms_fixed, mu_n_terms_estimate,
+#'   mu_raw_fixed, mu_n_sigma_fixed, mu_n_sigma_estimate,
+#'   mu_sigma_fixed`
 hierarchical_param_stan_data <- function(param_name, param_data,
                                          hierarchical_terms_fixed,
                                          hierarchical_sigmas_fixed,
                                          global_fit) {
   result <- list()
   result[[paste0(param_name, "_n_terms")]] <- param_data$n_terms
-  result[[paste0(param_name, "_n_sigma")]] <- param_data$n_re
   result[[paste0(param_name, "_re_start")]] <- array(param_data$re_start)
   result[[paste0(param_name, "_re_end")]] <- array(param_data$re_end)
[…]

### 2025-04-03 documentation/tmp_script.R

@@ -8,3 +8,5 @@ fit1a <- fit_model_simplified(runstep = "step1a",
 # functions to clean out
 ## check_nas
 ## hier data helpers
+
+

### 2025-04-03 R/hierarchical_param_stan_data.R
make fixing values little more robust, clean up

@@ -45,12 +45,15 @@ hierarchical_param_stan_data <- function(param_name, param_data,
 
     # extract relevant data
     globalfit_param_data = global_fit[[paste0(param_name, "_data")]]
-    # later on, these are used in order
-    # instead, use the indices as stored in the summary tibble
+    # later on, these are used in order in current version
+    # WIP: use the indices as stored in the summary tibble
     globalfit_param_post_summ_raw = global_fit$post_summ %>%
       filter(variable_no_index == paste0(param_name, "_raw"))
-    globalfit_param_post_summ_sigma = global_fit$post_summ %>%
-      filter(variable_no_index == paste0(param_name, "_sigma"))
+    #globalfit_param_post_summ_sigma = global_fit$post_summ %>%
+    #  filter(variable_no_index == paste0(param_name, "_sigma"))
+    is_matrix <- is_there_a_comma(globalfit_param_post_summ_raw$variable[1])
+    if (is_matrix)
+      stop("not yet implemented")
 
     # get number of fixed terms (etas): number of etas in
     # fixed hierarchical levels where that eta was in the global fit
@@ -69,7 +72,7 @@ hierarchical_param_stan_data <- function(param_name, param_data,
           return(nrow(global_index_subset))
         } else {
           # 0 or multiple matches
-          stop("We can't fix an eta, 0 or more than 1 match")
+          stop("We can't fix an eta, 0 (or more than 1) match")
         }
       }) |>
       sum()
@@ -92,7 +95,7 @@ hierarchical_param_stan_data <- function(param_name, param_data,
         index_global %>%
         filter(column == index_local$column[k] & level == index_local$level[k]) %>%
         pull(i)
-    }
+     }
 
     # number of sigmas to fix.
     # intercept is not included
@@ -101,16 +104,29 @@ hierarchical_param_stan_data <- function(param_name, param_data,
     result[[paste0(param_name, "_n_sigma_fixed")]] <- n_sigma_fixed
     result[[paste0(param_name, "_n_sigma_estimate")]] <- n_sigma - n_sigma_fixed
 
-
-
-    ### add estimates of the fixed terms, based on global fit
-    is_matrix <- is_there_a_comma(globalfit_param_post_summ_raw$variable[1])
+    # now plug in the values
     if (!is_matrix){
[…]

### 2025-04-04 documentation/hierarchicalmodels_seqruns.qmd
make fixing values little more robust, clean up
messy working start of k param vector set up

@@
 ```{r}
+ hierarchical_level     = c("intercept",  "subcluster", "iso")
 fit1a <- fit_model_simplified(runstep = "step1a",
-        hierarchical_level     = c("intercept",  "subcluster", "iso"),
+        hierarchical_level     =  hierarchical_level,
 
@@
 #### Produce and check outputs
-
-in the simpler model used, various mus are defined in the transformed parameters block so we can check those directly.
-(Note that this not (yet) true for the transition model because the creation of parameters takes place within the transition function, see notes in last section on this).
-
-here just checking the sigmas and first mu:
-
-```{r}
-
-fit <- fit1a
- parname_list <- list( c("mu_raw_estimate[1]",
-                          "mu_sigma_estimate"))
-#    plot_dirandname <- file.path(fit$output_dir, paste0(runname, "diagnostics.pdf"))
-#    pdf(plot_dirandname, width = 11, height = 11)
-    for (i in 1:length(parname_list)){
-      parname <- parname_list[[i]]
-      samp <- fit$samples$draws(parname )
-      #mcmc_trace(samp)
-      p <- mcmc_dens_overlay(samp)
-      # for more than 1 parameter
-      #mcmc_pairs(samp)
-      print(p)
-    }
-#    dev.off()
-```
 
@@
 
+# summary object
+
+going through what's happening to create the summary object
+
+```{r}
+#get_posterior_summaries_simplified(fit1a) 
+fit <- fit1a
+process_indicator_prefixes = c("")
+dm_indicator_prefixes = c("")
+process_params = c("mu_raw", "mu_sigma")
+dm_params = c("nonse")
+
+```
+
+```{r}
+## part to get the names of parameters to fix
+params_to_collect <- dplyr::bind_rows(
+  expand.grid(
+    prefix = process_indicator_prefixes,
+    param = process_params),
+  expand.grid(
+    prefix = dm_indicator_prefixes,
+    param = dm_params)
+) |>
+  dplyr::mutate(prefixed_param_name = paste0(prefix, param))
+
+# split parameters into those that are "fixable" in local fits and those that
+# are "unfixable"
+#  unfixable_param_names <- c("local_shrinkage_dm")
+# unfixable_param_names <- c("BLA")
+fixable_params_to_collect <- params_to_collect |>
+  #   dplyr::filter(!param %in% unfixable_param_names) |>
+  dplyr::pull(prefixed_param_name)
+fixable_params_to_collect 
+```
+
+get results for 1 parameter name
+
+```{r}
+
+get_posterior_summary_one_fixable_param(prefixed_param_name = fixable_params_to_collect[1], fit = fit)
+```
+
+when using a fit where things are fixed, fixed values and estimates are combined
+
+(this will be in test case for subnational, we could also add something like 1a and 1b)
+
+```{r}
+prefixed_param_name
+```
+
+how this summary is used... global fit is used when creaiting the stan data
+
+hier data does not depend on fixed values
+
+Set up hierarchical structures, folow the hierdata.qmd
+
+```{r}
+#hierarchical_level <- c("intercept",  "subcluster", "iso")
+area <- "iso"
+```
+
+Create district index for matching district and district index
+
+```{r}
+geo_unit_index <- get_geo_unit_index_data(dat,
+                                          hierarchical_levels = c(hierarchical_level), 
+                                          area = area)
+geo_unit_index
+```
+
+```{r}
+# use in fit_model:
+#hier_data <-  list()
+#hier_data[["mu_data"]] <- 
+# here we just work with one object:
+hier_data <- hierarchical_data(geo_unit_index, hierarchical_level)
+#names(hier_data[["mu_data"]])
+names(hier_data)
+hier_data
+
+```
+
+```{r}
+hier_stan_data <- hierarchical_param_stan_data(
+  global_fit = fit1a,
+  param_name ="mu",
+  param_data = hier_data,
+  hierarchical_terms_fixed = hierarchical_level[1:2], 
+  hierarchical_sigmas_fixed = hierarchical_level[1:3]
+)
+names(hier_stan_data)
+```
+
+use for non-hierarchical parameters, sigma_y here (vector for the full fpet model)
+
+```{r, eval = FALSE}
+ ##### Set up handing of data model hyperparameters ######
+  nonse_data <- list(
+    "fix_nonse" = as.integer(fix_nonse),
+    "nonse_fixed" = numeric(0))
+  if (fix_nonse) {
+    if (is.null(global_fit)) {
+      stop("fix_nonse was set to TRUE, but a global_fit was not provided.")
+    }
+    for (parname in "nonse"){
+      nonse_data[[paste0(parname, "_fixed")]] <- global_fit$post_summ %>% filter(variable == paste0(parname, "[1]")) %>% pull(median)
+    }
+  }# end fixing dm pars
+
+```
+
 # Related notes and to dos to come back to
@@
 -   Note on approach used to get this example together (can be followed when extending further): Start subfolder for simplified model and code, added current stan_model and fit_model function and simplify to produce just the toy example.
+
+# trace plots etc
+
+in the simpler model used, various mus are defined in the transformed parameters block so we can check those directly.
+(Note that this not (yet) true for the transition model because the creation of parameters takes place within the transition function, see notes in last section on this).
+
+here just checking the sigmas and first mu:
+
+```{r}
+
+fit <- fit1a
+ parname_list <- list( c("mu_raw_estimate[1]",
+                          "mu_sigma_estimate"))
+#    plot_dirandname <- file.path(fit$output_dir, paste0(runname, "diagnostics.pdf"))
+#    pdf(plot_dirandname, width = 11, height = 11)
+    for (i in 1:length(parname_list)){
+      parname <- parname_list[[i]]
+      samp <- fit$samples$draws(parname )
+      #mcmc_trace(samp)
+      p <- mcmc_dens_overlay(samp)
+      # for more than 1 parameter
+      #mcmc_pairs(samp)
+      print(p)
+    }
+#    dev.off()
+```

### 2025-04-04 documentation/hierarchicalmodels_seqruns_2par.qmd
messy working start of k param vector set up

---
title: "Hierarchical models and seq model fits"
author: "Leontine"
format: html
editor: visual
---

See 1parameter version first, this is a shortened one

```{r}
#| include: false
library(here) 
library(tidyverse) 
library(ggplot2) 
library(haven) 
library(stringr)
library(cmdstanr) 
library(bayesplot) # diagnostic plots 
devtools::load_all(here::here())
```

Read in data and preprocess as usual

```{r}
#| include: false
data_folder <- "data_raw" 
dat <- read_csv(here::here(data_folder, "coverage_data.csv"))
```

Let's use columns `cluster` and `subcluster` and `iso`:

```{r}
dat %>% select(iso, cluster, subcluster)
```

Summary of clusters and subclusters, after selecting one row per country

```{r}
dat %>%
  group_by(iso) %>%
  filter(year == max(year)) %>%
  select(country, iso, subcluster, cluster) %>%
  group_by(cluster) %>%
  summarize(nsubclusters = length(unique(subcluster)), ncountries = n())
```

# Model fitting

Easiest 2 parameter set up is to estimate the same mu twice.

(later also use 2 nonses)

global model fit

```{r}
 hierarchical_level     = c("intercept",  "subcluster", "iso")
fit1a <- fit_model_simplified(runstep = "step1a",
                                stan_file_path = file.path(here::here("inst/stan/", "hierfunctions_seq_2param.stan")),
                              mu2param = TRUE, 
        hierarchical_level     =  hierarchical_level,
                  survey_df = dat,
                  chains = 4)

```

create summary object

```{r}
fit1a$post_summ <- get_posterior_summaries_simplified(fit1a) %>%
  dplyr::rename(median = mean)
fit1a$post_summ
fit1a$post_summ %>% filter(variable %in% c("mu_raw[1,1]", "mu_raw[1,2]", "mu_raw[2,1]", "mu_raw[2,2]"))

```

do local fit... here we fix things but do get results for all countries

```{r}
#fit1a$samples <- NULL
fit_local <- fit_model_simplified(runstep = "local_national",
                                    stan_file_path = file.path(here::here("inst/stan/", "hierfunctions_seq_2param.stan")),
                              mu2param = TRUE, 
                  global_fit = fit1a,
                  survey_df = dat,
                  chains = 4)

```

check out extract_parameter_subhier function

```{r}
subhierarchy <- fit$hierarchical_level[1]
hierarchical_data = hierarchical_data(fit$geo_unit, fit$hierarchical_level)
parname = "mu"
fit_samples = fit$samples

```

```{r}
 start <- hierarchical_data$model_matrix$index %>%
    dplyr::filter(column == subhierarchy) %>%
    dplyr::pull(i) %>%
    min()

  end <- hierarchical_data$model_matrix$index %>%
    dplyr::filter(column == subhierarchy) %>%
    dplyr::pull(i) %>%
    max()

  # from bayestransition
  # a_star <- fit$samples$draws(c("a_star")) %>%
  #   tidybayes::spread_draws(a_star[j, i])
  #
  # a_star <- a_star %>%
  #   dplyr::group_by(.data$i, .data$.chain, .data$.iteration, .data$.draw) %>%
  #   dplyr::summarize(a_star =  list(.data[["a_star"]]))
  #

  pars <- c(glue::glue("{parname}_star"))
pars
  star <- fit_samples$draws(pars) %>%
    # for one param
    #tidybayes::spread_draws((!!sym(pars[1]))[i]) %>%
    #dplyr::group_by(.chain, .iteration, .draw) %>%
    tidybayes::spread_draws((!!sym(pars[1]))[i,k]) %>%
     dplyr::group_by(.data$k, .data$.chain, .data$.iteration, .data$.draw) %>%
    tidyr::nest() %>%
    dplyr::mutate(star = map(data, `[[`, glue::glue("{parname}_star"))) %>%
    dplyr::select(-data)
star

```

```{r}
  uniq <- unique(hierarchical_data$model_matrix$mat[, 1:end, drop = FALSE])

  titles <- c()
  for(i in 1:nrow(uniq)) {
    index <- rep(0, hierarchical_data$n_terms)
    index[1:end] <- uniq[i, 1:end]
    title <- hierarchical_data$model_matrix$index %>%
      dplyr::filter(i == last(which(index == 1))) %>%
      dplyr::pull(level)
    titles <- c(titles, title)

    star[[title]] = map_dbl(star[["star"]], function(star) {
      index %*% star
    })
  }

  star <- star %>%
    tidyr::pivot_longer(cols = all_of(titles)) %>%
    dplyr::ungroup() %>%
    dplyr::select(-star)
  star
```

```{r}
quantile_df <- function(x, probs = c(0.025, 0.5, 0.975)) {
  tibble(
    val = quantile(x, probs, na.rm = TRUE),
    quant = probs
  )
}

```

```{r}
star %>% select(name, value, k)  %>% reframe(quantile_df(value), .by = c(name, k))
```

```{r}
fit <- fit1a
mu <- list()
for(subhierarchy in fit$hierarchical_level) {
  mu[[subhierarchy]] <-
    extract_parameter_subhierarchical_simplified_kparam(
      hierarchical_data = hierarchical_data(fit$geo_unit, fit$hierarchical_level),
      subhierarchy = subhierarchy,
      parname = "mu",
      fit_samples = fit$samples)
}

```

```{r}
# mu[[]] has posterior samples for the subhierarchy level
# to get the summaries
res_global <- map(mu, function(tibble_samples) 
#  # for one param
#  tibble_samples %>% select(name, value)  %>% reframe(quantile_df(value), .by = name))
 # for k param
  tibble_samples %>% select(name, value, k)  %>% reframe(quantile_df(value), .by = c(name, k)))
res_global
```

```{r}
fit <- fit_local
mu <- list()
for(subhierarchy in fit$hierarchical_level) {
  mu[[subhierarchy]] <-
#    extract_parameter_subhierarchical_simplified (
  extract_parameter_subhierarchical_simplified_kparam(
    hierarchical_data = hierarchical_data(fit$geo_unit, fit$hierarchical_level),
      subhierarchy = subhierarchy,
      parname = "mu",
      fit_samples = fit$samples)
}
res_local <- map(mu, function(tibble_samples) 
#  tibble_samples %>% select(name, value)  %>% reframe(quantile_df(value), .by = name))
    tibble_samples %>% select(name, value, k)  %>% reframe(quantile_df(value), .by = c(name, k)))
res_local
```

checks that results from local model approximate those from global model

at subcluster level

```{r}
bind_rows(res_local$subcluster %>% mutate(model = "local"),
          res_global$subcluster %>% mutate(model = "global")) %>%
  group_by(name, model, k) %>% 
  reframe(y = val[quant == 0.5], ymin = val[quant == 0.025], ymax = val[quant == 0.975]) %>%
  ggplot(aes(y = y, x = name, color = model)) +
  geom_errorbar(aes(ymin = ymin, ymax = ymax)) +
  geom_point()

```

at country level

```{r}
bind_rows(res_local$iso %>% mutate(model = "local"),
          res_global$iso %>% mutate(model = "global")) %>%
  group_by(name, model, k) %>% 
  reframe(y = val[quant == 0.5], ymin = val[quant == 0.025], ymax = val[quant == 0.975]) %>%
  ggplot(aes(y = y, x = name, color = model)) +
  geom_errorbar(aes(ymin = ymin, ymax = ymax)) +
  geom_point()   
```

### 2025-04-04 documentation/hierdata.qmd
messy working start of k param vector set up

---
title: "hierarchical data"
author: "leontine"
format: html
editor: visual
---

Load relevant packages, including our own and data

```{r}
#| include: false
library(here) 
library(tidyverse) 
library(ggplot2) 
library(haven) 
library(stringr)
library(cmdstanr) 
library(bayesplot) # diagnostic plots 
devtools::load_all(here::here())
#data_folder <- "data_raw" 
# calling survey_df data here, as internal to fit_model
#data <- read_csv(here::here(data_folder, "coverage_data.csv"))
```

toy data set

```{r}
data <- tibble(subcluster = c("A", "A", "B", "B"), iso = c("iso1", "iso2", "iso3", "iso3"))
```

main info: hierarchical levels to use, and the level of interest

(level of interest does not need to be the lowest level, not clear how thoroughly tested)

```{r}
hierarchical_level <- c("intercept",  "subcluster", "iso")
area <- "iso"
```

Create district index for matching district and district index

```{r}
geo_unit_index <- get_geo_unit_index_data(data,
                                          hierarchical_levels = c(hierarchical_level), 
                                          area = area)
geo_unit_index
```

Set up hierarchical structures

```{r}
# use in fit_model:
#hier_data <-  list()
#hier_data[["mu_data"]] <- 
# here we just work with one object:
hier_data <- hierarchical_data(geo_unit_index, hierarchical_level)
#names(hier_data[["mu_data"]])
names(hier_data)
hier_data

```

for use in stan

```{r}
hier_stan_data <- hierarchical_param_stan_data(
  global_fit = NULL,
  param_name ="mu",
  param_data = hier_data #[["mu_data"]] #,
  # used only when fixing things 
#  hierarchical_terms_fixed = hierarchical_level_terms_fixed,
#  hierarchical_sigmas_fixed = hierarchical_level_sigmas_fixed
)
names(hier_stan_data)
```

```{r}
hier_stan_data$mu_n_terms # number of etas + global mu
length(unique(geo_unit_index$subcluster))
length(unique(geo_unit_index$iso))

```

```{r}
hier_stan_data$mu_n_sigma # update: no longer includes prior sd for intercept 
```

Let's illustrate how these objects are used in R and in the stan model.

For that, make up some values for the parameters based on

```         
"mu_n_terms"          "mu_n_sigma"
```

```{r}
#mu_raw refers to the vector with ALL $\gamma$s, including the global one.The ordering is global-allregional-allsubregional-country.
mu_raw <- seq(1, hier_stan_data$mu_n_terms)
mu_raw
#mu_sigma referred to the vector of stacked sigmas, excluding $s_{\mu, global}$ (in the #original package, it was called mu_sigma_raw).
mu_sigma <- seq(1, hier_stan_data$mu_n_sigma)
sprior <- 0.5

```

try to expose stan functions using cmdstanr (not rstan)

<https://mc-stan.org/cmdstanr/reference/model-method-expose_functions.html>

```{r}
stan_file <- here("inst/stan/hierfunctions_seq.stan")
mod <- cmdstan_model(stan_file, force_recompile = TRUE)  
mod$expose_functions(global = TRUE) # to add to global environment

```

#mu_star refers to the vector with (mu global and ALL $\eta$s). we get mu_stars from mu_raw \# and the vector ($s_{\mu, global}$ , mu_sigma), using the stan scale_blocks function.

```{r}
mu_star <- scale_blocks(mu_raw, c(sprior, mu_sigma), 
             start = hier_stan_data$mu_re_start,
             end = hier_stan_data$mu_re_end)
mu_star
```

mu refers to the final $\mu$s at the country level. Matrix multiplication is used to obtain the mu country vector from the mu_stars. The full matrix mu_model_matrix is a sparse matrix, with 1s and 0s (columns refer to $\mu_{global}$, the regional etas, the subregional etas, the country etas). The matrix multiplication is implemented using the stan function csr_matrix_times_vector, using a decomposition of mu_model_matrix into different parts U, V, W.

check if this is right

```{r}
mu <- hier_stan_data$mu_model_matrix%*%mu_star
mu
```

#

```{r}
hier_data$model_matrix
```

# Unsuccessful investigation into exposing stan build-in functions

Need to expose (compile) the stan math functions as well, eg for for sparse matrices. see here for info

<https://mc-stan.org/docs/functions-reference/sparse_matrix_operations.html#conversion-functions>

<https://stan-dev.r-universe.dev/articles/StanHeaders/stanmath.html>

<https://discourse.mc-stan.org/t/expose-stan-functions-with-stan-math-library-function/15771/6>

help(stanFunction, package = "StanHeaders")

first time around, i got an error

Error in sourceCpp(code = code, env = env, rebuild = rebuild, cacheDir = cacheDir, : Error 1 occurred building shared library.

Turns out that it may not work anyway:

looks like csr_extract_w would be tricky... (was getting errors anyway with something along lines of StanHeaders::stanFunction("csr_extract_w"))

<https://discourse.mc-stan.org/t/function-to-return-number-of-non-zero-elements-in-a-matrix/26123/7>

looks like we just want model_matrix\*vector

and can confirm if

csr_matrix_times_vector

```{r}
#csr_extract_w()
```

```{r}
# get_onehierparam(oneparam_sigma = c(sprior, mu_sigma),  
#                  oneparam_raw_fixed = mu_raw[1], #numeric(0), 
#                  oneparam_raw_estimate = mu_raw[-1],
#                  oneparam_re_start = hier_stan_data$mu_re_start,
#                  oneparam_re_end = hier_stan_data$mu_re_end, 
#                  C = nrow(geo_unit_index), 
#                  oneparam_n_terms = hier_stan_data$mu_n_terms, 
#                  oneparam_n_terms_fixed = 1, #hier_stan_data$mu_n_terms_fixed, 
#                  oneparam_n_terms_estimate = hier_stan_data$mu_n_terms_estimate -1,
#       oneparam_model_matrix_w= csr_extract_w(hier_stan_data$mu_model_matrix), 
#       oneparam_model_matrix_v = hier_stan_data$mu_model_matrix_v, 
#       oneparam_model_matrix_u = hier_stan_data$mu_model_matrix_u
#           )
```

### 2025-04-04 R/extract_parameter_subhierarchical.R

@@ -60,3 +60,55 @@ extract_parameter_subhierarchical_simplified <- function(
 
   return(star)
 }
+
+# repeated code to get this going
+extract_parameter_subhierarchical_simplified_kparam <- function(
+    hierarchical_data, # hierarchical_data, obtained from hierarchical_data(fit$geo_unit, fit$hierarchical_level)
+    subhierarchy, # selected hierarchical_level (example: "intercept" or "region")
+    parname, # selected parameter name (example: "mu")
+    fit_samples #fit$samples
+) {
+
+
+  start <- hierarchical_data$model_matrix$index %>%
+    dplyr::filter(column == subhierarchy) %>%
+    dplyr::pull(i) %>%
+    min()
+
+  end <- hierarchical_data$model_matrix$index %>%
+    dplyr::filter(column == subhierarchy) %>%
+    dplyr::pull(i) %>%
+    max()
+
+  pars <- c(glue::glue("{parname}_star"))
+
+  star <- fit_samples$draws(pars) %>%
+    tidybayes::spread_draws((!!sym(pars[1]))[i,k]) %>%
+    dplyr::group_by(.data$k, .data$.chain, .data$.iteration, .data$.draw) %>%
+    tidyr::nest() %>%
+    dplyr::mutate(star = map(data, `[[`, glue::glue("{parname}_star"))) %>%
+    dplyr::select(-data)
+
+  uniq <- unique(hierarchical_data$model_matrix$mat[, 1:end, drop = FALSE])
+
+  titles <- c()
+  for(i in 1:nrow(uniq)) {
+    index <- rep(0, hierarchical_data$n_terms)
+    index[1:end] <- uniq[i, 1:end]
+    title <- hierarchical_data$model_matrix$index %>%
+      dplyr::filter(i == last(which(index == 1))) %>%
+      dplyr::pull(level)
+    titles <- c(titles, title)
+
+    star[[title]] = map_dbl(star[["star"]], function(star) {
+      index %*% star
+    })
+  }
+
+  star <- star %>%
+    tidyr::pivot_longer(cols = all_of(titles)) %>%
+    dplyr::ungroup() %>%
+    dplyr::select(-star)
+
+  return(star)
+}

### 2025-04-04 R/fit_model_simplified.R

@@ -137,7 +137,9 @@ fit_model_simplified <- function(
   seed = 1234,
   refresh = 200,
   adapt_delta = 0.9,
-  max_treedepth = 14
+  max_treedepth = 14,
+  stan_file_path = file.path(here::here("inst/stan/", "hierfunctions_seq.stan")),
+  mu2param = FALSE
 ) {
 
   data <- survey_df # for now, just to keep the same name as in fpem
@@ -249,7 +251,6 @@ fit_model_simplified <- function(
 
   ##### Load model #####
   if (compile_model){
-    stan_file_path <- file.path(here::here("inst/stan/", "hierfunctions_seq.stan"))
     stan_model <- cmdstanr::cmdstan_model(
       stan_file = stan_file_path,
       force_recompile = force_recompile
@@ -302,6 +303,11 @@ fit_model_simplified <- function(
     geo_unit = array(ifelse(is.na(data$c), 0L, data$c)),
     y = array(data[[y]])
   )
+  # area = data[[area]]
+  if (mu2param){
+    # for 2 parameter model
+    stan_data[["k"]] <- 2
+  }
 
   ##### Set up hierarchical structures ######
   hier_data <- hier_stan_data  <- list()

### 2025-04-04 R/get_posterior_summaries.R

@@ -1,3 +1,25 @@
+
+# to do
+# consider udpate to prefixes, process vs dm set up etc...
+# different way to collect parameter names to get fixable_params_to_collect
+
+# gets vector of parnames to fix
+# calls get_posterior_summary_one_fixable_param for each of them
+# note that parameter has NO index with it...
+
+
+#' get_posterior_summaries_simplified
+#'
+#' @param fit # fit object to summarize
+#' @param process_indicator_prefixes # prefixes for parameters from process model
+#' @param dm_indicator_prefixes
+#' @param process_params
+#' @param dm_params
+#'
+#' @returns
+#' @export
+#'
+#' @examples
 get_posterior_summaries_simplified <- function(
     fit,
     process_indicator_prefixes = c(""),
@@ -5,6 +27,8 @@ get_posterior_summaries_simplified <- function(
     process_params = c("mu_raw", "mu_sigma"),
     dm_params = c("nonse")
 ) {
+
+  ## part to get the names of parameters to fix
   params_to_collect <- dplyr::bind_rows(
     expand.grid(
       prefix = process_indicator_prefixes,
@@ -23,6 +47,8 @@ get_posterior_summaries_simplified <- function(
     #   dplyr::filter(!param %in% unfixable_param_names) |>
     dplyr::pull(prefixed_param_name)
 
+  #####
+
   # unfixable_params_to_collect <- params_to_collect |>
   #   dplyr::filter(param %in% unfixable_param_names) |>
   #   dplyr::pull(prefixed_param_name)
@@ -32,30 +58,21 @@ get_posterior_summaries_simplified <- function(
   print(
     "Warning: in obtaining posterior summary in get_posterior_summary_one_fixable_param, we have hardcoded num_inds based on names of parameters."
   )
-  fixable_combined_summary <- purrr::map(
-    fixable_params_to_collect,
-    get_posterior_summary_one_fixable_param,
-    fit = fit
+  # new use of map with arguments to function
+  # seq(1,2) %>% map(\(x) sum(x, 2))
+  fixable_combined_summary <-
+    # purrr::map(
+    # fixable_params_to_collect,
+    # get_posterior_summary_one_fixable_param,
+    # fit = fit
+    fixable_params_to_collect %>%
+    purrr::map(\(x) get_posterior_summary_one_fixable_param(x, fit = fit)
   ) |>
     purrr::list_rbind()%>%
-    # LA added column wtith parname w/o []
+    # add column wtith parname w/o []
     dplyr::mutate(
[…]

### 2025-04-04 R/get_posterior_summary_one_fixable_param.R

@@ -1,8 +1,27 @@
 
+# what happens:
+# get estimate and/or fixed value
+# if there is a fixed value, need to use that
+# difficulty = figuring out what indices to use in summary
+# some messiness re mean or median
+
+
+
+#' Title
+#'
+#' @param fit fit object
+#' @param prefixed_param_name parameter name, can include an index
+#'
+#' @returns
+#' @export
+#'
+#' @examples
 get_posterior_summary_one_fixable_param <- function(
     fit,
     prefixed_param_name) {
-  # extract parameter summaries for estimated parameters
+
+
+  # extract parameter summaries for estimated parameter
   estimate_param_name <- paste0(prefixed_param_name, "_estimate")
   #print(estimate_param_name)
   if (variable_exists_in_draws(fit, estimate_param_name)) {
@@ -33,18 +52,16 @@ get_posterior_summary_one_fixable_param <- function(
   }
 
   # what is the dimension of the array we're expecting?
-  # param_inf <- fit$stan_model$variables()[["parameters"]][[estimate_param_name]]
-  # num_inds <- param_inf[["dimensions"]]
+  param_inf <- fit$stan_model$variables()[["parameters"]][[estimate_param_name]]
+  num_inds <- param_inf[["dimensions"]]
   # update la: when reading in a run, this `fit$stan_model$variables()` does not exist
   # it's used here just to get the dimension, 1 or 2
   # easiest to check in main function, so passing here as an argument
-  # hardcoding here
-  #warning
-  print("Warning: in obtaining posterior summary in get_posterior_summary_one_fixable_param, we have hardcoded num_inds based on names of parameters.")
-  num_inds = ifelse(prefixed_param_name %in% c("a_raw", "a_sigma", "d_a_raw", "d_a_sigma",
-                                               "z_a_raw", "z_a_sigma"), 2, 1)
-
-
+  # # hardcoding here
+  # #warning
+  # print("Warning: in obtaining posterior summary in get_posterior_summary_one_fixable_param, we have hardcoded num_inds based on names of parameters.")
+  # num_inds = ifelse(prefixed_param_name %in% c("a_raw", "a_sigma", "d_a_raw", "d_a_sigma",
+  #                                              "z_a_raw", "z_a_sigma"), 2, 1)
 
   # process variable names in param_estimate_summary, splitting
   # into columns var_name, i, and possibly j, where i and j

### 2025-04-04 R/hier_data_helpers.R

@@ -73,9 +73,25 @@ variable_exists_in_draws <- function(fit, var_name) {
 }
 
 
+
+#' split_var_name_and_index
 #' For a data frame of estimates with a column that contains combined info
 #' about a variable name and row and/or column indices, add columns containing
 #' just the variable name and the indices.
+#'
+#' @param estimate_df data frame with a column called `variable` that contains parname[i] or parname[i,j] entries
+#' @param num_inds number of indices in the variable name (used for 1 or 2)
+#'
+#' @returns data frame with the same columns as `estimate_df` plus columns containing
+#' just the variable name and the indices.
+#' @export
+#'
+#' @examples
+#' split_var_name_and_index(
+#' estimate_df = tibble(variable = c("a[1]", "a[2]")), num_inds = 1)
+#' split_var_name_and_index(
+#' estimate_df = tibble(variable = c("a[1,1]", "a[2,1]")), num_inds = 2)
+#'
 split_var_name_and_index <- function(estimate_df, num_inds) {
   # column names for indices, starting from "i"
   ind_names <- letters[seq(from = 9, length = num_inds)]
@@ -98,6 +114,14 @@ split_var_name_and_index <- function(estimate_df, num_inds) {
 
   return(estimate_df)
 }
+# # if we get it wrong
+# split_var_name_and_index(
+#   estimate_df = tibble(variable = c("a[1,1]", "a[2,1]")),
+#   num_inds = 1)
+# split_var_name_and_index(
+#   estimate_df = tibble(variable = c("a[1]", "a[2]")),
+#   num_inds = 2)
+
 
 #' Starting from an array of parameter values (of dimension 1 or 2),
 #' create a data frame with columns `variable` with the combined parameter name

### 2025-04-04 R/hierarchical_param_stan_data.R

@@ -51,13 +51,11 @@ hierarchical_param_stan_data <- function(param_name, param_data,
       filter(variable_no_index == paste0(param_name, "_raw"))
     #globalfit_param_post_summ_sigma = global_fit$post_summ %>%
     #  filter(variable_no_index == paste0(param_name, "_sigma"))
-    is_matrix <- is_there_a_comma(globalfit_param_post_summ_raw$variable[1])
-    if (is_matrix)
-      stop("not yet implemented")
 
     # get number of fixed terms (etas): number of etas in
     # fixed hierarchical levels where that eta was in the global fit
     # this code checks whether there are matches for all requested etas
+    # this code is the same for 1par or parvector
     n_terms_fixed <- purrr::map2_int(
       param_data$model_matrix$index$column,
       param_data$model_matrix$index$level,
@@ -105,8 +103,13 @@ hierarchical_param_stan_data <- function(param_name, param_data,
     result[[paste0(param_name, "_n_sigma_estimate")]] <- n_sigma - n_sigma_fixed
 
     # now plug in the values
+    is_matrix <- is_there_a_comma(globalfit_param_post_summ_raw$variable[1])
+    #if (is_matrix)
+    #  stop("not yet implemented")
+
     if (!is_matrix){
       results_sigma <- rep(NA, n_sigma_fixed)
+      # here index k used for etas [not dimension of parvector]
       for (k in 1:n_sigma_fixed){
         results_sigma[k] <- global_fit$post_summ %>%
           filter(variable == paste0(param_name, "_sigma[", k, "]")) %>%
@@ -121,12 +124,16 @@ hierarchical_param_stan_data <- function(param_name, param_data,
       result[[paste0(param_name, "_raw_fixed")]] <- results_eta # globalfit_param_post_summ_raw$median[indices_for_local]
       result[[paste0(param_name, "_sigma_fixed")]] <- results_sigma # globalfit_param_post_summ_sigma$median[seq_len(n_sigma_fixed)]
     } else {
-
-    } else {
-      # just loop over the 2nd index of the parameter!
-      # or make output for all a matrix (then do need to fix some things for 1param next steps)
-      #result[[paste0(param_name, "_raw_fixed")]] <- create_a_matrix(globalfit_param_post_summ_raw[, c("variable", "median")])[indices_for_local, , drop = FALSE]
[…]

### 2025-04-04 documentation/tmp_script.R

@@ -10,3 +10,40 @@ fit1a <- fit_model_simplified(runstep = "step1a",
 ## hier data helpers
 
 
+
+# overview of function calls
+# get the stan_datat
+data <- tibble(subcluster = c("A", "A", "B", "B"), iso = c("iso1", "iso2", "iso3", "iso3"))
+hierarchical_level <- c("intercept",  "subcluster", "iso")
+area <- "iso"
+geo_unit_index <- get_geo_unit_index_data(data,
+                                          hierarchical_levels = c(hierarchical_level),
+                                          area = area)
+hier_data <- hierarchical_data(geo_unit_index, hierarchical_level)
+# here the global fit can be used
+hier_stan_data <- hierarchical_param_stan_data(
+  global_fit = NULL,
+  param_name ="mu",
+  param_data = hier_data #[["mu_data"]] #,
+  # used only when fixing things
+  #  hierarchical_terms_fixed = hierarchical_level_terms_fixed,
+  #  hierarchical_sigmas_fixed = hierarchical_level_sigmas_fixed
+)
+
+# when working with a vector of parameters,
+# eg instead of mu[i] we have mu[i, k] for k>1
+# with same hierarchical model
+# then hier_stan_data changes only if we fix things (as we need to input mu[i,k]s as opposed to mu[i]s)
+hier_stan_data <- hierarchical_param_stan_data(
+  global_fit = NULL,
+  param_name ="mu",
+  param_num_indices = 2,
+  param_data = hier_data #[["mu_data"]] #,
+  # used only when fixing things
+  #  hierarchical_terms_fixed = hierarchical_level_terms_fixed,
+  #  hierarchical_sigmas_fixed = hierarchical_level_sigmas_fixed
+)
+
+
+
+

### 2025-04-04 documentation/hierarchicalmodels_seqruns.qmd
document function to get subhierarchy estimates

@@
 -   We clean up helper functions, so that they take in the mu_stars, and specific other stan_data, and are better documented.
-See function extract_parameter_subhierarchical_simplified in fit_model_simplified R script, and its usage below.
+See function extract_parameter_subhierarchical in fit_model_simplified R script, and its usage below.
+
+```{r}
+quantile_df <- function(x, probs = c(0.025, 0.5, 0.975)) {
+  tibble(
+    val = quantile(x, probs, na.rm = TRUE),
+    quant = probs
+  )
+}
+```
 
@@
   mu[[subhierarchy]] <-
-    extract_parameter_subhierarchical_simplified (
+    extract_parameter_subhierarchical (
       hierarchical_data = hierarchical_data(fit$geo_unit, fit$hierarchical_level),
@@
 # to get the summaries
-quantile_df <- function(x, probs = c(0.025, 0.5, 0.975)) {
-  tibble(
-    val = quantile(x, probs, na.rm = TRUE),
-    quant = probs
-  )
-}
+
 res_global <- map(mu, function(tibble_samples) 
@@
 ```{r}
-prefixed_param_name
+#prefixed_param_name
 ```

### 2025-04-04 documentation/hierarchicalmodels_seqruns_2par.qmd
document function to get subhierarchy estimates

@@
 ```{r}
+fit <- fit1a
 subhierarchy <- fit$hierarchical_level[1]
@@
   mu[[subhierarchy]] <-
-    extract_parameter_subhierarchical_simplified_kparam(
+    extract_parameter_subhierarchical(
       hierarchical_data = hierarchical_data(fit$geo_unit, fit$hierarchical_level),
@@
       parname = "mu",
-      fit_samples = fit$samples)
+      fit_samples = fit$samples,
+      morethan1param = TRUE)
 }
@@
   mu[[subhierarchy]] <-
-#    extract_parameter_subhierarchical_simplified (
-  extract_parameter_subhierarchical_simplified_kparam(
+#    extract_parameter_subhierarchical (
+  extract_parameter_subhierarchical(
     hierarchical_data = hierarchical_data(fit$geo_unit, fit$hierarchical_level),
@@
       parname = "mu",
-      fit_samples = fit$samples)
+      fit_samples = fit$samples,
+          morethan1param = TRUE)
+
 }

### 2025-04-04 R/extract_parameter_subhierarchical.R

@@ -1,55 +1,75 @@
 
-#' Title
+#' extract_parameter_subhierarchical
 #'
-#' @param hierarchical_data
-#' @param subhierarchy
-#' @param parname
-#' @param fit_samples
+#' @param hierarchical_data hierarchical_data, obtained from hierarchical_data(fit$geo_unit, fit$hierarchical_level)
+#' @param subhierarchy selected hierarchical_level (example: "intercept" or "region")
+#' @param parname selected parameter name (example: "mu")
+#' @param fit_samples fit$samples including parname_star
+#' @param morethan1param does parname refer to more than 1 parameter (a vector)
 #'
-#' @returns
+#' @returns a tibble with posterior samples of the selected parameter mu at the subhierarchy level.
+#' these mus are obtained by summing up all relevant etas
+#' variables are `name`, `value`, `.draw/iteration/chain'
+#' and `k` referring to parameter index if morethan1param = TRUE
 #' @export
 #'
 #' @examples
-extract_parameter_subhierarchical_simplified <- function(
-    hierarchical_data, # hierarchical_data, obtained from hierarchical_data(fit$geo_unit, fit$hierarchical_level)
-    subhierarchy, # selected hierarchical_level (example: "intercept" or "region")
-    parname, # selected parameter name (example: "mu")
-    fit_samples #fit$samples
+extract_parameter_subhierarchical <- function(
+    hierarchical_data,
+    subhierarchy,
+    parname,
+    fit_samples,
+    morethan1param = FALSE
+
 ) {
 
+  # not currently used
+  # start <- hierarchical_data$model_matrix$index %>%
+  #   dplyr::filter(column == subhierarchy) %>%
+  #   dplyr::pull(i) %>%
+  #   min()
 
-  start <- hierarchical_data$model_matrix$index %>%
-    dplyr::filter(column == subhierarchy) %>%
-    dplyr::pull(i) %>%
-    min()
-
+  # end is used to determine the number of columns in model matrix to be used
   end <- hierarchical_data$model_matrix$index %>%
     dplyr::filter(column == subhierarchy) %>%
     dplyr::pull(i) %>%
     max()
 
+  # get the param_stars
   pars <- c(glue::glue("{parname}_star"))
-
-  star <- fit_samples$draws(pars) %>%
[…]

### 2025-04-07 R/fit_model.R
update documentation

@@ -1,19 +1,20 @@
 
-# functions for simplified model fitting
 
 
-#' Simplified fitting function
+
+#' Fit model
 #'
 #' @param survey_df tibble with survey data
 #' @param y column name of outcome.
-#' @param area column name of the area of each observation (geounit, iso or subnational region)
+#' @param area column name of the area of each observation (geounit; iso or subnational region)
+#' @param area_select area name to use for local run (eg iso code or subnat region name)
 #'
 #' @param runstep type of run, currently one of XXX (see Details).
 #' @param global_fit optional object of class XX, used to obtain fixed
 #'   values to use for some parameters in the current fit (see Details).
-#' @param area_select area name to use for local run (eg iso code or subnat region name)
-#'
 #' @param hierarchical_level vector specifying hierarchical structure used for mu
+#' @param add_subnational_hierarchy level that's added to the hierarchy for subnational, defaults to 'subnat'
+#' @param mu_isvector Boolean, TRUE if mu is a vector, defaults to FALSE
 #'
 #' Settings for sampling
 #' @param chains number of chains to run
@@ -26,83 +27,41 @@
 #' @param adapt_delta target acceptance rate for the No-U-Turn Sampler
 #' @param max_treedepth maximum tree depth for the No-U-Turn Sampler
 #'
-#' @return
+#' @returns List that contains samples, stan_data, and other information relevant to model fit (arguments)
 #'
 #' @details
-#' The `fit_model_simplified` function fits the toy example for hierarchical models/seq fitting.
+#' The `fit_model` function fits the toy example for hierarchical models/seq fitting.
 #' The argument \code{runstep} determines the type of run to perform. The
 #' following run steps are supported:
 #' - "step1a": Fit the global model
 #' - "local_national": Fit the model to data from a single country, using a 1a fit.
+#' - "global_subnational": Fit the model to global database with subnational data, using a 1a fit.
+#' - "local_subnational": Fit the model to subnational data from a single country or region, using a global_subnational fit.
 #' This is also explained in the documentation folder.
 #'
 #' Details on hierarchical set ups used
-#' Several area-specific parameters of the fpemplus model have hierarchical priors
-#' assigned to them so that information can be shared between areas.
 #' The package allows the structure of the hierarchical prior to be configured by the user
 #' through the  \code{hierarchical_level} argument.
[…]

### 2025-04-07 R/get_posterior_summaries.R

@@ -10,7 +10,6 @@
 #' @returns tibble with variable (parname with index used in model) and postmean (posterior mean)
 #' @export
 #'
-#' @examples
 get_posterior_summaries <- function(
     fit,
     params = c("mu_raw", "mu_sigma", "nonse")

### 2025-04-07 R/inits.R

@@ -1,9 +1,11 @@
 
-# with or without inits, warnings can occur like pasted below
-# due to mean being sum of many mu_stars?
-# perhaps less warnings when using inits to keep mu_raws and mu_sigmas more constrained...
+# warnings can occur like pasted below
 # Chain XXX Informational Message: The current Metropolis proposal is about to be rejected because of the following issue:
 # Chain XXX Exception: normal_lpdf: Location parameter is inf, but must be finite!
+# ie mu is not finite..
+# meaning inf values are drawn for sum of mustars, so product (mu_raw* mu_sigma)
+# by setting inits, we avoid this for the initial values
+# HOWEVER they can still occur early on in the chain (not sure if adding a cap is good thing to do)
 
 
 init_fun <- function(chain_id, stan_data){

### 2025-04-07 documentation/hierarchicalmodels_seqruns.qmd
update quarto documents

--- 2025-04-03 documentation/hierarchicalmodels_seqruns.qmd
@@
 ---
-title: "Hierarchical models and seq model fits"
+title: "Hierarchical models and sequential model fits"
 author: "Leontine"
@@
 
-1.  how we set up a hierarchical model for a parameter;
+1.  how we set up a hierarchical model for a parameter; this code set up was introduced in <https://github.com/AlkemaLab/BayesTransitionModels>.
 
@@
 
-The current goal is to document the set up and include a simplified example of use, to help with code checks and to make it easier to do code updates.
-Longer term, we can see if we can pull out this related set of functions into a separate R package, to make it easier to use in other modeling exercises.
-
-The code set up was introduced in <https://github.com/AlkemaLab/BayesTransitionModels>
-
 # Set up
-
-Goal: create a simplified example that uses our data set and helper functions from the package (functions in R for processing, functions and code blocks in stan to set up the stan model).
 
@@
 library(cmdstanr)
+library(truncnorm) # for inits
 library(bayesplot) # diagnostic plots
@@
 dat <- read_csv(here::here(data_folder, "coverage_data.csv"))
+dat_subnat <- read_csv(here::here(data_folder, "coverage_data_subnat.csv"))
+
 ```
@@
 ```{r}
+hierarchical_level <- c("intercept",  "subcluster", "iso")
 dat %>% select(iso, cluster, subcluster)
@@
 
--   $\gamma \sim N(0,1)$ for each gamma (one global, $R$ region, $S$ subregions, $C$ countries.
+-   $\gamma \sim N(0,1)$ for each gamma (one global, $R$ region, $S$ subregions, $C$ countries).
 
@@
 ```{r}
-fit1a <- fit_model_simplified(runstep = "step1a",
-        hierarchical_level     = c("intercept",  "subcluster", "iso"),
-
+fit1a <- fit_model(runstep = "step1a",
+                              hierarchical_level     =  hierarchical_level,
                               survey_df = dat,
@@
 ```{r}
-fit1a$post_summ <- get_posterior_summaries_simplified(fit1a) %>%
-  dplyr::rename(median = mean)
+fit1a$post_summ <- get_posterior_summaries(fit1a)
 fit1a$post_summ
@@
 ```{r}
-#fit1a$samples <- NULL
-fit_local <- fit_model_simplified(runstep = "local_national",
+fit_local <- fit_model(runstep = "local_national",
                                   global_fit = fit1a,
@@
 
+#### Produce and check outputs
+
+get summaries, takes a little while. this creates, per fit, a list with each entry being one level, and the results (mu estimates) for that level
+
 ```{r}
-#fit1a$samples$summary()$variable
+res_global <- posterior_summary_hierparam(fit = fit1a, parname = "mu")
+res_local <- posterior_summary_hierparam(fit = fit_local, parname = "mu")
+
 ```
 
-#### Produce and check outputs
+Plots that show how values have been fixed in the local runs
 
-in the simpler model used, various mus are defined in the transformed parameters block so we can check those directly.
-(Note that this not (yet) true for the transition model because the creation of parameters takes place within the transition function, see notes in last section on this).
+```{r}
+bind_rows(res_local$subcluster %>% mutate(model = "local"),
+          res_global$subcluster %>% mutate(model = "global")) %>%
+  group_by(name, model) %>%
+  reframe(y = val[quant == 0.5], ymin = val[quant == 0.025], ymax = val[quant == 0.975]) %>%
+  ggplot(aes(y = y, x = name, color = model)) +
+  geom_errorbar(aes(ymin = ymin, ymax = ymax)) +
+  geom_point()
+
+```
+
+And a plot to show the difference in estimates between global and local runs
+
+```{r}
+bind_rows(res_local$iso %>% mutate(model = "local"),
+          res_global$iso %>% mutate(model = "global")) %>%
+  group_by(name, model) %>%
+  reframe(y = val[quant == 0.5], ymin = val[quant == 0.025], ymax = val[quant == 0.975]) %>%
+  ggplot(aes(y = y, x = name, color = model)) +
+  geom_errorbar(aes(ymin = ymin, ymax = ymax)) +
+  geom_point()
+```
+
+# Summary object
+
+Going through what's happening to create the summary object
+
+Set up hierarchical structures, folow the hierdata.qmd
+
+```{r}
+area <- "iso"
+```
+
+Create district index for matching district and district index
+
+```{r}
+geo_unit_index <- get_geo_unit_index_data(dat,
+                                          hierarchical_levels = c(hierarchical_level), 
+                                          area = area)
+geo_unit_index
+```
+
+```{r}
+hier_data <- hierarchical_data(geo_unit_index, hierarchical_level)
+names(hier_data)
+hier_data
+
+```
+
+This is where global fit is used to fix values
+
+```{r}
+hier_stan_data <- hierarchical_param_stan_data(
+  global_fit = fit1a,
+  param_name ="mu",
+  param_data = hier_data,
+  hierarchical_terms_fixed = hierarchical_level[1:2], 
+  hierarchical_sigmas_fixed = hierarchical_level[1:3]
+)
+names(hier_stan_data)
+```
+
+```{r}
+samples <- extract_parameter_subhierarchical(
+        hierarchical_data = hier_data,
+        subhierarchy = "iso",
+        parname = "mu",
+        fit_samples = fit1a$samples,
+        morethan1param = FALSE
+        )
+samples
+```
+
+# Trace plots etc
 
@@
 ```
-
-Re pulling out information at non-country levels (ie estimates for regions and subregions): The orginal code set up in <https://github.com/AlkemaLab/BayesTransitionModels> includes helper functions to pull out various outputs, for example, $\mu$s at the regional or subnational level.
-We moved some over and use them in the package here.
-However, they are not yet very user friendly given limited documentation, hardcoding of prior SDs for global means, and reliance on overall fit object.
-It is work in progress to clean this up further.
-What I did here for the simplified example (and will move over to the general package later):
-
--   To deal with the set up where things are fixed vs estimated, we let the stan model spit out the mu_stars, that's what the original helper functions use as inputs.
-
--   We clean up helper functions, so that they take in the mu_stars, and specific other stan_data, and are better documented.
-See function extract_parameter_subhierarchical_simplified in fit_model_simplified R script, and its usage below.
-
-```{r}
-fit <- fit1a
-# some notes, to understand updates between this code base and original one
-# and where to find info 
-#fit1a$country_index is now fit$geo_unit
-# see fit1a$hierarchical_level for info
-# hierarchical_data(fit$geo_unit, fit$hierarchical_level)
-# note that subhierarchies are called column in hierarchical data
-mu <- list()
-for(subhierarchy in fit$hierarchical_level) {
-  mu[[subhierarchy]] <-
-    extract_parameter_subhierarchical_simplified (
-      hierarchical_data = hierarchical_data(fit$geo_unit, fit$hierarchical_level),
-      subhierarchy = subhierarchy,
-      parname = "mu",
-      fit_samples = fit$samples)
-}
-# mu[[]] has posterior samples for the subhierarchy level
-# to get the summaries
-quantile_df <- function(x, probs = c(0.025, 0.5, 0.975)) {
-  tibble(
-    val = quantile(x, probs, na.rm = TRUE),
-    quant = probs
-  )
-}
-res_global <- map(mu, function(tibble_samples) 
-  tibble_samples %>% select(name, value)  %>% reframe(quantile_df(value), .by = name))
-res_global
-
-```
-
-```{r}
-fit <- fit_local
-mu <- list()
-for(subhierarchy in fit$hierarchical_level) {
-  mu[[subhierarchy]] <-
-    extract_parameter_subhierarchical_simplified (
-      hierarchical_data = hierarchical_data(fit$geo_unit, fit$hierarchical_level),
-      subhierarchy = subhierarchy,
-      parname = "mu",
-      fit_samples = fit$samples)
-}
-res_local <- map(mu, function(tibble_samples) 
-  tibble_samples %>% select(name, value)  %>% reframe(quantile_df(value), .by = name))
-res_local
-```
-
-checks that results from local model approximate those from global model
-
-at subcluster level
-
-```{r}
-bind_rows(res_local$subcluster %>% mutate(model = "local"),
-          res_global$subcluster %>% mutate(model = "global")) %>%
-  group_by(name, model) %>% 
-  reframe(y = val[quant == 0.5], ymin = val[quant == 0.025], ymax = val[quant == 0.975]) %>%
-  ggplot(aes(y = y, x = name, color = model)) +
-  geom_errorbar(aes(ymin = ymin, ymax = ymax)) +
-  geom_point()
-
-```
-
-at country level
-
-```{r}
-bind_rows(res_local$iso %>% mutate(model = "local"),
-          res_global$iso %>% mutate(model = "global")) %>%
-  group_by(name, model) %>% 
-  reframe(y = val[quant == 0.5], ymin = val[quant == 0.025], ymax = val[quant == 0.975]) %>%
-  ggplot(aes(y = y, x = name, color = model)) +
-  geom_errorbar(aes(ymin = ymin, ymax = ymax)) +
-  geom_point()   
-```
-
-# Related notes and to dos to come back to
-
--   To do next: clean code to fix parameters
-
--   How to get the mus in full model set up: in the simpler model used here, various mus are defined in the transformed parameters block so we can check those directly.
-Note that this not (yet) true for the full transition model because the creation of parameters takes place within the process model function.
-Options to obtain the mus in the full model are to
-
-    1.  add them to generated quantities: pull out mus with an `if {add_output_mus}` type statement, re-using functions/statements introduced in the simpler file here.
-
-    2.  rewrite transformed parameter block and pull mus out of the process model function (not sure if that slows things down when not used).
-
-    3.  create them in R (would need more work to get functions going, could consider exposing stan functions)
-
-    Options 1-2 seem easiest to LA.
-
--   This example does not yet get into estimation of parameter vectors and associated code for that (as we have for spline coefficients).
-Simplest way to set that up is to estimate the same parameter $\mu$ twice.
-
--   More general, on how to organize stan functions: Stan functions lived in a separate `include` folder in the original code base.
-In the fpet2 package we ended up including stan functions in the stan model directly.
-That inclusion was needed to ship a precompiled model with the package for use in the webtool (using the `instantiate` package).
-We can move functions back to an `include` subfolder in this documentation/package.
-
--   Note on approach used to get this example together (can be followed when extending further): Start subfolder for simplified model and code, added current stan_model and fit_model function and simplify to produce just the toy example.

### 2025-04-07 documentation/hierdata.qmd
update quarto documents

@@
 
-Load relevant packages, including our own and data
+#
+
+# Demonstration of hierarchical data with toy data set.
+
+Load relevant packages
 
@@
 devtools::load_all(here::here())
-#data_folder <- "data_raw" 
-# calling survey_df data here, as internal to fit_model
-#data <- read_csv(here::here(data_folder, "coverage_data.csv"))
 ```
 
-toy data set
+Toy data set
 
@@
 data <- tibble(subcluster = c("A", "A", "B", "B"), iso = c("iso1", "iso2", "iso3", "iso3"))
+data
 ```
 
-main info: hierarchical levels to use, and the level of interest
+We need to define the hierarchical levels to use, and the level of interest.
+Here we set level of interest to be equal to lowest level (iso), this is currently required (and the default setting in all runs in fit_model).
 
-(level of interest does not need to be the lowest level, not clear how thoroughly tested)
+(This does not need to be a requirement going forward but testing so far is based on it).
 
@@
 ```{r}
-# use in fit_model:
-#hier_data <-  list()
-#hier_data[["mu_data"]] <- 
-# here we just work with one object:
 hier_data <- hierarchical_data(geo_unit_index, hierarchical_level)
-#names(hier_data[["mu_data"]])
 names(hier_data)
@@
   global_fit = NULL,
-  param_name ="mu",
-  param_data = hier_data #[["mu_data"]] #,
+  param_name = "mu",
+  param_data = hier_data #,
   # used only when fixing things 
@@
 #mu_raw refers to the vector with ALL $\gamma$s, including the global one.The ordering is global-allregional-allsubregional-country.
-mu_raw <- seq(1, hier_stan_data$mu_n_terms)
+mu_raw <- seq(1, hier_stan_data$mu_raw_n_terms)
 mu_raw
@@
 ```{r}
-stan_file <- here("inst/stan/hierfunctions_seq.stan")
+stan_file <- here("inst/stan/hierfunctions_seq_muscalar.stan")
 mod <- cmdstan_model(stan_file, force_recompile = TRUE)  
@@
 
-#mu_star refers to the vector with (mu global and ALL $\eta$s). we get mu_stars from mu_raw \# and the vector ($s_{\mu, global}$ , mu_sigma), using the stan scale_blocks function.
+mu_star refers to the vector with (mu global and ALL $\eta$s). we get mu_stars from mu_raw \# and the vector ($s_{\mu, global}$ , mu_sigma), using the stan scale_blocks function.
+
+vector scale_blocks(vector mu_raw, int mu_raw_n_terms, vector mu_sigmawpriorsd, int n_mu_sigmawpriorsd, array\[\] int re_start, array\[\] int re_end) {
 
 ```{r}
-mu_star <- scale_blocks(mu_raw, c(sprior, mu_sigma), 
-             start = hier_stan_data$mu_re_start,
-             end = hier_stan_data$mu_re_end)
+mu_star <- scale_blocks(mu_raw = mu_raw, 
+                        mu_raw_n_terms = hier_stan_data$mu_raw_n_terms, 
+                        mu_sigmawpriorsd = c(sprior, mu_sigma),
+                        n_mu_sigmawpriorsd = hier_stan_data$mu_n_sigma +1, 
+             re_start = hier_stan_data$mu_re_start,
+             re_end = hier_stan_data$mu_re_end)
 mu_star
@@
 
-Need to expose (compile) the stan math functions as well, eg for for sparse matrices. see here for info
+If we want to use all stan functions, we need to expose (compile) the stan math functions as well, eg for for sparse matrices. see here for info
 
@@
 
-first time around, i got an error
+first time around, i got an error that I tried to fix
 
@@
 
-Turns out that it may not work anyway:
+But it turns out that it may not work anyway:
 
@@
 
-looks like we just want model_matrix\*vector
-
-and can confirm if
-
-csr_matrix_times_vector
+so moving forward in R with (sparse) model_matrix\*vector
 

### 2025-04-07 R/posterior_summary_hierparam.R

@@ -6,7 +6,7 @@
 #' @param parname selected parameter name (example: "mu")
 #' @param morethan1param does paramname refer to more than 1 parameter (a vector)
 #'
-#' @returns list with summaries of mu for each hierchical level (units with each level)
+#' @returns list with summaries of mu for each hierarchical level (units with each level)
 #'  these mus are obtained by summing up all relevant etas
 #' for morethan1param, k refers to the index
 #' @export

### 2025-04-07 documentation/differentusecases.R

@@ -30,13 +30,13 @@ hierarchical_level <- c("intercept",  "subcluster", "iso")
 ### use case 1: global then local national, all countries (in quarto too)
 
 devtools::load_all(here::here())
-fit1a <- fit_model_simplified(runstep = "step1a",
+fit1a <- fit_model(runstep = "step1a",
                               hierarchical_level     =  hierarchical_level,
                               survey_df = dat,
                               chains = 4)
 fit1a$post_summ <- get_posterior_summaries(fit1a)
 fit1a$post_summ
-fit_local <- fit_model_simplified(runstep = "local_national",
+fit_local <- fit_model(runstep = "local_national",
                                   global_fit = fit1a,
                                   survey_df = dat,
                                   chains = 4)
@@ -65,12 +65,12 @@ bind_rows(res_local$iso %>% mutate(model = "local"),
 iso_select <- "BFA"
 fit1a$samples <- NULL
 devtools::load_all(here::here())
-fit_local2 <- fit_model_simplified(runstep = "local_national",
+fit_local2 <- fit_model(runstep = "local_national",
                                   global_fit = fit1a,
                                   survey_df = dat %>% filter(iso == iso_select),
                                   chains = 4)
 # equivalent is using in area_select
-fit_local2 <- fit_model_simplified(runstep = "local_national",
+fit_local2 <- fit_model(runstep = "local_national",
                                    global_fit = fit1a,
                                    area_select = c(iso_select,"NER", "PER"),
                                    survey_df = dat,
@@ -100,7 +100,7 @@ bind_rows(res_local2$iso %>% mutate(model = "local"),
 ### use case 3: global then subnational global then local subnational
 #dat_subnat
 devtools::load_all(here::here())
-fit_globalsubnat <- fit_model_simplified(runstep = "global_subnational",
+fit_globalsubnat <- fit_model(runstep = "global_subnational",
                                          area = "subnat",
                               survey_df = dat_subnat,
                               global_fit = fit1a,
@@ -110,7 +110,7 @@ fit_globalsubnat$post_summ %>%
   filter(variable_no_index == "mu_sigma")
 fit1a$post_summ %>%
   filter(variable_no_index == "mu_sigma")
-fit_local <- fit_model_simplified(runstep = "local_subnational",
+fit_local <- fit_model(runstep = "local_subnational",
                                   area = "subnat",
                                   global_fit = fit_globalsubnat,
[…]

### 2025-04-07 documentation/tmp_script.R

@@ -1,49 +1,0 @@
-devtools::load_all(here::here())
-fit1a <- fit_model_simplified(runstep = "step1a",
-                              hierarchical_level     = c("intercept",  "subcluster", "iso"),
-
-                              survey_df = dat,
-                              chains = 4)
-
-# functions to clean out
-## check_nas
-## hier data helpers
-
-
-
-# overview of function calls
-# get the stan_datat
-data <- tibble(subcluster = c("A", "A", "B", "B"), iso = c("iso1", "iso2", "iso3", "iso3"))
-hierarchical_level <- c("intercept",  "subcluster", "iso")
-area <- "iso"
-geo_unit_index <- get_geo_unit_index_data(data,
-                                          hierarchical_levels = c(hierarchical_level),
-                                          area = area)
-hier_data <- hierarchical_data(geo_unit_index, hierarchical_level)
-# here the global fit can be used
-hier_stan_data <- hierarchical_param_stan_data(
-  global_fit = NULL,
-  param_name ="mu",
-  param_data = hier_data #[["mu_data"]] #,
-  # used only when fixing things
-  #  hierarchical_terms_fixed = hierarchical_level_terms_fixed,
-  #  hierarchical_sigmas_fixed = hierarchical_level_sigmas_fixed
-)
-
-# when working with a vector of parameters,
-# eg instead of mu[i] we have mu[i, k] for k>1
-# with same hierarchical model
-# then hier_stan_data changes only if we fix things (as we need to input mu[i,k]s as opposed to mu[i]s)
-hier_stan_data <- hierarchical_param_stan_data(
-  global_fit = NULL,
-  param_name ="mu",
-  param_num_indices = 2,
-  param_data = hier_data #[["mu_data"]] #,
-  # used only when fixing things
-  #  hierarchical_terms_fixed = hierarchical_level_terms_fixed,
-  #  hierarchical_sigmas_fixed = hierarchical_level_sigmas_fixed
-)
-
-
-
-

### 2025-04-07 documentation/hierarchicalmodels_seqruns.qmd
update for use in bayescoverage

@@
 ```{r}
-fit1a <- fit_model(runstep = "step1a",
+fit1a <- fit_model_localhierarchy(runstep = "step1a",
                               hierarchical_level     =  hierarchical_level,
@@
 ```{r}
-fit_local <- fit_model(runstep = "local_national",
+fit_local <- fit_model_localhierarchy(runstep = "local_national",
                                   global_fit = fit1a,

### 2025-04-07 documentation/hierdata.qmd
update for use in bayescoverage

@@
 We need to define the hierarchical levels to use, and the level of interest.
-Here we set level of interest to be equal to lowest level (iso), this is currently required (and the default setting in all runs in fit_model).
+Here we set level of interest to be equal to lowest level (iso), this is currently required (and the default setting in all runs in fit_model_localhierarchy).
 

### 2025-04-07 R/check_nas.R

@@ -6,3 +6,4 @@ check_nas <- function(data, column) {
     stop(glue::glue("There are NAs in column {column}."))
   }
 }
+

### 2025-04-07 R/fit_model.R

@@ -30,7 +30,7 @@
 #' @returns List that contains samples, stan_data, and other information relevant to model fit (arguments)
 #'
 #' @details
-#' The `fit_model` function fits the toy example for hierarchical models/seq fitting.
+#' The `fit_model_localhierarchy` function fits the toy example for hierarchical models/seq fitting.
 #' The argument \code{runstep} determines the type of run to perform. The
 #' following run steps are supported:
 #' - "step1a": Fit the global model
@@ -72,11 +72,10 @@
 #'
 #'
 #'
-#' @export
 #'
 #'
 #'
-fit_model <- function(
+fit_model_localhierarchy <- function(
   survey_df,
   y = "logit_indicator",
   area = "iso", # add default or set in function?

### 2025-04-07 R/posterior_summary_hierparam.R

@@ -120,6 +120,15 @@ extract_parameter_subhierarchical <- function(
 }
 
 
+#' Title
+#'
+#' @param x
+#' @param probs
+#'
+#' @returns
+#' @export
+#'
+#' @examples
 quantile_df <- function(x, probs = c(0.025, 0.5, 0.975)) {
   tibble(
     val = quantile(x, probs, na.rm = TRUE),

### 2025-04-07 documentation/differentusecases.R

@@ -30,13 +30,13 @@ hierarchical_level <- c("intercept",  "subcluster", "iso")
 ### use case 1: global then local national, all countries (in quarto too)
 
 devtools::load_all(here::here())
-fit1a <- fit_model(runstep = "step1a",
+fit1a <- fit_model_localhierarchy(runstep = "step1a",
                               hierarchical_level     =  hierarchical_level,
                               survey_df = dat,
                               chains = 4)
 fit1a$post_summ <- get_posterior_summaries(fit1a)
 fit1a$post_summ
-fit_local <- fit_model(runstep = "local_national",
+fit_local <- fit_model_localhierarchy(runstep = "local_national",
                                   global_fit = fit1a,
                                   survey_df = dat,
                                   chains = 4)
@@ -65,12 +65,12 @@ bind_rows(res_local$iso %>% mutate(model = "local"),
 iso_select <- "BFA"
 fit1a$samples <- NULL
 devtools::load_all(here::here())
-fit_local2 <- fit_model(runstep = "local_national",
+fit_local2 <- fit_model_localhierarchy(runstep = "local_national",
                                   global_fit = fit1a,
                                   survey_df = dat %>% filter(iso == iso_select),
                                   chains = 4)
 # equivalent is using in area_select
-fit_local2 <- fit_model(runstep = "local_national",
+fit_local2 <- fit_model_localhierarchy(runstep = "local_national",
                                    global_fit = fit1a,
                                    area_select = c(iso_select,"NER", "PER"),
                                    survey_df = dat,
@@ -100,7 +100,7 @@ bind_rows(res_local2$iso %>% mutate(model = "local"),
 ### use case 3: global then subnational global then local subnational
 #dat_subnat
 devtools::load_all(here::here())
-fit_globalsubnat <- fit_model(runstep = "global_subnational",
+fit_globalsubnat <- fit_model_localhierarchy(runstep = "global_subnational",
                                          area = "subnat",
                               survey_df = dat_subnat,
                               global_fit = fit1a,
@@ -110,7 +110,7 @@ fit_globalsubnat$post_summ %>%
   filter(variable_no_index == "mu_sigma")
 fit1a$post_summ %>%
   filter(variable_no_index == "mu_sigma")
-fit_local <- fit_model(runstep = "local_subnational",
+fit_local <- fit_model_localhierarchy(runstep = "local_subnational",
                                   area = "subnat",
[…]

### 2025-06-25 R/check_nas.R
update documentation

@@ -1,5 +1,14 @@
 
-# used in bayescoverage
+#' Check for missing values
+#'
+#' This function checks if there are any NAs in a specified column of a data frame.
+#'
+#' @param data tibble or data frame
+#' @param column column to check for NAs
+#'
+#' @returns stops if there are NAs in the column or if the column does not exist
+#' @export
+#'
 check_nas <- function(data, column) {
   if(!(column %in% names(data))) {
     stop(glue::glue("Could not find column {column}."))

### 2025-06-25 R/get_geo_unit_index_data.R

@@ -1,6 +1,8 @@
 
 
-#' get_geo_unit_index_data
+#' Get geo_unit_index_data
+#'
+#' Get unique hierarchical levels from survey data and assign index "c" based on area.
 #'
 #' @param data survey data of interest that contains the hierarchical_column_names
 #' @param hierarchical_levels vector that contains one of more specs of hierarchical_level (incl intercept)
@@ -10,7 +12,8 @@
 #' @export
 #'
 #' @examples
-#' survey_dat <- tibble(subcluster = c("A", "A", "B", "B"), iso = c("iso1", "iso2", "iso3", "iso3"))
+#' survey_dat <- tibble(subcluster = c("A", "A", "B", "B"),
+#'     iso = c("iso1", "iso2", "iso3", "iso4"))
 #' get_geo_unit_index_data(survey_dat,
 #'         hierarchical_levels = c("intercept", "subcluster", "iso"),
 #'         area = "iso")

### 2025-06-25 R/get_posterior_summaries.R

@@ -4,6 +4,8 @@
 
 #' get_posterior_summaries
 #'
+#' Get posterior summaries for a set of parameters
+#'
 #' @param fit fit object to summarize
 #' @param params vector of parnames to summarize (w/o index)
 #'
@@ -41,6 +43,8 @@ get_posterior_summaries <- function(
 
 #' get_posterior_summary_one_param
 #'
+#' Get posterior summary for one parameter
+#'
 #' @param fit fit object
 #' @param param_name parameter name w/o index
 #'
@@ -52,7 +56,6 @@ get_posterior_summaries <- function(
 #' @returns tibble with variable (parname with index used in model) and postmean (posterior mean)
 #' @export
 #'
-#' @examples
 get_posterior_summary_one_param <- function(
     fit,
     param_name) {

### 2025-06-25 R/hierarchical_data.R

@@ -1,7 +1,7 @@
 
 #' Hierarchical_data
 #'
-#' This is my description!
+#' Create hierarchical data structure for a hierarchical model.
 #'
 #' @param geo_unit_index_data tibble with unique hierarchical levels (one column for each level)
 #' @param hierarchical_level vector with hierarchical spec

### 2025-06-25 R/hierarchical_model_matrix.R

@@ -1,8 +1,9 @@
 
 
 #' hierarchical_model_matrix
+#' Create list with information including a model matrix, associated with the geo_unit_data and hierarchical_levels
 #'
-#' @param geo_unit_index_data tibble with unique hierarchical levels (one column for each level)
+#' @param geo_unit_index_data tibble with unique hierarchical levels (one column for each level, one row per combi)
 #' @param hierarchical_level vector with hierarchical spec from highest to lowest levels
 #'
 #' @returns list (modelmatrix, assign, index) associated with the geo_unit_data and hierarchical_levels
@@ -11,7 +12,6 @@
 #' index is a tibble with n_eta rows, with info for each eta in and columns i [index of eta], column [hierarchical level], and level [name of hierarchical level]
 #' @export
 #'
-#' @examples
 hierarchical_model_matrix <- function(geo_unit_index_data, hierarchical_level) {
   mat <- matrix(NA, nrow = nrow(geo_unit_index_data), ncol = 0)
   assign <- c()

### 2025-06-25 R/inits.R

@@ -1,13 +1,14 @@
 
-# warnings can occur like pasted below
-# Chain XXX Informational Message: The current Metropolis proposal is about to be rejected because of the following issue:
-# Chain XXX Exception: normal_lpdf: Location parameter is inf, but must be finite!
-# ie mu is not finite..
-# meaning inf values are drawn for sum of mustars, so product (mu_raw* mu_sigma)
-# by setting inits, we avoid this for the initial values
-# HOWEVER they can still occur early on in the chain (not sure if adding a cap is good thing to do)
-
-
+#' Set initial values
+#'
+#' This function sets the initial values for the parameters in a simple hierarchical model.
+#'
+#' @param chain_id the chain ID, used for setting the seed
+#' @param stan_data a list containing the Stan data, to obtain length of parameter vectors
+#'
+#' @returns a list of initial values for the model parameters
+#' @export
+#'
 init_fun_localhierarchy <- function(chain_id, stan_data){
   set.seed(chain_id)
   inits <- list()

### 2025-06-25 R/mu_raw.R

@@ -1,14 +1,16 @@
 
-#' Title
+#' Get mu_raw_labeled
 #'
-#' @param fit needs to include parname_raw_estimate
-#' @param parname selected parameter name (example: "mu")
-#' @param morethan1param does paramname refer to more than 1 parameter (a vector)
+#' This function extracts the mu_raw_estimate samples from a fitted model and
+#'  labels them with their hierarchical level.
+#'
+#' @param fit needs to include parname_raw_estimate (example: mu_raw)
+#' @param parname selected parameter name (example: mu)
+#' @param morethan1param does parname refer to more than 1 parameter (a vector)
 #'
 #' @returns tibble with samples of parname_raw_estimate, including info on hierarchical level
 #' @export
 #'
-#' @examples
 get_mu_raw_labeled <- function(fit, parname, morethan1param = FALSE){
   # mu_raw_estimate is saved directly so getting those is not difficult
   # the only challenge is to figure out what hier level they refer to

### 2025-06-25 R/posterior_summary_hierparam.R

@@ -2,6 +2,8 @@
 
 #' posterior_summary_hierparam
 #'
+#' Calculate posterior summaries for hierarchical parameters
+#'
 #' @param fit needs to include parname_star
 #' @param parname selected parameter name (example: "mu")
 #' @param morethan1param does paramname refer to more than 1 parameter (a vector)
@@ -9,10 +11,9 @@
 #'
 #' @returns list with summaries of mu for each hierarchical level (units with each level)
 #'  these mus are obtained by summing up all relevant etas
-#' for morethan1param, k refers to the index
+#' for morethan1param, each level has a list where k refers to the index
 #' @export
 #'
-#' @examples
 posterior_summary_hierparam <- function(fit, parname, morethan1param = FALSE,
                                         hierarchical_levels = fit$hierarchical_level){
   print("Calculating posterior summary for hierarchical parameters")
@@ -77,7 +78,6 @@ posterior_summary_hierparam <- function(fit, parname, morethan1param = FALSE,
 #' and `k` referring to parameter index if morethan1param = TRUE
 #' @export
 #'
-#' @examples
 extract_parameter_subhierarchical <- function(
     hierarchical_data,
     subhierarchy,
@@ -131,13 +131,14 @@ extract_parameter_subhierarchical <- function(
 
 #' Compute quantiles
 #'
-#' @param x
-#' @param probs
+#' This function computes quantiles for a given vector and returns them in a tibble format.
 #'
-#' @returns
+#' @param x vector with values
+#' @param probs vector of quantiles
+#'
+#' @returns tibble with the requested quantiles
 #' @export
 #'
-#' @examples
 quantile_df <- function(x, probs = c(0.025, 0.5, 0.975)) {
   tibble(
     val = quantile(x, probs, na.rm = TRUE),

### 2025-06-25 R/prior_post.R

@@ -1,11 +1,13 @@
 
 
-#' Plot estimates of mu raws
+#' Plot prior and posterior densities of mu_raw
 #'
-#' @param fit needs to include parname_raw_estimate and stan_data
-#' @param parname selected parameter name (example: "mu")
-#' @param morethan1param does paramname refer to more than 1 parameter (a vector)
-#' @param nresultsperpage number of results per page in summary plots
+#' This function plots the posterior densities of mu_raw parameters, with prior added.
+#'
+#' @param fit List that includes "parname"_raw_estimate and stan_data
+#' @param parname Selected parameter name (example: mu)
+#' @param morethan1param Logical, does parname refer to more than 1 parameter (a vector)
+#' @param nresultsperpage Number of results per page in summary plots
 #'
 #' @returns lists with list 'summary_plots' and list 'plots_allmuraw'.
 #' summary plots gives summary CIs, nresultsperpage at a time.
@@ -13,7 +15,6 @@
 #' If morethan1param = TRUE, then each list contains a list per parameter k
 #' #' @export
 #'
-#' @examples
 plot_mu_raw <- function(fit, parname, morethan1param = FALSE,
                         nresultsperpage = 30){
   helper_get_plots_allmuraw <- function(samples_tibble){
@@ -77,15 +78,16 @@ plot_mu_raw <- function(fit, parname, morethan1param = FALSE,
               summary_plots = summary_plots))
 }
 
-#' plot_prior_post_sigmas
+#' Plot prior and posteriors of sigmas from hierarchical models
 #'
-#' @param fit needs to include parname_sigma_estimate and stan_data
-#' @param parname selected parameter name (example: "mu")
+#' This function plots the prior and posterior densities of sigma_estimate parameters,
 #'
-#' @returns plot with density of sigma_estimate and prior added
+#' @param fit List, needs to include parname_sigma_estimate and stan_data
+#' @param parname Selected parameter name (example: "mu")
+#'
+#' @returns Plot with density of sigma_estimate and prior added
 #' @export
 #'
-#' @examples
 plot_prior_post_sigmas <- function(fit, parname){
   # sigmas, fine to look at all k's combined
   samp <- fit$samples$draws(paste0(parname, "_sigma_estimate"))

### 2025-06-25 documentation/example.qmd
more documentation, clean up example

--- 2025-04-07 documentation/hierarchicalmodels_seqruns.qmd
@@
 ```{r}
-fit1a <- fit_model(runstep = "step1a",
+fit_global <- fit_model_localhierarchy(runstep = "global_national",
                               hierarchical_level     =  hierarchical_level,
@@
 ```{r}
-fit1a$post_summ <- get_posterior_summaries(fit1a)
-fit1a$post_summ
+fit_global$post_summ <- get_posterior_summaries(fit_global)
+fit_global$post_summ
 
@@
 ```{r}
-fit_local <- fit_model(runstep = "local_national",
-                                  global_fit = fit1a,
+fit_local <- fit_model_localhierarchy(runstep = "local_national",
+                                  global_fit = fit_global,
                                   survey_df = dat,
@@
 ```{r}
-res_global <- posterior_summary_hierparam(fit = fit1a, parname = "mu")
+res_global <- posterior_summary_hierparam(fit = fit_global, parname = "mu")
 res_local <- posterior_summary_hierparam(fit = fit_local, parname = "mu")
@@
 
-Plots that show how values have been fixed in the local runs
+Plots that show outputs
 
 ```{r}
-bind_rows(res_local$subcluster %>% mutate(model = "local"),
-          res_global$subcluster %>% mutate(model = "global")) %>%
-  group_by(name, model) %>%
-  reframe(y = val[quant == 0.5], ymin = val[quant == 0.025], ymax = val[quant == 0.975]) %>%
-  ggplot(aes(y = y, x = name, color = model)) +
-  geom_errorbar(aes(ymin = ymin, ymax = ymax)) +
-  geom_point()
+plot_posterior_summaries(res = res_global, modelname1 = "global")
+```
+
+how values have been fixed in the local runs
+
+```{r}
+plot_posterior_summaries(res = res_local, 
+                         res2 = res_global,
+                         modelname2 = "global", modelname1 = "local")
 
@@
 
-And a plot to show the difference in estimates between global and local runs
+selecting specific levels
 
 ```{r}
-bind_rows(res_local$iso %>% mutate(model = "local"),
-          res_global$iso %>% mutate(model = "global")) %>%
-  group_by(name, model) %>%
-  reframe(y = val[quant == 0.5], ymin = val[quant == 0.025], ymax = val[quant == 0.975]) %>%
-  ggplot(aes(y = y, x = name, color = model)) +
-  geom_errorbar(aes(ymin = ymin, ymax = ymax)) +
-  geom_point()
+plot_posterior_summaries(res = res_local, 
+                         res2 = res_global,
+                         modelname2 = "global", modelname1 = "local",
+                        hierarchy_select = "subcluster")
+
+```
+
+Level and selected areas
+
+```{r}
+plot_posterior_summaries(res = res_local, 
+                         res2 = res_global,
+                         modelname2 = "global", modelname1 = "local",
+                        hierarchy_select = "iso",
+                        areas_select = res_local$iso$name[1:100])
+
 ```
@@
 hier_stan_data <- hierarchical_param_stan_data(
-  global_fit = fit1a,
-  param_name ="mu",
+  global_fit = fit_global,
+  param_name = "mu",
   param_data = hier_data,
@@
 
+# Prior-posts
+
+sigmas
+
 ```{r}
-samples <- extract_parameter_subhierarchical(
-        hierarchical_data = hier_data,
-        subhierarchy = "iso",
-        parname = "mu",
-        fit_samples = fit1a$samples,
-        morethan1param = FALSE
-        )
-samples
+plot_prior_post_sigmas(fit = fit_globalsubnat, parname = "mu")
+
 ```
 
-# Trace plots etc
+create plots for mu_raws
 
-here just checking the sigmas and first mu:
+```{r}
+plots <- plot_mu_raw(fit = fit_global, parname = "mu")
+# a list with plots
+```
+
+summary plots gives summaries, 30 at a time
+
+```{r}
+plots[["summary_plots"]][[1]]
+
+```
+
+specific prior-post plots for one parameter are in plots\[\["plots_allmuraw"\]\]
+
+```{r}
+plots[["plots_allmuraw"]][[1]]
+```
+
+Simple trace plots etc
 
@@
 
-fit <- fit1a
+fit <- fit_global
  parname_list <- list( c("mu_raw_estimate[1]",

### 2025-06-25 documentation/hierdata.qmd
more documentation, clean up example

@@
 ```{r}
-hier_stan_data$mu_n_terms # number of etas + global mu
+hier_stan_data$mu_raw_n_terms # number of etas + global mu
 length(unique(geo_unit_index$subcluster))
@@
 
-# Unsuccessful investigation into exposing stan build-in functions
+# Notes on unsuccessful investigation into exposing stan build-in functions
 

### 2025-06-25 R/fit_model.R

@@ -2,19 +2,21 @@
 
 
 
-#' Fit model
+#' Fit model for localhierarchy model
+#'
+#' This function fits the localhierarchy model
 #'
 #' @param survey_df tibble with survey data
 #' @param y column name of outcome.
-#' @param area column name of the area of each observation (geounit; iso or subnational region)
+#' @param area column name of the area of each observation (used as geounit; iso or subnational region)
 #' @param area_select area name to use for local run (eg iso code or subnat region name)
 #'
 #' @param runstep type of run, currently one of XXX (see Details).
-#' @param global_fit optional object of class XX, used to obtain fixed
+#' @param global_fit optional global fit object, used to obtain fixed
 #'   values to use for some parameters in the current fit (see Details).
 #' @param hierarchical_level vector specifying hierarchical structure used for mu
 #' @param add_subnational_hierarchy level that's added to the hierarchy for subnational, defaults to 'subnat'
-#' @param mu_isvector Boolean, TRUE if mu is a vector, defaults to FALSE
+#' @param mu_isvector Logical, TRUE if mu is a vector, defaults to FALSE
 #'
 #' Settings for sampling
 #' @param chains number of chains to run
@@ -33,10 +35,10 @@
 #' The `fit_model_localhierarchy` function fits the toy example for hierarchical models/seq fitting.
 #' The argument \code{runstep} determines the type of run to perform. The
 #' following run steps are supported:
-#' - "step1a": Fit the global model
-#' - "local_national": Fit the model to data from a single country, using a 1a fit.
-#' - "global_subnational": Fit the model to global database with subnational data, using a 1a fit.
-#' - "local_subnational": Fit the model to subnational data from a single country or region, using a global_subnational fit.
+#' * "global_national": Fit the global model.
+#' * "local_national": Fit the model to data from a single country, using a global_national fit.
+#' * "global_subnational": Fit the model to global database with subnational data, using a global_national fit.
+#' * "local_subnational": Fit the model to subnational data from a single country or region, using a global_subnational fit.
 #' This is also explained in the documentation folder.
 #'
 #' Details on hierarchical set ups used
@@ -102,11 +104,10 @@ fit_model_localhierarchy <- function(
     stop("No hierarchical structure supplied. See the hierarchical_level argument.")
   }
 
[…]

### 2025-06-25 R/hierarchical_param_stan_data.R

@@ -1,4 +1,6 @@
 
+#' Stan data for hierarchical parameters
+#'
 #' Set up data for Stan related to hierarchical parameters
 #'
 #' @param param_name The name of the parameter we are working with, e.g. "mu"
@@ -79,8 +81,6 @@ hierarchical_param_stan_data <- function(param_name, param_data,
     result[[paste0(param_name, "_raw_n_terms_estimate")]] <-
       param_data$n_terms - n_terms_fixed
 
-
-
     # get indices in global for local vector of fixed parameter values
     # in the local model, the fixed etas are before the to-be-estimated ones
     # (this is correct when sticking to an ordering of levels from higher order to lower order
@@ -133,9 +133,9 @@ hierarchical_param_stan_data <- function(param_name, param_data,
       result[[paste0(param_name, "_sigma_fixed")]] <-
         create_a_matrix(globalfit_param_post_summ_sigma[, c("variable", "postmean")])[seq_len(n_sigma_fixed), , drop=FALSE]
     }
-    print("We are fixing the following parameters:")
-    print(result[[paste0(param_name, "_raw_fixed")]])
-    print(result[[paste0(param_name, "_sigma_fixed")]])
+    #print("We are fixing the following parameters:")
+    #print(result[[paste0(param_name, "_raw_fixed")]])
+    #print(result[[paste0(param_name, "_sigma_fixed")]])
     return(result)
   }
 }

### 2025-06-25 R/plot_posterior_summaries.R

@@ -5,8 +5,8 @@
 #'
 #' @param res output from posterior_summary_hierparam
 #' @param hierarchy_select optional, what hierarchical level to show? if NULL, all levels are shown
-#' @param areas_select optional: areas to filter by, allowed only if one hierarchical level is selected
-#' @param res2 optional, output from posterior_summary_hierparam
+#' @param areas_select optional: specific areas in a level to filter by, allowed only if one hierarchical level is selected
+#' @param res2 optional, output from posterior_summary_hierparam for comparison
 #' @param modelname1 label for res
 #' @param modelname2 label for res2
 #' @param k_select optional, if res contains k, which k values to show? if NULL, all k values are shown

### 2025-06-25 documentation/differentusecases.R

@@ -27,7 +27,7 @@ hierarchical_level <- c("intercept",  "subcluster", "iso")
 
 ### use case 1: global then local national, all countries (in quarto too)
 
-fit1a <- fit_model_localhierarchy(runstep = "step1a",
+fit1a <- fit_model_localhierarchy(runstep = "global_national",
                               hierarchical_level     =  hierarchical_level,
                               survey_df = dat,
                               chains = 4)

### 2025-06-25 documentation/usecase_morethan1param.R

@@ -26,7 +26,7 @@ hierarchical_level <- c("intercept",  "subcluster", "iso")
 # use case 4 needed? invariance to re-ordering of input data in use case 3
 
 
-fit1a_mult <- fit_model_localhierarchy(runstep = "step1a",
+fit1a_mult <- fit_model_localhierarchy(runstep = "global_national",
                                        mu_isvector = TRUE,
                                        hierarchical_level     =  hierarchical_level,
                                        survey_df = dat,

### 2025-06-25 documentation/example.qmd
add subnational to example quarto

@@
 ```
+# Subnational estimation
+
+## Equations 
+We extend the model with an additional level of hierarchy, here subnational regions. The simplest set up is as follows:
+
+-   Parameter to estimate = $\mu_r^{(subnat)}$ = subnational-region-specific coverage
+
+-   We assume that $\mu_r^{(subnat)} =  \mu_{c[r]} + \eta_{r}^{(subnat)}$,  with $\eta_r^{(subnat)} \sim N(0, \sigma_{\mu, across-subnat}^2)$, where $\sigma_{\mu, across-subnat}^2$ refers to across-subnational-region variance.
+The subnational $\eta$ can be parametrized as before, i.e.
+$\eta = \sigma\cdot \gamma$ where $\gamma \sim N(0,1)$.
+
+- We can consider a data model for data observed at the subnational level as follows: $y_i^{(region)} \sim N(\mu_{r[i]}^{(subnat)}, \sigma_y^{2(subnat)})$.
+Here $\sigma_y^{2(subnat)}$ is the subnational error variance which is likely to be larger than the national-level variance due to larger sampling errors at the subnational level.
+(Note that in real examples, sampling error is calculated for an observation and included in the variance term, to capture differences between national and subnational level observations).
+
+- Data on national level can be used as well, by aggregation subnational estimates. This is left out of this package but can be found in fpet2025 and BayesCoverage.
+
+## Model fitting
+In model fitting, we can fit a global subnational model using subnational data from different countries and without fixing any parameters.
+Alternatively, we can fix parameters based on fits to national level data.
+This second approach is helpful when subnational level data are more limited as compared to national level data.
+
+We illustrate the sequential approach where subnational estimates are obtained using information from a global national fit. The model that is fitted is as follows:
+
+- We set $\mu_r^{(subnat)} =  \mu_{c[r]} + \eta_{r}^{(subnat)}$, and set $\mu_{c[r]} = \hat{\mu}_{c[r]}, the posterior mean from a fit to national level data.
+(The model could also be fitted using fixed values up to one level above the country level, to introduce a hierarchical model for the subnational parameters).
+
+- In this simplified example, we set $\sigma_y^{2(subnat)} = \sigma_y^2$.
+
+We fit a global subnational model with these settings 
+```{r}
+fit_globalsubnat <- fit_model_localhierarchy(runstep = "global_subnational",
+                             area = "subnat",
+
+                              survey_df = dat_subnat,
+                              global_fit = fit_global,
+                              chains = 4)
+
+```
+
+Here only one new sigma is estimated:
+```{r}
+plot_prior_post_sigmas(fit = fit_globalsubnat, parname = "mu")
+
+```
+
+We can obtain the posterior summaries 
+```{r}
+fit_globalsubnat$post_summ <- get_posterior_summaries(fit_globalsubnat)
+
+```
+
+Note that values of sigmas for higher levels are the same in the global national and subnational fit.
+However, mu_raws may be different if different countries were included or ordered differently.
+(this is documented inand reproducable with fit_global$geo_unit_index )
+```{r}
+fit_globalsubnat$post_summ %>%
+  filter(variable_no_index == "mu_sigma")
+fit1a$post_summ %>%
+  filter(variable_no_index == "mu_sigma")
+```
+
+fitting a local subnational model 
+(note that if the data are from a new country, and national mu's were fixed, this approach does not work. See xxx for a fix)
+```{r}
+fit_local <- fit_model_localhierarchy(runstep = "local_subnational",
+                                  area = "subnat",
+                                  global_fit = fit_globalsubnat,
+                                  survey_df = dat_subnat,
+                                  chains = 4)
+```
+
+checking summaries 
+```{r}
+res_globalsubnat <- posterior_summary_hierparam(fit = fit_globalsubnat, parname = "mu")
+res_localsubnat <- posterior_summary_hierparam(fit = fit_local, parname = "mu")
+
+```
+
+comparison plots, local subnational and various other fits  
+```{r}
+plot_posterior_summaries(res = res_localsubnat, res2 = res_global,
+                              modelname2 = "global national", modelname1 = "local",
+                         hierarchy_select = "subcluster")
+```
+
+The fixed country estimates that were used: 
+```{r}
+plot_posterior_summaries(res = res_localsubnat, 
+                         res2 = res_global,
+                         modelname2 = "global national", modelname1 = "local",
+                         hierarchy_select = "iso")
+```

### 2025-07-22 documentation/example.qmd
continue cleaning example quarto

@@
 library(bayesplot) # diagnostic plots
+library(posterior) # else error for as_draws_list
 devtools::load_all(here::here())
@@
 names(hier_data)
-hier_data
+# hier_data # not printing, model matrix is too large, see toy example instead
 
@@
 ```{r}
-plot_prior_post_sigmas(fit = fit_globalsubnat, parname = "mu")
+plot_prior_post_sigmas(fit = fit_global, parname = "mu")
 
@@
 However, mu_raws may be different if different countries were included or ordered differently.
-(this is documented inand reproducable with fit_global$geo_unit_index )
+(this is documented in and reproducable with fit_global$geo_unit_index )
 ```{r}
@@
   filter(variable_no_index == "mu_sigma")
-fit1a$post_summ %>%
+fit_global$post_summ %>%
   filter(variable_no_index == "mu_sigma")

### 2025-07-22 R/posterior_summary_hierparam.R

@@ -55,10 +55,10 @@ posterior_summary_hierparam <- function(fit, parname, morethan1param = FALSE,
   }
   if (!morethan1param){
     res <- map(mu, function(tibble_samples)
-      tibble_samples %>% select(name, value)  %>% reframe(quantile_df(value), .by = name))
+      tibble_samples %>% select(name, value)  %>% reframe(summary_df(value), .by = name))
   } else {
     res <- map(mu, function(tibble_samples)
-      tibble_samples %>% select(name, value, k)  %>% reframe(quantile_df(value), .by = c(name, k)))
+      tibble_samples %>% select(name, value, k)  %>% reframe(summary_df(value), .by = c(name, k)))
   }
   return(res)
 }
@@ -143,5 +143,24 @@ quantile_df <- function(x, probs = c(0.025, 0.5, 0.975)) {
   tibble(
     val = quantile(x, probs, na.rm = TRUE),
     quant = probs
+  )
+}
+
+#' Compute summaries
+#'
+#' This function computes posterior summaries for a given vector and returns them in a tibble format.
+#'
+#' @param x vector with values
+#'
+#' @returns tibble with 2.5th quantile, posterior mean, and 97.5th quantile, and quant = c(0.025, 0.5, 0.975)
+#'
+#' @export
+#'
+summary_df <- function(x) {
+  tibble(
+    val = c(quantile(x, 0.025, na.rm = TRUE),
+            mean(x, na.rm = TRUE),
+            quantile(x, 0.975, na.rm = TRUE)),
+    quant = c(0.025, 0.5, 0.975)
   )
 }

### 2025-08-22 R/fit_model.R
cleaning

@@ -73,9 +73,7 @@
 #'
 #' @importFrom cmdstanr cmdstan_model write_stan_file
 #' @importFrom tibble tibble
-#' @importFrom splines bs
 #' @import dplyr
-#' @importFrom readr read_file
 #' @importFrom stringr str_replace_all
 #'
 #'
@@ -96,7 +94,7 @@ fit_model_localhierarchy <- function(
   mu_isvector = FALSE, # TRUE if mu is a vector
   # settings for sampling
   chains = 4,
-  iter_sampling = 200,
+  iter_sampling = 300,
   iter_warmup = 150,
   compile_model  = TRUE, force_recompile = FALSE,
   seed = 1234,
@@ -143,17 +141,17 @@ fit_model_localhierarchy <- function(
       print("We fix data model parameters.")
       fix_nonse <- TRUE
       if (runstep %in% c("global_subnational") | use_globalsubnat_fromnat) {
-        print("For subnational global run, or in a local subnat run with a global fit from a national run, we add a level for subnational hierarchical settings ")
+        print("We add a level for subnational hierarchical settings ")
         hierarchical_level <- c(hierarchical_level, add_subnational_hierarchy)
       }
       if (runstep %in% c("global_subnational")){
-        print("For hierarchical terms, we fix things up to the 2nd-lowest or 3rd level (here 2nd is used).")
-        hierarchical_level_terms_fixed = hierarchical_level[1:(length(hierarchical_level)-1)]
+        print("For hierarchical terms, we fix things up to the 2nd-lowest level (here 3rd is used).")
+        hierarchical_level_terms_fixed = hierarchical_level[1:(length(hierarchical_level)-2)]
         print("For sigma terms, we fix up to 2nd-lowest level.")
         hierarchical_level_sigmas_fixed = hierarchical_level[1:(length(hierarchical_level)-1)]
       } else {
-        print("For hierarchical terms, we fix things up to the 2nd-lowest level.")
-        hierarchical_level_terms_fixed = hierarchical_level[1:(length(hierarchical_level)-1)]
+        print("For hierarchical terms, we fix things up to the 2nd-lowest level (here 3rd is used).")
+        hierarchical_level_terms_fixed = hierarchical_level[1:(length(hierarchical_level)-2)]
         print("For sigma terms, we fix up to lowest level.")
         hierarchical_level_sigmas_fixed = hierarchical_level[1:(length(hierarchical_level))]
       }

### 2025-08-22 R/simulate.R

@@ -1,4 +1,4 @@
-#' Simulate nested multilevel data and etas
+#' Simulate multilevel data
 #'
 #' Simulates a nested multilevel data structure with etas at each level.
 #' Lowest level can be used as observational error.
@@ -7,17 +7,26 @@
 #' @param n_units_perlevel A vector specifying the number of units at each level (default: rep(25, n_levels)).
 #' @param sigma_perlevel A vector specifying the standard deviation of etas at each level (default: rep(1, n_levels)).
 #' @param mu_global Global mean for the outcome variable (default: 0).
+#' @param add_data Logical indicating whether to add observations (default: TRUE). If TRUE, adds column y.
+#' @param n_data Number of observations to add at the lowest level (default: 10).
+#' @param sigma_y Standard deviation of the added observations (default: 0.05)
+#' @param add_data1levelup Logical indicating whether to add observations one level up
+#' e.g. at national level if lowest level is subnational (default: FALSE).
 #' @param seed Random seed for reproducibility (default: 1234).
 #'
-#' @returns A tibble with columns for each level, etas, and the outcome variable.
+#' @returns A tibble with columns for each level, mu_global, etas, and their total labeled y
 #' @export
 #'
-#' @examples
-simulate_etas_nested_multilevel_data <- function(n_levels = 3,
-                                            n_units_perlevel = rep(25, n_levels),
-                                            sigma_perlevel = rep(1, n_levels),
-                                            mu_global = 0,
-                                            seed = 1234) {
+simulate_multilevel_data <- function(n_levels = 3,
+                                      n_units_perlevel = rep(25, n_levels),
+                                      sigma_perlevel = rep(1, n_levels),
+                                      mu_global = 0,
+                                      add_data = TRUE,
+                                      n_data = 10,
+                                      sigma_y = 0.05,
+                                      add_data1levelup = FALSE,
+                                      seed = 12345
+                                     ) {
   set.seed(seed)
 
   if (length(n_units_perlevel) != n_levels || length(sigma_perlevel) != n_levels) {
[…]

### 2025-08-22 documentation/differentusecases.R

@@ -1,234 +1,0 @@
-
-# different use cases
-library(here)
-library(tidyverse)
-library(ggplot2)
-library(haven)
-library(stringr)
-library(cmdstanr)
-library(truncnorm) # for inits
-library(bayesplot) # diagnostic plots
-library(posterior)
-devtools::load_all(here::here())
-data_folder <- "data_raw"
-
-# national data
-dat <- read_csv(here::here(data_folder, "coverage_data.csv"))
-dat_subnat <- read_csv(here::here(data_folder, "coverage_data_subnat.csv"))
-hierarchical_level <- c("intercept",  "subcluster", "iso")
-
-### overview
-# use case 1: global then local national, all countries (in quarto too)
-# use case 2: global then local national, 1 country
-# use case 3: global then subnational global then local subnational
-### extended to also be able to produce subnational estimates for countries that were
-# included in global national run but not in global subnational run
-# use case 4 needed? invariance to re-ordering of input data in use case 3
-
-### use case 1: global then local national, all countries (in quarto too)
-
-fit1a <- fit_model_localhierarchy(runstep = "global_national",
-                              hierarchical_level     =  hierarchical_level,
-                              survey_df = dat,
-                              chains = 4)
-fit1a$post_summ <- get_posterior_summaries(fit1a)
-fit1a$post_summ
-fit_local <- fit_model_localhierarchy(runstep = "local_national",
-                                  global_fit = fit1a,
-                                  survey_df = dat,
-                                  chains = 4)
-
-# check the mu_raws
-devtools::load_all(here::here())
-plots <- plot_mu_raw(fit = fit1a, parname = "mu")
-# a list with plots
-# summary plots gives summaries, 30 at a time
-plots[["summary_plots"]][[1]]
-# specific plots for one parameter are in plots[["plots_allmuraw"]]
-plots[["plots_allmuraw"]][[1]]
-plots[["plots_allmuraw"]][grepl("intercept", names(plots[["plots_allmuraw"]]))]
-plots[["plots_allmuraw"]][grepl("subcluster", names(plots[["plots_allmuraw"]]))][[1]]
-
-# plot the sigmas
-plot_prior_post_sigmas(fit = fit1a, parname = "mu")
-
-plots <- plot_mu_raw(fit = fit_local, parname = "mu")
-plots[[1]][[1]]
-plots[[2]][[1]]
-
-# the total mus
[…]

### 2025-08-22 documentation/simulate_nested.R

@@ -1,105 +1,0 @@
-
-library(here)
-library(tidyverse)
-library(ggplot2)
-library(haven)
-library(stringr)
-library(cmdstanr)
-library(truncnorm) # for inits
-library(bayesplot) # diagnostic plots
-library(posterior) # else error for as_draws_list
-devtools::load_all(here::here())
-
-sim_data <- simulate_etas_nested_multilevel_data(
-  n_levels = 3,
-  n_units_perlevel = c(30, 20, 10),
-  sigma_perlevel = c(1, 0.5, 0.3)
-)
-sim_data
-table(sim_data$level1)         # 30 subcluster
-table(sim_data$level2)         # 20 countries
-table(sim_data$level3)         # 10 observations in each country
-dim(sim_data)                  # 6000 rows, 5 columns (3 levels + eta + y)
-sim_data %>% filter(level2 %in% c("1_1", "1_2"))
-
-hierarchical_level <- c("intercept",  "level1", "level2")
-sim_data %>%
-  group_by(level2) %>%
-  group_by(level1) %>%
-  summarize(nlevel1 = length(unique(level1)), ncountries = length(unique(level2)))
-
-
-area <- "level2"
-#Create district index for matching district and district index
-geo_unit_index <- get_geo_unit_index_data(sim_data,
-                                          hierarchical_levels = c(hierarchical_level),
-                                          area = area)
-geo_unit_index
-hier_data <- hierarchical_data(geo_unit_index, hierarchical_level)
-names(hier_data)
-# what data entries does it expect?
-dat <- sim_data %>%
-  dplyr::rename(y = total_effect)
-dat
-
-fit_global <- fit_model_localhierarchy(runstep = "global_national",
-                                       hierarchical_level     =  hierarchical_level,
-                                       area = area,
-                                       survey_df = dat,
-                                       y = "y",
-                                       chains = 4)
-
-#create summary object
-fit_global$post_summ <- get_posterior_summaries(fit_global)
-fit_global$post_summ
-
-#do local fit... here we fix things but do get results for all countries
-hier_stan_data <- hierarchical_param_stan_data(
-  global_fit = fit_global,
-  param_name = "mu",
-  param_data = hier_data,
-  hierarchical_terms_fixed = hierarchical_level[1:2],
-  hierarchical_sigmas_fixed = hierarchical_level[1:3]
-)
-names(hier_stan_data)
-#get a local fit
[…]

### 2025-08-22 documentation/usecase_morethan1param.R

@@ -1,193 +1,0 @@
-
-# different use cases
-library(here)
-library(tidyverse)
-library(ggplot2)
-library(haven)
-library(stringr)
-library(cmdstanr)
-library(truncnorm) # for inits
-library(bayesplot) # diagnostic plots
-library(posterior)
-devtools::load_all(here::here())
-data_folder <- "data_raw"
-
-# national data
-dat <- read_csv(here::here(data_folder, "coverage_data.csv"))
-dat_subnat <- read_csv(here::here(data_folder, "coverage_data_subnat.csv"))
-hierarchical_level <- c("intercept",  "subcluster", "iso")
-
-### overview, for more than 1 param
-# use case 1: global then local national, all countries (in quarto too)
-# use case 2: global then local national, 1 country
-# use case 3: global then subnational global then local subnational
-### extended to also be able to produce subnational estimates for countries that were
-# included in global national run but not in global subnational run
-# use case 4 needed? invariance to re-ordering of input data in use case 3
-
-
-fit1a_mult <- fit_model_localhierarchy(runstep = "global_national",
-                                       mu_isvector = TRUE,
-                                       hierarchical_level     =  hierarchical_level,
-                                       survey_df = dat,
-                                       chains = 4)
-
-# check the mu_raws
-plots <- plot_mu_raw(fit = fit1a_mult, parname = "mu", morethan1param = TRUE)
-# a list with plots
-# summary plots is a list of dimension k, for each k, it gives summaries, 30 at a time
-plots[["summary_plots"]][[1]][[1]]
-plots[["summary_plots"]][[2]][[1]]
-plots[["summary_plots"]][[3]][[1]]
-# specific plots for one parameter are in plots[["plots_allmuraw"]][[k_select]]
-names(plots[["plots_allmuraw"]][[1]])
-plots[["plots_allmuraw"]][[1]][[1]]
-plots[["plots_allmuraw"]][grepl("intercept", names(plots[["plots_allmuraw"]]))]
-plots[["plots_allmuraw"]][grepl("subcluster", names(plots[["plots_allmuraw"]]))][[1]]
-
-# plot the sigmas
-# these are still fine even when ordered because they are defined in transformed params block
-plot_prior_post_sigmas(fit = fit1a_mult, parname = "mu")
-
[…]

### 2025-08-22 paper/paper.qmd
add paper

---
title: "`localHierarchy`: an R package to facilitate fitting of global and local Bayesian hierarchical models"
author: "Leontine, Herb, Shauna, Evan"
format: 
  pdf:
    latex-engine: pdflatex    # lualatex or xelatex or pdflatex
    includes:
      in-header: preamble.tex 
editor: visual
execute:
  cache: true

abstract: | 
  Bayesian hierarchical models are widely used in global health estimation, where data availability may vary across national or subnational populations. Such models are typically fitted to a global data base, e.g., to produce national-level estimates for all countries in the world or in some region. To facilitate analysis at a local level, models are often desirable that are informed by global models but can be fitted to just a subset of the data, e.g., data for one country. We refer to such models as local models; models that are derived from global hierarchical models but adapted such that they can be fitted to the data from one population alone.

  We present the `localHierarchy` R package, which provides functionality for fitting Bayesian hierarchical models in settings where both global and local estimation is required. The package provides R functions and Stan model components to support global modeling, in which all parameters in hierarchical models are estimated, and local modeling, where parameters are estimated for only one or a small number of populations, using fixed values from a global model fit.

  This article presents a practical introduction to the package for the applied user and illustrates the package’s functionality through several example use cases.
date: last-modified
bibliography: references.bib
number-sections: true
---

```{r}
#| include: false
library(here)
library(tidyverse)
library(ggplot2)
library(haven)
library(stringr)
library(cmdstanr)
library(truncnorm) # for inits
library(bayesplot) # diagnostic plots
library(posterior) # else error for as_draws_list
devtools::load_all(here::here())
```

# Introduction

to finish:

-   context: global health estimation exercises, use of bayesian hier models, examples

-   need for local models

    -   We want local Bayesian models to allow users to do runs on their own laptops/get results quickly, use data from own country only, not to get updates when other country is updated.

    -   Examples of where local models are used: maternal, fp. where they could be used ....

-   challenges with global-local combis: no easy set up to keep code in synch between global and local model, to pass parameters from global model on to local model

-   this paper: introduce a package to facilitate development and fitting of combined global-local models. Models fitted in Stan.

# Methods

The modeling approach introduced in `localhierarchy` was motivated by the hierarchical structures associated with national and subnational level estimation. These structures are illustrated in Figure 1, showing how countries are grouped within regions, regions are grouped in major world regions.

In this section, we introduce the main set up used for global and local hierarchical models in this context of national estimation. We use general notation and a specific example focused on national level estimation. Applications and an extension to subnational estimation are included in the use cases.

Figure 1: Illustration of hierarchical grouping of countries in world regions.

## Hierarchical model specification

Suppose we want to estimate some population parameter of interest $\mu_r$ for a population indexed by $r$. For example, $\mu$ could refer to some population health indicator such as service coverage or morbidity or mortality-related outcomes. Index $r$ refers to the population in a region indexed by $r$, for example, a subnational region, a country, or a group of countries. We assume that $\mu$ is unconstrained (possibly after a transformation has been applied) and consider the following general hierarchical model set up, see Figure 2: \begin{align*}
\mu_{r^{(l)}}^{(l)} \mid \mu^{(l-1)}_{r^\prime}, \sigma^{(l)}_\theta &\sim N(\mu^{(l-1)}_{r^{(l-1)}}, \sigma^{(l)}_\mu), \text{ for } l > 0,
\end{align*} where $\mu_{r^{(l)}}^{(l)}$ ... Here, $\mu^{(l-1)}_{r^\prime}$ is the parameter value at one level higher in the hierarchy for the region $r^\prime$ that contains unit $r$. The parameter $\sigma_\mu^{(l)}$ for $l > 0$ describes variability of values of the parameter $\mu$ across geographic units at level $l$ of the hierarchy. At the top level of the hierarchy, $l = 0$, only a single parameter is estimated.

Figure 2: General specification of a hierarchical model.

As a specific example, we consider the estimation of country-specific parameters, where countries are organized into subregions of countries as illustrated in Figure 1. In this setting, using alternative notation for the same set up to indicate the level by name, the hierarchical model can be written as follows:

\begin{align}
\mu_c|\mu_{s[c]}^{(subregion)}, \sigma_{\mu, country} &\sim N(\mu_{s[c]}^{(subregion)}, \sigma_{\mu, country}^2), \label{eq-countryhier} \\
\mu_s^{(subregion)}|\mu_{w[s]}^{(region)}, \sigma_{\mu, subregion}^2 &\sim N(\mu_{r[s]}^{(region)}, \sigma_{\mu, subregion}^2),\nonumber \\
\mu_w^{(region)}|\mu^{(global)}, \sigma_{\mu, regions}^2 &\sim N(\mu^{(global)}, \sigma_{\mu, region}^2), \nonumber
\end{align} where $c =1, ..., C$ refers to the country index, $s[c] = 1, .., S$ to the subregion index of country $c$, and $w[s[c]] = 1, .., W$ to the world region index of subregion $s[c]$. The parameter $\mu^{(global)}$ is the global mean for the parameter of interest, e.g., a global mean for service coverage or mortality. The parameters $\sigma_{\mu, country}$, $\sigma_{\mu, subregions}$, and $\sigma_{\mu, regions}$ describe variability in the parameter of interest across countries, subregions, and regions, respectively.

<!-- Here, if $\mu^{(global)}$ has prior $\mu^{(global)} \sim N(0, s_{\mu, global}^2)$, then $$\mu_c =  s_{global}\cdot \gamma_{global} + \sigma_{\mu, region}\cdot \gamma_{r[s[c]]]}^{(region)} + \sigma_{\mu, subregion}\cdot\gamma_{s[c]}^{(subregion)} + \sigma_{\mu, country}\cdot\gamma_{\mu, c},$$ with $\gamma \sim N(0,1)$ for each gamma (one global, $R$ regions, $S$ subregions, $C$ countries). -->

## Local hierarchical models

We use the term local hierarchical models to refer to a hierarchical model in which a subset of parameters have been fixed at posterior mean estimates obtained from a fit of the model to a larger data set. A typical workflow for local estimation of national level outcomes is illustrated in Figure 3. For national estimation using the model given by Eq.\~\ref{eq-countryhier}, the user starts by fitting a global model, using a global data base and estimating all parameters in a hierarchical model. In a local model, data from one country are used and parameters that are not country-specific are fixed using parameter estimates from the global model fit.

Figure 3: Workflow for global and local national modeling.

A local model is derived from a global one by fixing parameters. Continuing with the specific example, an example local model to produce country estimates using data from one country alone has the following set up:

$$\mu_c \sim N(\hat{\mu}_{s[c]}^{(subregion)}, \hat{\sigma}_{\mu, country}^2),$$

replacing the hierarchical mean and variances by point estimates. Details are given in the implementation section.

The `localhierachy` package provides a general implementation of local models by facilitating the fixing of subsets of parameters. Continuing with the general notation introduced above, in the package, a local model for one parameter of level $l$ is derived by fixing $\sigma^{(k)}$ for $k=1, ..., l$ and fixing $\mu_{r\prime}{(k)}$ for $k = 1, .., l-1$ at posterior means derived from a global fit. More generally, if a model was fitted using data up to level $l$, the user can create a local model by fixing $\sigma^{(k)}$ for $k \leq l$ as long as, for a given fixed level $f$, variance parameters at levels $f' < f$ have been fixed. Similarly, the user can fix $\mu$'s for for $k \leq l$ as long as, for a given fixed level $f$, $\mu$s at levels $f' < f$ have been fixed and if the $\sigma$'s for the levels up to and include $f$, i.e., $f' \leq f$, have been fixed.

## Implementation

The `localhierarchy` R package contains R functions and example Stan model files to fit global and local hierarchical models. The package is set up to allow for users to incorporate the global-local hierarchy set up into their own project. For a given parameter and hierarchical model specification, the package functions enable processing of data to produce Stan input data for global and local models, Stan model components that work for both global as well as local models, and R functions for postprocessing of model outputs to check results and produce summary files of global models for use in local models.

In a typical work flow, a user first fits a global model, followed by local ones. Worked examples are given in the use case section. The global and local models in these examples are fitted using the `localhierarchy` wrapper function `fit_model_localhierarchy`, using its arguments to specify a hierarchical model, data, and, for local fits, pass on outputs from the global model. The wrapper function calls R subfunctions to process data and prepare Stan input data, read in a Stan model, fit the model, and to postprocess the results.

The package includes a user-friendly specification of hierarchical model structures, and associated specification of fixed parameters for local runs. A worked example is given in the first use case. Briefly, if a user wants to specify a hierarchical model like the one introduced above, with levels world, regions, subregion, and country, <!-- country parameters $\mu_c =  s_{global}\cdot \gamma_{global} + \sigma_{\mu, region}\cdot \gamma_{r[s[c]]]}^{(region)} + \sigma_{\mu, subregion}\cdot\gamma_{s[c]}^{(subregion)} + \sigma_{\mu, country}\cdot\gamma_{\mu, c}$,  --> using a data set with information on these groupings, with naming `region`, `subregion`, and `country`, in the R functions, the user can specify the hierarchical model as follows \[Q for herb: refs/credit to be given for this set up to other work?\]:

```{r, eval = FALSE}
hierarchical_level <- c("intercept",  "region", "subregion", "country")
```

The first level is the intercept, which is the global mean, followed by the region, subregion, and country levels. The user can then specify a local model for one country by fixing parameters at different levels, using the `hierarchical_sigmas_fixed` and `hierarchical_terms_fixed` arguments in the `fit_model_localhierarchy` function. For example, if the $\sigma$s are to be fixed up to the country level and if the $\mu$ terms are to be fixed up to the subregion, the user can encode this as follows:

```{r, eval = FALSE}
hierarchical_sigmas_fixed <- c("intercept",  "region", "subregion", "country")
hierarchical_terms_fixed <- c("intercept",  "region", "subregion")
```

Note that even though there is no $\sigma$ at the intercept level, this term is always included in the `hierarchical_sigmas_fixed` vector.

The parametrization used in Stan is a non-centered parametrization. In general notation, for $l>0$, non-centered parametrization refers to writing mean parameters as deviations from higher-level group means: \begin{align*}
\mu_r^{(l)} &= \mu^{(l-1)} + \sigma^{(l)} \cdot \gamma_{r^\prime}^{(l-1)},
\end{align*} where $\gamma_r^{(l)} \sim N(0, 1)$. Applied sequentially, and if $\mu^{(0)} \sim N(0, s)$, we get \begin{align*}
\mu_r^{(l)} &= s\cdot\gamma^{(0)} + \sum_{k=1}^{l} \sigma^{(k)} \cdot \gamma_{r^\prime}^{(k)},
\end{align*} where $r^\prime$ now refers more generally to the index of the unit containing the lower-level group. For the specific example of country parameters, the non-centered parametrization is given by $$\mu_c =  s_{global}\cdot {\gamma}_{global} + {\sigma}_{\mu, region}\cdot {\gamma}_{r[s[c]]]}^{(region)} + {\sigma}_{\mu, subregion}\cdot{\gamma}_{s[c]}^{(subregion)} + {\sigma}_{\mu, country}\cdot\gamma_{c}.$$

Implementation is based on a model matrix. For our model given by the equation above, the following matrix multiplication is used to obtain the vector of $\mu_c$s: $$(\mu_1, ..., \mu_c) = M\cdot (\eta_1, ..., \eta_{1+R+S+C}),$$ where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_{1+R+S+C})$ refers to the vector of the product of the standard deviation terms and $\gamma$s, i.e., $\gamma_1 = (s_{global}\cdot \gamma_{global}, \gamma_2 = \sigma_{\mu, region}\cdot\gamma_{\mu, 1}^{(region)},..., \gamma_{1+W+S+C} = \sigma_{\mu, country}\cdot\gamma_{\mu, C}$. When estimating a subset of parameters in a local model, fixed values of $\sigma$s and $\gamma$s by point estimates (posterior means) from a global fit. For example, for a national model, we get $$\mu_c =  s_{global}\cdot \hat{\gamma}_{global} + \hat{\sigma}_{\mu, region}\cdot \hat{\gamma}_{r[s[c]]]}^{(region)} + \hat{\sigma}_{\mu, subregion}\cdot\hat{\gamma}_{s[c]}^{(subregion)} + \hat{\sigma}_{\mu, country}\cdot\gamma_{c}.$$

## Operation

\[to finish, LA took mostly from fpemlocal paper\] `localhierarchy` is a publicly available R package stored on Github. For usage, an R (R Core Team, 2020) installation (≥) and a cmdStan (xxx) installation (≥xxx) are required. R can be downloaded on CRAN. Stan install \[add info\] xxx . The `localhierarchy` package can be installed with one of two methods: (1) direct install from Github repository using the `install_ github()` API from the devtools package (Wickham et al., 2020); (2) install from the package binary using the base R function `install.package()`. Package dependencies are listed in the package DESCRIPTION file and will be automatically installed upon installing the main package. There are no minimum RAM, CPU, or HARDDRIVE requirements apart from what is necessary to store model runs, which varies case-by-case.

# Use cases

Example 1 covers national estimation while example 2 illustrated subnational estimation. Both examples consider one parameter. The package also includes functionality for parameter vectors. We illustrate the package functionality using a simple example with simulated data, introduced in the following section.

## Simulation set up

The simulation set up is summarized in Figure 4. It covers both national and subnational parameters and data. While for national level estimation alone, the subnational settings are not needed, we add them here to create a set up that can serve to illustrate both national as well as subnational-level estimation.

Figure 4: Diagram of simulation settings. Top: Hierarchical model for generating country and subnational parameters. Center: Data generating mechanism for national data, illustrated for the first country. Bottom: Data generating mechanism for subnational data, illustrated for the first subnational region.

The top box in Figure 4 indicates the hierarchical model considered. For national level estimation, we consider countries to be grouped in the world, i.e., $\mu_c|\mu^{(global)}, \sigma_{\mu, country} \sim N(\mu^{(global)}, \sigma_{\mu, country}^2)$. For subnational estimation, an additional level is added, with $\mu_r^{(subnat)} =  \mu_{c[r]} + \eta_{r}^{(subnat)}$, $\eta_{r}^{(subnat)}|\sigma_{\mu, subnat} \sim N(0, \sigma_{\mu, subnat})$.

We chose simple normal densities as data generating mechanisms in the simulation, to focus the modeling on the hierarchical model and the `localhierarchy` functionality, as opposed to the data model aspects of modeling. In practice, the data model can be set by the user to reflect actual data generating mechanisms, for example, by using a binomial likelihood, considering transformed outcomes, or accounting for sampling errors for observations that were obtained from a complex survey design.

For generating national level data (box 2), we assume a normal likelihood with constant variance: $$y_i^{(country)}|\mu_{c[i]}, \sigma_y^2 \sim N(\mu_{c[i]}, \sigma_y^2),$$ where $c[i]$ refers to the country of country observation $i$. Again, the use of a normal likelihood with constant variance is for illustration purposes only. In practice, the data model can be set by the user to reflect the observational error. We also generate subnational data following a normal distribution: $$y^{(subnat)}_i|\mu^{(subnat)}_{g[i]}, \sigma_y^2 \sim N(\mu^{(subnat)}_{r[i]}, \sigma_y^2),$$ where $g[i]$ refers to the subnational region of subnational observation $i$. In this simulation set up, we generate subnational data independently of national-level data, and we use the same constant data variance at the national and subnational levels. In real examples, data variances are likely to differ (for example, because subnational sampling errors are larger at the subnational level) and national and subnational data may be derived from the same source. We discuss an example of those settings in the case study.

We create simulated data sets using the `simulate_multilevel_data` function. We simulate a set of model parameters, setting $\mu^{(global)} = 0$, $\sigma_{\mu,country} = 0.5$, $\sigma_{\mu}^{(subnat)} = 0.2$, creating 30 countries with 20 subnational regions in each country. We then simulate subnational data, using $\sigma_y = 0.1$ and generating 10 observations per region. Finally, the function adds national data as well using the argument `add_data1levelup`, to output the subnational and national level data sets:

```{r}
dat_all <- simulate_multilevel_data(
  n_levels = 2,
  n_units_perlevel = c(30, 20),
  sigma_perlevel = c(0.5, 0.2),
  mu_global = 0,
  add_data = TRUE,
  n_data = 10,
  sigma_y = 0.1,
  add_data1levelup = TRUE)
dat_subnat <- dat_all$data
dat_nat <- dat_all$data_1levelup
```

In the simulated data, level1 refers to the country level and level2 refers to the subnational level. The data sets contain the simulated parameters (`eta`) and data (`y`):

```{r}
dat_subnat %>% select(level1, level2, mu_global, level1_eta, level2_eta, y)
dat_nat %>% select(level1,  mu_global, level1_eta, y)
```

## Use case 1: National estimation

### Model fitting

We can fit global and local models using the `fit_model_localhierarchy` wrapper function, as shown later in this section. We first provide implementation details, using the main functions of the package and excerpts from the Stan model.

For model fitting, the first step is to specify a hierarchical structure, using `intercept` and names included in the survey data. For national level estimation, `level1` is the name used in the data:

```{r}
hierarchical_level <- c("intercept",  "level1")
```

Input data are processed using the function `get_geo_unit_index_data`. This function obtains unique hierarchical groupings from the input data and assigns index `c` to each unique grouping:

```{r}
geo_unit_index <- get_geo_unit_index_data(dat_nat,
                                  hierarchical_levels = hierarchical_level,
                                  area = "level1")
geo_unit_index
```

The second step in the processing is to create a hierarchical data object, using the function `hierarchical_data`. This function outputs a list with various components associated with the parametrization of the hierarchical parameters, used in the Stan model and for further post-processing:

```{r}
hier_data <- hierarchical_data(geo_unit_index, hierarchical_level)
names(hier_data)
```

Elements include the model matrix and other elements are used to make a link between columns in the model matrix, or equivalently, indices in the vector $\eta$, and the hierarchical groupings. The model matrix $M$, as introduced above, is as follows:

```{r}
dim(hier_data$model_matrix$matrix)
hier_data$model_matrix$matrix[1:10, 1:10]
```

All this information is passed to the function `hierarchical_param_stan_data` to produce inputs needed in the Stan model fitting. This function outputs a named list with Stan input data relevant to the hierarchical set up:

```{r}
hier_stan_data <- hierarchical_param_stan_data(
  param_name = "mu",
  param_data = hier_data
)
names(hier_stan_data)
```

For example, in this setting, we only have to estimate one $\sigma_{\mu}$, and no $\sigma_{\mu}$ is fixed, as recorded in the `n_sigma_fixed` and `n_sigma_estimate` elements of the Stan input data:

```{r}
hier_stan_data$mu_n_sigma_fixed
hier_stan_data$mu_n_sigma_estimate
```

Excerpts from the Stan model are copied below to introduce model specification. The same Stan model is used for both global and local models. The first excerpt starts at the end in terms of model specification and concerns the specification of the hierarchical parameters $\mu_c$, using the `localhierarchy` Stan function `get_mu`:

```{r, eval = FALSE}
data {
  matrix[n_geounit, mu_raw_n_terms] mu_model_matrix;
}

transformed data {
  vector[rows(csr_extract_w(mu_model_matrix))] mu_model_matrix_w     
              = csr_extract_w(mu_model_matrix);
  array[size(csr_extract_v(mu_model_matrix))] int mu_model_matrix_v  
              = csr_extract_v(mu_model_matrix);
  array[size(csr_extract_u(mu_model_matrix))] int mu_model_matrix_u  
              = csr_extract_u(mu_model_matrix);
}

transformed parameters {
 vector[n_geounit] mu = get_mu(
    mu_star,
    mu_raw_n_terms,
    n_geounit,
    mu_model_matrix_w, mu_model_matrix_v, mu_model_matrix_u);
}
```

Here we use that $(\mu_1, ..., \mu_c) = M\cdot (\eta_1, ..., \eta_c)$, where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_c)$ refers to the vector with products of standard deviation and $\gamma$ terms. The function `get_mu` calculates the vector of $\mu_c$s from the vector of $\eta$s (`mu_star` in the Stan model) and a decomposition of $M$ into different parts $U$, $V$, $W$. This implementation uses the sparse representation of the model matrix $M$ with Stan functions `csr_matrix_times_vector` and `crs_extract`.

The second Stan model excerpt concerns the specification of the $\eta$s (`mu_star`). This vector is obtained by multiplying the vector of $\gamma$s, which is referred to as `mu_raw` in the Stan model by their respective standard deviation, i.e., the vector containing $s_{\mu, global}$ (labeled `mu_scalarprior_mean` in the Stan model) and the vector of stacked $\sigma$s (labeled `mu_sigma` in the Stan model). To allow for global and local models, additions `_estimate` and `_fixed` are used to allow for splitting the `mu_raw` vector into two parts, one for fixed and one for estimated parameters. Similarly, vector `mu_sigma` is also split into two parts, one part for fixed and one part for estimated parameters. Relevant Stan code is as follows:

```{r, eval = FALSE}
data {
  // terms of hierarchical model
  int<lower=0> mu_raw_n_terms;
  int<lower=0> mu_raw_n_terms_fixed;
  int<lower=0> mu_raw_n_terms_estimate;
  int<lower=0> mu_n_sigma;
  int<lower=0> mu_n_sigma_fixed;
  int<lower=0> mu_n_sigma_estimate;
  array[mu_n_sigma + 1] int<lower=1, upper=mu_raw_n_terms> mu_re_start;
  array[mu_n_sigma + 1] int<lower=1, upper=mu_raw_n_terms> mu_re_end;
  // values of fixed parameters
  vector[mu_raw_n_terms_fixed] mu_raw_fixed;
  vector<lower=verysmallnumber>[mu_n_sigma_fixed] mu_sigma_fixed;
  // prior parameters
  real<lower = verysmallnumber> mu_scalarprior_sd;
  real mu_scalarprior_mean;
  real<lower = verysmallnumber> mu_prior_sd_sigma_estimate;
}

parameters {
  vector[mu_raw_n_terms_estimate] mu_raw_estimate;
  vector<lower=verysmallnumber>[mu_n_sigma_estimate] mu_sigma_estimate;
}

transformed parameters {
 vector[mu_raw_n_terms] mu_star = get_mu_star(
    mu_n_sigma, mu_n_sigma_fixed, mu_n_sigma_estimate,
    mu_sigma_fixed, mu_sigma_estimate,
    mu_scalarprior_mean,
    mu_scalarprior_sd,
    mu_raw_n_terms, mu_raw_n_terms_fixed, mu_raw_n_terms_estimate,
    mu_raw_fixed, mu_raw_estimate,
    mu_re_start, mu_re_end);
}
model {
  to_vector(mu_raw_estimate) ~ std_normal();
  to_vector(mu_sigma_estimate) ~ normal(0, mu_prior_sd_sigma_estimate);
}
```

In the model code, the data block contains the input data for the $\gamma$s and $\sigma$s. The `localhierarchy` Stan function `get_mu_star` is used to obtain the $\eta$s. Its arguments also include prior parameters `mu_scalarprior_mean` and `mu_scalarprior_sd`, which refer to the (rescaled) prior mean and standard deviation of $\mu^{(global)}$. The model block contains the priors for the $\gamma$s and $\sigma_{\mu}$s, given by \begin{align*}
\mu^{(global)} &\sim N(0,1),\\
\sigma_{\mu} &\sim \text{Half-Normal}(0, s_{\mu}).
\end{align*} All prior parameters are added as input data to the Stan model.

<!-- In fit_model function, prior parameters are set for the specific parname: stan_data\[\[paste0(parname, "\_scalarprior_sd")\]\] \<- etc. The prior sd parameters are used in plotting. For the prior on the intercept, the following transformation is used: mu_global = prior sd\*(gamma + mu_prior_mean), such that gamma = mu_global/sd - mu_prior_mean. (not doing the /sd would be easier for specification going forward?) -->

We can use the function `fit_model_localhierarchy` to fit the model as specified so far. The argument `runstep` is used to define the type of model, setting it to `global_national` results in fitting a model to all data without fixing any parameters. Other arguments include the hierarchical level specification, the area, and survey data:

```{r}
global_fit <- fit_model_localhierarchy(runstep = "global_national",
                                       hierarchical_level = hierarchical_level,
                                       area = "level1",
                                       survey_df = dat_nat)
```

The fitting returns posterior samples and other meta information about the fit.

```{r}
names(global_fit)
```

The `post_summ` object contains posterior means for relevant model parameters. For example:

```{r}
global_fit$post_summ %>% 
  filter(variable_no_index == "mu_sigma")
```

We can do a local fit by passing on the global fit object to the wrapper function and including data for countries of interest. Here we do local runs for all countries, by including the entire data set, and specify what to fix using the `runstep` argument:

```{r}
fit_local <- fit_model_localhierarchy(runstep = "local_national",
                                      global_fit = global_fit,
                                      area = "level1",
                                      survey_df = dat_nat)
```

When doing a local national run with `runstep = local_national`, the hierarchical levels are taken from the global fit, and non-country-specific parameters are fixed. This is indicated in the print statements above. In the wrapper function, relevant code excerpts for `local_national` runs are as follows:

```{r}
hierarchical_level <- global_fit$hierarchical_level
hierarchical_level_terms_fixed <- 
  hierarchical_level[1:(length(hierarchical_level)-1)]
hierarchical_level_sigmas_fixed <- 
  hierarchical_level[1:(length(hierarchical_level))]
```

When doing a local fit, the same R functions and Stan model is used as introduced above but Stan input data is updated to fix a subset of parameters. Specifically, when producing Stan input data, the following function call is used:

```{r}
hier_stan_data_local <- hierarchical_param_stan_data(
  param_name = "mu",
  param_data = hier_data, 
  # information to fix parameters 
  global_fit = global_fit,
  hierarchical_terms_fixed = hierarchical_level_terms_fixed, 
  hierarchical_sigmas_fixed = hierarchical_level_sigmas_fixed
)
```

where the arguments `hierarchical_terms_fixed` and `hierarchical_sigmas_fixed` are used to fix the $\mu$ at the intercept-level and the $\sigma$s up to and including the country-level. Indeed, when comparing the outputs of this function call to the one for a global fit, we find that the country $\sigma$ is fixed at the value from the global fit:

```{r}
hier_stan_data_local$mu_n_sigma_fixed
hier_stan_data_local$mu_n_sigma_estimate
hier_stan_data_local$mu_sigma_fixed
```

In addition to fixing parameters of hierarchical models, other non-country-specific parameters can be fixed as well. In this example, $\sigma_y$ is a global parameter that is fixed at the posterior mean from the global run for use in local runs.

### Processing model fits

There are several functions to produce and check outputs. The function `posterior_summary_hierparam` is used to summarize posterior draws for hierarchical parameters, to produce summaries of $\mu$ at any level of interest. It takes a fitted model object and a parameter name as input and returns a list with each entry referring one level, and the results (e.g., estimates for $\mu$ estimates) for that level. Here we obtain these results for the global and local model:

```{r, message=FALSE, warning=FALSE}
res_global <- posterior_summary_hierparam(fit = global_fit, parname = "mu")
res_local <- posterior_summary_hierparam(fit = fit_local, parname = "mu")
```

Example output for the global model is as follows:

```{r}
res_local$level1
```

Above column `quant` refers to posterior means and upper and lower bounds of 95% credible intervals. We can create plots to show outputs, for example for the first 4 countries:

```{r}
plot_posterior_summaries(res = res_global, modelname1 = "global",
                          hierarchy_select = "level1",
                          areas_select = unique(res_local$level1$name)[1:4]
                                      )$level1
```

By plotting results from global and local models, we can check how estimates compare and how values have been fixed in the local runs. For example, here we see how country estimates compare between global and local models:

```{r}
plot_posterior_summaries(res = res_local, modelname1 = "local",
                         res2 = res_global, modelname2 = "global", 
                         hierarchy_select = "level1",
                         areas_select = unique(res_local$level1$name)[1:4]
                                      )$level1
```

Estimates are similar, as expected.

When visualizing higher levels, we see how parameters were fixed in the local model:

```{r}
plot_posterior_summaries(res = res_local, modelname1 = "local",
                         res2 = res_global, modelname2 = "global")$intercept 
```

We can also create plots for the $\gamma$s (`mu_raw`s):

```{r, message=FALSE, warning=FALSE}
plots <- plot_mu_raw(fit = global_fit, parname = "mu")
```

This gives a list with plots, including summary plots and plots with priors and posteriors. For example, the figures below show the first 30 $\gamma$s, followed by the prior-posterior densities for the first $\gamma$ (referring to the intercept).

```{r}
plots[["summary_plots"]][[1]]
```

```{r}
plots[["plots_allmuraw"]][[1]]
```

Plots with prior and posterior densities of the $\sigma$s can be produced as well:

```{r}
plot_prior_post_sigmas(fit = global_fit, parname = "mu")
```

## Use case 2: Subnational estimation

### Workflow for nationally-informed subnational estimation
For many global health applications, subnational estimation is of interest to inform policy and programmatic decision making at a more local level. If data are limited, models can be considered to share information across regions. We consider here the use of hierarchical models for a parameter of interest. The `localhierarchy` package can be used to fit subnational hierarchical models, following the same principles as for national estimation or through an updated approach. Just as for national-level hierarchical modeling, the starting point for subnational hierarchical modeling is to specify a hierarchical modeling structure for the subnational regions, such as the one from Figure 1. When following the same principles as for national-level hierarchical modeling, a global model can be fitted to a global data base of subnational data, followed by local models for subnational regions.

In addition to a standard approach to subnational modeling, functionality in the `localhierarchy` package can also be used to fit subnational models that are informed by national level data and estimation, referred to here as "nationally-informed subnational estimation (NISE)". This approach is useful when subnational data are limited, for example, in terms of the number of countries with subnational data, the number of subnational regions with data, the number of observations per subnational region, or due to uncertainty associated with subnational data. An example is given in the case study on family planning estimation, where national level data are available for a large number of country-years while a global subnational data set is more limited in terms of the number of country-years included.

The approach followed in NISE is illustrated in Figure 4, for the hierarchical structure from Figure 1. In the NISE approach, the first step is to fit a model to a global data base with national-level data, to obtain estimates of model parameters up to and including national level parameters. This national-level model fit is used in a global subnational fit to a global data base of subnational data. Specifically, in the global subnational fitting step, national level parameters are fixed at point estimates from the global national fit, i.e., $\hat{\sigma}_{\mu, l}$ and $\mu_{\cdot}^{(l)}$ for level $l$ up to the country level. Finally, in a 3rd step, local models can be fitted, using fixed values from the global subnational fit. In the remainder of this section, we illustrate this approach, continuing with the simulated data from the earlier example. Note that national level data can also inform subnational estimates through inclusion in subnational global or local fits, as indicated in the NISE diagram in Figure 4. This is discussed further in the case study.

Figure 5: Workflow for nationally-informed global and local subnational modeling.

### Modeling

A NISE model is fitted as follows:

```{r}
area_subnat <- "level2"
global_fitsubnat_seq <- fit_model_localhierarchy(
   runstep = "global_subnational",
   area = area_subnat, # subnational area name 
   survey_df = dat_subnat, # pass on subnat data
   # for NISE:
   global_fit = global_fit, # global national fit
   add_subnational_hierarchy = area_subnat
   # what level to add to the hierarchy, compared to the 
   # global national model 
   )
```

Through the `add_subnational_hierarchy` argument, and as shown in the printed statements, the hierarchical structure is now given by intercept $\rightarrow$ country $\rightarrow$ subnational level. By default, the world mean and across-country variance are fixed. Code excerpts to do so are as follows:

```{r, eval = FALSE}
hierarchical_level <- global_fit$hierarchical_level
hierarchical_level <- c(hierarchical_level, add_subnational_hierarchy)
hierarchical_level_terms_fixed <-
  hierarchical_level[1:(length(hierarchical_level)-2)]
hierarchical_level_sigmas_fixed <-
  hierarchical_level[1:(length(hierarchical_level)-1)]
```

By comparing the fit summaries from the global national and global subnational fits for the $\sigma$s, we see that, as expected, one additional $\sigma_{\mu}$ is estimated in the subnational model (to capture across-subnational region variability) and the $\sigma_{\mu}$ at the higher country level is the same in the global national and subnational fit:

```{r}
global_fit$post_summ %>%
  filter(str_detect(variable_no_index, "sigma"))
global_fitsubnat_seq$post_summ %>%
  filter(str_detect(variable_no_index, "sigma"))
```

We can use the subnational global fit to do a local subnational fit, using the `runstep = "local_subnational"` setting:

```{r}
fit_local_seq <- fit_model_localhierarchy(runstep = "local_subnational",
                                   area = area_subnat,
                                   survey_df = dat_subnat,
                                   global_fit = global_fitsubnat_seq)
```

Similarly to the local national run, in subnational local runs, hierarchical levels and what is fixed is set in the wrapper function; we estimate subnational parameters only. Point estimates in the subnational global fit include point estimates from the national global run as well as point estimates from the global subnational fit. By default, parameters that are duplicated between the two fits, e.g., world region or country means, are taken from the national global run. This is implemented as such to avoid the situation where local runs would fail for countries that were not included in the subnational global fitting, i.e., when country parameters were fixed and missing from the subnational fit. If the fixed parameters differ between national and subnational global fits, i.e., if country means are re-estimated in the global subnational fit as in this example, and if the user wants to use these fixed parameter estimates from the subnational global fit, i.e., if the user wants to use the fixed country estimates from the global national run, the argument `use_globalsubnat_fromnat` can be used when calling the `fit_model_localhierarchy` function, see documentation.

We can create and check summaries of subnational fits using the same functions for national fits. Below we compare the subnational global fit with the local subnational fit. The results are comparable, as expected.

```{r, message=FALSE, warning=FALSE}
res_globalsubnat_seq <- posterior_summary_hierparam(
  fit = global_fitsubnat_seq,
  parname = "mu")
res_local_seq <- posterior_summary_hierparam(
  fit = fit_local_seq, 
  parname = "mu")
```

```{r}
plot_posterior_summaries(res = res_local_seq, modelname1 = "local",
                         res2 = res_globalsubnat_seq, modelname2 = "global",
                        hierarchy_select = "level2",
                        areas_select = unique(res_local_seq$level2$name)[1:30]
                          )$level2
```

<!-- could add back the direct subnat fit here too: two local subnational fits: one created from the sequential approach introduced above, and one create from a subnational global fit that did not use national level information. The results are comparable, as expected.  -->

# Usage in other projects

The `localhierarchy` package is used in different projects, including the Family Planning Estimation Tool (FPET) and national and subnational modeling of coverage indicators for the Countdown initiative. In these projects, the goal is to estimate and create short-term forecasts of a set of indicators at the national and subnational levels. For example, in FPET, the goal is to produce estimates with uncertainty for modern contraceptive use from 1970 until 2030. Bayesian statistical models are used that parametrize how indicators change with time and hierarchical models are incorpored in such temporal models. A workflow of global-local estimation is followed such that countries can produce estimates and forecasts based on data for their respective countries. In these applications, `localhierarchy` functions are either used as is or adapted for the application. Stan models include the specification of the hierarchical models and Stan functions `localhierarchy`.

TO DO: finish a summary of one application, probably fpet, to present

-   summary of modeling approach used and how hier model is incorporated (ie transition model, parameters of longer term trend are estimated hierarchically)

    <!-- Model indicators demand and demand satisfied. -->

    <!-- Hierarchical models are used to capture country differences in the transitions, i.e., the long-term process by which these indicators change. Specific parameters are the rate of change (cpatured by spline regression coefficients), the timing, and asymptote. -->

-   perhaps a brief note about data models used

-   workflow used: global model fitting, then country workshops for local fitting, sensitivity analyses

    -   to decide whether to add the first step of estimating longer term trends here

        <!-- In the global national model, 2 steps are taken as well: we first estimate long terms trends, subcluster parameters, followed by country parameters. -->

-   how additional data can be used in local models: introduce routine data

-   subnational modeling:

    -   summary to show that subnat global data base is much more limited than global national one

        <!-- Global national data base has xx obs for married and xxx for unmarried women. Subnational data were processed for xxx countries at the time of model development. -->

    -   summary of process/workflow

        <!-- at subnational global fitting, we fix country parameters, effectovely introducing informative priors centered at natiional-data informed means. -->

    -   bottom-up estimation, how national level data is used in subnational modeling. draft text below

-   \[probably not\] fp workflow additions (fitting for married vs unmarried, combination in all-women model)

Draft text on bottom-up estimation: Typically, a national level indicator $\mu_c$ for country $c$ is a weighted mean of the subnational estimates, i.e. $\mu_c = \sum_s 1(c[s] = c) w_s \mu_{s}^{(subnat)}$, where the weights $w_s$ are based on population size or a different demographic indicator in subnational region $s$. For example, if the outcome refers to a coverage proportion, weights are based on the relevant population size in each subnational region. National level estimates that are obtained from subnational estimates are referred to as bottom-up estimates. Bottom-up estimates can be produced as a by-product of subnational estimates in a model fit where estimates for all regions in a country are produced. In settings where bottom-up estimates can be produced, national data can be considered for inclusion in model fitting. Specifically, national level data can inform the bottom-up estimate and hence indirectly, the subnational estimates. This is useful in settings where a national survey may not be available at the subnational level.

<!-- Bottom-up estimates are typically close to but not equal to national estimates $\hat{\mu}_c$, as different data sources and modeling assumptions are used when directly estimating national level indicators using national-level data versus when constructing bottom-up estimates produced from subnational data. -->

# Conclusions

# Additional details

##

### 2025-09-10 paper/paper.qmd
move intro and application to overleaf

@@
 # Introduction
-
-to finish:
-
--   context: global health estimation exercises, use of bayesian hier models, examples
-
--   need for local models
-
-    -   We want local Bayesian models to allow users to do runs on their own laptops/get results quickly, use data from own country only, not to get updates when other country is updated.
-
-    -   Examples of where local models are used: maternal, fp. where they could be used ....
-
--   challenges with global-local combis: no easy set up to keep code in synch between global and local model, to pass parameters from global model on to local model
-
--   this paper: introduce a package to facilitate development and fitting of combined global-local models. Models fitted in Stan.
+see overleaf
 

### 2025-09-10 paper/paper.qmd
move usage

@@
 
-The `localhierarchy` package is used in different projects, including the Family Planning Estimation Tool (FPET) and national and subnational modeling of coverage indicators for the Countdown initiative.
-In these projects, the goal is to estimate and create short-term forecasts of a set of indicators at the national and subnational levels.
-For example, in FPET, the goal is to produce estimates with uncertainty for modern contraceptive use from 1970 until 2030.
-Bayesian statistical models are used that parametrize how indicators change with time and hierarchical models are incorpored in such temporal models.
-A workflow of global-local estimation is followed such that countries can produce estimates and forecasts based on data for their respective countries.
-In these applications, `localhierarchy` functions are either used as is or adapted for the application.
-Stan models include the specification of the hierarchical models and Stan functions `localhierarchy`.
-
-TO DO: finish a summary of one application, probably fpet, to present
-
--   summary of modeling approach used and how hier model is incorporated (ie transition model, parameters of longer term trend are estimated hierarchically)
-
-    <!-- Model indicators demand and demand satisfied. -->
-
-    <!-- Hierarchical models are used to capture country differences in the transitions, i.e., the long-term process by which these indicators change.
-Specific parameters are the rate of change (cpatured by spline regression coefficients), the timing, and asymptote.
--->
-
--   perhaps a brief note about data models used
-
--   workflow used: global model fitting, then country workshops for local fitting, sensitivity analyses
-
-    -   to decide whether to add the first step of estimating longer term trends here
-
-        <!-- In the global national model, 2 steps are taken as well: we first estimate long terms trends, subcluster parameters, followed by country parameters. -->
-
--   how additional data can be used in local models: introduce routine data
-
--   subnational modeling:
-
-    -   summary to show that subnat global data base is much more limited than global national one
-
-        <!-- Global national data base has xx obs for married and xxx for unmarried women. Subnational data were processed for xxx countries at the time of model development. -->
-
-    -   summary of process/workflow
-
-        <!-- at subnational global fitting, we fix country parameters, effectovely introducing informative priors centered at natiional-data informed means. -->
-
-    -   bottom-up estimation, how national level data is used in subnational modeling. draft text below
-
--   \[probably not\] fp workflow additions (fitting for married vs unmarried, combination in all-women model)
-
-Draft text on bottom-up estimation: Typically, a national level indicator $\mu_c$ for country $c$ is a weighted mean of the subnational estimates, i.e.
-$\mu_c = \sum_s 1(c[s] = c) w_s \mu_{s}^{(subnat)}$, where the weights $w_s$ are based on population size or a different demographic indicator in subnational region $s$.
-For example, if the outcome refers to a coverage proportion, weights are based on the relevant population size in each subnational region.
-National level estimates that are obtained from subnational estimates are referred to as bottom-up estimates.
-Bottom-up estimates can be produced as a by-product of subnational estimates in a model fit where estimates for all regions in a country are produced.
-In settings where bottom-up estimates can be produced, national data can be considered for inclusion in model fitting.
-Specifically, national level data can inform the bottom-up estimate and hence indirectly, the subnational estimates.
-This is useful in settings where a national survey may not be available at the subnational level.
-
-<!-- Bottom-up estimates are typically close to but not equal to national estimates $\hat{\mu}_c$, as different data sources and modeling assumptions are used when directly estimating national level indicators using national-level data versus when constructing bottom-up estimates produced from subnational data.
--->
+see overleaf
 

### 2025-09-10 README.Rmd
update documentation

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# localHierarchy

The `localHierarchy` R package provides functionality for fitting Bayesian hierarchical models in settings where both global and local estimation is required. The package provides R functions and Stan model components to support global modeling, in which all parameters in hierarchical models are estimated, and local modeling, where parameters are estimated for only one or a small number of populations, using fixed values from a global model fit.

## Installation

Dependency on `cmdstanr:`

Instructions for installing `cmdstanr` are available in their [Getting started](https://mc-stan.org/cmdstanr/articles/cmdstanr.html) guide.

Install `localHierarchy` from Github:
`remotes::install_github("AlkemaLab/localhierarchy")`

## Examples

to do: add from paper

### 2025-09-10 R/hier_data_helpers.R

@@ -32,7 +32,7 @@ is_there_a_comma <- function(variable){
 #' relevant information
 #'
 #' @param post_summ tibble with columns `variable` and `postmean`,
-#' variable examples are 'a[1,1]', 'a[2,1]', etc.
+#' variable examples are `a[1,1]`, `a[2,1]`, etc.
 #'
 #' @returns matrix with rows and columns corresponding to the indices in the variable
 #' @export
@@ -89,7 +89,7 @@ variable_exists_in_draws <- function(fit, var_name) {
 #' about a variable name and row and/or column indices, add columns containing
 #' just the variable name and the indices.
 #'
-#' @param estimate_df data frame with a column called `variable` that contains parname[i] or parname[i,j] entries
+#' @param estimate_df data frame with a column called `variable` that contains `parname[i]` or `parname[i,j]` entries
 #' @param num_inds number of indices in the variable name (used for 1 or 2)
 #'
 #' @returns data frame with the same columns as `estimate_df` plus columns containing

### 2025-09-10 R/hierarchical_model_matrix.R

@@ -1,15 +1,22 @@
 
 
 #' hierarchical_model_matrix
+#'
 #' Create list with information including a model matrix, associated with the geo_unit_data and hierarchical_levels
 #'
 #' @param geo_unit_index_data tibble with unique hierarchical levels (one column for each level, one row per combi)
 #' @param hierarchical_level vector with hierarchical spec from highest to lowest levels
 #'
-#' @returns list (modelmatrix, assign, index) associated with the geo_unit_data and hierarchical_levels
-#' in modelmatrix each column refers to one eta (and nrows is the number of lowest level units).
-#' assign is length of etas, with 0 for intercept, then an hierarchical level index (starting at 2) for each eta level
-#' index is a tibble with n_eta rows, with info for each eta in and columns i [index of eta], column [hierarchical level], and level [name of hierarchical level]
+#' @return A list with components `modelmatrix`, `assign`, and `index`:
+#'   - `modelmatrix`: A matrix where each column refers to one eta (number of rows equals number of lowest level units).
+#'   - `assign`: Integer vector of length equal to the number of etas; `0` for the intercept, then a hierarchical level index (starting at 2) for each eta level.
+#'   - `index`: A tibble with `n_eta` rows, and columns:
+#'       - `i`: Index of eta
+#'       - `column`: Hierarchical level
+#'       - `level`: Name of hierarchical level
+#'
+#' All components are associated with the provided `geo_unit_data` and `hierarchical_levels`.
+#'
 #' @export
 #'
 hierarchical_model_matrix <- function(geo_unit_index_data, hierarchical_level) {

### 2025-09-12 paper/paper.qmd
Updated paper with typographical edits; standardized package name on 'localhierarchy'

@@
 ---
-title: "`localHierarchy`: an R package to facilitate fitting of global and local Bayesian hierarchical models"
+title: "`localhierarchy`: an R package to facilitate fitting of global and local Bayesian hierarchical models"
 author: "Leontine, Herb, Shauna, Evan"
@@
     latex-engine: pdflatex    # lualatex or xelatex or pdflatex
-    includes:
-      in-header: preamble.tex 
+    include-in-header: 
+      - file: preamble.tex 
 editor: visual
@@
   Bayesian hierarchical models are widely used in global health estimation, where data availability may vary across national or subnational populations.
-Such models are typically fitted to a global data base, e.g., to produce national-level estimates for all countries in the world or in some region.
-To facilitate analysis at a local level, models are often desirable that are informed by global models but can be fitted to just a subset of the data, e.g., data for one country.
-We refer to such models as local models; models that are derived from global hierarchical models but adapted such that they can be fitted to the data from one population alone.
+Such models are typically fitted to a global database, e.g., to produce national-level estimates for all countries in the world or in some region.
+To facilitate analysis at a local level, models are often desirable that are informed by global models but can be fitted to just a subset of the data, such as data for one country.
+We refer to such models as local models: models that are derived from global hierarchical models but adapted such that they can be fitted to the data from one population alone.
 
   We present the `localHierarchy` R package, which provides functionality for fitting Bayesian hierarchical models in settings where both global and local estimation is required.
-The package provides R functions and Stan model components to support global modeling, in which all parameters in hierarchical models are estimated, and local modeling, where parameters are estimated for only one or a small number of populations, using fixed values from a global model fit.
+The package provides R functions and Stan model components to support global modeling, in which all parameters in hierarchical models are estimated, and local modeling, in which parameters are estimated for only one or a small number of populations, using fixed values from a global model fit.
 
@@
 # Introduction
+
 see overleaf
@@
 The modeling approach introduced in `localhierarchy` was motivated by the hierarchical structures associated with national and subnational level estimation.
-These structures are illustrated in Figure 1, showing how countries are grouped within regions, regions are grouped in major world regions.
+These structures are illustrated in Figure 1, showing how countries are grouped within regions, and regions are grouped in major world regions.
 
-In this section, we introduce the main set up used for global and local hierarchical models in this context of national estimation.
+In this section, we introduce the principle setup used for global and local hierarchical models in the context of national estimation.
 We use general notation and a specific example focused on national level estimation.
@@
 Index $r$ refers to the population in a region indexed by $r$, for example, a subnational region, a country, or a group of countries.
-We assume that $\mu$ is unconstrained (possibly after a transformation has been applied) and consider the following general hierarchical model set up, see Figure 2: \begin{align*}
+We assume that $\mu$ is unconstrained (possibly after a transformation has been applied) and consider the following general hierarchical model setup, see Figure 2: \begin{align*}
   \mu_{r^{(l)}}^{(l)} \mid \mu^{(l-1)}_{r^\prime}, \sigma^{(l)}_\theta &\sim N(\mu^{(l-1)}_{r^{(l-1)}}, \sigma^{(l)}_\mu), \text{ for } l > 0,
-\end{align*} where $\mu_{r^{(l)}}^{(l)}$ ...
+\end{align*} 
+where $\mu_{r^{(l)}}^{(l)}$ ...
 Here, $\mu^{(l-1)}_{r^\prime}$ is the parameter value at one level higher in the hierarchy for the region $r^\prime$ that contains unit $r$.
@@
 
-As a specific example, we consider the estimation of country-specific parameters, where countries are organized into subregions of countries as illustrated in Figure 1.
-In this setting, using alternative notation for the same set up to indicate the level by name, the hierarchical model can be written as follows:
+As a specific example, we consider the estimation of country-specific parameters, where countries are organized into subregions of countries (illustrated in Figure 1).
+In this setting, using alternative notation for the same setup to indicate the level by name, the hierarchical model can be written as follows:
 
-\begin{align}
-\mu_c|\mu_{s[c]}^{(subregion)}, \sigma_{\mu, country} &\sim N(\mu_{s[c]}^{(subregion)}, \sigma_{\mu, country}^2), \label{eq-countryhier} \\
-\mu_s^{(subregion)}|\mu_{w[s]}^{(region)}, \sigma_{\mu, subregion}^2 &\sim N(\mu_{r[s]}^{(region)}, \sigma_{\mu, subregion}^2),\nonumber \\
-\mu_w^{(region)}|\mu^{(global)}, \sigma_{\mu, regions}^2 &\sim N(\mu^{(global)}, \sigma_{\mu, region}^2), \nonumber
-\end{align} where $c =1, ..., C$ refers to the country index, $s[c] = 1, .., S$ to the subregion index of country $c$, and $w[s[c]] = 1, .., W$ to the world region index of subregion $s[c]$.
-The parameter $\mu^{(global)}$ is the global mean for the parameter of interest, e.g., a global mean for service coverage or mortality.
-The parameters $\sigma_{\mu, country}$, $\sigma_{\mu, subregions}$, and $\sigma_{\mu, regions}$ describe variability in the parameter of interest across countries, subregions, and regions, respectively.
+\begin{alignat}{2}
+  \mu_c &\mid \mu_{s[c]}^{(\subregion)}, \sigma_{\mu, \country} &&\sim N(\mu_{s[c]}^{(\subregion)}, \sigma_{\mu, \country}^2), \label{eq-countryhier} \\
+  \mu_s^{(\subregion)} &\mid \mu_{w[s]}^{(\region)}, \sigma_{\mu, \subregion}^2 &&\sim N(\mu_{r[s]}^{(\region)}, \sigma_{\mu, \subregion}^2), \nonumber \\
+  \mu_w^{(\region)} &\mid \mu^{(\glob)}, \sigma_{\mu, \region}^2 &&\sim N(\mu^{(\glob)}, \sigma_{\mu, \region}^2), \nonumber
+\end{alignat} 
+where $c =1,\dots,C$ refers to the country index, $s[c] = 1, .., S$ to the subregion index of country $c$, and $w[s[c]] = 1, .., W$ to the region index of subregion $s[c]$.
+The parameter $\mu^{(\glob)}$ is the global mean for the parameter of interest, e.g., a global mean for service coverage or mortality.
+The parameters $\sigma_{\mu, \country}$, $\sigma_{\mu, \subregion}$, and $\sigma_{\mu, \region}$ describe variability in the parameter of interest across countries, subregions, and regions, respectively.
 
-<!-- Here, if $\mu^{(global)}$ has prior $\mu^{(global)} \sim N(0, s_{\mu, global}^2)$, then $$\mu_c =  s_{global}\cdot \gamma_{global} + \sigma_{\mu, region}\cdot \gamma_{r[s[c]]]}^{(region)} + \sigma_{\mu, subregion}\cdot\gamma_{s[c]}^{(subregion)} + \sigma_{\mu, country}\cdot\gamma_{\mu, c},$$ with $\gamma \sim N(0,1)$ for each gamma (one global, $R$ regions, $S$ subregions, $C$ countries).
+<!-- Here, if $\mu^{(\glob)}$ has prior $\mu^{(\glob)} \sim N(0, s_{\mu, global}^2)$, then $$\mu_c =  s_{global}\cdot \gamma_{global} + \sigma_{\mu, region}\cdot \gamma_{r[s[c]]]}^{(
+egion)} + \sigma_{\mu, subregion}\cdot\gamma_{s[c]}^{(\subregion)} + \sigma_{\mu, country}\cdot\gamma_{\mu, c},$$ with $\gamma \sim N(0,1)$ for each gamma (one global, $R$ regions, $S$ subregions, $C$ countries).
 -->
@@
 A local model is derived from a global one by fixing parameters.
-Continuing with the specific example, an example local model to produce country estimates using data from one country alone has the following set up:
+Continuing with the specific example, an example local model to produce country estimates using data from one country alone has the following setup:
 
-$$\mu_c \sim N(\hat{\mu}_{s[c]}^{(subregion)}, \hat{\sigma}_{\mu, country}^2),$$
+$$\mu_c \sim N(\hat{\mu}_{s[c]}^{(\subregion)}, \hat{\sigma}_{\mu, \country}^2),$$
 
@@
 The `localhierarchy` R package contains R functions and example Stan model files to fit global and local hierarchical models.
-The package is set up to allow for users to incorporate the global-local hierarchy set up into their own project.
+The package is set up to allow for users to incorporate the global-local hierarchy setup into their own project.
 For a given parameter and hierarchical model specification, the package functions enable processing of data to produce Stan input data for global and local models, Stan model components that work for both global as well as local models, and R functions for postprocessing of model outputs to check results and produce summary files of global models for use in local models.
@@
 The global and local models in these examples are fitted using the `localhierarchy` wrapper function `fit_model_localhierarchy`, using its arguments to specify a hierarchical model, data, and, for local fits, pass on outputs from the global model.
-The wrapper function calls R subfunctions to process data and prepare Stan input data, read in a Stan model, fit the model, and to postprocess the results.
+The wrapper function calls R functions to process data and prepare Stan input data, read in a Stan model, fit the model, and to postprocess the results.
 
@@
 A worked example is given in the first use case.
-Briefly, if a user wants to specify a hierarchical model like the one introduced above, with levels world, regions, subregion, and country, <!-- country parameters $\mu_c =  s_{global}\cdot \gamma_{global} + \sigma_{\mu, region}\cdot \gamma_{r[s[c]]]}^{(region)} + \sigma_{\mu, subregion}\cdot\gamma_{s[c]}^{(subregion)} + \sigma_{\mu, country}\cdot\gamma_{\mu, c}$,  --> using a data set with information on these groupings, with naming `region`, `subregion`, and `country`, in the R functions, the user can specify the hierarchical model as follows \[Q for herb: refs/credit to be given for this set up to other work?\]:
+Briefly, if a user wants to specify a hierarchical model like the one introduced above, with the levels world, regions, subregion, and country, <!-- country parameters $\mu_c =  s_{global}\cdot \gamma_{global} + \sigma_{\mu, region}\cdot \gamma_{r[s[c]]]}^{(
+egion)} + \sigma_{\mu, subregion}\cdot\gamma_{s[c]}^{(\subregion)} + \sigma_{\mu, country}\cdot\gamma_{\mu, c}$,  --> using a dataset with information on these groupings, with naming `region`, `subregion`, and `country`, in the R functions, the user can specify the hierarchical model as follows \[Q for herb: refs/credit to be given for this set up to other work?\]:
 
@@
 
-The parametrization used in Stan is a non-centered parametrization.
-In general notation, for $l>0$, non-centered parametrization refers to writing mean parameters as deviations from higher-level group means: \begin{align*}
+The parameterization used in Stan is a non-centered parameterization.
+In general notation, for $l>0$, non-centered parameterization refers to writing mean parameters as deviations from higher-level group means:
+\begin{align*}
   \mu_r^{(l)} &= \mu^{(l-1)} + \sigma^{(l)} \cdot \gamma_{r^\prime}^{(l-1)},
-\end{align*} where $\gamma_r^{(l)} \sim N(0, 1)$. Applied sequentially, and if $\mu^{(0)} \sim N(0, s)$, we get \begin{align*}
+\end{align*} 
+where $\gamma_r^{(l)} \sim N(0, 1)$. Applied sequentially, and if $\mu^{(0)} \sim N(0, s)$, we get 
+\begin{align*}
   \mu_r^{(l)} &= s\cdot\gamma^{(0)} + \sum_{k=1}^{l} \sigma^{(k)} \cdot \gamma_{r^\prime}^{(k)},
-\end{align*} where $r^\prime$ now refers more generally to the index of the unit containing the lower-level group.
-For the specific example of country parameters, the non-centered parametrization is given by $$\mu_c =  s_{global}\cdot {\gamma}_{global} + {\sigma}_{\mu, region}\cdot {\gamma}_{r[s[c]]]}^{(region)} + {\sigma}_{\mu, subregion}\cdot{\gamma}_{s[c]}^{(subregion)} + {\sigma}_{\mu, country}\cdot\gamma_{c}.$$
+\end{align*}
+where $r^\prime$ now refers more generally to the index of the unit containing the lower-level group. For the specific example of country parameters, the non-centered parameterization is given by 
+\begin{align*}
+  \mu_c =  s_{\glob} \cdot {\gamma}_{\glob} + {\sigma}_{\mu, \region}\cdot {\gamma}_{r[s[c]]]}^{(\region)} + {\sigma}_{\mu, \subregion}\cdot{\gamma}_{s[c]}^{(\subregion)} + {\sigma}_{\mu, \country}\cdot\gamma_{c}.
+\end{align*}
 
 Implementation is based on a model matrix.
-For our model given by the equation above, the following matrix multiplication is used to obtain the vector of $\mu_c$s: $$(\mu_1, ..., \mu_c) = M\cdot (\eta_1, ..., \eta_{1+R+S+C}),$$ where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_{1+R+S+C})$ refers to the vector of the product of the standard deviation terms and $\gamma$s, i.e., $\gamma_1 = (s_{global}\cdot \gamma_{global}, \gamma_2 = \sigma_{\mu, region}\cdot\gamma_{\mu, 1}^{(region)},..., \gamma_{1+W+S+C} = \sigma_{\mu, country}\cdot\gamma_{\mu, C}$.
+For our model given by the equation above, the following matrix multiplication is used to obtain the vector of $\mu_c$s: $$(\mu_1, ..., \mu_c) = M\cdot (\eta_1, ..., \eta_{1+R+S+C}),$$ where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_{1+R+S+C})$ refers to the vector of the product of the standard deviation terms and $\gamma$s, i.e., $\gamma_1 = (s_{\glob}\cdot \gamma_{\glob}, \gamma_2 = \sigma_{\mu, \region} \cdot \gamma_{\mu, 1}^{(\region)},..., \gamma_{1+W+S+C} = \sigma_{\mu, \country}\cdot\gamma_{\mu, C}$.
 When estimating a subset of parameters in a local model, fixed values of $\sigma$s and $\gamma$s by point estimates (posterior means) from a global fit.
-For example, for a national model, we get $$\mu_c =  s_{global}\cdot \hat{\gamma}_{global} + \hat{\sigma}_{\mu, region}\cdot \hat{\gamma}_{r[s[c]]]}^{(region)} + \hat{\sigma}_{\mu, subregion}\cdot\hat{\gamma}_{s[c]}^{(subregion)} + \hat{\sigma}_{\mu, country}\cdot\gamma_{c}.$$
+For example, for a national model, we get
+\begin{align*}
+  \mu_c = s_{\glob} \cdot \hat{\gamma}_{\glob} + 
+          \hat{\sigma}_{\mu, \region} \cdot \hat{\gamma}_{r[s[c]]]}^{(\region)} + 
+          \hat{\sigma}_{\mu, \subregion} \cdot \hat{\gamma}_{s[c]}^{(\subregion)} + 
+          \hat{\sigma}_{\mu, \country} \cdot \gamma_{c}.
+\end{align*}
 
@@
 
-## Simulation set up
+## Simulation setup
 
-The simulation set up is summarized in Figure 4.
+The simulation setup is summarized in Figure 4.
 It covers both national and subnational parameters and data.
-While for national level estimation alone, the subnational settings are not needed, we add them here to create a set up that can serve to illustrate both national as well as subnational-level estimation.
+While for national level estimation alone, the subnational settings are not needed, we add them here to create a setup that can serve to illustrate both national as well as subnational-level estimation.
 
@@
 The top box in Figure 4 indicates the hierarchical model considered.
-For national level estimation, we consider countries to be grouped in the world, i.e., $\mu_c|\mu^{(global)}, \sigma_{\mu, country} \sim N(\mu^{(global)}, \sigma_{\mu, country}^2)$.
-For subnational estimation, an additional level is added, with $\mu_r^{(subnat)} =  \mu_{c[r]} + \eta_{r}^{(subnat)}$, $\eta_{r}^{(subnat)}|\sigma_{\mu, subnat} \sim N(0, \sigma_{\mu, subnat})$.
+For national level estimation, we consider countries to be grouped in the world, i.e., $\mu_c|\mu^{(\glob)}, \sigma_{\mu, \country} \sim N(\mu^{(\glob)}, \sigma_{\mu, \country}^2)$.
+For subnational estimation, an additional level is added, with $\mu_r^{(\subnat)} = \mu_{c[r]} + \eta_{r}^{(\subnat)}$, $\eta_{r}^{(\subnat)}|\sigma_{\mu, \subnat} \sim N(0, \sigma_{\mu, \subnat})$.
 
-We chose simple normal densities as data generating mechanisms in the simulation, to focus the modeling on the hierarchical model and the `localhierarchy` functionality, as opposed to the data model aspects of modeling.
+We chose simple normal densities as data generating mechanisms in the simulation, to focus the modeling on the hierarchical model and the `localhierarchy` functionality as opposed to the data model aspects of modeling.
 In practice, the data model can be set by the user to reflect actual data generating mechanisms, for example, by using a binomial likelihood, considering transformed outcomes, or accounting for sampling errors for observations that were obtained from a complex survey design.
 
-For generating national level data (box 2), we assume a normal likelihood with constant variance: $$y_i^{(country)}|\mu_{c[i]}, \sigma_y^2 \sim N(\mu_{c[i]}, \sigma_y^2),$$ where $c[i]$ refers to the country of country observation $i$.
+For generating national level data (box 2), we assume a normal likelihood with constant variance: $$y_i^{(\country)}|\mu_{c[i]}, \sigma_y^2 \sim N(\mu_{c[i]}, \sigma_y^2),$$ where $c[i]$ refers to the country of country observation $i$.
 Again, the use of a normal likelihood with constant variance is for illustration purposes only.
@@
 We also generate subnational data following a normal distribution: $$y^{(subnat)}_i|\mu^{(subnat)}_{g[i]}, \sigma_y^2 \sim N(\mu^{(subnat)}_{r[i]}, \sigma_y^2),$$ where $g[i]$ refers to the subnational region of subnational observation $i$.
-In this simulation set up, we generate subnational data independently of national-level data, and we use the same constant data variance at the national and subnational levels.
+In this simulation setup, we generate subnational data independently of national-level data, and we use the same constant data variance at the national and subnational levels.
 In real examples, data variances are likely to differ (for example, because subnational sampling errors are larger at the subnational level) and national and subnational data may be derived from the same source.
@@
 
-We create simulated data sets using the `simulate_multilevel_data` function.
-We simulate a set of model parameters, setting $\mu^{(global)} = 0$, $\sigma_{\mu,country} = 0.5$, $\sigma_{\mu}^{(subnat)} = 0.2$, creating 30 countries with 20 subnational regions in each country.
+Simulated data can be generated using  `simulate_multilevel_data` function.
+We simulate a set of model parameters, setting $\mu^{(\glob)} = 0$, $\sigma_{\mu,\country} = 0.5$, $\sigma_{\mu}^{(\subnat)} = 0.2$, creating 30 countries with 20 subnational regions in each country.
 We then simulate subnational data, using $\sigma_y = 0.1$ and generating 10 observations per region.
@@
   sigma_y = 0.1,
-  add_data1levelup = TRUE)
+  add_data1levelup = TRUE
+)
+
 dat_subnat <- dat_all$data
@@
 
-In the simulated data, level1 refers to the country level and level2 refers to the subnational level. The data sets contain the simulated parameters (`eta`) and data (`y`):
+In the simulated data, `level1` refers to the country level and `level2` refers to the subnational level. The data sets contain the simulated parameters (`eta`) and data (`y`):
 
 ```{r}
-dat_subnat %>% select(level1, level2, mu_global, level1_eta, level2_eta, y)
-dat_nat %>% select(level1,  mu_global, level1_eta, y)
+dat_subnat |> select(level1, level2, mu_global, level1_eta, level2_eta, y)
+dat_nat |> select(level1,  mu_global, level1_eta, y)
 ```
@@
 ```{r}
-geo_unit_index <- get_geo_unit_index_data(dat_nat,
+geo_unit_index <- get_geo_unit_index_data(
+  dat_nat,
   hierarchical_levels = hierarchical_level,
-                                  area = "level1")
+  area = "level1"
+)
 geo_unit_index
@@
 The second step in the processing is to create a hierarchical data object, using the function `hierarchical_data`.
-This function outputs a list with various components associated with the parametrization of the hierarchical parameters, used in the Stan model and for further post-processing:
+This function outputs a list with various components associated with the parameterization of the hierarchical parameters, used in the Stan model and for further post-processing:
 
@@
 All this information is passed to the function `hierarchical_param_stan_data` to produce inputs needed in the Stan model fitting.
-This function outputs a named list with Stan input data relevant to the hierarchical set up:
+This function outputs a named list with Stan input data relevant to the hierarchical setup:
 
@@
 
-For example, in this setting, we only have to estimate one $\sigma_{\mu}$, and no $\sigma_{\mu}$ is fixed, as recorded in the `n_sigma_fixed` and `n_sigma_estimate` elements of the Stan input data:
+For example, in this setting, we only have to estimate one $\sigma_{\mu}$, and no $\sigma_{\mu}$ is fixed, as recorded in the `n_sigma_estimate` and `n_sigma_fixed` elements of the Stan input data:
 
 ```{r}
+hier_stan_data$mu_n_sigma_estimate
 hier_stan_data$mu_n_sigma_fixed
-hier_stan_data$mu_n_sigma_estimate
 ```
 
-Excerpts from the Stan model are copied below to introduce model specification.
+Excerpts from the Stan model are copied below to introduce the model specification.
 The same Stan model is used for both global and local models.
@@
 
-Here we use that $(\mu_1, ..., \mu_c) = M\cdot (\eta_1, ..., \eta_c)$, where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_c)$ refers to the vector with products of standard deviation and $\gamma$ terms.
+Here we use that $(\mu_1, ..., \mu_c) = M \cdot (\eta_1, ..., \eta_c)$, where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_c)$ refers to the vector with products of standard deviation and $\gamma$ terms.
 The function `get_mu` calculates the vector of $\mu_c$s from the vector of $\eta$s (`mu_star` in the Stan model) and a decomposition of $M$ into different parts $U$, $V$, $W$.
@@
 The `localhierarchy` Stan function `get_mu_star` is used to obtain the $\eta$s.
-Its arguments also include prior parameters `mu_scalarprior_mean` and `mu_scalarprior_sd`, which refer to the (rescaled) prior mean and standard deviation of $\mu^{(global)}$.
-The model block contains the priors for the $\gamma$s and $\sigma_{\mu}$s, given by \begin{align*}
-\mu^{(global)} &\sim N(0,1),\\
+Its arguments also include prior parameters `mu_scalarprior_mean` and `mu_scalarprior_sd`, which refer to the (rescaled) prior mean and standard deviation of $\mu^{(\glob)}$.
+The model block contains the priors for the $\gamma$s and $\sigma_{\mu}$s, given by
+\begin{align*}
+  \mu^{(\glob)} &\sim N(0,1),\\
   \sigma_{\mu} &\sim \text{Half-Normal}(0, s_{\mu}).
-\end{align*} All prior parameters are added as input data to the Stan model.
+\end{align*} 
+All prior parameters are added as input data to the Stan model.
 
@@
 ```{r}
-global_fit <- fit_model_localhierarchy(runstep = "global_national",
+global_fit <- fit_model_localhierarchy(
+  runstep = "global_national",
   hierarchical_level = hierarchical_level,
   area = "level1",
-                                       survey_df = dat_nat)
+  survey_df = dat_nat
+)
 ```
@@
 ```{r}
-global_fit$post_summ %>% 
+global_fit$post_summ |> 
   filter(variable_no_index == "mu_sigma")
@@
 
-We can do a local fit by passing on the global fit object to the wrapper function and including data for countries of interest.
+A local fit can be esetimated by passing on the global fit object to the wrapper function and including data for countries of interest.
 Here we do local runs for all countries, by including the entire data set, and specify what to fix using the `runstep` argument:
@@
 ```{r}
-fit_local <- fit_model_localhierarchy(runstep = "local_national",
+fit_local <- fit_model_localhierarchy(
+  runstep = "local_national",
   global_fit = global_fit,
   area = "level1",
-                                      survey_df = dat_nat)
+  survey_df = dat_nat
+)
 ```
@@
 
-Above column `quant` refers to posterior means and upper and lower bounds of 95% credible intervals. We can create plots to show outputs, for example for the first 4 countries:
+The column `quant` above refers to posterior means and upper and lower bounds of 95% credible intervals. We can create plots to show outputs, for example for the first 4 countries:
 
 ```{r}
-plot_posterior_summaries(res = res_global, modelname1 = "global",
+plots <- plot_posterior_summaries(
+  res = res_global, 
+  modelname1 = "global",
   hierarchy_select = "level1",
   areas_select = unique(res_local$level1$name)[1:4]
-                                      )$level1
+)
+
+plots$level1
 ```
@@
 ```{r}
-plot_posterior_summaries(res = res_local, modelname1 = "local",
-                         res2 = res_global, modelname2 = "global", 
+plots <- plot_posterior_summaries(
+  res = res_local, 
+  modelname1 = "local",
+  res2 = res_global, 
+  modelname2 = "global", 
   hierarchy_select = "level1",
   areas_select = unique(res_local$level1$name)[1:4]
-                                      )$level1
+)
+
+plots$level1
 ```
@@
 ```{r}
-plot_posterior_summaries(res = res_local, modelname1 = "local",
-                         res2 = res_global, modelname2 = "global")$intercept 
+plots <- plot_posterior_summaries(
+  res = res_local, 
+  modelname1 = "local",
+  res2 = res_global, 
+  modelname2 = "global"
+)
+
+plots$intercept 
 ```
@@
 ### Workflow for nationally-informed subnational estimation
+
 For many global health applications, subnational estimation is of interest to inform policy and programmatic decision making at a more local level.
 If data are limited, models can be considered to share information across regions.
-We consider here the use of hierarchical models for a parameter of interest.
+We consider here the use of hierarchical models for an indicator of interest.
 The `localhierarchy` package can be used to fit subnational hierarchical models, following the same principles as for national estimation or through an updated approach.
@@
 
-In addition to a standard approach to subnational modeling, functionality in the `localhierarchy` package can also be used to fit subnational models that are informed by national level data and estimation, referred to here as "nationally-informed subnational estimation (NISE)".
+In addition to a standard approach to subnational modeling, functionality in the `localhierarchy` package can also be used to fit subnational models that are informed by national level data and estimation, referred to here as _nationally-informed subnational estimation_ (NISE).
 This approach is useful when subnational data are limited, for example, in terms of the number of countries with subnational data, the number of subnational regions with data, the number of observations per subnational region, or due to uncertainty associated with subnational data.
@@
    global_fit = global_fit, # global national fit
-   add_subnational_hierarchy = area_subnat
    # what level to add to the hierarchy, compared to the 
    # global national model 
+   add_subnational_hierarchy = area_subnat
   )
@@
 ```{r}
-global_fit$post_summ %>%
+global_fit$post_summ |>
   filter(str_detect(variable_no_index, "sigma"))
-global_fitsubnat_seq$post_summ %>%
+global_fitsubnat_seq$post_summ |>
   filter(str_detect(variable_no_index, "sigma"))
@@
 ```{r}
-fit_local_seq <- fit_model_localhierarchy(runstep = "local_subnational",
+fit_local_seq <- fit_model_localhierarchy(
+  runstep = "local_subnational",
   area = area_subnat,
   survey_df = dat_subnat,
-                                   global_fit = global_fitsubnat_seq)
+  global_fit = global_fitsubnat_seq
+)
 ```
 
-Similarly to the local national run, in subnational local runs, hierarchical levels and what is fixed is set in the wrapper function; we estimate subnational parameters only.
+Similarly to the local national run, in subnational local runs, hierarchical levels and which parameters are fixed is set in the wrapper function; we estimate subnational parameters only.
 Point estimates in the subnational global fit include point estimates from the national global run as well as point estimates from the global subnational fit.
@@
   fit = global_fitsubnat_seq,
-  parname = "mu")
+  parname = "mu"
+)
+
 res_local_seq <- posterior_summary_hierparam(
   fit = fit_local_seq, 
-  parname = "mu")
+  parname = "mu"
+)
 ```
@@
 ```{r}
-plot_posterior_summaries(res = res_local_seq, modelname1 = "local",
+plots <- plot_posterior_summaries(
+  res = res_local_seq, 
+  modelname1 = "local",
   res2 = res_globalsubnat_seq, modelname2 = "global",
@@
   areas_select = unique(res_local_seq$level2$name)[1:30]
-                          )$level2
+)
+
+plots$level2
 ```

### 2025-09-12 paper/paper.qmd
update case studies, add diagrams

@@
 library(posterior) # else error for as_draws_list
+library(tidybayes)
 devtools::load_all(here::here())
@@
 
-see overleaf
+<!-- % context: global health estimation exercises, use of bayesian hier models, examples -->
+In global health estimation, Bayesian hierarchical models have become a cornerstone methodology, enabling the integration of diverse data sources to produce reliable estimates across multiple populations and regions.
+For example, they are used to estimate family planning indicators for the FP2030 initiative (ref) and maternal mortality by the United Nations Interagency Group (ref).
+TO DO.
+In these efforts, national or subnational estimates are often required for population-periods with limited data availability.
+Bayesian hierarchical models—whether used directly or for specific components within larger modeling frameworks—are valuable in these contexts because they facilitate the sharing of information across populations.
+As a result, global modeling exercises typically fit Bayesian hierarchical models to comprehensive global databases, producing national-level estimates for all countries worldwide.
+
+<!-- % introduce local models and how they comapre to global model, why they are needed -->
+To facilitate local-level analysis, it is often desirable to use models that are informed by global models but can be fitted to just a subset of the data, such as data from a single country.
+We refer to these as local models: models derived from global hierarchical models but adapted to be fitted using data from only one population.
+Beyond allowing model fitting to smaller, specific datasets, local models offer practical advantages such as faster computation, making it feasible for users to perform analyses quickly on their own computers.
+Additionally, local models enable estimates for different populations to be updated independently, for example, when new data become available.
+This approach has been used successfully in global estimation exercises for family planning and maternal mortality (ref, ref).
+
+Since local models are usually derived from global models, a key challenge in using both for global health estimation is synchronizing codebases and data flows between these approaches.
+This is particularly important for modeling approaches that involve hierarchical estimation of multiple parameters, where the hierarchical structures may differ across parameters.
+
+This paper introduces the \texttt{localhierarchy} R package designed to address these challenges by supporting the development and fitting of Bayesian hierarchical models in settings requiring both global and local estimation.
+The package provides R functions and Stan functions and model blocks
+<!-- citep{stan2023}  -->
+to facilitate specification of hierarchical models, to fit full global models, and to subsequently fit local models using fixed parameter values inferred from the global analysis.
+Through this framework, users can conveniently perform local inference informed by global context, streamlining workflows in global health estimation tasks.
+
+This article presents a practical introduction to the \texttt{localhierarchy} package along with illustrative case studies demonstrating its applicability.
+The article is organized as follows.
+The next section introduces the modeling approach and workflow used for estimating family planning indicators, motivating the need for the \texttt{localhierarchy} functionality of fitting global and local hierarchical models.
+The methods section introduces the package' functionality for the estimation of one parameter at the national and subnational level.
+Finally, we return to the case study of family planning estimation to present illustrative results of the application of the \texttt{localhierarchy} setup in this context.
+
+# The Family Planning Estimation Tool (FPET)
+
+The Family Planning Estimation Tool (FPET) is widely used to estimate key family planning indicators in countries and regions.
+Key indicators include modern contraceptive use, demand for family planning, and demand satisfied, referring to as the ratio of modern use to demand.
+Data on such indicators is available from household surveys and routine data systems.
+FPET estimates are obtained from fitting a Bayesian model to available data.
+An illustration of FPET estimates and available data is shown in Figure 1 for countries ccc and xxx.
+
+At its core, the Bayesian FPET model employs transition models to characterize the long-term trajectories of family planning indicators over time within a population.
+The transition model describes changes in an indicator in terms of three principal parameters: the rate of change, the timing of the transition, and the asymptotic level (the maximum attainable level for an indicator).
+To allow for population-specific variation in how family planning indicators evolve, the parameters of the transition model are population-specific.
+Given limited country-specific information on some parameters, Bayesian hierarchical models are used to share information between populations.
+For example, the asymptotic level of demand is not yet known for the countries in Figure 1.
+Data from other countries inform this estimate using a hierarchical structure as summarized in Figure 2.
+Estimates of asymptotes for a selection of countries is shown in Figure 3.
+
+<!-- These structures are illustrated in Figure 1, showing how countries are grouped within regions, and regions are grouped in major world regions. -->
+
+```{=latex}
+\begin{figure}[h]
+\begin{adjustbox}{max size={\textwidth}{\textheight},center}
+%\resizebox{\textwidth}{\textwidth}{
+\begin{tikzpicture}[%scale=0.8,
+  box/.style={rectangle, draw, rounded corners, minimum width=2.6cm, minimum height=1cm, align=center},
+  level0/.style={box, fill=world!40},
+  level1/.style={box, fill=blue!20},
+  level2/.style={box, fill=purple!20},
+  level3/.style={box, fill=country!40, minimum width=2.2cm},
+    level4/.style={box, fill=data!20, minimum width=2.2cm},
+  arrow/.style={-{Latex[length=3mm]}, thick},
+  faintarrow/.style={-{Latex[length=2mm]}, thick, opacity=0.4},
+  background/.style 2 args={rectangle, rounded corners, draw=none, fill=#1, fill opacity=0.15, inner sep=0.5cm, fit=#2}
+]
+
+% --- Level 0 ---
+\node[level0] (mu0) {World};
+
+% --- Level 1 ---
+\node[level1, below left=1cm and 3cm of mu0] (mu11) {Sub-Saharan Africa};
+\node[level1, right=of mu11] (mu12) {South-East Asia};
+\node[level1, right=of mu12] (dots1) {$\dots$};
+% --- Level 2 ---
+\node[level2, below left=1cm and 0cm of mu11] (mu21) {Western Africa};
+\node[level2, right=of mu21] (mu22) {Eastern Africa};
+\node[level2, right=of mu22] (dots2) {$\dots$};
+% % --- Level 3 ---
+\node[level3, below left=1cm and 0cm of mu21] (mu31) {Senegal};
+\node[level3, right=of mu31] (mu32) {Burkina Faso};
+\node[level3, right=of mu32] (dots3) {$\dots$};
+% % % --- Level 4 ---
+% \node[level4, below left=1cm and 0cm of mu31] (mu41) {Data point 1};
+% \node[level4, right=of mu41] (mu42) {Data point 2};
+% \node[level4, right=of mu42] (dots4) {$\dots$};
+% --- Background blocks ---
+\begin{pgfonlayer}{background}
+  \node[background={world!40}{(mu0)}] {};
+  \node[background={blue!20}{(mu11)(mu12)(dots1)}] (bg1) {};
+  \node[background={pink!60}{(mu21)(mu22)(dots2) }] (bg2) {};
+ \node[background={country!40}{(mu31)(mu32)(dots3)}] (bg3) {};
+ % \node[background={data!40}{(mu41)(mu42)(dots4)}] (bg4) {};
+\end{pgfonlayer}
+
+% Sigma labels inside background blocks
+\node[anchor=north west] at (bg1.east) {Major world regions};
+\node[anchor=north west] at (bg2.east) {Smaller world regions};
+\node[anchor=north west] at (bg3.east) {Countries};
+%\node[anchor=north west] at (bg4.east) %{Country data};
+% --- Arrows ---
+\draw[arrow] (mu0.south) -- (mu11.north);
+\draw[arrow] (mu0.south) -- (mu12.north);
+\draw[arrow] (mu0.south) -- (dots1.north);
+
+\draw[arrow] (mu11.south) -- (mu21.north);
+\draw[arrow] (mu11.south) -- (mu22.north);
+\draw[arrow] (mu11.south) -- (dots2.north);
+
+\draw[arrow] (mu21.south) -- (mu31.north);
+\draw[arrow] (mu21.south) -- (mu32.north);
+\draw[arrow] (mu21.south) -- (dots3.north);
+
+% \draw[arrow] (mu31.south) -- (mu41.north);
+% \draw[arrow] (mu31.south) -- (mu42.north);
+% \draw[arrow] (mu31.south) -- (dots4.north);
+\end{tikzpicture}%}
+\end{adjustbox}
+\caption{Illustration of hierarchical grouping of countries in world regions.}
+\end{figure}
+```
+
+The workflow used in FPET to produce country-level estimates is as follows.
+Firstly, the full Bayesian model is fitted to a global data base of survey data.
+Data base compilation is carried out by the United Nations Population Division (ref) and the Track20 project (ref).
+The global model fit provides information on model parameters, including hierarchical parameters, e.g., estimates of asymptotes.
+Subsequently, a local model can be used to produce FP estimates based on the data from one country, using information from the global model fit (explained further below).
+As part of the Track20 project, local model fitting is performed by country monitoring and evaluation officers using an online tool (ref).
+At this step, in addition to fitting models to country survey data, as done in the global modeling, users can incorporate additional data sources such as routine information system records (e.g., the number of contraceptive commodities supplied, add REF).
 
@@
 
-The modeling approach introduced in `localhierarchy` was motivated by the hierarchical structures associated with national and subnational level estimation.
-These structures are illustrated in Figure 1, showing how countries are grouped within regions, and regions are grouped in major world regions.
-
+The modeling approach introduced in `localhierarchy` was motivated by application like FPET that depend on hierarchical structures associated with national and subnational level estimation, requiring global and local models to be fit.
 In this section, we introduce the principle setup used for global and local hierarchical models in the context of national estimation.
@@
 Applications and an extension to subnational estimation are included in the use cases.
-
-Figure 1: Illustration of hierarchical grouping of countries in world regions.
 
@@
 
-Figure 2: General specification of a hierarchical model.
+```{=latex}
+\begin{figure}[h]
+\begin{adjustbox}{max size={\textwidth}{\textheight},center}
+%\resizebox{\textwidth}{\textwidth}{
+\begin{tikzpicture}[
+  box/.style={rectangle, draw, rounded corners, minimum width=2.6cm, minimum height=1cm, align=center},
+  level0/.style={box, fill=yellow!40},
+  level1/.style={box, fill=blue!20},
+  level2/.style={box, fill=purple!20},
+  level3/.style={box, fill=orange!40, minimum width=2.2cm},
+   level4/.style={box, fill=green!20, minimum width=2.2cm},
+  arrow/.style={-{Latex[length=3mm]}, thick},
+  faintarrow/.style={-{Latex[length=2mm]}, thick, opacity=0.4},
+  background/.style 2 args={rectangle, rounded corners, draw=none, fill=#1, fill opacity=0.15, inner sep=0.5cm, fit=#2}
+]
 
-As a specific example, we consider the estimation of country-specific parameters, where countries are organized into subregions of countries (illustrated in Figure 1).
+% --- Level 0 ---
+\node[level0] (mu0) {$\mu^{(0)} \sim \mathcal{N}(0,s_\mu^2)$};
+
+% --- Level 1 ---
+\node[level1, below left=2cm and 3cm of mu0] (mu11) {$\mu^{(1)}_1 \sim \mathcal{N}(\mu^{(0)},\sigma_{\mu,1}^2)$};
+\node[level1, right=of mu11] (mu12) {$\mu^{(1)}_2$};
+\node[level1, right=of mu12] (dots1) {$\dots$};
+
+% --- Level 2 ---
+\node[level2, below left=2cm and 0cm of mu11] (mu21) {$\mu^{(2)}_1 \sim \mathcal{N}(\mu^{(1)}_1,\sigma_{\mu,2}^2)$};
+\node[level2, right=of mu21] (mu22) {$\mu^{(2)}_2$};
+\node[level2, right=of mu22] (dots2) {$\dots$};
+
+% % --- Level 3 ---
+\node[level3, below left=2cm and 0cm of mu21] (mu31) {$\mu^{(3)}_1 \sim \mathcal{N}(\mu^{(2)}_1,\sigma_{\mu,3}^2)$};
+\node[level3, right=of mu31] (mu32) {$\mu^{(3)}_2$};
+\node[level3, right=of mu32] (dots3) {$\dots$};
+
+% % % --- Level 4 ---
+% \node[level4, below left=2cm and 0cm of mu31] (mu41) {$\mu^{(4)}_1 \sim \mathcal{N}(\mu^{(3)}_1,\sigma_{\mu,4}^2)$};
+% \node[level4, right=of mu41] (mu42) {$\mu^{(4)}_2$};
+% \node[level4, right=of mu42] (dots4) {$\dots$};
+
+% --- Background blocks ---
+\begin{pgfonlayer}{background}
+  \node[background={yellow!40}{(mu0)}] {};
+  \node[background={blue!20}{(mu11)(mu12)(dots1)}] (bg1) {};
+  \node[background={pink!60}{(mu21)(mu22)(dots2) }] (bg2) {};
+ \node[background={orange!40}{(mu31)(mu32)(dots3)}] (bg3) {};
+% \node[background={green!40}{(mu41)(mu42)(dots4)}] (bg4) {};
+\end{pgfonlayer}
+
+% Sigma labels inside background blocks
+\node[anchor=north west] at (bg1.east) {$\sigma_{\mu,1}$};
+\node[anchor=north west] at (bg2.east) {$\sigma_{\mu,2}$};
+\node[anchor=north west] at (bg3.east) {$\sigma_{\mu,3}$};
+%\node[anchor=north west] at (bg4.east) {$\sigma_{\mu,4}$};
+% --- Arrows ---
+\draw[arrow] (mu0.south) -- (mu11.north);
+\draw[arrow] (mu0.south) -- (mu12.north);
+\draw[arrow] (mu0.south) -- (dots1.north);
+
+\draw[arrow] (mu11.south) -- (mu21.north);
+\draw[arrow] (mu11.south) -- (mu22.north);
+\draw[arrow] (mu11.south) -- (dots2.north);
+
+\draw[arrow] (mu21.south) -- (mu31.north);
+\draw[arrow] (mu21.south) -- (mu32.north);
+\draw[arrow] (mu21.south) -- (dots3.north);
+
+% \draw[arrow] (mu31.south) -- (mu41.north);
+% \draw[arrow] (mu31.south) -- (mu42.north);
+% \draw[arrow] (mu31.south) -- (dots4.north);
+\end{tikzpicture}
+\end{adjustbox}
+\caption{General specification of a hierarchical model.}
+\end{figure}
+```
+
+As a specific example, we consider the estimation of country-specific parameters, where countries are organized into subregions of countries (illustrated in Figure 2).
 In this setting, using alternative notation for the same setup to indicate the level by name, the hierarchical model can be written as follows:
@@
 
-Figure 3: Workflow for global and local national modeling.
+```{=latex}
+\begin{figure}[h]
+\begin{adjustbox}{max size={\textwidth}{\textheight},center}
+\begin{tikzpicture}[
+  box/.style={rectangle, draw, thick, rounded corners, minimum width=6cm, align=left, inner sep=6pt},
+  databox/.style={box, fill=data!20},
+  modelbox/.style={box, fill=param!20},
+  arrowdata/.style={-{Latex[length=3mm]}, thick, data},
+  arrowparam/.style={-{Latex[length=3mm]}, thick, param!80},
+  font=\sffamily
+]
+% --- Nodes ---
+\node[databox] (globaldata) {Global data base of country data};
+\node[modelbox, below=1cm of globaldata] (globalmodel) {%
+  Global national modeling:\\
+  Estimate all parameters in hierarchical model, e.g., \\
+\(
+  \begin{aligned}
+\mu_w^{(region)} &\sim N(\mu^{(global)}, \sigma_{\mu, region}^2),\\
+\mu_s^{(subregion)} &\sim N(\mu_{w[s]}^{(region)}, \sigma_{\mu, subregion}^2), \\
+\mu_c^{(country)} &\sim N(\mu_{s[c]}^{(subregion)}, \sigma_{\mu, country}^2),  \end{aligned}
+\)
+};
+\node[modelbox, below=1cm of globalmodel] (localmodel) {%
+  Local national modeling:\\
+  Estimate national parameters only, \\i.e., $\mu_c^{(country)}$ with $\mu_c^{(country)} \sim N(\hat{\mu}_{s[c]}^{(subregion)}, \hat{\sigma}_{\mu, country}^2).$
+};
+
+% --- Data arrow (down then right) ---
+\draw[arrowdata] (globaldata.south) -- (globalmodel.north);
+\draw[arrowdata] (globaldata.east) ++(0,0) -- ++(4.5,0) |- (localmodel.east)
+node[midway,above] {Data for one country};
+
+% --- Parameter arrow ---
+\draw[arrowparam] (globalmodel.south) -- (localmodel.north);
+
+% --- Legend ---
+% \node[arrowparam, below=1.2cm of localmodel.west, anchor=west] (leg1) {}; 
+% \node[right=0.3cm of leg1] {\textcolor{param}{$\rightarrow$ Parameter estimates}};
+\node[below=1cm of localmodel](parameter) {\textcolor{param}{$\rightarrow$ Parameter estimates}};
+\node[right=1cm of parameter] {\textcolor{data}{$\rightarrow$ Data}};
+\end{tikzpicture}
+\end{adjustbox}
+\caption{Workflow for global and local national modeling. }
+\end{figure}
+```
 
@@
 
-# Use cases
+# Use case: National estimation
 
-Example 1 covers national estimation while example 2 illustrated subnational estimation.
-Both examples consider one parameter.
-The package also includes functionality for parameter vectors.
-We illustrate the package functionality using a simple example with simulated data, introduced in the following section.
+We present a detailed use case of the `localhierarchy` package using simulated data, followed by an application of that functionality in FPET.
 
-## Simulation setup
+## Detailed example using simulated data
 
-The simulation setup is summarized in Figure 4.
-It covers both national and subnational parameters and data.
-While for national level estimation alone, the subnational settings are not needed, we add them here to create a setup that can serve to illustrate both national as well as subnational-level estimation.
+### Simulation setup
+The simulation setup is summarized in Figure 4, panels A and B. Simulation of subnational data is discussed in a later section.
 
-Figure 4: Diagram of simulation settings.
+```{=latex}
+\begin{figure}[h]
+\fbox{\begin{adjustbox}{max size={\textwidth}{\textheight},center}
+\begin{tikzpicture}[
+  box/.style={rectangle, draw, rounded corners, minimum width=2cm, minimum height=1cm, align=center},
+  level0/.style={box, fill=yellow!40},
+  level1/.style={box, fill=orange!40},
+  level2/.style={box, fill=green!20},
+  arrow/.style={-{Latex[length=3mm]}, thick},
+  faintarrow/.style={-{Latex[length=2mm]}, thick, opacity=0.4},
+  background/.style 2 args={rectangle, rounded corners, draw=none, fill=#1, fill opacity=0.15, inner sep=0.5cm, fit=#2}
+]
+
+% --- Level 0 ---
+\node[level0] (mu0) {$\mu^{(0)} \sim \mathcal{N}(0,s_\mu^2)$};
+
+% --- Level 1 ---
+\node[level1, below left=1.5cm and 3cm of mu0] (mu11) {$\mu^{(1)}_1 \sim \mathcal{N}(\mu^{(0)},\sigma_{\mu,1}^2)$};
+\node[level1, right=of mu11] (mu12) {$\mu^{(1)}_2$};
+\node[level1, right=of mu12] (dots1) {$\dots$};
+% --- Level 2 ---
+\node[level2, below left=1.5cm and 0cm of mu11] (mu21) {$\mu^{(2)}_1 \sim \mathcal{N}(\mu^{(1)}_1,\sigma_{\mu,2}^2)$};
+\node[level2, right=of mu21] (mu22) {$\mu^{(2)}_2$};
+\node[level2, right=of mu22] (dots2) {$\dots$};
+% --- Background blocks ---
+\begin{pgfonlayer}{background}
+  \node[background={yellow!40}{(mu0)}] {};
+  \node[background={orange!40}{(mu11)(mu12)(dots1)}] (bg1) {};
+  \node[background={green!40}{(mu21)(mu22)(dots2) }] (bg2) {};
+\end{pgfonlayer}
+
+% Sigma labels inside background blocks
+\node[anchor=north west] at (bg1.east) {$\sigma_{\mu,1}$, country};
+\node[anchor=north west] at (bg2.east) {$\sigma_{\mu,2}$, subnational region};
+% --- Arrows ---
+\draw[arrow] (mu0.south) -- (mu11.north);
+\draw[arrow] (mu0.south) -- (mu12.north);
+\draw[arrow] (mu0.south) -- (dots1.north);
+
+\draw[arrow] (mu11.south) -- (mu21.north);
+\draw[arrow] (mu11.south) -- (mu22.north);
+\draw[arrow] (mu11.south) -- (dots2.north);
+\end{tikzpicture}
+\end{adjustbox}}
+
+% local model nat
+\fbox{\begin{adjustbox}{max size={\textwidth}{\textheight},center}
+\begin{tikzpicture}[
+  box/.style={rectangle, draw, rounded corners, minimum width=2cm, minimum height=1cm, align=center},
+  level0/.style={box, fill=yellow!40},
+  level1/.style={box, fill=orange!40},
+  level2/.style={box, fill=green!20},
+    data/.style={box, fill=gray!20},
+  arrow/.style={-{Latex[length=3mm]}, thick},
+  faintarrow/.style={-{Latex[length=2mm]}, thick, opacity=0.4},
+  background/.style 2 args={rectangle, rounded corners, draw=none, fill=#1, fill opacity=0.15, inner sep=0.5cm, fit=#2}
+]
+
+\node[level1] (mu0) {$\mu^{(1)}_1$};
+
+% --- Level 1 ---
+\node[data, below left=1.5cm and 3cm of mu0] (mu11) {$y^{(country)}_1 \sim \mathcal{N}(\mu^{(1)}_1,\sigma_{y}^2)$};
+\node[data, right=of mu11] (mu12) {$y^{(country)}_2$};
+\node[data, right=of mu12] (dots1) {$\dots$};
+% --- Background blocks ---
+\begin{pgfonlayer}{background}
+  \node[background={gray!40}{(mu11)(mu12)(dots1)}] (bg1) {};
+\end{pgfonlayer}
+
+% Sigma labels inside background blocks
+\node[anchor=north west] at (bg1.east) {$\sigma_{y}$, country data for first country};
+
+% --- Arrows ---
+\draw[arrow] (mu0.south) -- (mu11.north);
+\draw[arrow] (mu0.south) -- (mu12.north);
+\draw[arrow] (mu0.south) -- (dots1.north);
+\end{tikzpicture}
+\end{adjustbox}}
+
+% local model subnat
+\fbox{
+\begin{adjustbox}{max size={\textwidth}{\textheight},center}
+\begin{tikzpicture}[
+  box/.style={rectangle, draw, rounded corners, minimum width=2cm, minimum height=1cm, align=center},
+  level0/.style={box, fill=yellow!40},
+  level1/.style={box, fill=orange!40},
+  level2/.style={box, fill=green!20},
+    data/.style={box, fill=gray!20},
+  arrow/.style={-{Latex[length=3mm]}, thick},
+  faintarrow/.style={-{Latex[length=2mm]}, thick, opacity=0.4},
+  background/.style 2 args={rectangle, rounded corners, draw=none, fill=#1, fill opacity=0.15, inner sep=0.5cm, fit=#2}
+]
+
+% keep same colors so now color of level 1 at top 
+\node[level2] (mu0) {$\mu^{(2)}_1$};
+
+% --- Level 1 ---
+\node[data, below left=1.5cm and 3cm of mu0] (mu11) {$y^{(subnat)}_1 \sim \mathcal{N}(\mu^{(1)}_1,\sigma_{y}^2)$};
+\node[data, right=of mu11] (mu12) {$y^{(subnat)}_2$};
+\node[data, right=of mu12] (dots1) {$\dots$};
+% --- Background blocks ---
+\begin{pgfonlayer}{background}
+  \node[background={gray!40}{(mu11)(mu12)(dots1)}] (bg1) {};
+\end{pgfonlayer}
+
+% Sigma labels inside background blocks
+\node[anchor=north west] at (bg1.east) {$\sigma_{y}$, subnational data for first region};
+
+% --- Arrows ---
+\draw[arrow] (mu0.south) -- (mu11.north);
+\draw[arrow] (mu0.south) -- (mu12.north);
+\draw[arrow] (mu0.south) -- (dots1.north);
+\end{tikzpicture}
+\end{adjustbox}}
+\caption{Diagram of simulation settings.
 Top: Hierarchical model for generating country and subnational parameters.
 Center: Data generating mechanism for national data, illustrated for the first country.
-Bottom: Data generating mechanism for subnational data, illustrated for the first subnational region.
+Bottom: Data generating mechanism for subnational data, illustrated for the first subnational region.}
+\end{figure}
+\clearpage
+```
 
-The top box in Figure 4 indicates the hierarchical model considered.
+Panel A in Figure 4 indicates the hierarchical model considered.
 For national level estimation, we consider countries to be grouped in the world, i.e., $\mu_c|\mu^{(\glob)}, \sigma_{\mu, \country} \sim N(\mu^{(\glob)}, \sigma_{\mu, \country}^2)$.
-For subnational estimation, an additional level is added, with $\mu_r^{(\subnat)} = \mu_{c[r]} + \eta_{r}^{(\subnat)}$, $\eta_{r}^{(\subnat)}|\sigma_{\mu, \subnat} \sim N(0, \sigma_{\mu, \subnat})$.
-
+Panel B introduces the likelihood function for national data, given by a normal likelihood with constant variance: $$y_i^{(\country)}|\mu_{c[i]}, \sigma_y^2 \sim N(\mu_{c[i]}, \sigma_y^2),$$ where $c[i]$ refers to the country of country observation $i$.
 We chose simple normal densities as data generating mechanisms in the simulation, to focus the modeling on the hierarchical model and the `localhierarchy` functionality as opposed to the data model aspects of modeling.
@@
 
-For generating national level data (box 2), we assume a normal likelihood with constant variance: $$y_i^{(\country)}|\mu_{c[i]}, \sigma_y^2 \sim N(\mu_{c[i]}, \sigma_y^2),$$ where $c[i]$ refers to the country of country observation $i$.
-Again, the use of a normal likelihood with constant variance is for illustration purposes only.
-In practice, the data model can be set by the user to reflect the observational error.
-We also generate subnational data following a normal distribution: $$y^{(subnat)}_i|\mu^{(subnat)}_{g[i]}, \sigma_y^2 \sim N(\mu^{(subnat)}_{r[i]}, \sigma_y^2),$$ where $g[i]$ refers to the subnational region of subnational observation $i$.
-In this simulation setup, we generate subnational data independently of national-level data, and we use the same constant data variance at the national and subnational levels.
-In real examples, data variances are likely to differ (for example, because subnational sampling errors are larger at the subnational level) and national and subnational data may be derived from the same source.
-We discuss an example of those settings in the case study.
-
-Simulated data can be generated using  `simulate_multilevel_data` function.
-We simulate a set of model parameters, setting $\mu^{(\glob)} = 0$, $\sigma_{\mu,\country} = 0.5$, $\sigma_{\mu}^{(\subnat)} = 0.2$, creating 30 countries with 20 subnational regions in each country.
-We then simulate subnational data, using $\sigma_y = 0.1$ and generating 10 observations per region.
-Finally, the function adds national data as well using the argument `add_data1levelup`, to output the subnational and national level data sets:
+Simulated data were generated using  `simulate_multilevel_data` function.
+We simulate a set of model parameters, setting $\mu^{(\glob)} = 0$, and $\sigma_{\mu,\country} = 0.5$, creating 30 countries, and national data, using $\sigma_y = 0.1$:
 
@@
 )
-
-dat_subnat <- dat_all$data
 dat_nat <- dat_all$data_1levelup
@@
 
-In the simulated data, `level1` refers to the country level and `level2` refers to the subnational level. The data sets contain the simulated parameters (`eta`) and data (`y`):
+In the simulated data, `level1` refers to the country level. The data sets contain the simulated parameters (`eta`) and data (`y`):
 
 ```{r}
-dat_subnat |> select(level1, level2, mu_global, level1_eta, level2_eta, y)
 dat_nat |> select(level1,  mu_global, level1_eta, y)
 ```
-
-## Use case 1: National estimation
 
@@
 
-## Use case 2: Subnational estimation
+## Application: national estimation of family planning indicators using FPET
 
-### Workflow for nationally-informed subnational estimation
+TO ADD
+
+# Use case 2: Subnational estimation
+
+## Workflow for nationally-informed subnational estimation
 
@@
 The `localhierarchy` package can be used to fit subnational hierarchical models, following the same principles as for national estimation or through an updated approach.
-Just as for national-level hierarchical modeling, the starting point for subnational hierarchical modeling is to specify a hierarchical modeling structure for the subnational regions, such as the one from Figure 1.
+Just as for national-level hierarchical modeling, the starting point for subnational hierarchical modeling is to specify a hierarchical modeling structure for the subnational regions, such as the one from Figure 2.
 When following the same principles as for national-level hierarchical modeling, a global model can be fitted to a global data base of subnational data, followed by local models for subnational regions.
@@
 
-The approach followed in NISE is illustrated in Figure 4, for the hierarchical structure from Figure 1.
+The approach followed in NISE is illustrated in Figure 4, for the hierarchical structure from Figure 2.
 In the NISE approach, the first step is to fit a model to a global data base with national-level data, to obtain estimates of model parameters up to and including national level parameters.
@@
 Finally, in a 3rd step, local models can be fitted, using fixed values from the global subnational fit.
-In the remainder of this section, we illustrate this approach, continuing with the simulated data from the earlier example.
-Note that national level data can also inform subnational estimates through inclusion in subnational global or local fits, as indicated in the NISE diagram in Figure 4.
-This is discussed further in the case study.
+In the remainder of this section, we illustrate this approach using simulated data, and present an application of this approach for family planning estimation.
 
-Figure 5: Workflow for nationally-informed global and local subnational modeling.
+```{=latex}
+\begin{figure}[h]
+\begin{adjustbox}{max size={\textwidth}{\textheight},center}
+\begin{tikzpicture}[
+  box/.style={rectangle, draw, thick, rounded corners, minimum width=2.5cm, align=left, inner sep=6pt},
+  eqbox/.style={box},
+  databox/.style={box, fill=data!20},
+  modelbox/.style={box, fill=param!20},
+  arrowdata/.style={-{Latex[length=3mm]}, thick, data},
+  arrowparam/.style={-{Latex[length=3mm]}, thick, param!80},
+  font=\sffamily
+]
+% --- Nodes ---
+\node[eqbox](model){
+Example subnat. hierarchical model:\\
+\(
+  \begin{aligned}
+\mu_w^{(region)} &\sim N(\mu^{(global)}, \sigma_{\mu, region}^2),\\
+\mu_s^{(subregion)} &\sim N(\mu_{w[s]}^{(region)}, \sigma_{\mu, subregion}^2), \\
+\mu_c^{(country)} &\sim N(\mu_{s[c]}^{(subregion)}, \sigma_{\mu, country}^2),  \\
+\mu_r^{(subnat)} &\sim N(\mu_{r[c]}^{(country)}, \sigma_{\mu, subnat}^2),  \\
+\end{aligned}
+\)
+};
+
+\node[databox, right=5cm of model.north] (globaldata) {Global data base of country data};
+\node[modelbox, below=1cm of globaldata] (globalmodel) {%
+  Global national modeling:\\
+  Estimate all parameters in hierarchical \\ model 
+  $\leq$ country level 
+};
+\node[modelbox, below=1cm of globalmodel] (subglobalmodel) {%
+  Global subnational modeling:\\
+  Estimate subset of parameters in \\ hierarchical model, 
+  i.e., $\sigma_{\mu, subnat}$
+};
+\node[databox, left=1cm of subglobalmodel] (globalsubdata) {Global data base\\ of subnational data};
+\node[modelbox, below=1cm of subglobalmodel] (localmodel) {%
+  Local subnational modeling:\\
+  Estimate subnational parameters only, \\i.e., $\mu_r$ in $\mu_r \sim N(\hat{\mu}_{c[r]}^{(country)}, \hat{\sigma}_{\mu, subnat}^2).$
+};
+
+% --- Data arrow  ---
+\draw[arrowdata] (globaldata.south) -- (globalmodel.north);
+\draw[arrowdata] (globaldata.east) ++(0,0) -- ++(1.5,0) |- (localmodel.east)
+node[midway,above] {};
+\draw[arrowdata] (globaldata.east) ++(0,0) -- ++(1.5,0) |- (subglobalmodel.east)
+node[midway,above] {};
+\draw[arrowdata] (globalsubdata.east) -- (subglobalmodel.west);
+\draw[arrowdata] (globalsubdata.south) -- (localmodel.west)
+node[midway,above] {};
+
+% --- Parameter arrow ---
+\draw[arrowparam] (globalmodel.south) -- (subglobalmodel.north);
+\draw[arrowparam] (subglobalmodel.south) -- (localmodel.north);
+
+% --- Legend ---
+% \node[arrowparam, below=1.2cm of localmodel.west, anchor=west] (leg1) {}; 
+% \node[right=0.3cm of leg1] {\textcolor{param}{$\rightarrow$ Parameter estimates}};
+\node[below=1.2cm of localmodel.west](parameter) {\textcolor{param}{$\rightarrow$ Parameter estimates}};
+\node[right=1cm of parameter] {\textcolor{data}{$\rightarrow$ Data}};
+\end{tikzpicture}
+\end{adjustbox}
+\caption{Workflow for nationally-informed global and local subnational modeling. }
+\end{figure}
+\clearpage
+```
+
+## Detailed example of subnational estimation using simulated data
+
+### Simulation setup
+The simulation setup for subnational parameters and data is summarized in Figure 4, Panels A-B-C.
+As explained above for national estimation, Panel A Figure 4 indicates the hierarchical model considered.
+For subnational estimation, an additional level is added, with $\mu_r^{(\subnat)} = \mu_{c[r]} + \eta_{r}^{(\subnat)}$, $\eta_{r}^{(\subnat)}|\sigma_{\mu, \subnat} \sim N(0, \sigma_{\mu, \subnat})$.
+
+As explained earlier as well, we chose simple normal densities as data generating mechanisms in the simulation, to focus the modeling on the hierarchical model and the `localhierarchy` functionality as opposed to the data model aspects of modeling.
+Subnational data follows a normal distribution with constant variance: $$y^{(subnat)}_i|\mu^{(subnat)}_{g[i]}, \sigma_y^2 \sim N(\mu^{(subnat)}_{r[i]}, \sigma_y^2),$$ where $g[i]$ refers to the subnational region of subnational observation $i$.
+In this simulation setup, for simplicity, we generate subnational data independently of national-level data, and we use the same constant data variance at the national and subnational levels.
+In real examples, data variances are likely to differ (for example, because subnational sampling errors are larger at the subnational level) and national and subnational data may be derived from the same source.
+We discuss an example of those settings in the case study.
+
+Simulated data at the subnational level was generated using the   `simulate_multilevel_data` function described above.
+For subnational levels, we set  $\sigma_{\mu}^{(\subnat)} = 0.2$, and we created 20 subnational regions in each country and 10 observations per region.
+In the simulated data, `level1` refers to the country level and `level2` refers to the subnational level.
+The data sets contain the simulated parameters (`eta`) and data (`y`):
+
+```{r}
+dat_subnat <- dat_all$data
+dat_subnat |> select(level1, level2, mu_global, level1_eta, level2_eta, y)
+```
 
@@
 
-# Usage in other projects
+## Application: subnational estimation of family planning indicators using FPET
 
-see overleaf
+For subnational estimation, the workflow used in FPET mirrors the approach described in this paper: a global national model is fitted first, followed by a global subnational model, and finally by local fits.
+This sequential approach, in which the subnational global model is based on the national global model, was motivated by differences in data availability between national and subnational global databases.
+For example, when estimating family planning indicators for married women of reproductive age, the global national database contains XX observations from XXX countries, whereas the subnational database includes YY observations from YYY countries.
+
+ADD
 

### 2025-09-18 paper/paper.qmd
update paper and add fpet applications

@@
 title: "`localhierarchy`: an R package to facilitate fitting of global and local Bayesian hierarchical models"
-author: "Leontine, Herb, Shauna, Evan"
+author: "Leontine, Shauna, Evan, Herb"
 format: 
@@
 
-  This article presents a practical introduction to the package for the applied user and illustrates the package’s functionality through several example use cases.
+  This article presents a practical introduction to the package for the applied user and illustrates the package’s functionality through examples for national and subnational estimation.
 date: last-modified
@@
 library(tidybayes)
+library(patchwork)
 devtools::load_all(here::here())
@@
 <!-- % context: global health estimation exercises, use of bayesian hier models, examples -->
+
 In global health estimation, Bayesian hierarchical models have become a cornerstone methodology, enabling the integration of diverse data sources to produce reliable estimates across multiple populations and regions.
-For example, they are used to estimate family planning indicators for the FP2030 initiative (ref) and maternal mortality by the United Nations Interagency Group (ref).
-TO DO.
+For example, they are used to estimate family planning indicators for the FP2030 initiative (ref), low birth weight (ref), and maternal mortality by the United Nations Interagency Group (ref).
 In these efforts, national or subnational estimates are often required for population-periods with limited data availability.
@@
 <!-- % introduce local models and how they comapre to global model, why they are needed -->
+
 To facilitate local-level analysis, it is often desirable to use models that are informed by global models but can be fitted to just a subset of the data, such as data from a single country.
@@
 This paper introduces the \texttt{localhierarchy} R package designed to address these challenges by supporting the development and fitting of Bayesian hierarchical models in settings requiring both global and local estimation.
-The package provides R functions and Stan functions and model blocks
-<!-- citep{stan2023}  -->
-to facilitate specification of hierarchical models, to fit full global models, and to subsequently fit local models using fixed parameter values inferred from the global analysis.
+The package provides R functions and Stan functions and model blocks <!-- citep{stan2023}  --> to facilitate specification of hierarchical models, to fit full global models, and to subsequently fit local models using fixed parameter values inferred from the global analysis.
 Through this framework, users can conveniently perform local inference informed by global context, streamlining workflows in global health estimation tasks.
@@
 FPET estimates are obtained from fitting a Bayesian model to available data.
-An illustration of FPET estimates and available data is shown in Figure 1 for countries ccc and xxx.
-
-At its core, the Bayesian FPET model employs transition models to characterize the long-term trajectories of family planning indicators over time within a population.
+An illustration of FPET estimates and available data is shown in @fig-localfits for three countries in Western Africa.
+
+```{r}
+#| fig-height: 9
+#| fig-width: 7
+#| echo: false
+#| label: fig-localfits
+#| fig-cap: "Data, estimates, and forecasts from 1970 until 2030 for FP indicators for married women in Benin (BEN), Gambia (GMB), and Ghana (GHA).
+Graphs show mCPR, unmet need for modern methods, demand, and demand satisfied with modern methods.
+Survey data are shown in colored dots with 95% confidence intervals that reflect total error variance.
+FPET estimates are shown in black, with grey shaded areas representing 90% credible intervals."
+plots <- readRDS("outputfpet_localfits.rds")
+plots[[1]] /  plots[[2]]/ plots[[3]] 
+```
+
+At its core, the Bayesian FPET model employs transition models for demand and demand satisfied to characterize the long-term trajectories of family planning indicators over time within a population, illustrated in @fig-transition.
 The transition model describes changes in an indicator in terms of three principal parameters: the rate of change, the timing of the transition, and the asymptotic level (the maximum attainable level for an indicator).
 To allow for population-specific variation in how family planning indicators evolve, the parameters of the transition model are population-specific.
-Given limited country-specific information on some parameters, Bayesian hierarchical models are used to share information between populations.
-For example, the asymptotic level of demand is not yet known for the countries in Figure 1.
-Data from other countries inform this estimate using a hierarchical structure as summarized in Figure 2.
-Estimates of asymptotes for a selection of countries is shown in Figure 3.
-
-<!-- These structures are illustrated in Figure 1, showing how countries are grouped within regions, and regions are grouped in major world regions. -->
+
+![Illustration of transition model used in the modeling of family planning indicators. TO DO (if kept): make specific to this text, ie name asymptote, level](fig/transition.png){#fig-transition}
+
+Given limited country-specific information on some parameters, Bayesian hierarchical models are used to share information between populations, using hierarchical structures such as the one illustrated in Figure \ref{fig-hierstruc}.
+Specifically for family planning estimation and forecasting, a hierarchical model is used to estimate the asymptote of the transition of demand from low to high values, and the level of demand in a reference year.
+@fig-level shows estimates of the level in the reference year at the country level for countries in Western Africa, and for the groupings used in the hierarchical model.
+@fig-asympt shows estimates for the asymptote.
+These figures show variation in levels among countries in Western Africa, informed by country-specific data.
+Estimates of asymptotes are similar across countries in Western Africa because countries have not yet reached the end phase of their transition yet; the asymptotes in Western African countries are informed by the estimate of the asymptote in sub-Saharan Africa through the hierarchical model.
+
+<!-- figure hier structure  -->
 
@@
 \end{adjustbox}
-\caption{Illustration of hierarchical grouping of countries in world regions.}
+\caption{Illustration of hierarchical grouping of countries in world regions.}\label{fig-hierstruc}
 \end{figure}
+```
+
+```{r, echo = FALSE}
+plots <- readRDS("outputfpet_params.rds")
+```
+
+```{r}
+#| fig-height: 9
+#| fig-width: 7
+#| echo: false
+#| label: fig-level
+#| fig-cap: "Estimates of the level of demand for countries in Western Africa and for higher levels used in the hierarchical model (smaller world regions and major world regions).
+Point estimates and 95% credible intervals are shown.
+TO DO: fix formatting, show country names"
+plots$plot_iso_omega /  plots$plot_reg_omega[[3]]/ plots$plot_reg_omega[[2]]
+```
+
+```{r}
+#| fig-height: 9
+#| fig-width: 7
+#| echo: false
+#| label: fig-asympt
+#| fig-cap: "Estimates of the asymptote for demand for countries in Western Africa and for higher levels used in the hierarchical model (major world regions).
+Point estimates and 95% credible intervals are shown.
+TO DO: fix formatting, show country names"
+plots$plot_iso_ptilde /  plots$plot_reg_ptilde[[2]] 
 ```
@@
 The global model fit provides information on model parameters, including hierarchical parameters, e.g., estimates of asymptotes.
-Subsequently, a local model can be used to produce FP estimates based on the data from one country, using information from the global model fit (explained further below).
-As part of the Track20 project, local model fitting is performed by country monitoring and evaluation officers using an online tool (ref).
-At this step, in addition to fitting models to country survey data, as done in the global modeling, users can incorporate additional data sources such as routine information system records (e.g., the number of contraceptive commodities supplied, add REF).
+Subsequently, a local model can be used to produce FP estimates based on the data from one country, using information from the global model fit.The technical details for this approach are given in the next section.
+We return to the application of estimating family planning indicators in Section 4.2.
+
+\clearpage
 
@@
 We assume that $\mu$ is unconstrained (possibly after a transformation has been applied) and consider the following general hierarchical model setup, see Figure 2: \begin{align*}
-  \mu_{r^{(l)}}^{(l)} \mid \mu^{(l-1)}_{r^\prime}, \sigma^{(l)}_\theta &\sim N(\mu^{(l-1)}_{r^{(l-1)}}, \sigma^{(l)}_\mu), \text{ for } l > 0,
-\end{align*} 
-where $\mu_{r^{(l)}}^{(l)}$ ...
+  \mu_{r^{(l)}}^{(l)} \mid \mu^{(l-1)}_{r^\prime}, \sigma^{(l)}_\theta &\sim N(\mu^{(l-1)}_{r^{(l-1)}}, \sigma^{(l)}), \text{ for } l > 0,
+\end{align*} where $\mu_{r^{(l)}}^{(l)}$ ...
 Here, $\mu^{(l-1)}_{r^\prime}$ is the parameter value at one level higher in the hierarchy for the region $r^\prime$ that contains unit $r$.
-The parameter $\sigma_\mu^{(l)}$ for $l > 0$ describes variability of values of the parameter $\mu$ across geographic units at level $l$ of the hierarchy.
+The parameter $\sigma^{(l)}$ for $l > 0$ describes variability of values of the parameter $\mu$ across geographic units at level $l$ of the hierarchy.
 At the top level of the hierarchy, $l = 0$, only a single parameter is estimated.
@@
 % --- Level 0 ---
-\node[level0] (mu0) {$\mu^{(0)} \sim \mathcal{N}(0,s_\mu^2)$};
+\node[level0] (mu0) {$\mu^{(0)} \sim \mathcal{N}(0,s^2)$};
 
 % --- Level 1 ---
-\node[level1, below left=2cm and 3cm of mu0] (mu11) {$\mu^{(1)}_1 \sim \mathcal{N}(\mu^{(0)},\sigma_{\mu,1}^2)$};
+\node[level1, below left=2cm and 3cm of mu0] (mu11) {$\mu^{(1)}_1 \sim \mathcal{N}(\mu^{(0)},\sigma_{1}^2)$};
 \node[level1, right=of mu11] (mu12) {$\mu^{(1)}_2$};
@@
 % --- Level 2 ---
-\node[level2, below left=2cm and 0cm of mu11] (mu21) {$\mu^{(2)}_1 \sim \mathcal{N}(\mu^{(1)}_1,\sigma_{\mu,2}^2)$};
+\node[level2, below left=2cm and 0cm of mu11] (mu21) {$\mu^{(2)}_1 \sim \mathcal{N}(\mu^{(1)}_1,\sigma_{2}^2)$};
 \node[level2, right=of mu21] (mu22) {$\mu^{(2)}_2$};
@@
 % % --- Level 3 ---
-\node[level3, below left=2cm and 0cm of mu21] (mu31) {$\mu^{(3)}_1 \sim \mathcal{N}(\mu^{(2)}_1,\sigma_{\mu,3}^2)$};
+\node[level3, below left=2cm and 0cm of mu21] (mu31) {$\mu^{(3)}_1 \sim \mathcal{N}(\mu^{(2)}_1,\sigma_{3}^2)$};
 \node[level3, right=of mu31] (mu32) {$\mu^{(3)}_2$};
@@
 % % % --- Level 4 ---
-% \node[level4, below left=2cm and 0cm of mu31] (mu41) {$\mu^{(4)}_1 \sim \mathcal{N}(\mu^{(3)}_1,\sigma_{\mu,4}^2)$};
+% \node[level4, below left=2cm and 0cm of mu31] (mu41) {$\mu^{(4)}_1 \sim \mathcal{N}(\mu^{(3)}_1,\sigma_{4}^2)$};
 % \node[level4, right=of mu41] (mu42) {$\mu^{(4)}_2$};
@@
 % Sigma labels inside background blocks
-\node[anchor=north west] at (bg1.east) {$\sigma_{\mu,1}$};
-\node[anchor=north west] at (bg2.east) {$\sigma_{\mu,2}$};
-\node[anchor=north west] at (bg3.east) {$\sigma_{\mu,3}$};
-%\node[anchor=north west] at (bg4.east) {$\sigma_{\mu,4}$};
+\node[anchor=north west] at (bg1.east) {$\sigma_{1}$};
+\node[anchor=north west] at (bg2.east) {$\sigma_{2}$};
+\node[anchor=north west] at (bg3.east) {$\sigma_{3}$};
+%\node[anchor=north west] at (bg4.east) {$\sigma_{4}$};
 % --- Arrows ---
@@
 \begin{alignat}{2}
-  \mu_c &\mid \mu_{s[c]}^{(\subregion)}, \sigma_{\mu, \country} &&\sim N(\mu_{s[c]}^{(\subregion)}, \sigma_{\mu, \country}^2), \label{eq-countryhier} \\
-  \mu_s^{(\subregion)} &\mid \mu_{w[s]}^{(\region)}, \sigma_{\mu, \subregion}^2 &&\sim N(\mu_{r[s]}^{(\region)}, \sigma_{\mu, \subregion}^2), \nonumber \\
-  \mu_w^{(\region)} &\mid \mu^{(\glob)}, \sigma_{\mu, \region}^2 &&\sim N(\mu^{(\glob)}, \sigma_{\mu, \region}^2), \nonumber
-\end{alignat} 
-where $c =1,\dots,C$ refers to the country index, $s[c] = 1, .., S$ to the subregion index of country $c$, and $w[s[c]] = 1, .., W$ to the region index of subregion $s[c]$.
+  \mu_c &\mid \mu_{s[c]}^{(\subregion)}, \sigma_{ \country} &&\sim N(\mu_{s[c]}^{(\subregion)}, \sigma_{ \country}^2), \label{eq-countryhier} \\
+  \mu_s^{(\subregion)} &\mid \mu_{w[s]}^{(\region)}, \sigma_{ \subregion}^2 &&\sim N(\mu_{r[s]}^{(\region)}, \sigma_{ \subregion}^2), \nonumber \\
+  \mu_w^{(\region)} &\mid \mu^{(\glob)}, \sigma_{ \region}^2 &&\sim N(\mu^{(\glob)}, \sigma_{ \region}^2), \nonumber
+\end{alignat} where $c =1,\dots,C$ refers to the country index, $s[c] = 1, .., S$ to the subregion index of country $c$, and $w[s[c]] = 1, .., W$ to the region index of subregion $s[c]$.
 The parameter $\mu^{(\glob)}$ is the global mean for the parameter of interest, e.g., a global mean for service coverage or mortality.
-The parameters $\sigma_{\mu, \country}$, $\sigma_{\mu, \subregion}$, and $\sigma_{\mu, \region}$ describe variability in the parameter of interest across countries, subregions, and regions, respectively.
-
-<!-- Here, if $\mu^{(\glob)}$ has prior $\mu^{(\glob)} \sim N(0, s_{\mu, global}^2)$, then $$\mu_c =  s_{global}\cdot \gamma_{global} + \sigma_{\mu, region}\cdot \gamma_{r[s[c]]]}^{(
-egion)} + \sigma_{\mu, subregion}\cdot\gamma_{s[c]}^{(\subregion)} + \sigma_{\mu, country}\cdot\gamma_{\mu, c},$$ with $\gamma \sim N(0,1)$ for each gamma (one global, $R$ regions, $S$ subregions, $C$ countries).
+The parameters $\sigma_{ \country}$, $\sigma_{ \subregion}$, and $\sigma_{ \region}$ describe variability in the parameter of interest across countries, subregions, and regions, respectively.
+
+```{=html}
+<!-- Here, if $\mu^{(\glob)}$ has prior $\mu^{(\glob)} \sim N(0, s_{\mu, global}^2)$, then $$\mu_c =  s_{global}\cdot \gamma_{global} + \sigma_{ region}\cdot \gamma_{r[s[c]]]}^{(
+egion)} + \sigma_{ subregion}\cdot\gamma_{s[c]}^{(\subregion)} + \sigma_{ country}\cdot\gamma_{ c},$$ with $\gamma \sim N(0,1)$ for each gamma (one global, $R$ regions, $S$ subregions, $C$ countries).
 -->
+```
 
@@
   \begin{aligned}
-\mu_w^{(region)} &\sim N(\mu^{(global)}, \sigma_{\mu, region}^2),\\
-\mu_s^{(subregion)} &\sim N(\mu_{w[s]}^{(region)}, \sigma_{\mu, subregion}^2), \\
-\mu_c^{(country)} &\sim N(\mu_{s[c]}^{(subregion)}, \sigma_{\mu, country}^2),  \end{aligned}
+\mu_w^{(region)} &\sim N(\mu^{(global)}, \sigma_{ region}^2),\\
+\mu_s^{(subregion)} &\sim N(\mu_{w[s]}^{(region)}, \sigma_{ subregion}^2), \\
+\mu_c^{(country)} &\sim N(\mu_{s[c]}^{(subregion)}, \sigma_{ country}^2),  \end{aligned}
 \)
@@
   Local national modeling:\\
-  Estimate national parameters only, \\i.e., $\mu_c^{(country)}$ with $\mu_c^{(country)} \sim N(\hat{\mu}_{s[c]}^{(subregion)}, \hat{\sigma}_{\mu, country}^2).$
+  Estimate national parameters only, \\i.e., $\mu_c^{(country)}$ with $\mu_c^{(country)} \sim N(\hat{\mu}_{s[c]}^{(subregion)}, \hat{\sigma}_{ country}^2).$
 };
@@
 
-$$\mu_c \sim N(\hat{\mu}_{s[c]}^{(\subregion)}, \hat{\sigma}_{\mu, \country}^2),$$
+$$\mu_c \sim N(\hat{\mu}_{s[c]}^{(\subregion)}, \hat{\sigma}_{ \country}^2),$$
 
@@
 A worked example is given in the first use case.
-Briefly, if a user wants to specify a hierarchical model like the one introduced above, with the levels world, regions, subregion, and country, <!-- country parameters $\mu_c =  s_{global}\cdot \gamma_{global} + \sigma_{\mu, region}\cdot \gamma_{r[s[c]]]}^{(
-egion)} + \sigma_{\mu, subregion}\cdot\gamma_{s[c]}^{(\subregion)} + \sigma_{\mu, country}\cdot\gamma_{\mu, c}$,  --> using a dataset with information on these groupings, with naming `region`, `subregion`, and `country`, in the R functions, the user can specify the hierarchical model as follows \[Q for herb: refs/credit to be given for this set up to other work?\]:
+Briefly, if a user wants to specify a hierarchical model like the one introduced above, with the levels world, regions, subregion, and country, <!-- country parameters $\mu_c =  s_{global}\cdot \gamma_{global} + \sigma_{ region}\cdot \gamma_{r[s[c]]]}^{(
+egion)} + \sigma_{ subregion}\cdot\gamma_{s[c]}^{(\subregion)} + \sigma_{ country}\cdot\gamma_{ c}$,  --> using a dataset with information on these groupings, with naming `region`, `subregion`, and `country`, in the R functions, the user can specify the hierarchical model as follows \[Q for herb: refs/credit to be given for this set up to other work?\]:
 
@@
 The parameterization used in Stan is a non-centered parameterization.
-In general notation, for $l>0$, non-centered parameterization refers to writing mean parameters as deviations from higher-level group means:
-\begin{align*}
+In general notation, for $l>0$, non-centered parameterization refers to writing mean parameters as deviations from higher-level group means: \begin{align*}
   \mu_r^{(l)} &= \mu^{(l-1)} + \sigma^{(l)} \cdot \gamma_{r^\prime}^{(l-1)},
-\end{align*} 
-where $\gamma_r^{(l)} \sim N(0, 1)$. Applied sequentially, and if $\mu^{(0)} \sim N(0, s)$, we get 
-\begin{align*}
+\end{align*} where $\gamma_r^{(l)} \sim N(0, 1)$. Applied sequentially, and if $\mu^{(0)} \sim N(0, s)$, we get \begin{align*}
   \mu_r^{(l)} &= s\cdot\gamma^{(0)} + \sum_{k=1}^{l} \sigma^{(k)} \cdot \gamma_{r^\prime}^{(k)},
-\end{align*}
-where $r^\prime$ now refers more generally to the index of the unit containing the lower-level group. For the specific example of country parameters, the non-centered parameterization is given by 
-\begin{align*}
+\end{align*} where $r^\prime$ now refers more generally to the index of the unit containing the lower-level group.
+For the specific example of country parameters, the non-centered parameterization is given by \begin{align*}
   \mu_c =  s_{\glob} \cdot {\gamma}_{\glob} + {\sigma}_{\mu, \region}\cdot {\gamma}_{r[s[c]]]}^{(\region)} + {\sigma}_{\mu, \subregion}\cdot{\gamma}_{s[c]}^{(\subregion)} + {\sigma}_{\mu, \country}\cdot\gamma_{c}.
@@
 Implementation is based on a model matrix.
-For our model given by the equation above, the following matrix multiplication is used to obtain the vector of $\mu_c$s: $$(\mu_1, ..., \mu_c) = M\cdot (\eta_1, ..., \eta_{1+R+S+C}),$$ where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_{1+R+S+C})$ refers to the vector of the product of the standard deviation terms and $\gamma$s, i.e., $\gamma_1 = (s_{\glob}\cdot \gamma_{\glob}, \gamma_2 = \sigma_{\mu, \region} \cdot \gamma_{\mu, 1}^{(\region)},..., \gamma_{1+W+S+C} = \sigma_{\mu, \country}\cdot\gamma_{\mu, C}$.
+For our model given by the equation above, the following matrix multiplication is used to obtain the vector of $\mu_c$s: $$(\mu_1, ..., \mu_c) = M\cdot (\eta_1, ..., \eta_{1+R+S+C}),$$ where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_{1+R+S+C})$ refers to the vector of the product of the standard deviation terms and $\gamma$s, i.e., $\gamma_1 = (s_{\glob}\cdot \gamma_{\glob}, \gamma_2 = \sigma_{ \region} \cdot \gamma_{ 1}^{(\region)},..., \gamma_{1+W+S+C} = \sigma_{ \country}\cdot\gamma_{ C}$.
 When estimating a subset of parameters in a local model, fixed values of $\sigma$s and $\gamma$s by point estimates (posterior means) from a global fit.
-For example, for a national model, we get
-\begin{align*}
+For example, for a national model, we get \begin{align*}
   \mu_c = s_{\glob} \cdot \hat{\gamma}_{\glob} + 
-          \hat{\sigma}_{\mu, \region} \cdot \hat{\gamma}_{r[s[c]]]}^{(\region)} + 
-          \hat{\sigma}_{\mu, \subregion} \cdot \hat{\gamma}_{s[c]}^{(\subregion)} + 
-          \hat{\sigma}_{\mu, \country} \cdot \gamma_{c}.
+          \hat{\sigma}_{ \region} \cdot \hat{\gamma}_{r[s[c]]]}^{(\region)} + 
+          \hat{\sigma}_{ \subregion} \cdot \hat{\gamma}_{s[c]}^{(\subregion)} + 
+          \hat{\sigma}_{ \country} \cdot \gamma_{c}.
 \end{align*}
@@
 ### Simulation setup
+
 The simulation setup is summarized in Figure 4, panels A and B. Simulation of subnational data is discussed in a later section.
@@
 % --- Level 0 ---
-\node[level0] (mu0) {$\mu^{(0)} \sim \mathcal{N}(0,s_\mu^2)$};
+\node[level0] (mu0) {$\mu^{(0)} \sim \mathcal{N}(0,s^2)$};
 
 % --- Level 1 ---
-\node[level1, below left=1.5cm and 3cm of mu0] (mu11) {$\mu^{(1)}_1 \sim \mathcal{N}(\mu^{(0)},\sigma_{\mu,1}^2)$};
+\node[level1, below left=1.5cm and 3cm of mu0] (mu11) {$\mu^{(1)}_1 \sim \mathcal{N}(\mu^{(0)},\sigma_{1}^2)$};
 \node[level1, right=of mu11] (mu12) {$\mu^{(1)}_2$};
@@
 % --- Level 2 ---
-\node[level2, below left=1.5cm and 0cm of mu11] (mu21) {$\mu^{(2)}_1 \sim \mathcal{N}(\mu^{(1)}_1,\sigma_{\mu,2}^2)$};
+\node[level2, below left=1.5cm and 0cm of mu11] (mu21) {$\mu^{(2)}_1 \sim \mathcal{N}(\mu^{(1)}_1,\sigma_{2}^2)$};
 \node[level2, right=of mu21] (mu22) {$\mu^{(2)}_2$};
@@
 % Sigma labels inside background blocks
-\node[anchor=north west] at (bg1.east) {$\sigma_{\mu,1}$, country};
-\node[anchor=north west] at (bg2.east) {$\sigma_{\mu,2}$, subnational region};
+\node[anchor=north west] at (bg1.east) {$\sigma_{1}$, country};
+\node[anchor=north west] at (bg2.east) {$\sigma_{2}$, subnational region};
 % --- Arrows ---
@@
 Panel A in Figure 4 indicates the hierarchical model considered.
-For national level estimation, we consider countries to be grouped in the world, i.e., $\mu_c|\mu^{(\glob)}, \sigma_{\mu, \country} \sim N(\mu^{(\glob)}, \sigma_{\mu, \country}^2)$.
+For national level estimation, we consider countries to be grouped in the world, i.e., $\mu_c|\mu^{(\glob)}, \sigma_{ \country} \sim N(\mu^{(\glob)}, \sigma_{ \country}^2)$.
 Panel B introduces the likelihood function for national data, given by a normal likelihood with constant variance: $$y_i^{(\country)}|\mu_{c[i]}, \sigma_y^2 \sim N(\mu_{c[i]}, \sigma_y^2),$$ where $c[i]$ refers to the country of country observation $i$.
@@
 Simulated data were generated using `simulate_multilevel_data` function.
-We simulate a set of model parameters, setting $\mu^{(\glob)} = 0$, and $\sigma_{\mu,\country} = 0.5$, creating 30 countries, and national data, using $\sigma_y = 0.1$:
+We simulate a set of model parameters, setting $\mu^{(\glob)} = 0$, and $\sigma_{\country} = 0.5$, creating 30 countries, and national data, using $\sigma_y = 0.1$:
 
@@
 Its arguments also include prior parameters `mu_scalarprior_mean` and `mu_scalarprior_sd`, which refer to the (rescaled) prior mean and standard deviation of $\mu^{(\glob)}$.
-The model block contains the priors for the $\gamma$s and $\sigma_{\mu}$s, given by
-\begin{align*}
+The model block contains the priors for the $\gamma$s and $\sigma_{\mu}$s, given by \begin{align*}
   \mu^{(\glob)} &\sim N(0,1),\\
   \sigma_{\mu} &\sim \text{Half-Normal}(0, s_{\mu}).
-\end{align*} 
-All prior parameters are added as input data to the Stan model.
+\end{align*} All prior parameters are added as input data to the Stan model.
 
@@
 
-## Application: national estimation of family planning indicators using FPET
-
-TO ADD
-
+## Application: National estimation of family planning indicators using FPET
+
+For family planning estimation, a local model fitting approach was motivated by the need to enable country monitoring and evaluation (M&E) officers to produce estimates for their countries, as part of the Track20 project (ref).
+The Track20 project works to build the capacity of countries to generate and use data to inform their programming, improve impact, and accelerate progress toward their FP goals.
+One of the pillars of this approach is the cultivation of a network of M&E officers dedicated to increasing the quality and use of FP data.
+We developed the family planning estimation tool (FPET), as a local version of a global family planning model, for use by the M&E officers.
+
+For local estimation of family planning indicators, the Bayesian model described in section 2 is used.
+Specifically, hierarchical models are used for estimation of transition model parameters.
+The workflow used for estimating family planning indicators at the national level is based on model fitting in several steps, including the estimation of long-term trends in the first step, the estimation of all model parameters related to modern use in a second step, the addition of information on the use of traditional methods of contraception in a third step, followed by local model fitting to produce estimates for all indicators of interest.
+The `localhierarchy` functionality is incorporated in the family planning modeling code base such that all steps are run with the same code base, and parameters can be passed from one model fit to another.
+Full details on this workflow and model specification are described elsewhere (ref).
+As an illustration, @fig-allfits shows global and local fits for modern use, demand, and demand satisfied for selected countries in Western Africa.
+The global fit in red corresponds closely to the local fits, shown in green.
+
+```{r}
+#| fig-height: 9
+#| fig-width: 7
+#| echo: false
+#| label: fig-allfits
+#| fig-cap: "Data, estimates, and forecasts from 1970 until 2030 for FP indicators for married women in in Benin (BEN), Gambia (GMB), and Ghana (GHA).
+FPET estimates are shown in in colored lines (red for global, green for local), with shaded areas representing 90% credible intervals.
+Survey data are shown in colored dots with 95% confidence intervals that reflect total error variance."
+plots <- readRDS("outputfpet_localglobalfits.rds")
+plots[[1]] /  plots[[2]]/ plots[[3]] # Stack three plots vertically
+```
+\clearpage 
 # Use case 2: Subnational estimation
@@
 
-In addition to a standard approach to subnational modeling, functionality in the `localhierarchy` package can also be used to fit subnational models that are informed by national level data and estimation, referred to here as _nationally-informed subnational estimation_ (NISE).
+In addition to a standard approach to subnational modeling, functionality in the `localhierarchy` package can also be used to fit subnational models that are informed by national level data and estimation, referred to here as *nationally-informed subnational estimation* (NISE).
 This approach is useful when subnational data are limited, for example, in terms of the number of countries with subnational data, the number of subnational regions with data, the number of observations per subnational region, or due to uncertainty associated with subnational data.
@@
 This national-level model fit is used in a global subnational fit to a global data base of subnational data.
-Specifically, in the global subnational fitting step, national level parameters are fixed at point estimates from the global national fit, i.e., $\hat{\sigma}_{\mu, l}$ and $\mu_{\cdot}^{(l)}$ for level $l$ up to the country level.
+Specifically, in the global subnational fitting step, national level parameters are fixed at point estimates from the global national fit, i.e., $\hat{\sigma}_{ l}$ and $\mu_{\cdot}^{(l)}$ for level $l$ up to the country level.
 Finally, in a 3rd step, local models can be fitted, using fixed values from the global subnational fit.
@@
   \begin{aligned}
-\mu_w^{(region)} &\sim N(\mu^{(global)}, \sigma_{\mu, region}^2),\\
-\mu_s^{(subregion)} &\sim N(\mu_{w[s]}^{(region)}, \sigma_{\mu, subregion}^2), \\
-\mu_c^{(country)} &\sim N(\mu_{s[c]}^{(subregion)}, \sigma_{\mu, country}^2),  \\
-\mu_r^{(subnat)} &\sim N(\mu_{r[c]}^{(country)}, \sigma_{\mu, subnat}^2),  \\
+\mu_w^{(region)} &\sim N(\mu^{(global)}, \sigma_{ region}^2),\\
+\mu_s^{(subregion)} &\sim N(\mu_{w[s]}^{(region)}, \sigma_{ subregion}^2), \\
+\mu_c^{(country)} &\sim N(\mu_{s[c]}^{(subregion)}, \sigma_{ country}^2),  \\
+\mu_r^{(subnat)} &\sim N(\mu_{r[c]}^{(country)}, \sigma_{ subnat}^2),  \\
 \end{aligned}
@@
   Estimate subset of parameters in \\ hierarchical model, 
-  i.e., $\sigma_{\mu, subnat}$
+  i.e., $\sigma_{ subnat}$
 };
@@
   Local subnational modeling:\\
-  Estimate subnational parameters only, \\i.e., $\mu_r$ in $\mu_r \sim N(\hat{\mu}_{c[r]}^{(country)}, \hat{\sigma}_{\mu, subnat}^2).$
+  Estimate subnational parameters only, \\i.e., $\mu_r$ in $\mu_r \sim N(\hat{\mu}_{c[r]}^{(country)}, \hat{\sigma}_{ subnat}^2).$
 };
@@
 ### Simulation setup
+
 The simulation setup for subnational parameters and data is summarized in Figure 4, Panels A-B-C.
 As explained above for national estimation, Panel A Figure 4 indicates the hierarchical model considered.
-For subnational estimation, an additional level is added, with $\mu_r^{(\subnat)} = \mu_{c[r]} + \eta_{r}^{(\subnat)}$, $\eta_{r}^{(\subnat)}|\sigma_{\mu, \subnat} \sim N(0, \sigma_{\mu, \subnat})$.
+For subnational estimation, an additional level is added, with $\mu_r^{(\subnat)} = \mu_{c[r]} + \eta_{r}^{(\subnat)}$, $\eta_{r}^{(\subnat)}|\sigma_{ \subnat} \sim N(0, \sigma_{ \subnat})$.
 
@@
 
-## Application: subnational estimation of family planning indicators using FPET
-
-For subnational estimation, the workflow used in FPET mirrors the approach described in this paper: a global national model is fitted first, followed by a global subnational model, and finally by local fits.
-This sequential approach, in which the subnational global model is based on the national global model, was motivated by differences in data availability between national and subnational global databases.
-For example, when estimating family planning indicators for married women of reproductive age, the global national database contains XX observations from XXX countries, whereas the subnational database includes YY observations from YYY countries.
-
-ADD
+## Application: Subnational estimation of family planning indicators using FPET
+
+Subnational estimation of family planning indicators is of interest to country M&E officers for informing family planning programming.
+To that end, the Track20 project compiled survey data at the subnational level over 35 low- and middle income countries.
+
+For subnational estimation, the workflow used mirrors the NISE approach described in this paper: global modeling of national-level indicators is carried out first, followed by  subnational model fitting using a global data base, and finally by local fits.
+This sequential approach, in which the global subnational model is based on global national modeling, was motivated by differences in data availability between national and subnational global databases: When estimating family planning indicators for married women of reproductive age, the global national database contains observations from 198 countries, whereas the global subnational database includes observations from 35 countries only.
+
+The figures below show illustrations of regional fits for Benin.
+The subnational model is fitted to all regional data from the country.
+@fig-subnatoverview illustrates fits in selected regions and differences across regions in 2024.
+
+```{r}
+#| fig-height: 9
+#| fig-width: 7
+#| echo: false
+#| label: fig-subnatfits
+#| fig-cap: "Data, estimates, and forecasts from 1970 until 2030 for FP indicators for married women in subnational regions in Benin.
+Graphs show mCPR, unmet need for modern methods, demand, and demand satisfied with modern methods.
+Survey data are shown in colored dots with 95% confidence intervals that reflect total error variance.
+FPET estimates are shown in black, with grey shaded areas representing 90% credible intervals.
+"
+plots <- readRDS("outputfpet_plotssubnat_BEN.rds")
+plots[["Oueme"]]$married / plots[["Collines"]]$married/ plots[["Donga"]]$married
+```
+
+```{r}
+#| echo: false
+#| label: fig-subnatoverview
+#| fig-cap: "Regional estimates for Benin for modern contraceptive use in 2024. Point estimates and 95% credible intervals shown."
+plots[["comp"]] 
+```
+
+<!-- The figures below show illustrations of regional fits for Cote d'Ivoire.
+The subnational model is fitted to all data from the country, here the combination of regional data based on the most recent DHS survey, and national data for other survey-years (not shown in regional plots).
+This results in estimates for recent years based on all available information.
+An example, estimates of modern contraceptive use in 2024, is shown as well.
+-->
 
 # Conclusions 
-
-# Additional details
-
-##
+Bayesian hierarchical models are widely used in global health estimation, where data availability may vary across national or subnational populations.
+For analysis at a local level, local models are often of interest, referring to models that are derived from global hierarchical models but adapted such that they can be fitted to the data from one population alone.
+To that end, we introduced the `localHierarchy` R package, which provides functionality for fitting Bayesian hierarchical models in settings where both global and local estimation is required.
+
+We have used the `localHierarchy` functionality in different projects.
+We discussed the estimation of family planning indicators, where country M&E officers use the local model to produce estimates at the national and subnational level.
+In this area, M&E officers have successfully used local models to produce national estimates since 2013 [LA to check year].
+Availability of a local modeling tool has enabled users to do additional analyses and check sensitivity of estimates to specific data sources or data quality assumptions.
+In addition to fitting the model to survey data, users of the local family planning model can also incorporate additional data from routine information systems  (e.g., the number of contraceptive commodities supplied, add REF).
+These use cases illustrate the value of local models to enable local analysis.
+
+More recently, we have used the `localHierarchy` approach and functionality for estimating coverage indicators for the Countdown to 2030 initiative (ref).
+By developing the `localHierarchy` package and providing use cases in this paper, we hope that this work can be useful for other global health modelers, to consider local modeling in their own work.

### 2025-09-29 paper/paper.qmd
comments evan

@@
 
-Since local models are usually derived from global models, a key challenge in using both for global health estimation is synchronizing codebases and data flows between these approaches.
-This is particularly important for modeling approaches that involve hierarchical estimation of multiple parameters, where the hierarchical structures may differ across parameters.
+A key challenge in the use of local models in global health estimation is synchronizing codebases and data flows between global and local estimation approaches.
+Synchronizing codebases and data flows is important because the fitting of local models relies on parameter estimates from global models, hence synchronization is important to ensure consistency across global and local models.
+Synchronization can be particularly challenging for modeling approaches that involve hierarchical estimation of multiple parameters, where the hierarchical structures may differ across parameters.
 
@@
 The next section introduces the modeling approach and workflow used for estimating family planning indicators, motivating the need for the \texttt{localhierarchy} functionality of fitting global and local hierarchical models.
-The methods section introduces the package' functionality for the estimation of one parameter at the national and subnational level.
+The methods section introduces the package's functionality for the estimation of one parameter at the national and subnational levels.
 Finally, we return to the case study of family planning estimation to present illustrative results of the application of the \texttt{localhierarchy} setup in this context.
@@
 The Family Planning Estimation Tool (FPET) is widely used to estimate key family planning indicators in countries and regions.
-Key indicators include modern contraceptive use, demand for family planning, and demand satisfied, referring to as the ratio of modern use to demand.
+Key indicators include modern contraceptive use, demand for family planning, and demand satisfied, referring to the ratio of modern use to demand.
 Data on such indicators is available from household surveys and routine data systems.
@@
 These figures show variation in levels among countries in Western Africa, informed by country-specific data.
-Estimates of asymptotes are similar across countries in Western Africa because countries have not yet reached the end phase of their transition yet; the asymptotes in Western African countries are informed by the estimate of the asymptote in sub-Saharan Africa through the hierarchical model.
+Estimates of asymptotes are similar across countries in Western Africa because countries have not yet reached the end phase of their transition; the asymptotes in Western African countries are informed by the estimate of the asymptote in sub-Saharan Africa through the hierarchical model.
 
@@
 The global model fit provides information on model parameters, including hierarchical parameters, e.g., estimates of asymptotes.
-Subsequently, a local model can be used to produce FP estimates based on the data from one country, using information from the global model fit.The technical details for this approach are given in the next section.
+Subsequently, a local model can be used to produce FP estimates based on the data from one country, using information from the global model fit.
+The technical details for this approach are given in the next section.
 We return to the application of estimating family planning indicators in Section 4.2.
@@
 
-The modeling approach introduced in `localhierarchy` was motivated by application like FPET that depend on hierarchical structures associated with national and subnational level estimation, requiring global and local models to be fit.
+The modeling approach introduced in `localhierarchy` was motivated by applications like FPET that depend on hierarchical structures associated with national and subnational level estimation, requiring global and local models to be fit.
 In this section, we introduce the principle setup used for global and local hierarchical models in the context of national estimation.
@@
 
-Suppose we want to estimate some population parameter of interest $\mu_r$ for a population indexed by $r$.
-For example, $\mu$ could refer to some population health indicator such as service coverage or morbidity or mortality-related outcomes.
-Index $r$ refers to the population in a region indexed by $r$, for example, a subnational region, a country, or a group of countries.
-We assume that $\mu$ is unconstrained (possibly after a transformation has been applied) and consider the following general hierarchical model setup, see Figure 2: \begin{align*}
-  \mu_{r^{(l)}}^{(l)} \mid \mu^{(l-1)}_{r^\prime}, \sigma^{(l)}_\theta &\sim N(\mu^{(l-1)}_{r^{(l-1)}}, \sigma^{(l)}), \text{ for } l > 0,
-\end{align*} where $\mu_{r^{(l)}}^{(l)}$ ...
-Here, $\mu^{(l-1)}_{r^\prime}$ is the parameter value at one level higher in the hierarchy for the region $r^\prime$ that contains unit $r$.
-The parameter $\sigma^{(l)}$ for $l > 0$ describes variability of values of the parameter $\mu$ across geographic units at level $l$ of the hierarchy.
+Suppose we want to estimate a population parameter $\mu_r^{(l)}$ of interest, referring to a population health indicator for region $r$ at level $l$ of the hierarchy.
+For example, the indicator could refer to some population health indicator such as service coverage or morbidity or mortality-related outcomes.
+For a given level $l$, index $r$ refers to the population in a particular region indexed by $r$, for example, a subnational region, a country, or a group of countries.
+We assume that $\mu$ is unconstrained (possibly after a transformation has been applied) and consider the following general hierarchical model setup, see Figure 6:
+\begin{align*}
+  \mu_{r}^{(l)} \mid \mu^{(l-1)}_{r^\prime}, \sigma_l} &\sim N(\mu^{(l-1)}_{r^\prime}, \sigma_l^2), \text{ for } l > 0,
+\end{align*} where $\mu^{(l-1)}_{r^\prime}$ is the parameter value at one level higher in the hierarchy for the region $r^\prime$ that contains unit $r$.
+The parameter $\sigma_l$ for $l > 0$ describes variability of values of the parameter $\mu$ across geographic units at level $l$ of the hierarchy.
 At the top level of the hierarchy, $l = 0$, only a single parameter is estimated.
@@
 % --- Level 0 ---
-\node[level0] (mu0) {$\mu^{(0)} \sim \mathcal{N}(0,s^2)$};
+\node[level0] (mu0) {$\mu^{(0)} \sim \mathcal{N}(0,s_0^2)$};
 
@@
 \begin{alignat}{2}
-  \mu_c &\mid \mu_{s[c]}^{(\subregion)}, \sigma_{ \country} &&\sim N(\mu_{s[c]}^{(\subregion)}, \sigma_{ \country}^2), \label{eq-countryhier} \\
-  \mu_s^{(\subregion)} &\mid \mu_{w[s]}^{(\region)}, \sigma_{ \subregion}^2 &&\sim N(\mu_{r[s]}^{(\region)}, \sigma_{ \subregion}^2), \nonumber \\
+  \mu_c^{(\country)}^{(\country)} &\mid \mu_{s[c]}^{(\subregion)}, \sigma_{ \country}^2 &&\sim N(\mu_{s[c]}^{(\subregion)}, \sigma_{ \country}^2), \label{eq-countryhier} \\
+  \mu_s^{(\subregion)} &\mid \mu_{w[s]}^{(\region)}, \sigma_{ \subregion}^2 &&\sim N(\mu_{w[s]}^{(\region)}, \sigma_{ \subregion}^2), \nonumber \\
   \mu_w^{(\region)} &\mid \mu^{(\glob)}, \sigma_{ \region}^2 &&\sim N(\mu^{(\glob)}, \sigma_{ \region}^2), \nonumber
@@
 We use the term local hierarchical models to refer to a hierarchical model in which a subset of parameters have been fixed at posterior mean estimates obtained from a fit of the model to a larger data set.
-A typical workflow for local estimation of national level outcomes is illustrated in Figure 3.
-For national estimation using the model given by Eq.\~\ref{eq-countryhier}, the user starts by fitting a global model, using a global data base and estimating all parameters in a hierarchical model.
+A typical workflow for local estimation of national level outcomes is illustrated in Figure 7.
+For national estimation using the model given by Eq.\ref{eq-countryhier}, the user starts by fitting a global model, using a global data base and estimating all parameters in a hierarchical model.
 In a local model, data from one country are used and parameters that are not country-specific are fixed using parameter estimates from the global model fit.
@@
   Local national modeling:\\
-  Estimate national parameters only, \\i.e., $\mu_c^{(country)}$ with $\mu_c^{(country)} \sim N(\hat{\mu}_{s[c]}^{(subregion)}, \hat{\sigma}_{ country}^2).$
+  Estimate national parameters only, \\
+  fix parameter estimates at higher levels of the hierarchy \\i.e., $\mu_c^{(country)}$ with $\mu_c^{(country)} \sim N(\hat{\mu}_{s[c]}^{(subregion)}, \hat{\sigma}_{ country}^2).$
 };
@@
 
-$$\mu_c \sim N(\hat{\mu}_{s[c]}^{(\subregion)}, \hat{\sigma}_{ \country}^2),$$
+$$\mu_c^{(\country)} \sim N(\hat{\mu}_{s[c]}^{(\subregion)}, \hat{\sigma}_{ \country}^2),$$
 
@@
 Continuing with the general notation introduced above, in the package, a local model for one parameter of level $l$ is derived by fixing $\sigma^{(k)}$ for $k=1, ..., l$ and fixing $\mu_{r\prime}{(k)}$ for $k = 1, .., l-1$ at posterior means derived from a global fit.
-More generally, if a model was fitted using data up to level $l$, the user can create a local model by fixing $\sigma^{(k)}$ for $k \leq l$ as long as, for a given fixed level $f$, variance parameters at levels $f' < f$ have been fixed.
-Similarly, the user can fix $\mu$'s for for $k \leq l$ as long as, for a given fixed level $f$, $\mu$s at levels $f' < f$ have been fixed and if the $\sigma$'s for the levels up to and include $f$, i.e., $f' \leq f$, have been fixed.
+More generally, if a model was fitted using data up to level $l$, the user can create a local model by fixing $\sigma^{(k)}$ at all levels $k$ up to some chosen level $K_{\sigma}$.
+Similarly, the $\mu^{(k)}$ parameters can be fixed at all levels $k$ up to a level $K_{\mu} \leq K_{\sigma}$.
 
@@
 A worked example is given in the first use case.
-Briefly, if a user wants to specify a hierarchical model like the one introduced above, with the levels world, regions, subregion, and country, <!-- country parameters $\mu_c =  s_{global}\cdot \gamma_{global} + \sigma_{ region}\cdot \gamma_{r[s[c]]]}^{(
+Briefly, if a user wants to specify a hierarchical model like the one introduced above, with the levels world, regions, subregion, and country, <!-- country parameters $\mu_c^{(\country)} =  s_{global}\cdot \gamma_{global} + \sigma_{ region}\cdot \gamma_{r[s[c]]]}^{(
 egion)} + \sigma_{ subregion}\cdot\gamma_{s[c]}^{(\subregion)} + \sigma_{ country}\cdot\gamma_{ c}$,  --> using a dataset with information on these groupings, with naming `region`, `subregion`, and `country`, in the R functions, the user can specify the hierarchical model as follows \[Q for herb: refs/credit to be given for this set up to other work?\]:
@@
 In general notation, for $l>0$, non-centered parameterization refers to writing mean parameters as deviations from higher-level group means: \begin{align*}
-  \mu_r^{(l)} &= \mu^{(l-1)} + \sigma^{(l)} \cdot \gamma_{r^\prime}^{(l-1)},
-\end{align*} where $\gamma_r^{(l)} \sim N(0, 1)$. Applied sequentially, and if $\mu^{(0)} \sim N(0, s)$, we get \begin{align*}
-  \mu_r^{(l)} &= s\cdot\gamma^{(0)} + \sum_{k=1}^{l} \sigma^{(k)} \cdot \gamma_{r^\prime}^{(k)},
+  \mu_r^{(l)} &= \mu^{(l-1)} + \sigma_l \cdot \gamma_{r}^{(l)},
+\end{align*} where $\gamma_r^{(l)} \sim N(0, 1)$. Applied sequentially, and if $\mu^{(0)} \sim N(0,s_{\glob})$ with prior standard deviation $s_{\glob}$, we get \begin{align*}
+  \mu_r^{(l)} &= s_{\glob}\cdot\gamma^{(0)} + \sum_{k=1}^{l} \sigma_k \cdot \gamma_{r^\prime}^{(k)},
 \end{align*} where $r^\prime$ now refers more generally to the index of the unit containing the lower-level group.
 For the specific example of country parameters, the non-centered parameterization is given by \begin{align*}
-  \mu_c =  s_{\glob} \cdot {\gamma}_{\glob} + {\sigma}_{\mu, \region}\cdot {\gamma}_{r[s[c]]]}^{(\region)} + {\sigma}_{\mu, \subregion}\cdot{\gamma}_{s[c]}^{(\subregion)} + {\sigma}_{\mu, \country}\cdot\gamma_{c}.
+  \mu_c^{(\country)} =  s_{\glob} \cdot {\gamma}^{(\glob)} + {\sigma}_{\region}\cdot {\gamma}_{w[s[c]]]}^{(\region)} + {\sigma}_{\subregion}\cdot{\gamma}_{s[c]}^{(\subregion)} + {\sigma}_{\country}\cdot\gamma_{c}^{(\country)}.
 \end{align*}
@@
 Implementation is based on a model matrix.
-For our model given by the equation above, the following matrix multiplication is used to obtain the vector of $\mu_c$s: $$(\mu_1, ..., \mu_c) = M\cdot (\eta_1, ..., \eta_{1+R+S+C}),$$ where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_{1+R+S+C})$ refers to the vector of the product of the standard deviation terms and $\gamma$s, i.e., $\gamma_1 = (s_{\glob}\cdot \gamma_{\glob}, \gamma_2 = \sigma_{ \region} \cdot \gamma_{ 1}^{(\region)},..., \gamma_{1+W+S+C} = \sigma_{ \country}\cdot\gamma_{ C}$.
-When estimating a subset of parameters in a local model, fixed values of $\sigma$s and $\gamma$s by point estimates (posterior means) from a global fit.
+For our model given by the equation above, the following matrix multiplication is used to obtain the vector of $\mu_c^{(\country)}$s: $$(\mu_1^{(\country)}, ..., \mu_c^{(\country)}) = M\cdot (\eta_1, ..., \eta_{1+R+S+C})',$$ where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_{1+R+S+C})$ refers to the vector of the product of the standard deviation terms and $\gamma$s, i.e., $\eta_1 = s_{\glob}\cdot \gamma_{\glob}, \eta_2 = \sigma_{ \region} \cdot \gamma_{ 1}^{(\region)}, ..., \eta_{1+W+S+C} = \sigma_{ \country}\cdot\gamma_{ C}^{(\country)}$.
+When estimating a subset of parameters in a local model, the values of $\sigma$s and $\gamma$s are fixed to point estimates (posterior means) from a global fit.
 For example, for a national model, we get \begin{align*}
   \mu_c = s_{\glob} \cdot \hat{\gamma}_{\glob} + 
-          \hat{\sigma}_{ \region} \cdot \hat{\gamma}_{r[s[c]]]}^{(\region)} + 
+          \hat{\sigma}_{ \region} \cdot \hat{\gamma}_{w[s[c]]]}^{(\region)} + 
           \hat{\sigma}_{ \subregion} \cdot \hat{\gamma}_{s[c]}^{(\subregion)} + 
-          \hat{\sigma}_{ \country} \cdot \gamma_{c}.
+          \hat{\sigma}_{ \country} \cdot \gamma_{c}^{(\country)}.
 \end{align*}
@@
 
-The simulation setup is summarized in Figure 4, panels A and B. Simulation of subnational data is discussed in a later section.
+The simulation setup is summarized in Figure 8, panels A and B. Simulation of subnational data is discussed in a later section.
 
@@
 % --- Level 0 ---
-\node[level0] (mu0) {$\mu^{(0)} \sim \mathcal{N}(0,s^2)$};
+\node[level0] (mu0) {$\mu^{(\glob)} \sim \mathcal{N}(0,s_\glob^2)$};
 
 % --- Level 1 ---
-\node[level1, below left=1.5cm and 3cm of mu0] (mu11) {$\mu^{(1)}_1 \sim \mathcal{N}(\mu^{(0)},\sigma_{1}^2)$};
-\node[level1, right=of mu11] (mu12) {$\mu^{(1)}_2$};
+\node[level1, below left=1.5cm and 3cm of mu0] (mu11) {$\mu^{(\country)}_1 \sim \mathcal{N}(\mu^{(\glob)},\sigma_{\country}^2)$};
+\node[level1, right=of mu11] (mu12) {$\mu^{(\country)}_2$};
 \node[level1, right=of mu12] (dots1) {$\dots$};
 % --- Level 2 ---
-\node[level2, below left=1.5cm and 0cm of mu11] (mu21) {$\mu^{(2)}_1 \sim \mathcal{N}(\mu^{(1)}_1,\sigma_{2}^2)$};
+\node[level2, below left=1.5cm and 0cm of mu11] (mu21) {$\mu^{(\subnat)}_1 \sim \mathcal{N}(\mu^{(\country)}_1,\sigma_{\subnat}^2)$};
 \node[level2, right=of mu21] (mu22) {$\mu^{(2)}_2$};
@@
 % Sigma labels inside background blocks
-\node[anchor=north west] at (bg1.east) {$\sigma_{1}$, country};
-\node[anchor=north west] at (bg2.east) {$\sigma_{2}$, subnational region};
+\node[anchor=north west] at (bg1.east) {$\sigma_{\country}$, country};
+\node[anchor=north west] at (bg2.east) {$\sigma_{\subnat}$, subnational region};
 % --- Arrows ---
@@
 % --- Level 1 ---
-\node[data, below left=1.5cm and 3cm of mu0] (mu11) {$y^{(country)}_1 \sim \mathcal{N}(\mu^{(1)}_1,\sigma_{y}^2)$};
+\node[data, below left=1.5cm and 3cm of mu0] (mu11) {$y^{(country)}_1 \sim \mathcal{N}(\mu^{(\country)}_1,\sigma_{y}^2)$};
 \node[data, right=of mu11] (mu12) {$y^{(country)}_2$};
@@
 % --- Level 1 ---
-\node[data, below left=1.5cm and 3cm of mu0] (mu11) {$y^{(subnat)}_1 \sim \mathcal{N}(\mu^{(1)}_1,\sigma_{y}^2)$};
+\node[data, below left=1.5cm and 3cm of mu0] (mu11) {$y^{(subnat)}_1 \sim \mathcal{N}(\mu^{(\subnat)}_1,\sigma_{y}^2)$};
 \node[data, right=of mu11] (mu12) {$y^{(subnat)}_2$};
@@
 Center: Data generating mechanism for national data, illustrated for the first country.
-Bottom: Data generating mechanism for subnational data, illustrated for the first subnational region.}
+Bottom: Data generating mechanism for subnational data, illustrated for the first subnational region (used only in Section 5).}
 \end{figure}
@@
 Panel A in Figure 4 indicates the hierarchical model considered.
-For national level estimation, we consider countries to be grouped in the world, i.e., $\mu_c|\mu^{(\glob)}, \sigma_{ \country} \sim N(\mu^{(\glob)}, \sigma_{ \country}^2)$.
-Panel B introduces the likelihood function for national data, given by a normal likelihood with constant variance: $$y_i^{(\country)}|\mu_{c[i]}, \sigma_y^2 \sim N(\mu_{c[i]}, \sigma_y^2),$$ where $c[i]$ refers to the country of country observation $i$.
+For national level estimation, we consider countries to be grouped in the world, i.e., $\mu_c^{(\country)}|\mu^{(\glob)}, \sigma_{ \country} \sim N(\mu^{(\glob)}, \sigma_{ \country}^2)$.
+Panel B introduces the likelihood function for national data, given by a normal likelihood with constant variance: $$y_i^{(\country)}|\mu_{c[i]}^{(\country)}, \sigma_y^2 \sim N(\mu_{c[i]}^{(\country)}, \sigma_y^2),$$ where $c[i]$ refers to the country of country observation $i$.
 We chose simple normal densities as data generating mechanisms in the simulation, to focus the modeling on the hierarchical model and the `localhierarchy` functionality as opposed to the data model aspects of modeling.
@@
 Simulated data were generated using `simulate_multilevel_data` function.
-We simulate a set of model parameters, setting $\mu^{(\glob)} = 0$, and $\sigma_{\country} = 0.5$, creating 30 countries, and national data, using $\sigma_y = 0.1$:
+We simulate a set of model parameters, setting $\mu^{(\glob)} = 0$, and $\sigma_{\country} = 0.5$, creating 30 countries, and national data, using $\sigma_y = 0.1$.
+The function call is as follows (note that the function also generates parameters and data at the subnational level, these are used in Section 5):
 
@@
   n_units_perlevel = c(30, 20),
-  sigma_perlevel = c(0.5, 0.2),
+  sigma_perlevel = c(0.5, 0.2), # national and subnational level
   mu_global = 0,
@@
   n_data = 10,
-  sigma_y = 0.1,
+  sigma_y = 0.1, # used for national and subnational data
   add_data1levelup = TRUE
@@
 
-In the simulated data, `level1` refers to the country level. The data sets contain the simulated parameters (`eta`) and data (`y`):
+In the simulated data, `level1` refers to the country level. The data sets contain the simulated parameters (`eta`; the deviation away from the global mean) and data (`y`):
 
@@
 
-Input data are processed using the function `get_geo_unit_index_data`. This function obtains unique hierarchical groupings from the input data and assigns index `c` to each unique grouping:
+Input data are processed using the function `get_geo_unit_index_data`. This function obtains unique hierarchical groupings from the input data and assigns an index `c` to each unique grouping:
 
@@
 
-Excerpts from the Stan model are copied below to introduce the model specification.
-The same Stan model is used for both global and local models.
+We can use the function `fit_model_localhierarchy` to fit the model as specified so far.
+The argument `runstep` is used to define the type of model, setting it to `"global_national"` results in fitting a model to all data without fixing any parameters.
+Other arguments include the hierarchical level specification, the area, and survey data:
+
+```{r}
+global_fit <- fit_model_localhierarchy(
+  runstep = "global_national",
+  hierarchical_level = hierarchical_level,
+  area = "level1",
+  survey_df = dat_nat
+)
+```
+
+When doing model fitting, the same Stan model is used for both global and local models. Excerpts from the Stan model are included in Appendix 1.
+
+The fitting returns posterior samples and other meta information about the fit.
+
+```{r}
+names(global_fit)
+```
+
+The `post_summ` object contains posterior means for relevant model parameters. For example:
+
+```{r}
+global_fit$post_summ |> 
+  filter(variable_no_index == "mu_sigma")
+```
+
+A local fit can be esetimated by passing on the global fit object to the wrapper function and including data for countries of interest.
+Here we do local runs for all countries, by including the entire data set, and specify what to fix using the `runstep` argument:
+
+```{r}
+fit_local <- fit_model_localhierarchy(
+  runstep = "local_national",
+  global_fit = global_fit,
+  area = "level1",
+  survey_df = dat_nat
+)
+```
+
+When doing a local national run with `runstep = "local_national"`, the hierarchical levels are taken from the global fit, and non-country-specific parameters are fixed.
+This is indicated in the print statements above.
+In the wrapper function, relevant code excerpts for `"local_national"` runs are as follows:
+
+```{r}
+hierarchical_level <- global_fit$hierarchical_level
+hierarchical_level_terms_fixed <- 
+  hierarchical_level[1:(length(hierarchical_level)-1)]
+hierarchical_level_sigmas_fixed <- 
+  hierarchical_level[1:(length(hierarchical_level))]
+```
+
+When doing a local fit, the same R functions and Stan model are used as introduced above but Stan input data is updated to fix a subset of parameters.
+Specifically, when producing Stan input data, the following function call is used:
+
+```{r}
+hier_stan_data_local <- hierarchical_param_stan_data(
+  param_name = "mu",
+  param_data = hier_data, 
+  # information to fix parameters 
+  global_fit = global_fit,
+  hierarchical_terms_fixed = hierarchical_level_terms_fixed, 
+  hierarchical_sigmas_fixed = hierarchical_level_sigmas_fixed
+)
+```
+
+where the arguments `hierarchical_terms_fixed` and `hierarchical_sigmas_fixed` are used to fix the $\mu$ at the intercept-level and the $\sigma$s up to and including the country-level.
+Indeed, when comparing the outputs of this function call to the one for a global fit, we find that the country $\sigma$ is fixed at the value from the global fit:
+
+```{r}
+hier_stan_data_local$mu_n_sigma_fixed
+hier_stan_data_local$mu_n_sigma_estimate
+hier_stan_data_local$mu_sigma_fixed
+```
+
+In addition to fixing parameters of hierarchical models, other non-country-specific parameters can be fixed as well.
+In this example, $\sigma_y$ is a global parameter that is fixed at the posterior mean from the global run for use in local runs.
+
+### Processing model fits
+
+There are several functions to produce and check outputs.
+The function `posterior_summary_hierparam` is used to summarize posterior draws for hierarchical parameters, to produce summaries of $\mu$ at any level of interest.
+It takes a fitted model object and a parameter name as input and returns a list with each entry referring one level, and the results (e.g., estimates for $\mu$ estimates) for that level.
+Here we obtain these results for the global and local model:
+
+```{r, message=FALSE, warning=FALSE}
+res_global <- posterior_summary_hierparam(fit = global_fit, parname = "mu")
+res_local <- posterior_summary_hierparam(fit = fit_local, parname = "mu")
+```
+
+Example output for the global model is as follows:
+
+```{r}
+res_local$level1
+```
+
+The column `quant` above refers to posterior means and upper and lower bounds of 95% credible intervals. We can create plots to show outputs, for example for the first 4 countries:
+
+```{r}
+plots <- plot_posterior_summaries(
+  res = res_global, 
+  modelname1 = "global",
+  hierarchy_select = "level1",
+  areas_select = unique(res_local$level1$name)[1:4]
+)
+
+plots$level1
+```
+
+By plotting results from global and local models, we can check how estimates compare and how values have been fixed in the local runs.
+For example, here we see how country estimates compare between global and local models:
+
+```{r}
+plots <- plot_posterior_summaries(
+  res = res_local, 
+  modelname1 = "local",
+  res2 = res_global, 
+  modelname2 = "global", 
+  hierarchy_select = "level1",
+  areas_select = unique(res_local$level1$name)[1:4]
+)
+
+plots$level1
+```
+
+Estimates are similar, as expected.
+
+When visualizing higher levels, we see how parameters were fixed in the local model:
+
+```{r}
+plots <- plot_posterior_summaries(
+  res = res_local, 
+  modelname1 = "local",
+  res2 = res_global, 
+  modelname2 = "global"
+)
+
+plots$intercept 
+```
+
+We can also create plots for the $\gamma$s (`mu_raw`s):
+
+```{r, message=FALSE, warning=FALSE}
+plots <- plot_mu_raw(fit = global_fit, parname = "mu")
+```
+
+This gives a list with plots, including summary plots and plots with priors and posteriors.
+For example, the figures below show the first 30 $\gamma$s, followed by the prior-posterior densities for the first $\gamma$ (referring to the intercept).
+
+```{r}
+plots[["summary_plots"]][[1]]
+```
+
+```{r}
+plots[["plots_allmuraw"]][[1]]
+```
+
+Plots with prior and posterior densities of the $\sigma$s can be produced as well:
+
+```{r}
+plot_prior_post_sigmas(fit = global_fit, parname = "mu")
+```
+
+## Application: National estimation of family planning indicators using FPET
+
+For family planning estimation, a local model fitting approach was motivated by the need to enable country monitoring and evaluation (M&E) officers to produce estimates for their countries, as part of the Track20 project (ref).
+The Track20 project works to build the capacity of countries to generate and use data to inform their programming, improve impact, and accelerate progress toward their FP goals.
+One of the pillars of this approach is the cultivation of a network of M&E officers dedicated to increasing the quality and use of FP data.
+We developed the family planning estimation tool (FPET) as a local version of a global family planning model, for use by the M&E officers.
+
+For local estimation of family planning indicators, the Bayesian model described in section 2 is used.
+Specifically, hierarchical models are used for estimation of transition model parameters.
+The workflow used for estimating family planning indicators at the national level is based on model fitting in several steps, including the estimation of long-term trends in the first step, the estimation of all model parameters related to modern use in a second step, the addition of information on the use of traditional methods of contraception in a third step, followed by local model fitting to produce estimates for all indicators of interest.
+The `localhierarchy` functionality is incorporated in the family planning modeling code base such that all steps are run with the same code base, and parameters can be passed from one model fit to another.
+Full details on this workflow and model specification are described elsewhere (ref).
+As an illustration, @fig-allfits shows global and local fits for modern use, demand, and demand satisfied for selected countries in Western Africa.
+The global fit in red corresponds closely to the local fits, shown in green.
+
+```{r}
+#| fig-height: 9
+#| fig-width: 7
+#| echo: false
+#| label: fig-allfits
+#| fig-cap: "Data, estimates, and forecasts from 1970 until 2030 for FP indicators for married women in in Benin (BEN), Gambia (GMB), and Ghana (GHA).
+FPET estimates are shown in in colored lines (red for global, green for local), with shaded areas representing 90% credible intervals.
+Survey data are shown in colored dots with 95% confidence intervals that reflect total error variance."
+plots <- readRDS("outputfpet_localglobalfits.rds")
+plots[[1]] /  plots[[2]]/ plots[[3]] # Stack three plots vertically
+```
+
+\clearpage
+
+# Use case 2: Subnational estimation
+
+## Workflow for nationally-informed subnational estimation
+
+For many global health applications, subnational estimation is of interest to inform policy and programmatic decision making at a more local level.
+If data are limited, models can be considered to share information across regions.
+We consider here the use of hierarchical models for an indicator of interest.
+The `localhierarchy` package can be used to fit subnational hierarchical models, following the same principles as for national estimation or through an updated approach.
+Just as for national-level hierarchical modeling, the starting point for subnational hierarchical modeling is to specify a hierarchical modeling structure for the subnational regions, such as the one from Figure 3.
+When following the same principles as for national-level hierarchical modeling, a global model can be fitted to a global data base of subnational data, followed by local models for subnational regions.
+
+In addition to a standard approach to subnational modeling, functionality in the `localhierarchy` package can also be used to fit subnational models that are informed by national level data and estimation, referred to here as *nationally-informed subnational estimation* (NISE).
+This approach is useful when subnational data are limited, for example, in terms of the number of countries with subnational data, the number of subnational regions with data, the number of observations per subnational region, or due to uncertainty associated with subnational data.
+An example is given in the case study on family planning estimation, where national level data are available for a large number of country-years while a global subnational data set is more limited in terms of the number of country-years included.
+
+The approach followed in NISE is illustrated in Figure 4, for the hierarchical structure from Figure 2.
+In the NISE approach, the first step is to fit a model to a global data base with national-level data, to obtain estimates of model parameters up to and including national level parameters.
+This national-level model fit is used in a global subnational fit to a global data base of subnational data.
+Specifically, in the global subnational fitting step, national level parameters are fixed at point estimates from the global national fit, i.e., $\hat{\sigma}_{ l}$ and $\mu_{\cdot}^{(l)}$ for level $l$ up to the country level.
+Finally, in a 3rd step, local models can be fitted, using fixed values from the global subnational fit.
+In the remainder of this section, we illustrate this approach using simulated data, and present an application of this approach for family planning estimation.
+
+```{=latex}
+\begin{figure}[h]
+\begin{adjustbox}{max size={\textwidth}{\textheight},center}
+\begin{tikzpicture}[
+  box/.style={rectangle, draw, thick, rounded corners, minimum width=2.5cm, align=left, inner sep=6pt},
+  eqbox/.style={box},
+  databox/.style={box, fill=data!20},
+  modelbox/.style={box, fill=param!20},
+  arrowdata/.style={-{Latex[length=3mm]}, thick, data},
+  arrowparam/.style={-{Latex[length=3mm]}, thick, param!80},
+  font=\sffamily
+]
+% --- Nodes ---
+\node[eqbox](model){
+Example subnat. hierarchical model:\\
+\(
+  \begin{aligned}
+\mu_w^{(\region)} &\sim N(\mu^{(\glob)}, \sigma_{ \region}^2),\\
+\mu_s^{(\subregion)} &\sim N(\mu_{w[s]}^{(\region)}, \sigma_{ \subregion}^2), \\
+\mu_c^{(\country)} &\sim N(\mu_{s[c]}^{(\subregion)}, \sigma_{ \country}^2),  \\
+\mu_r^{(\subnat)} &\sim N(\mu_{c[r]}^{(\country)}, \sigma_{ \subnat}^2),  \\
+\end{aligned}
+\)
+};
+
+\node[databox, right=5cm of model.north] (globaldata) {Global data base of country data};
+\node[modelbox, below=1cm of globaldata] (globalmodel) {%
+  Global national modeling:\\
+  Estimate all parameters in hierarchical \\ model 
+  $\leq$ country level 
+};
+\node[modelbox, below=1cm of globalmodel] (subglobalmodel) {%
+  Global subnational modeling:\\
+  Estimate subset of parameters in \\ hierarchical model, 
+  i.e., $\sigma_{ subnat}$
+};
+\node[databox, left=1cm of subglobalmodel] (globalsubdata) {Global data base\\ of subnational data};
+\node[modelbox, below=1cm of subglobalmodel] (localmodel) {%
+  Local subnational modeling:\\
+  Estimate subnational parameters only, \\i.e., $\mu_r^{(subnat)}$ in $\mu_r^{(subnat)} \sim N(\hat{\mu}_{c[r]}^{(country)}, \hat{\sigma}_{ subnat}^2).$
+};
+
+% --- Data arrow  ---
+\draw[arrowdata] (globaldata.south) -- (globalmodel.north);
+\draw[arrowdata] (globaldata.east) ++(0,0) -- ++(1.5,0) |- (localmodel.east)
+node[midway,above] {};
+\draw[arrowdata] (globaldata.east) ++(0,0) -- ++(1.5,0) |- (subglobalmodel.east)
+node[midway,above] {};
+\draw[arrowdata] (globalsubdata.east) -- (subglobalmodel.west);
+\draw[arrowdata] (globalsubdata.south) -- (localmodel.west)
+node[midway,above] {};
+
+% --- Parameter arrow ---
+\draw[arrowparam] (globalmodel.south) -- (subglobalmodel.north);
+\draw[arrowparam] (subglobalmodel.south) -- (localmodel.north);
+
+% --- Legend ---
+% \node[arrowparam, below=1.2cm of localmodel.west, anchor=west] (leg1) {}; 
+% \node[right=0.3cm of leg1] {\textcolor{param}{$\rightarrow$ Parameter estimates}};
+\node[below=1.2cm of localmodel.west](parameter) {\textcolor{param}{$\rightarrow$ Parameter estimates}};
+\node[right=1cm of parameter] {\textcolor{data}{$\rightarrow$ Data}};
+\end{tikzpicture}
+\end{adjustbox}
+\caption{Workflow for nationally-informed global and local subnational modeling. }
+\end{figure}
+\clearpage
+```
+
+## Detailed example of subnational estimation using simulated data
+
+### Simulation setup
+
+The simulation setup for subnational parameters and data is summarized in Figure 8, Panels A-B-C.
+As explained above for national estimation, Panel A indicates the hierarchical model considered.
+For subnational estimation, an additional level is added, with $\mu_r^{(\subnat)} = \mu_{c[r]}^{(\country)} + \eta_{r}^{(\subnat)}$, $\eta_{r}^{(\subnat)}|\sigma_{ \subnat} \sim N(0, \sigma_{ \subnat})$.
+
+As explained earlier as well, we chose simple normal densities as data generating mechanisms in the simulation, to focus the modeling on the hierarchical model and the `localhierarchy` functionality as opposed to the data model aspects of modeling.
+Subnational data follows a normal distribution with constant variance: $$y^{(subnat)}_i|\mu^{(subnat)}_{r[i]}, \sigma_y^2 \sim N(\mu^{(subnat)}_{r[i]}, \sigma_y^2),$$ where $r[i]$ refers to the subnational region of subnational observation $i$.
+In this simulation setup, for simplicity, we generate subnational data independently of national-level data, and we use the same constant data variance at the national and subnational levels.
+In real examples, data variances are likely to differ (for example, because subnational sampling errors are larger at the subnational level) and national and subnational data may be derived from the same source.
+We discuss an example of those settings in the case study.
+
+Simulated data at the subnational level was generated using the `simulate_multilevel_data` function described above.
+For subnational levels, we set $\sigma_{\subnat} = 0.2$, and we created 20 subnational regions in each country and 10 observations per region.
+In the simulated data, `level1` refers to the country level and `level2` refers to the subnational level.
+The data sets contain the simulated parameters (`eta`) and data (`y`):
+
+```{r}
+dat_subnat <- dat_all$data
+dat_subnat |> select(level1, level2, mu_global, level1_eta, level2_eta, y)
+```
+
+### Modeling
+
+A NISE model is fitted as follows:
+
+```{r}
+area_subnat <- "level2"
+global_fitsubnat_seq <- fit_model_localhierarchy(
+   runstep = "global_subnational",
+   area = area_subnat, # subnational area name 
+   survey_df = dat_subnat, # pass on subnat data
+   # for NISE:
+   global_fit = global_fit, # global national fit
+   # what level to add to the hierarchy, compared to the 
+   # global national model 
+   add_subnational_hierarchy = area_subnat
+  )
+```
+
+Through the `add_subnational_hierarchy` argument, and as shown in the printed statements, the hierarchical structure is now given by intercept $\rightarrow$ country $\rightarrow$ subnational level.
+By default, the world mean and across-country variance are fixed.
+Code excerpts to do so are as follows:
+
+```{r, eval = FALSE}
+hierarchical_level <- global_fit$hierarchical_level
+hierarchical_level <- c(hierarchical_level, add_subnational_hierarchy)
+hierarchical_level_terms_fixed <-
+  hierarchical_level[1:(length(hierarchical_level)-2)]
+hierarchical_level_sigmas_fixed <-
+  hierarchical_level[1:(length(hierarchical_level)-1)]
+```
+
+By comparing the fit summaries from the global national and global subnational fits for the $\sigma$s, we see that, as expected, one additional $\sigma_{\mu}$ is estimated in the subnational model (to capture variability across subnational regions) and the $\sigma_{\mu}$ at the higher country level is the same in the global national and subnational fit:
+
+```{r}
+global_fit$post_summ |>
+  filter(str_detect(variable_no_index, "sigma"))
+global_fitsubnat_seq$post_summ |>
+  filter(str_detect(variable_no_index, "sigma"))
+```
+
+We can use the subnational global fit to do a local subnational fit, using the `runstep = "local_subnational"` setting:
+
+```{r}
+fit_local_seq <- fit_model_localhierarchy(
+  runstep = "local_subnational",
+  area = area_subnat,
+  survey_df = dat_subnat,
+  global_fit = global_fitsubnat_seq
+)
+```
+
+Similarly to the local national run, in subnational local runs, hierarchical levels and which parameters are fixed is set in the wrapper function; we estimate subnational parameters only.
+Point estimates in the subnational global fit include point estimates from the national global run as well as point estimates from the global subnational fit.
+By default, parameters that are duplicated between the two fits, e.g., world region or country means, are taken from the national global run.
+This is implemented as such to avoid the situation where local runs would fail for countries that were not included in the subnational global fitting, i.e., when country parameters were fixed and missing from the subnational fit.
+Alternatively, if the user wants to use parameter estimates from the subnational global fit in these settings, the argument `use_globalsubnat_fromnat` can be used when calling the `fit_model_localhierarchy` function, see documentation.
+<!-- If the fixed parameters differ between national and subnational global fits, i.e., if country means are re-estimated in the global subnational fit as in this example, and if the user wants to use these fixed parameter estimates from the subnational global fit, i.e., if the user wants to use the fixed country estimates from the global national run, the argument `use_globalsubnat_fromnat` can be used when calling the `fit_model_localhierarchy` function, see documentation.
+-->
+
+We can create and check summaries of subnational fits using the same functions for national fits.
+Below we compare the subnational global fit with the local subnational fit.
+The results are comparable, as expected.
+
+```{r, message=FALSE, warning=FALSE}
+res_globalsubnat_seq <- posterior_summary_hierparam(
+  fit = global_fitsubnat_seq,
+  parname = "mu"
+)
+
+res_local_seq <- posterior_summary_hierparam(
+  fit = fit_local_seq, 
+  parname = "mu"
+)
+```
+
+```{r}
+plots <- plot_posterior_summaries(
+  res = res_local_seq, 
+  modelname1 = "local",
+  res2 = res_globalsubnat_seq, modelname2 = "global",
+  hierarchy_select = "level2",
+  areas_select = unique(res_local_seq$level2$name)[1:30]
+)
+
+plots$level2
+```
+
+<!-- could add back the direct subnat fit here too: two local subnational fits: one created from the sequential approach introduced above, and one create from a subnational global fit that did not use national level information.
+The results are comparable, as expected.
+-->
+
+## Application: Subnational estimation of family planning indicators using FPET
+
+Subnational estimation of family planning indicators is of interest to country M&E officers for informing family planning programming.
+To that end, the Track20 project compiled survey data at the subnational level over 35 low- and middle income countries.
+
+For subnational estimation, the workflow used mirrors the NISE approach described in this paper: global modeling of national-level indicators is carried out first, followed by subnational model fitting using a global data base, and finally by local fits.
+This sequential approach, in which the global subnational model is based on global national modeling, was motivated by differences in data availability between national and subnational global databases: When estimating family planning indicators for married women of reproductive age, the global national database contains observations from 198 countries, whereas the global subnational database includes observations from 35 countries only.
+
+The figures below show illustrations of regional fits for Benin.
+The subnational model is fitted to all regional data from the country.
+@fig-subnatoverview illustrates fits in selected regions and differences across regions in 2024.
+
+```{r}
+#| fig-height: 9
+#| fig-width: 7
+#| echo: false
+#| label: fig-subnatfits
+#| fig-cap: "Data, estimates, and forecasts from 1970 until 2030 for FP indicators for married women in subnational regions in Benin.
+Graphs show mCPR, unmet need for modern methods, demand, and demand satisfied with modern methods.
+Survey data are shown in colored dots with 95% confidence intervals that reflect total error variance.
+FPET estimates are shown in black, with grey shaded areas representing 90% credible intervals.
+"
+plots <- readRDS("outputfpet_plotssubnat_BEN.rds")
+plots[["Oueme"]]$married / plots[["Collines"]]$married/ plots[["Donga"]]$married
+```
+
+```{r}
+#| echo: false
+#| label: fig-subnatoverview
+#| fig-cap: "Regional estimates for Benin for modern contraceptive use in 2024. Point estimates and 95% credible intervals shown."
+plots[["comp"]] 
+```
+
+<!-- The figures below show illustrations of regional fits for Cote d'Ivoire.
+The subnational model is fitted to all data from the country, here the combination of regional data based on the most recent DHS survey, and national data for other survey-years (not shown in regional plots).
+This results in estimates for recent years based on all available information.
+An example, estimates of modern contraceptive use in 2024, is shown as well.
+-->
+
+# Conclusions
+
+Bayesian hierarchical models are widely used in global health estimation, where data availability may vary across national or subnational populations.
+For analysis at a local level, local models are often of interest, referring to models that are derived from global hierarchical models but adapted such that they can be fitted to the data from one population alone.
+To that end, we introduced the `localHierarchy` R package, which provides functionality for fitting Bayesian hierarchical models in settings where both global and local estimation is required.
+
+We have used the `localHierarchy` functionality in different projects.
+We discussed the estimation of family planning indicators, where country M&E officers use the local model to produce estimates at the national and subnational level.
+In this area, M&E officers have successfully used local models to produce national estimates since 2014.
+Availability of a local modeling tool has enabled users to do additional analyses and check sensitivity of estimates to specific data sources or data quality assumptions.
+In addition to fitting the model to survey data, users of the local family planning model can also incorporate additional data from routine information systems (e.g., the number of contraceptive commodities supplied, add REF).
+These use cases illustrate the value of local models to enable local analysis.
+
+More recently, we have used the `localHierarchy` approach and functionality for estimating coverage indicators for the Countdown to 2030 initiative (ref).
+By developing the `localHierarchy` package and providing use cases in this paper, we hope that this work can be useful for other global health modelers, to consider local modeling in their own work.
+
+# Appendix
+
+## Appendix I: Excerpts from the Stan model
+
+The same Stan model is used for both global and local models. Excerpts are given in this section to explain key parts.
+
 The first excerpt starts at the end in terms of model specification and concerns the specification of the hierarchical parameters $\mu_c$, using the `localhierarchy` Stan function `get_mu`:
@@
 
-Here we use that $(\mu_1, ..., \mu_c) = M \cdot (\eta_1, ..., \eta_c)$, where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_c)$ refers to the vector with products of standard deviation and $\gamma$ terms.
-The function `get_mu` calculates the vector of $\mu_c$s from the vector of $\eta$s (`mu_star` in the Stan model) and a decomposition of $M$ into different parts $U$, $V$, $W$.
+Here we use that $(\mu_1^{(country)}, ..., \mu_C^{(country)}) = M \cdot (\eta_1, ..., \eta_C)$, where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_c)$ refers to the vector with products of standard deviation and $\gamma$ terms.
+The function `get_mu` calculates the vector of $\mu_c^{(country)}$s from the vector of $\eta$s (`mu_star` in the Stan model) and a decomposition of $M$ into different parts $U$, $V$, $W$.
 This implementation uses the sparse representation of the model matrix $M$ with Stan functions `csr_matrix_times_vector` and `crs_extract`.
@@
 The second Stan model excerpt concerns the specification of the $\eta$s (`mu_star`).
-This vector is obtained by multiplying the vector of $\gamma$s, which is referred to as `mu_raw` in the Stan model by their respective standard deviation, i.e., the vector containing $s_{\mu, global}$ (labeled `mu_scalarprior_mean` in the Stan model) and the vector of stacked $\sigma$s (labeled `mu_sigma` in the Stan model).
+This vector is obtained by multiplying the vector of $\gamma$s, which is referred to as `mu_raw` in the Stan model by their respective standard deviation, i.e., the vector containing $s_{\glob}$ (labeled `mu_scalarprior_mean` in the Stan model) and the vector of stacked $\sigma$s (labeled `mu_sigma` in the Stan model).
 To allow for global and local models, additions `_estimate` and `_fixed` are used to allow for splitting the `mu_raw` vector into two parts, one for fixed and one for estimated parameters.
@@
 Its arguments also include prior parameters `mu_scalarprior_mean` and `mu_scalarprior_sd`, which refer to the (rescaled) prior mean and standard deviation of $\mu^{(\glob)}$.
-The model block contains the priors for the $\gamma$s and $\sigma_{\mu}$s, given by \begin{align*}
+The model block contains the priors for the $\gamma$s and $\sigma_{k}$s, given by
+\begin{align*}
   \mu^{(\glob)} &\sim N(0,1),\\
-  \sigma_{\mu} &\sim \text{Half-Normal}(0, s_{\mu}).
+  \sigma_{k} &\sim \text{Half-Normal}(0, s_{\sigma}).
 \end{align*} All prior parameters are added as input data to the Stan model.
@@
 (not doing the /sd would be easier for specification going forward?) -->
-
-We can use the function `fit_model_localhierarchy` to fit the model as specified so far.
-The argument `runstep` is used to define the type of model, setting it to `global_national` results in fitting a model to all data without fixing any parameters.
-Other arguments include the hierarchical level specification, the area, and survey data:
-
-```{r}
-global_fit <- fit_model_localhierarchy(
-  runstep = "global_national",
-  hierarchical_level = hierarchical_level,
-  area = "level1",
-  survey_df = dat_nat
-)
-```
-
-The fitting returns posterior samples and other meta information about the fit.
-
-```{r}
-names(global_fit)
-```
-
-The `post_summ` object contains posterior means for relevant model parameters. For example:
-
-```{r}
-global_fit$post_summ |> 
-  filter(variable_no_index == "mu_sigma")
-```
-
-A local fit can be esetimated by passing on the global fit object to the wrapper function and including data for countries of interest.
-Here we do local runs for all countries, by including the entire data set, and specify what to fix using the `runstep` argument:
-
-```{r}
-fit_local <- fit_model_localhierarchy(
-  runstep = "local_national",
-  global_fit = global_fit,
-  area = "level1",
-  survey_df = dat_nat
-)
-```
-
-When doing a local national run with `runstep = local_national`, the hierarchical levels are taken from the global fit, and non-country-specific parameters are fixed.
-This is indicated in the print statements above.
-In the wrapper function, relevant code excerpts for `local_national` runs are as follows:
-
-```{r}
-hierarchical_level <- global_fit$hierarchical_level
-hierarchical_level_terms_fixed <- 
-  hierarchical_level[1:(length(hierarchical_level)-1)]
-hierarchical_level_sigmas_fixed <- 
-  hierarchical_level[1:(length(hierarchical_level))]
-```
-
-When doing a local fit, the same R functions and Stan model is used as introduced above but Stan input data is updated to fix a subset of parameters.
-Specifically, when producing Stan input data, the following function call is used:
-
-```{r}
-hier_stan_data_local <- hierarchical_param_stan_data(
-  param_name = "mu",
-  param_data = hier_data, 
-  # information to fix parameters 
-  global_fit = global_fit,
-  hierarchical_terms_fixed = hierarchical_level_terms_fixed, 
-  hierarchical_sigmas_fixed = hierarchical_level_sigmas_fixed
-)
-```
-
-where the arguments `hierarchical_terms_fixed` and `hierarchical_sigmas_fixed` are used to fix the $\mu$ at the intercept-level and the $\sigma$s up to and including the country-level.
-Indeed, when comparing the outputs of this function call to the one for a global fit, we find that the country $\sigma$ is fixed at the value from the global fit:
-
-```{r}
-hier_stan_data_local$mu_n_sigma_fixed
-hier_stan_data_local$mu_n_sigma_estimate
-hier_stan_data_local$mu_sigma_fixed
-```
-
-In addition to fixing parameters of hierarchical models, other non-country-specific parameters can be fixed as well.
-In this example, $\sigma_y$ is a global parameter that is fixed at the posterior mean from the global run for use in local runs.
-
-### Processing model fits
-
-There are several functions to produce and check outputs.
-The function `posterior_summary_hierparam` is used to summarize posterior draws for hierarchical parameters, to produce summaries of $\mu$ at any level of interest.
-It takes a fitted model object and a parameter name as input and returns a list with each entry referring one level, and the results (e.g., estimates for $\mu$ estimates) for that level.
-Here we obtain these results for the global and local model:
-
-```{r, message=FALSE, warning=FALSE}
-res_global <- posterior_summary_hierparam(fit = global_fit, parname = "mu")
-res_local <- posterior_summary_hierparam(fit = fit_local, parname = "mu")
-```
-
-Example output for the global model is as follows:
-
-```{r}
-res_local$level1
-```
-
-The column `quant` above refers to posterior means and upper and lower bounds of 95% credible intervals. We can create plots to show outputs, for example for the first 4 countries:
-
-```{r}
-plots <- plot_posterior_summaries(
-  res = res_global, 
-  modelname1 = "global",
-  hierarchy_select = "level1",
-  areas_select = unique(res_local$level1$name)[1:4]
-)
-
-plots$level1
-```
-
-By plotting results from global and local models, we can check how estimates compare and how values have been fixed in the local runs.
-For example, here we see how country estimates compare between global and local models:
-
-```{r}
-plots <- plot_posterior_summaries(
-  res = res_local, 
-  modelname1 = "local",
-  res2 = res_global, 
-  modelname2 = "global", 
-  hierarchy_select = "level1",
-  areas_select = unique(res_local$level1$name)[1:4]
-)
-
-plots$level1
-```
-
-Estimates are similar, as expected.
-
-When visualizing higher levels, we see how parameters were fixed in the local model:
-
-```{r}
-plots <- plot_posterior_summaries(
-  res = res_local, 
-  modelname1 = "local",
-  res2 = res_global, 
-  modelname2 = "global"
-)
-
-plots$intercept 
-```
-
-We can also create plots for the $\gamma$s (`mu_raw`s):
-
-```{r, message=FALSE, warning=FALSE}
-plots <- plot_mu_raw(fit = global_fit, parname = "mu")
-```
-
-This gives a list with plots, including summary plots and plots with priors and posteriors.
-For example, the figures below show the first 30 $\gamma$s, followed by the prior-posterior densities for the first $\gamma$ (referring to the intercept).
-
-```{r}
-plots[["summary_plots"]][[1]]
-```
-
-```{r}
-plots[["plots_allmuraw"]][[1]]
-```
-
-Plots with prior and posterior densities of the $\sigma$s can be produced as well:
-
-```{r}
-plot_prior_post_sigmas(fit = global_fit, parname = "mu")
-```
-
-## Application: National estimation of family planning indicators using FPET
-
-For family planning estimation, a local model fitting approach was motivated by the need to enable country monitoring and evaluation (M&E) officers to produce estimates for their countries, as part of the Track20 project (ref).
-The Track20 project works to build the capacity of countries to generate and use data to inform their programming, improve impact, and accelerate progress toward their FP goals.
-One of the pillars of this approach is the cultivation of a network of M&E officers dedicated to increasing the quality and use of FP data.
-We developed the family planning estimation tool (FPET), as a local version of a global family planning model, for use by the M&E officers.
-
-For local estimation of family planning indicators, the Bayesian model described in section 2 is used.
-Specifically, hierarchical models are used for estimation of transition model parameters.
-The workflow used for estimating family planning indicators at the national level is based on model fitting in several steps, including the estimation of long-term trends in the first step, the estimation of all model parameters related to modern use in a second step, the addition of information on the use of traditional methods of contraception in a third step, followed by local model fitting to produce estimates for all indicators of interest.
-The `localhierarchy` functionality is incorporated in the family planning modeling code base such that all steps are run with the same code base, and parameters can be passed from one model fit to another.
-Full details on this workflow and model specification are described elsewhere (ref).
-As an illustration, @fig-allfits shows global and local fits for modern use, demand, and demand satisfied for selected countries in Western Africa.
-The global fit in red corresponds closely to the local fits, shown in green.
-
-```{r}
-#| fig-height: 9
-#| fig-width: 7
-#| echo: false
-#| label: fig-allfits
-#| fig-cap: "Data, estimates, and forecasts from 1970 until 2030 for FP indicators for married women in in Benin (BEN), Gambia (GMB), and Ghana (GHA).
-FPET estimates are shown in in colored lines (red for global, green for local), with shaded areas representing 90% credible intervals.
-Survey data are shown in colored dots with 95% confidence intervals that reflect total error variance."
-plots <- readRDS("outputfpet_localglobalfits.rds")
-plots[[1]] /  plots[[2]]/ plots[[3]] # Stack three plots vertically
-```
-\clearpage 
-# Use case 2: Subnational estimation
-
-## Workflow for nationally-informed subnational estimation
-
-For many global health applications, subnational estimation is of interest to inform policy and programmatic decision making at a more local level.
-If data are limited, models can be considered to share information across regions.
-We consider here the use of hierarchical models for an indicator of interest.
-The `localhierarchy` package can be used to fit subnational hierarchical models, following the same principles as for national estimation or through an updated approach.
-Just as for national-level hierarchical modeling, the starting point for subnational hierarchical modeling is to specify a hierarchical modeling structure for the subnational regions, such as the one from Figure 2.
-When following the same principles as for national-level hierarchical modeling, a global model can be fitted to a global data base of subnational data, followed by local models for subnational regions.
-
-In addition to a standard approach to subnational modeling, functionality in the `localhierarchy` package can also be used to fit subnational models that are informed by national level data and estimation, referred to here as *nationally-informed subnational estimation* (NISE).
-This approach is useful when subnational data are limited, for example, in terms of the number of countries with subnational data, the number of subnational regions with data, the number of observations per subnational region, or due to uncertainty associated with subnational data.
-An example is given in the case study on family planning estimation, where national level data are available for a large number of country-years while a global subnational data set is more limited in terms of the number of country-years included.
-
-The approach followed in NISE is illustrated in Figure 4, for the hierarchical structure from Figure 2.
-In the NISE approach, the first step is to fit a model to a global data base with national-level data, to obtain estimates of model parameters up to and including national level parameters.
-This national-level model fit is used in a global subnational fit to a global data base of subnational data.
-Specifically, in the global subnational fitting step, national level parameters are fixed at point estimates from the global national fit, i.e., $\hat{\sigma}_{ l}$ and $\mu_{\cdot}^{(l)}$ for level $l$ up to the country level.
-Finally, in a 3rd step, local models can be fitted, using fixed values from the global subnational fit.
-In the remainder of this section, we illustrate this approach using simulated data, and present an application of this approach for family planning estimation.
-
-```{=latex}
-\begin{figure}[h]
-\begin{adjustbox}{max size={\textwidth}{\textheight},center}
-\begin{tikzpicture}[
-  box/.style={rectangle, draw, thick, rounded corners, minimum width=2.5cm, align=left, inner sep=6pt},
-  eqbox/.style={box},
-  databox/.style={box, fill=data!20},
-  modelbox/.style={box, fill=param!20},
-  arrowdata/.style={-{Latex[length=3mm]}, thick, data},
-  arrowparam/.style={-{Latex[length=3mm]}, thick, param!80},
-  font=\sffamily
-]
-% --- Nodes ---
-\node[eqbox](model){
-Example subnat. hierarchical model:\\
-\(
-  \begin{aligned}
-\mu_w^{(region)} &\sim N(\mu^{(global)}, \sigma_{ region}^2),\\
-\mu_s^{(subregion)} &\sim N(\mu_{w[s]}^{(region)}, \sigma_{ subregion}^2), \\
-\mu_c^{(country)} &\sim N(\mu_{s[c]}^{(subregion)}, \sigma_{ country}^2),  \\
-\mu_r^{(subnat)} &\sim N(\mu_{r[c]}^{(country)}, \sigma_{ subnat}^2),  \\
-\end{aligned}
-\)
-};
-
-\node[databox, right=5cm of model.north] (globaldata) {Global data base of country data};
-\node[modelbox, below=1cm of globaldata] (globalmodel) {%
-  Global national modeling:\\
-  Estimate all parameters in hierarchical \\ model 
-  $\leq$ country level 
-};
-\node[modelbox, below=1cm of globalmodel] (subglobalmodel) {%
-  Global subnational modeling:\\
-  Estimate subset of parameters in \\ hierarchical model, 
-  i.e., $\sigma_{ subnat}$
-};
-\node[databox, left=1cm of subglobalmodel] (globalsubdata) {Global data base\\ of subnational data};
-\node[modelbox, below=1cm of subglobalmodel] (localmodel) {%
-  Local subnational modeling:\\
-  Estimate subnational parameters only, \\i.e., $\mu_r$ in $\mu_r \sim N(\hat{\mu}_{c[r]}^{(country)}, \hat{\sigma}_{ subnat}^2).$
-};
-
-% --- Data arrow  ---
-\draw[arrowdata] (globaldata.south) -- (globalmodel.north);
-\draw[arrowdata] (globaldata.east) ++(0,0) -- ++(1.5,0) |- (localmodel.east)
-node[midway,above] {};
-\draw[arrowdata] (globaldata.east) ++(0,0) -- ++(1.5,0) |- (subglobalmodel.east)
-node[midway,above] {};
-\draw[arrowdata] (globalsubdata.east) -- (subglobalmodel.west);
-\draw[arrowdata] (globalsubdata.south) -- (localmodel.west)
-node[midway,above] {};
-
-% --- Parameter arrow ---
-\draw[arrowparam] (globalmodel.south) -- (subglobalmodel.north);
-\draw[arrowparam] (subglobalmodel.south) -- (localmodel.north);
-
-% --- Legend ---
-% \node[arrowparam, below=1.2cm of localmodel.west, anchor=west] (leg1) {}; 
-% \node[right=0.3cm of leg1] {\textcolor{param}{$\rightarrow$ Parameter estimates}};
-\node[below=1.2cm of localmodel.west](parameter) {\textcolor{param}{$\rightarrow$ Parameter estimates}};
-\node[right=1cm of parameter] {\textcolor{data}{$\rightarrow$ Data}};
-\end{tikzpicture}
-\end{adjustbox}
-\caption{Workflow for nationally-informed global and local subnational modeling. }
-\end{figure}
-\clearpage
-```
-
-## Detailed example of subnational estimation using simulated data
-
-### Simulation setup
-
-The simulation setup for subnational parameters and data is summarized in Figure 4, Panels A-B-C.
-As explained above for national estimation, Panel A Figure 4 indicates the hierarchical model considered.
-For subnational estimation, an additional level is added, with $\mu_r^{(\subnat)} = \mu_{c[r]} + \eta_{r}^{(\subnat)}$, $\eta_{r}^{(\subnat)}|\sigma_{ \subnat} \sim N(0, \sigma_{ \subnat})$.
-
-As explained earlier as well, we chose simple normal densities as data generating mechanisms in the simulation, to focus the modeling on the hierarchical model and the `localhierarchy` functionality as opposed to the data model aspects of modeling.
-Subnational data follows a normal distribution with constant variance: $$y^{(subnat)}_i|\mu^{(subnat)}_{g[i]}, \sigma_y^2 \sim N(\mu^{(subnat)}_{r[i]}, \sigma_y^2),$$ where $g[i]$ refers to the subnational region of subnational observation $i$.
-In this simulation setup, for simplicity, we generate subnational data independently of national-level data, and we use the same constant data variance at the national and subnational levels.
-In real examples, data variances are likely to differ (for example, because subnational sampling errors are larger at the subnational level) and national and subnational data may be derived from the same source.
-We discuss an example of those settings in the case study.
-
-Simulated data at the subnational level was generated using the `simulate_multilevel_data` function described above.
-For subnational levels, we set $\sigma_{\mu}^{(\subnat)} = 0.2$, and we created 20 subnational regions in each country and 10 observations per region.
-In the simulated data, `level1` refers to the country level and `level2` refers to the subnational level.
-The data sets contain the simulated parameters (`eta`) and data (`y`):
-
-```{r}
-dat_subnat <- dat_all$data
-dat_subnat |> select(level1, level2, mu_global, level1_eta, level2_eta, y)
-```
-
-### Modeling
-
-A NISE model is fitted as follows:
-
-```{r}
-area_subnat <- "level2"
-global_fitsubnat_seq <- fit_model_localhierarchy(
-   runstep = "global_subnational",
-   area = area_subnat, # subnational area name 
-   survey_df = dat_subnat, # pass on subnat data
-   # for NISE:
-   global_fit = global_fit, # global national fit
-   # what level to add to the hierarchy, compared to the 
-   # global national model 
-   add_subnational_hierarchy = area_subnat
-  )
-```
-
-Through the `add_subnational_hierarchy` argument, and as shown in the printed statements, the hierarchical structure is now given by intercept $\rightarrow$ country $\rightarrow$ subnational level.
-By default, the world mean and across-country variance are fixed.
-Code excerpts to do so are as follows:
-
-```{r, eval = FALSE}
-hierarchical_level <- global_fit$hierarchical_level
-hierarchical_level <- c(hierarchical_level, add_subnational_hierarchy)
-hierarchical_level_terms_fixed <-
-  hierarchical_level[1:(length(hierarchical_level)-2)]
-hierarchical_level_sigmas_fixed <-
-  hierarchical_level[1:(length(hierarchical_level)-1)]
-```
-
-By comparing the fit summaries from the global national and global subnational fits for the $\sigma$s, we see that, as expected, one additional $\sigma_{\mu}$ is estimated in the subnational model (to capture across-subnational region variability) and the $\sigma_{\mu}$ at the higher country level is the same in the global national and subnational fit:
-
-```{r}
-global_fit$post_summ |>
-  filter(str_detect(variable_no_index, "sigma"))
-global_fitsubnat_seq$post_summ |>
-  filter(str_detect(variable_no_index, "sigma"))
-```
-
-We can use the subnational global fit to do a local subnational fit, using the `runstep = "local_subnational"` setting:
-
-```{r}
-fit_local_seq <- fit_model_localhierarchy(
-  runstep = "local_subnational",
-  area = area_subnat,
-  survey_df = dat_subnat,
-  global_fit = global_fitsubnat_seq
-)
-```
-
-Similarly to the local national run, in subnational local runs, hierarchical levels and which parameters are fixed is set in the wrapper function; we estimate subnational parameters only.
-Point estimates in the subnational global fit include point estimates from the national global run as well as point estimates from the global subnational fit.
-By default, parameters that are duplicated between the two fits, e.g., world region or country means, are taken from the national global run.
-This is implemented as such to avoid the situation where local runs would fail for countries that were not included in the subnational global fitting, i.e., when country parameters were fixed and missing from the subnational fit.
-If the fixed parameters differ between national and subnational global fits, i.e., if country means are re-estimated in the global subnational fit as in this example, and if the user wants to use these fixed parameter estimates from the subnational global fit, i.e., if the user wants to use the fixed country estimates from the global national run, the argument `use_globalsubnat_fromnat` can be used when calling the `fit_model_localhierarchy` function, see documentation.
-
-We can create and check summaries of subnational fits using the same functions for national fits.
-Below we compare the subnational global fit with the local subnational fit.
-The results are comparable, as expected.
-
-```{r, message=FALSE, warning=FALSE}
-res_globalsubnat_seq <- posterior_summary_hierparam(
-  fit = global_fitsubnat_seq,
-  parname = "mu"
-)
-
-res_local_seq <- posterior_summary_hierparam(
-  fit = fit_local_seq, 
-  parname = "mu"
-)
-```
-
-```{r}
-plots <- plot_posterior_summaries(
-  res = res_local_seq, 
-  modelname1 = "local",
-  res2 = res_globalsubnat_seq, modelname2 = "global",
-  hierarchy_select = "level2",
-  areas_select = unique(res_local_seq$level2$name)[1:30]
-)
-
-plots$level2
-```
-
-<!-- could add back the direct subnat fit here too: two local subnational fits: one created from the sequential approach introduced above, and one create from a subnational global fit that did not use national level information.
-The results are comparable, as expected.
--->
-
-## Application: Subnational estimation of family planning indicators using FPET
-
-Subnational estimation of family planning indicators is of interest to country M&E officers for informing family planning programming.
-To that end, the Track20 project compiled survey data at the subnational level over 35 low- and middle income countries.
-
-For subnational estimation, the workflow used mirrors the NISE approach described in this paper: global modeling of national-level indicators is carried out first, followed by  subnational model fitting using a global data base, and finally by local fits.
-This sequential approach, in which the global subnational model is based on global national modeling, was motivated by differences in data availability between national and subnational global databases: When estimating family planning indicators for married women of reproductive age, the global national database contains observations from 198 countries, whereas the global subnational database includes observations from 35 countries only.
-
-The figures below show illustrations of regional fits for Benin.
-The subnational model is fitted to all regional data from the country.
-@fig-subnatoverview illustrates fits in selected regions and differences across regions in 2024.
-
-```{r}
-#| fig-height: 9
-#| fig-width: 7
-#| echo: false
-#| label: fig-subnatfits
-#| fig-cap: "Data, estimates, and forecasts from 1970 until 2030 for FP indicators for married women in subnational regions in Benin.
-Graphs show mCPR, unmet need for modern methods, demand, and demand satisfied with modern methods.
-Survey data are shown in colored dots with 95% confidence intervals that reflect total error variance.
-FPET estimates are shown in black, with grey shaded areas representing 90% credible intervals.
-"
-plots <- readRDS("outputfpet_plotssubnat_BEN.rds")
-plots[["Oueme"]]$married / plots[["Collines"]]$married/ plots[["Donga"]]$married
-```
-
-```{r}
-#| echo: false
-#| label: fig-subnatoverview
-#| fig-cap: "Regional estimates for Benin for modern contraceptive use in 2024. Point estimates and 95% credible intervals shown."
-plots[["comp"]] 
-```
-
-<!-- The figures below show illustrations of regional fits for Cote d'Ivoire.
-The subnational model is fitted to all data from the country, here the combination of regional data based on the most recent DHS survey, and national data for other survey-years (not shown in regional plots).
-This results in estimates for recent years based on all available information.
-An example, estimates of modern contraceptive use in 2024, is shown as well.
--->
-
-# Conclusions 
-Bayesian hierarchical models are widely used in global health estimation, where data availability may vary across national or subnational populations.
-For analysis at a local level, local models are often of interest, referring to models that are derived from global hierarchical models but adapted such that they can be fitted to the data from one population alone.
-To that end, we introduced the `localHierarchy` R package, which provides functionality for fitting Bayesian hierarchical models in settings where both global and local estimation is required.
-
-We have used the `localHierarchy` functionality in different projects.
-We discussed the estimation of family planning indicators, where country M&E officers use the local model to produce estimates at the national and subnational level.
-In this area, M&E officers have successfully used local models to produce national estimates since 2013 [LA to check year].
-Availability of a local modeling tool has enabled users to do additional analyses and check sensitivity of estimates to specific data sources or data quality assumptions.
-In addition to fitting the model to survey data, users of the local family planning model can also incorporate additional data from routine information systems  (e.g., the number of contraceptive commodities supplied, add REF).
-These use cases illustrate the value of local models to enable local analysis.
-
-More recently, we have used the `localHierarchy` approach and functionality for estimating coverage indicators for the Countdown to 2030 initiative (ref).
-By developing the `localHierarchy` package and providing use cases in this paper, we hope that this work can be useful for other global health modelers, to consider local modeling in their own work.

### 2025-09-30 paper/paper.qmd
misc paper editing, refs

@@
 ---
-title: "`localhierarchy`: an R package to facilitate fitting of global and local Bayesian hierarchical models"
-author: "Leontine, Shauna, Evan, Herb"
+title: ""
+author: ""
 format:
@@
     latex-engine: pdflatex    # lualatex or xelatex or pdflatex
+    include-before-body: title.tex
     include-in-header: 
@@
   This article presents a practical introduction to the package for the applied user and illustrates the package’s functionality through examples for national and subnational estimation.
-date: last-modified
-bibliography: references.bib
+bibliography: bibliography.bib
 number-sections: true
@@
 In global health estimation, Bayesian hierarchical models have become a cornerstone methodology, enabling the integration of diverse data sources to produce reliable estimates across multiple populations and regions.
-For example, they are used to estimate family planning indicators for the FP2030 initiative (ref), low birth weight (ref), and maternal mortality by the United Nations Interagency Group (ref).
+For example, they are used to estimate family planning indicators for the FP2030 initiative [@alkema_2024evol], incidence of low birth weight [@okwaraji_national_2024], abortion and unintended pregnancies rates [@bearak_unintended_2020], and maternal mortality by the United Nations Interagency Group [@peterson_bayesian_2024].
 In these efforts, national or subnational estimates are often required for population-periods with limited data availability.
@@
 Additionally, local models enable estimates for different populations to be updated independently, for example, when new data become available.
-This approach has been used successfully in global estimation exercises for family planning and maternal mortality (ref, ref).
+This approach has been used successfully in global estimation exercises for family planning and maternal mortality [@alkema_2024evol; @peterson_bayesian_2024].
 
@@
 
-```{r}
-#| fig-height: 9
-#| fig-width: 7
-#| echo: false
-#| label: fig-localfits
-#| fig-cap: "Data, estimates, and forecasts from 1970 until 2030 for FP indicators for married women in Benin (BEN), Gambia (GMB), and Ghana (GHA).
+![Data, estimates, and forecasts from 1970 until 2030 for FP indicators for married women in Benin (BEN), Gambia (GMB), and Ghana (GHA).
 Graphs show mCPR, unmet need for modern methods, demand, and demand satisfied with modern methods.
 Survey data are shown in colored dots with 95% confidence intervals that reflect total error variance.
-FPET estimates are shown in black, with grey shaded areas representing 90% credible intervals."
-plots <- readRDS("outputfpet_localfits.rds")
-plots[[1]] /  plots[[2]]/ plots[[3]] 
-```
-
-At its core, the Bayesian FPET model employs transition models for demand and demand satisfied to characterize the long-term trajectories of family planning indicators over time within a population, illustrated in @fig-transition.
+FPET estimates are shown in black, with grey shaded areas representing 90% credible intervals.](fig/fp_local.jpg){#fig-localfits}
+
+<!-- ```{r} -->
+
+<!-- #| fig-height: 9 -->
+
+<!-- #| fig-width: 7 -->
+
+<!-- #| echo: false -->
+
+<!-- #| label: fig-localfits -->
+
+<!-- #| fig-cap: "Data, estimates, and forecasts from 1970 until 2030 for FP indicators for married women in Benin (BEN), Gambia (GMB), and Ghana (GHA).
+Graphs show mCPR, unmet need for modern methods, demand, and demand satisfied with modern methods.
+Survey data are shown in colored dots with 95% confidence intervals that reflect total error variance.
+FPET estimates are shown in black, with grey shaded areas representing 90% credible intervals." -->
+
+<!-- plots <- readRDS("outputfpet_localfits.rds") -->
+
+<!-- plots[[1]] /  plots[[2]]/ plots[[3]]  -->
+
+<!-- ``` -->
+
+At its core, the Bayesian FPET model employs transition models for demand and demand satisfied to characterize the long-term trajectories of family planning indicators over time within a population [@susmann_flexible_2025], illustrated in @fig-transition.
 The transition model describes changes in an indicator in terms of three principal parameters: the rate of change, the timing of the transition, and the asymptotic level (the maximum attainable level for an indicator).
@@
 
-![Illustration of transition model used in the modeling of family planning indicators. TO DO (if kept): make specific to this text, ie name asymptote, level](fig/transition.png){#fig-transition}
+![Schematic illustration of transition model used in the modeling of family planning indicators, modified from @susmann_flexible_2025.
+The transition model describes changes in an indicator (here $\eta_{c,t}$ in terms of three principal parameters: the rate of change, the level in a reference year (here labeled with $\Omega_c$), and the asymptotic level (the maximum attainable level for an indicator (here labeled with $\lambda_c$).](fig/transition.png){#fig-transition width="476"}
 
@@
 
-```{r, echo = FALSE}
-plots <- readRDS("outputfpet_params.rds")
-```
-
-```{r}
-#| fig-height: 9
-#| fig-width: 7
-#| echo: false
-#| label: fig-level
-#| fig-cap: "Estimates of the level of demand for countries in Western Africa and for higher levels used in the hierarchical model (smaller world regions and major world regions).
+<!-- ```{r, echo = FALSE} -->
+
+<!-- plots <- readRDS("outputfpet_params.rds") -->
+
+<!-- ``` -->
+
+<!-- ```{r} -->
+
+<!-- #| fig-height: 9 -->
+
+<!-- #| fig-width: 7 -->
+
+<!-- #| echo: false -->
+
+<!-- #| label: fig-level -->
+
+<!-- #| fig-cap: "Estimates of the level of demand for countries in Western Africa and for higher levels used in the hierarchical model (smaller world regions and major world regions).
 Point estimates and 95% credible intervals are shown.
-TO DO: fix formatting, show country names"
-plots$plot_iso_omega /  plots$plot_reg_omega[[3]]/ plots$plot_reg_omega[[2]]
-```
-
-```{r}
-#| fig-height: 9
-#| fig-width: 7
-#| echo: false
-#| label: fig-asympt
-#| fig-cap: "Estimates of the asymptote for demand for countries in Western Africa and for higher levels used in the hierarchical model (major world regions).
+" -->
+
+<!-- plots$plot_reg_omega[[2]]/ plots$plot_reg_omega[[3]]/  plots$plot_iso_omega -->
+
+<!-- ``` -->
+
+<!-- ```{r} -->
+
+<!-- #| fig-height: 9 -->
+
+<!-- #| fig-width: 7 -->
+
+<!-- #| echo: false -->
+
+<!-- #| label: fig-asympt -->
+
+<!-- #| fig-cap: "Estimates of the asymptote for demand for countries in Western Africa and for higher levels used in the hierarchical model (major world regions).
 Point estimates and 95% credible intervals are shown.
-TO DO: fix formatting, show country names"
-plots$plot_iso_ptilde /  plots$plot_reg_ptilde[[2]] 
-```
+" -->
+
+<!-- plots$plot_reg_ptilde[[2]]  / plots$plot_iso_ptilde -->
+
+<!-- ``` -->
+
+![Estimates of the level of demand for countries in Western Africa and for higher levels used in the hierarchical model (smaller world regions and major world regions).
+Point estimates and 95% credible intervals are shown.](fig/hier_level.jpg){#fig-level width="685"}
+
+![Estimates of the asymptote for demand for countries in Western Africa and for higher levels used in the hierarchical model (major world regions).
+Point estimates and 95% credible intervals are shown.](fig/hier_asymptote.jpg){#fig-asympt}
 
@@
 Firstly, the full Bayesian model is fitted to a global data base of survey data.
-Data base compilation is carried out by the United Nations Population Division (ref) and the Track20 project (ref).
+Data base compilation is carried out by the United Nations Population Division (see [UNPD world contraceptive use website](https://www.un.org/development/desa/pd/world-contraceptive-use)) and the Track20 project (see [www.track20.org](https://www.track20.org/)).
 The global model fit provides information on model parameters, including hierarchical parameters, e.g., estimates of asymptotes.
@@
 For a given level $l$, index $r$ refers to the population in a particular region indexed by $r$, for example, a subnational region, a country, or a group of countries.
-We assume that $\mu$ is unconstrained (possibly after a transformation has been applied) and consider the following general hierarchical model setup, see Figure 6:
-\begin{align*}
-  \mu_{r}^{(l)} \mid \mu^{(l-1)}_{r^\prime}, \sigma_l} &\sim N(\mu^{(l-1)}_{r^\prime}, \sigma_l^2), \text{ for } l > 0,
+We assume that $\mu$ is unconstrained (possibly after a transformation has been applied) and consider the following general hierarchical model setup, see Figure 6: \begin{align*}
+  \mu_{r}^{(l)} \mid \mu^{(l-1)}_{r^\prime}, \sigma_l &\sim N(\mu^{(l-1)}_{r^\prime}, \sigma_l^2), \text{ for } l > 0,
 \end{align*} where $\mu^{(l-1)}_{r^\prime}$ is the parameter value at one level higher in the hierarchy for the region $r^\prime$ that contains unit $r$.
@@
 \begin{alignat}{2}
-  \mu_c^{(\country)}^{(\country)} &\mid \mu_{s[c]}^{(\subregion)}, \sigma_{ \country}^2 &&\sim N(\mu_{s[c]}^{(\subregion)}, \sigma_{ \country}^2), \label{eq-countryhier} \\
-  \mu_s^{(\subregion)} &\mid \mu_{w[s]}^{(\region)}, \sigma_{ \subregion}^2 &&\sim N(\mu_{w[s]}^{(\region)}, \sigma_{ \subregion}^2), \nonumber \\
+  \mu_c^{(\country)} &\mid \mu_{s[c]}^{(\subregion)}, \sigma_{ \country}^2 &&\sim N(\mu_{s[c]}^{(\subregion)}, \sigma_{ \country}^2), \label{eq-countryhier} \\
+  \mu_s^{(\subregion)} &\mid \mu_{w[s]}^{(\region)}, \sigma_{ \subregion}^2 &&\sim N(\mu_{w[s]}^{(\region)}, \sigma_{\subregion}^2), \nonumber \\
   \mu_w^{(\region)} &\mid \mu^{(\glob)}, \sigma_{ \region}^2 &&\sim N(\mu^{(\glob)}, \sigma_{ \region}^2), \nonumber
@@
 
-\[to finish, LA took mostly from fpemlocal paper\] `localhierarchy` is a publicly available R package stored on Github.
-For usage, an R (R Core Team, 2020) installation (≥) and a cmdStan (xxx) installation (≥xxx) are required.
-R can be downloaded on CRAN.
-Stan install \[add info\] xxx .
-The `localhierarchy` package can be installed with one of two methods: (1) direct install from Github repository using the `install_ github()` API from the devtools package (Wickham et al., 2020); (2) install from the package binary using the base R function `install.package()`.
+`localhierarchy` is a publicly available R package stored on Github.
+For usage, installation of R (≥4.5.1), Stan (≥2.37.0), and CmdStanR (≥ 0.9.0) are required [@WeloveR; @stan2025; @cmdstanr_2025].
+The `localhierarchy` package can be installed from the Github repository.
 Package dependencies are listed in the package DESCRIPTION file and will be automatically installed upon installing the main package.
 There are no minimum RAM, CPU, or HARDDRIVE requirements apart from what is necessary to store model runs, which varies case-by-case.
+
+<!-- The `localhierarchy` package can be installed with one of two methods: (1) direct install from Github repository using the `install_ github()` API from the devtools package (Wickham et al., 2020); (2) install from the package binary using the base R function `install.package()`.
+Package dependencies are listed in the package DESCRIPTION file and will be automatically installed upon installing the main package.
+There are no minimum RAM, CPU, or HARDDRIVE requirements apart from what is necessary to store model runs, which varies case-by-case.
+-->
 
@@
 
-\node[level1] (mu0) {$\mu^{(1)}_1$};
+\node[level1] (mu0) {$\mu^{(\country)}_1$};
 
@@
 % keep same colors so now color of level 1 at top 
-\node[level2] (mu0) {$\mu^{(2)}_1$};
+\node[level2] (mu0) {$\mu^{(\subnat)}_1$};
 
@@
 
-For family planning estimation, a local model fitting approach was motivated by the need to enable country monitoring and evaluation (M&E) officers to produce estimates for their countries, as part of the Track20 project (ref).
+For family planning estimation, a local model fitting approach was motivated by the need to enable country monitoring and evaluation (M&E) officers to produce estimates for their countries, as part of the Track20 project [@alkema_2024evol].
 The Track20 project works to build the capacity of countries to generate and use data to inform their programming, improve impact, and accelerate progress toward their FP goals.
@@
 The `localhierarchy` functionality is incorporated in the family planning modeling code base such that all steps are run with the same code base, and parameters can be passed from one model fit to another.
-Full details on this workflow and model specification are described elsewhere (ref).
+Full details on this workflow and model specification are described elsewhere [@alkema_2024evol].
 As an illustration, @fig-allfits shows global and local fits for modern use, demand, and demand satisfied for selected countries in Western Africa.
@@
 
-```{r}
-#| fig-height: 9
-#| fig-width: 7
-#| echo: false
-#| label: fig-allfits
-#| fig-cap: "Data, estimates, and forecasts from 1970 until 2030 for FP indicators for married women in in Benin (BEN), Gambia (GMB), and Ghana (GHA).
+![Data, estimates, and forecasts from 1970 until 2030 for FP indicators for married women in in Benin (BEN), Gambia (GMB), and Ghana (GHA).
 FPET estimates are shown in in colored lines (red for global, green for local), with shaded areas representing 90% credible intervals.
-Survey data are shown in colored dots with 95% confidence intervals that reflect total error variance."
-plots <- readRDS("outputfpet_localglobalfits.rds")
-plots[[1]] /  plots[[2]]/ plots[[3]] # Stack three plots vertically
-```
+Survey data are shown in colored dots with 95% confidence intervals that reflect total error variance.](fig/fpet_localglobal.jpg){#fig-allfits}
+
+<!-- ```{r} -->
+
+<!-- #| fig-height: 9 -->
+
+<!-- #| fig-width: 7 -->
+
+<!-- #| echo: false -->
+
+<!-- #| label: fig-allfits -->
+
+<!-- #| fig-cap: "Data, estimates, and forecasts from 1970 until 2030 for FP indicators for married women in in Benin (BEN), Gambia (GMB), and Ghana (GHA).
+FPET estimates are shown in in colored lines (red for global, green for local), with shaded areas representing 90% credible intervals.
+Survey data are shown in colored dots with 95% confidence intervals that reflect total error variance." -->
+
+<!-- plots <- readRDS("outputfpet_localglobalfits.rds") -->
+
+<!-- plots[[1]] /  plots[[2]]/ plots[[3]] # Stack three plots vertically -->
+
+<!-- ``` -->
 
@@
 
-The figures below show illustrations of regional fits for Benin.
+Figure @fig-subnatfits show illustrations of regional fits for Benin.
 The subnational model is fitted to all regional data from the country.
@@
 
-```{r}
-#| fig-height: 9
-#| fig-width: 7
-#| echo: false
-#| label: fig-subnatfits
-#| fig-cap: "Data, estimates, and forecasts from 1970 until 2030 for FP indicators for married women in subnational regions in Benin.
+![Data, estimates, and forecasts from 1970 until 2030 for FP indicators for married women in subnational regions in Benin.
+Graphs show mCPR, unmet need for modern methods, demand, and demand satisfied with modern methods.
+Survey data are shown in colored dots with 95% confidence intervals that reflect total error variance.
+FPET estimates are shown in black, with grey shaded areas representing 90% credible intervals.](fig/fpet_subnat.jpg){#fig-subnatfits}
+
+![Regional estimates for Benin for modern contraceptive use in 2024. Point estimates and 95% credible intervals shown.](fig/subnat_overview.jpg){#fig-subnatoverview}
+
+<!-- ```{r} -->
+
+<!-- #| fig-height: 9 -->
+
+<!-- #| fig-width: 7 -->
+
+<!-- #| echo: false -->
+
+<!-- #| label: fig-subnatfits -->
+
+<!-- #| fig-cap: "Data, estimates, and forecasts from 1970 until 2030 for FP indicators for married women in subnational regions in Benin.
 Graphs show mCPR, unmet need for modern methods, demand, and demand satisfied with modern methods.
@@
 FPET estimates are shown in black, with grey shaded areas representing 90% credible intervals.
-"
-plots <- readRDS("outputfpet_plotssubnat_BEN.rds")
-plots[["Oueme"]]$married / plots[["Collines"]]$married/ plots[["Donga"]]$married
-```
-
-```{r}
-#| echo: false
-#| label: fig-subnatoverview
-#| fig-cap: "Regional estimates for Benin for modern contraceptive use in 2024. Point estimates and 95% credible intervals shown."
-plots[["comp"]] 
-```
+" -->
+
+<!-- plots <- readRDS("outputfpet_plotssubnat_BEN.rds") -->
+
+<!-- plots[["Oueme"]]$married / plots[["Collines"]]$married/ plots[["Donga"]]$married -->
+
+<!-- ``` -->
+
+<!-- ```{r} -->
+
+<!-- #| echo: false -->
+
+<!-- #| label: fig-subnatoverview -->
+
+<!-- #| fig-cap: "Regional estimates for Benin for modern contraceptive use in 2024. Point estimates and 95% credible intervals shown." -->
+
+<!-- plots[["comp"]]  -->
+
+<!-- ``` -->
 
@@
 Availability of a local modeling tool has enabled users to do additional analyses and check sensitivity of estimates to specific data sources or data quality assumptions.
-In addition to fitting the model to survey data, users of the local family planning model can also incorporate additional data from routine information systems (e.g., the number of contraceptive commodities supplied, add REF).
+In addition to fitting the model to survey data, users of the local family planning model can also incorporate additional data from routine information systems (e.g., the number of contraceptive commodities supplied, see [@mooney_emu]).
 These use cases illustrate the value of local models to enable local analysis.
 
-More recently, we have used the `localHierarchy` approach and functionality for estimating coverage indicators for the Countdown to 2030 initiative (ref).
+More recently, we have used the `localHierarchy` approach and functionality for estimating coverage indicators for the *Countdown to 2030 for Women’s, Children’s and Adolescents’ Health initiative.* This initiative focuses on “tracking progress of life-saving interventions for Reproductive, Maternal, Newborn, Child and Adolescent Health and Nutrition” in low- and middle-income countries, see [www.
+countdown2030.org](https://www.%20countdown2030.org/).
 By developing the `localHierarchy` package and providing use cases in this paper, we hope that this work can be useful for other global health modelers, to consider local modeling in their own work.
@@
 Its arguments also include prior parameters `mu_scalarprior_mean` and `mu_scalarprior_sd`, which refer to the (rescaled) prior mean and standard deviation of $\mu^{(\glob)}$.
-The model block contains the priors for the $\gamma$s and $\sigma_{k}$s, given by
-\begin{align*}
+The model block contains the priors for the $\gamma$s and $\sigma_{k}$s, given by \begin{align*}
   \mu^{(\glob)} &\sim N(0,1),\\
@@
 (not doing the /sd would be easier for specification going forward?) -->
+
+# References

### 2025-09-30 paper/title.tex
misc paper editing, refs

\title{localhierarchy: an R package to facilitate fitting of global and local Bayesian hierarchical models}

\author[1]{Leontine Alkema\thanks{
This work was supported, in whole or in part, by the Bill \& Melinda Gates Foundation (INV-00844 and OPP1192802).
Under the grant conditions of the Foundation, a Creative Commons Attribution 4.0 Generic License has already been assigned to the Author Accepted Manuscript version that might arise from this submission.
Contact: lalkema@umass.edu.}}
\affil[1]{\textit{Department of Biostatistics and Epidemiology, %School of Public Health and Health Sciences,
University of Massachusetts Amherst, %Amherst, %Massachusetts,
USA}}

\author[2]{Shauna Mooney}
\affil[2]{\textit{Hamilton Institute and Department of Mathematics and Statistics, Maynooth University, Ireland}}

\author[1]{Evan Ray}

\author[3]{Herbert Susmann}
\affil[3]{\textit{Division of Biostatistics, Department of Population Health, NYU, % Grossman School of Medicine, %New York, %New York,
USA}}
\date{\today}
\maketitle

### 2025-09-30 R/check_nas.R
documentation fixes

@@ -7,7 +7,6 @@
 #' @param column column to check for NAs
 #'
 #' @returns stops if there are NAs in the column or if the column does not exist
-#' @export
 #'
 check_nas <- function(data, column) {
   if(!(column %in% names(data))) {

### 2025-09-30 R/get_posterior_summaries.R

@@ -2,9 +2,9 @@
 
 
 
-#' get_posterior_summaries
+#' Posterior summaries for use in local models
 #'
-#' Get posterior summaries for a set of parameters
+#' Get posterior summaries for a set of parameters for use in local model
 #'
 #' @param fit fit object to summarize
 #' @param params vector of parnames to summarize (w/o index)

### 2025-09-30 R/hierarchical_param_stan_data.R

@@ -20,6 +20,9 @@
 #'   mu_model_matrix, mu_n_terms_fixed, mu_n_terms_estimate,
 #'   mu_raw_fixed, mu_n_sigma_fixed, mu_n_sigma_estimate,
 #'   mu_sigma_fixed`
+#'
+#' @export
+#'
 hierarchical_param_stan_data <- function(param_name,
                                          param_data,
                                          global_fit = NULL,

### 2025-09-30 R/inits.R

@@ -7,7 +7,6 @@
 #' @param stan_data a list containing the Stan data, to obtain length of parameter vectors
 #'
 #' @returns a list of initial values for the model parameters
-#' @export
 #'
 init_fun_localhierarchy <- function(chain_id, stan_data){
   set.seed(chain_id)

### 2025-09-30 R/posterior_summary_hierparam.R

@@ -137,7 +137,6 @@ extract_parameter_subhierarchical <- function(
 #' @param probs vector of quantiles
 #'
 #' @returns tibble with the requested quantiles
-#' @export
 #'
 quantile_df <- function(x, probs = c(0.025, 0.5, 0.975)) {
   tibble(
@@ -154,7 +153,6 @@ quantile_df <- function(x, probs = c(0.025, 0.5, 0.975)) {
 #'
 #' @returns tibble with 2.5th quantile, posterior mean, and 97.5th quantile, and quant = c(0.025, 0.5, 0.975)
 #'
-#' @export
 #'
 summary_df <- function(x) {
   tibble(

### 2025-09-30 README.Rmd
cleaning

@@
 
-Dependency on `cmdstanr:`
+-   Dependency on `cmdstanr:` Instructions for installing `cmdstanr` are available in their [Getting started](https://mc-stan.org/cmdstanr/articles/cmdstanr.html) guide.
 
-Instructions for installing `cmdstanr` are available in their [Getting started](https://mc-stan.org/cmdstanr/articles/cmdstanr.html) guide.
-
-Install `localHierarchy` from Github:
-`remotes::install_github("AlkemaLab/localhierarchy")`
+-   Install `localHierarchy` from Github: `remotes::install_github("AlkemaLab/localhierarchy")`
 

### 2025-10-07 paper/paper.qmd
edits shauna

@@
 In global health estimation, Bayesian hierarchical models have become a cornerstone methodology, enabling the integration of diverse data sources to produce reliable estimates across multiple populations and regions.
-For example, they are used to estimate family planning indicators for the FP2030 initiative [@alkema_2024evol], incidence of low birth weight [@okwaraji_national_2024], abortion and unintended pregnancies rates [@bearak_unintended_2020], and maternal mortality by the United Nations Interagency Group [@peterson_bayesian_2024].
+For example, they are used to estimate and forecast family planning indicators for the FP2030 initiative [@alkema_2024evol], incidence of low birth weight [@okwaraji_national_2024], abortion and unintended pregnancies rates [@bearak_unintended_2020], and maternal mortality by the United Nations Interagency Group [@peterson_bayesian_2024].
 In these efforts, national or subnational estimates are often required for population-periods with limited data availability.
@@
 This paper introduces the \texttt{localhierarchy} R package designed to address these challenges by supporting the development and fitting of Bayesian hierarchical models in settings requiring both global and local estimation.
-The package provides R functions and Stan functions and model blocks <!-- citep{stan2023}  --> to facilitate specification of hierarchical models, to fit full global models, and to subsequently fit local models using fixed parameter values inferred from the global analysis.
+The package provides R functions and Stan functions and model blocks [@stan2025] to facilitate specification of hierarchical models, to fit full global models, and to subsequently fit local models using fixed parameter values inferred from the global analysis.
 Through this framework, users can conveniently perform local inference informed by global context, streamlining workflows in global health estimation tasks.
@@
 
-# The Family Planning Estimation Tool (FPET)
+# The Family Planning Estimation Tool (FPET) {#sec-fpet}
 
 The Family Planning Estimation Tool (FPET) is widely used to estimate key family planning indicators in countries and regions.
-Key indicators include modern contraceptive use, demand for family planning, and demand satisfied, referring to the ratio of modern use to demand.
+Key indicators include modern contraceptive use, demand for family planning, and demand satisfied, referring to the ratio of modern contraceptive use to demand.
 Data on such indicators is available from household surveys and routine data systems.
@@
 Given limited country-specific information on some parameters, Bayesian hierarchical models are used to share information between populations, using hierarchical structures such as the one illustrated in Figure \ref{fig-hierstruc}.
-Specifically for family planning estimation and forecasting, a hierarchical model is used to estimate the asymptote of the transition of demand from low to high values, and the level of demand in a reference year.
+Specifically for family planning estimation and forecasting, a hierarchical model is used to estimate the level of demand in a reference year and the asymptote of the transition of demand from low to high values.
 @fig-level shows estimates of the level in the reference year at the country level for countries in Western Africa, and for the groupings used in the hierarchical model.
@@
 
-![Estimates of the level of demand for countries in Western Africa and for higher levels used in the hierarchical model (smaller world regions and major world regions).
+![Estimates of the level of demand for major world regions (top), smaller world regions (middle) and countries in Western Africa (bottom).
 Point estimates and 95% credible intervals are shown.](fig/hier_level.jpg){#fig-level width="685"}
 
-![Estimates of the asymptote for demand for countries in Western Africa and for higher levels used in the hierarchical model (major world regions).
+![Estimates of the asymptote for demand for major world regions (top) and countries in Western Africa (bottom).
 Point estimates and 95% credible intervals are shown.](fig/hier_asymptote.jpg){#fig-asympt}
@@
 The technical details for this approach are given in the next section.
-We return to the application of estimating family planning indicators in Section 4.2.
+We return to the application of estimating family planning indicators in @sec-fpetnat.
 
@@
 For a given level $l$, index $r$ refers to the population in a particular region indexed by $r$, for example, a subnational region, a country, or a group of countries.
-We assume that $\mu$ is unconstrained (possibly after a transformation has been applied) and consider the following general hierarchical model setup, see Figure 6: \begin{align*}
+We assume that $\mu$ is unconstrained (possibly after a transformation has been applied) and consider the following general hierarchical model setup, see Figure \ref{fig-hiermodel}:
+
+\begin{align*}
   \mu_{r}^{(l)} \mid \mu^{(l-1)}_{r^\prime}, \sigma_l &\sim N(\mu^{(l-1)}_{r^\prime}, \sigma_l^2), \text{ for } l > 0,
@@
 \end{adjustbox}
-\caption{General specification of a hierarchical model.}
+\caption{General specification of a hierarchical model.}\label{fig-hiermodel}
 \end{figure}
@@
 
-As a specific example, we consider the estimation of country-specific parameters, where countries are organized into subregions of countries (illustrated in Figure 2).
+As a specific example, we consider the estimation of country-specific parameters, where countries are organized into subregions of countries (illustrated in Figure \ref{fig-hiermodel}).
 In this setting, using alternative notation for the same setup to indicate the level by name, the hierarchical model can be written as follows:
@@
 We use the term local hierarchical models to refer to a hierarchical model in which a subset of parameters have been fixed at posterior mean estimates obtained from a fit of the model to a larger data set.
-A typical workflow for local estimation of national level outcomes is illustrated in Figure 7.
+A typical workflow for local estimation of national level outcomes is illustrated in Figure \ref{fig-workflow}.
 For national estimation using the model given by Eq.\ref{eq-countryhier}, the user starts by fitting a global model, using a global data base and estimating all parameters in a hierarchical model.
@@
 \end{adjustbox}
-\caption{Workflow for global and local national modeling. }
+\caption{Workflow for global and local national modeling. }\label{fig-workflow}
 \end{figure}
@@
 
-replacing the hierarchical mean and variances by point estimates. Details are given in the implementation section.
+replacing the hierarchical mean and variances by point estimates $\hat{\mu}_{s[c]}^{(\subregion)}$ and $\hat{\sigma}_{ \country}^2$, respectively. Details are given in the implementation section.
 
 The `localhierachy` package provides a general implementation of local models by facilitating the fixing of subsets of parameters.
-Continuing with the general notation introduced above, in the package, a local model for one parameter of level $l$ is derived by fixing $\sigma^{(k)}$ for $k=1, ..., l$ and fixing $\mu_{r\prime}{(k)}$ for $k = 1, .., l-1$ at posterior means derived from a global fit.
-More generally, if a model was fitted using data up to level $l$, the user can create a local model by fixing $\sigma^{(k)}$ at all levels $k$ up to some chosen level $K_{\sigma}$.
+Continuing with the general notation introduced above, in the package, a local model for one parameter of level $l$ is derived by fixing $\sigma_k$ for $k=1, ..., l$ and fixing $\mu_{\cdot}^{(k)}$ for $k = 1, .., l-1$ at posterior means derived from a global fit.
+More generally, if a model was fitted using data up to level $l$, the user can create a local model by fixing $\sigma_k$ at all levels $k$ up to some chosen level $K_{\sigma}$.
 Similarly, the $\mu^{(k)}$ parameters can be fixed at all levels $k$ up to a level $K_{\mu} \leq K_{\sigma}$.
 
-## Implementation
+## Implementation {#sec-implementation}
 
@@
 In a typical work flow, a user first fits a global model, followed by local ones.
-Worked examples are given in the use case section.
+Worked examples are given in the use cases in @sec-nationalusecase and @sec-subnatusecase.
 The global and local models in these examples are fitted using the `localhierarchy` wrapper function `fit_model_localhierarchy`, using its arguments to specify a hierarchical model, data, and, for local fits, pass on outputs from the global model.
@@
 The package includes a user-friendly specification of hierarchical model structures, and associated specification of fixed parameters for local runs.
-A worked example is given in the first use case.
-Briefly, if a user wants to specify a hierarchical model like the one introduced above, with the levels world, regions, subregion, and country, <!-- country parameters $\mu_c^{(\country)} =  s_{global}\cdot \gamma_{global} + \sigma_{ region}\cdot \gamma_{r[s[c]]]}^{(
-egion)} + \sigma_{ subregion}\cdot\gamma_{s[c]}^{(\subregion)} + \sigma_{ country}\cdot\gamma_{ c}$,  --> using a dataset with information on these groupings, with naming `region`, `subregion`, and `country`, in the R functions, the user can specify the hierarchical model as follows \[Q for herb: refs/credit to be given for this set up to other work?\]:
+A worked example is given in the first use case (@sec-nationalusecase).
+Briefly, if a user wants to specify a hierarchical model with the levels world, regions, subregion, and country, <!-- country parameters $\mu_c^{(\country)} =  s_{global}\cdot \gamma_{global} + \sigma_{ region}\cdot \gamma_{r[s[c]]]}^{(
+egion)} + \sigma_{ subregion}\cdot\gamma_{s[c]}^{(\subregion)} + \sigma_{ country}\cdot\gamma_{ c}$,  --> using a dataset with information on these groupings, with naming `region`, `subregion`, and `country`, in the R functions, the user can specify the hierarchical model as follows:
 
@@
 The user can then specify a local model for one country by fixing parameters at different levels, using the `hierarchical_sigmas_fixed` and `hierarchical_terms_fixed` arguments in the `fit_model_localhierarchy` function.
-For example, if the $\sigma$s are to be fixed up to the country level and if the $\mu$ terms are to be fixed up to the subregion, the user can encode this as follows:
+For example, if the $\sigma$ parameters are to be fixed up to the country level and if the $\mu$ terms are to be fixed up to the subregion, the user can encode this as follows:
 
@@
 Implementation is based on a model matrix.
-For our model given by the equation above, the following matrix multiplication is used to obtain the vector of $\mu_c^{(\country)}$s: $$(\mu_1^{(\country)}, ..., \mu_c^{(\country)}) = M\cdot (\eta_1, ..., \eta_{1+R+S+C})',$$ where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_{1+R+S+C})$ refers to the vector of the product of the standard deviation terms and $\gamma$s, i.e., $\eta_1 = s_{\glob}\cdot \gamma_{\glob}, \eta_2 = \sigma_{ \region} \cdot \gamma_{ 1}^{(\region)}, ..., \eta_{1+W+S+C} = \sigma_{ \country}\cdot\gamma_{ C}^{(\country)}$.
+For our model given by the equation above, the following matrix multiplication is used to obtain the vector of $\mu_c^{(\country)}$s: $$(\mu_1^{(\country)}, ..., \mu_c^{(\country)}) = M\cdot (\eta_1, ..., \eta_{1+R+S+C})',$$ where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_{1+R+S+C})$ refers to the vector of the product of the standard deviation terms and $\gamma$ terms, i.e., $\eta_1 = s_{\glob}\cdot \gamma_{\glob}, \eta_2 = \sigma_{ \region} \cdot \gamma_{ 1}^{(\region)}, ..., \eta_{1+W+S+C} = \sigma_{ \country}\cdot\gamma_{ C}^{(\country)}$.
 When estimating a subset of parameters in a local model, the values of $\sigma$s and $\gamma$s are fixed to point estimates (posterior means) from a global fit.
@@
 
-# Use case: National estimation
+# Use case: National estimation {#sec-nationalusecase}
 
@@
 
-The simulation setup is summarized in Figure 8, panels A and B. Simulation of subnational data is discussed in a later section.
+The simulation setup is summarized in Figure \ref{fig-simusetup}, top and center panels. Simulation of subnational data is discussed in a later section.
 
@@
 Center: Data generating mechanism for national data, illustrated for the first country.
-Bottom: Data generating mechanism for subnational data, illustrated for the first subnational region (used only in Section 5).}
+Bottom: Data generating mechanism for subnational data, illustrated for the first subnational region (used only in Section 5).}\label{fig-simusetup}
 \end{figure}
@@
 
-Panel A in Figure 4 indicates the hierarchical model considered.
+The top panel in Figure \ref{fig-simusetup} indicates the hierarchical model considered.
 For national level estimation, we consider countries to be grouped in the world, i.e., $\mu_c^{(\country)}|\mu^{(\glob)}, \sigma_{ \country} \sim N(\mu^{(\glob)}, \sigma_{ \country}^2)$.
-Panel B introduces the likelihood function for national data, given by a normal likelihood with constant variance: $$y_i^{(\country)}|\mu_{c[i]}^{(\country)}, \sigma_y^2 \sim N(\mu_{c[i]}^{(\country)}, \sigma_y^2),$$ where $c[i]$ refers to the country of country observation $i$.
+The center panel introduces the likelihood function for national data, given by a normal likelihood with constant variance: $$y_i^{(\country)}|\mu_{c[i]}^{(\country)}, \sigma_y^2 \sim N(\mu_{c[i]}^{(\country)}, \sigma_y^2),$$ where $c[i]$ refers to the country of country observation $i$.
 We chose simple normal densities as data generating mechanisms in the simulation, to focus the modeling on the hierarchical model and the `localhierarchy` functionality as opposed to the data model aspects of modeling.
@@
 Elements include the model matrix and other elements are used to make a link between columns in the model matrix, or equivalently, indices in the vector $\eta$, and the hierarchical groupings.
-The model matrix $M$, as introduced above, is as follows:
+The model matrix $M$, as introduced in @sec-implementation, is as follows:
 
@@
 
-All this information is passed to the function `hierarchical_param_stan_data` to produce inputs needed in the Stan model fitting.
+All of this information is passed to the function `hierarchical_param_stan_data` to produce inputs needed in the Stan model fitting.
 This function outputs a named list with Stan input data relevant to the hierarchical setup:
@@
 We can use the function `fit_model_localhierarchy` to fit the model as specified so far.
+Details are included in the appendix (@sec-helpfilefitmodel).
 The argument `runstep` is used to define the type of model, setting it to `"global_national"` results in fitting a model to all data without fixing any parameters.
-Other arguments include the hierarchical level specification, the area, and survey data:
+Other arguments include the hierarchical level specification, the area, and survey data.
+When calling the function, print statements are used to inform the user of key points related to the model fit:
 
@@
 
-A local fit can be esetimated by passing on the global fit object to the wrapper function and including data for countries of interest.
+A local fit can be estimated by passing on the global fit object to the wrapper function and including data for countries of interest.
 Here we do local runs for all countries, by including the entire data set, and specify what to fix using the `runstep` argument:
@@
 
-There are several functions to produce and check outputs.
+The package includes several functions to produce and check outputs.
 The function `posterior_summary_hierparam` is used to summarize posterior draws for hierarchical parameters, to produce summaries of $\mu$ at any level of interest.
@@
 
-The column `quant` above refers to posterior means and upper and lower bounds of 95% credible intervals. We can create plots to show outputs, for example for the first 4 countries:
+The column `name` includes the country names, given by simple indicares in this simulation.
+The column `quant` above refers to posterior means and upper and lower bounds of 95% credible intervals.
+We can create plots to show outputs, for example for the first 4 countries:
 
@@
 
-## Application: National estimation of family planning indicators using FPET
+## Application: National estimation of family planning indicators using FPET {#sec-fpetnat}
 
@@
 
-For local estimation of family planning indicators, the Bayesian model described in section 2 is used.
+For local estimation of family planning indicators, the Bayesian model described in @sec-fpet is used.
 Specifically, hierarchical models are used for estimation of transition model parameters.
-The workflow used for estimating family planning indicators at the national level is based on model fitting in several steps, including the estimation of long-term trends in the first step, the estimation of all model parameters related to modern use in a second step, the addition of information on the use of traditional methods of contraception in a third step, followed by local model fitting to produce estimates for all indicators of interest.
+The workflow used for estimating family planning indicators at the national level is based on model fitting in several steps, including the estimation of long-term trends in the first step, the estimation of all model parameters related to modern contraceptive use in a second step, the addition of information on the use of traditional methods of contraception in a third step, followed by local model fitting to produce estimates for all indicators of interest.
 The `localhierarchy` functionality is incorporated in the family planning modeling code base such that all steps are run with the same code base, and parameters can be passed from one model fit to another.
 Full details on this workflow and model specification are described elsewhere [@alkema_2024evol].
-As an illustration, @fig-allfits shows global and local fits for modern use, demand, and demand satisfied for selected countries in Western Africa.
+As an illustration, @fig-allfits shows global and local fits for modern contraceptive use, demand, and demand satisfied for selected countries in Western Africa.
 The global fit in red corresponds closely to the local fits, shown in green.
@@
 
-# Use case 2: Subnational estimation
+# Use case 2: Subnational estimation {#sec-subnatusecase}
 
@@
 The `localhierarchy` package can be used to fit subnational hierarchical models, following the same principles as for national estimation or through an updated approach.
-Just as for national-level hierarchical modeling, the starting point for subnational hierarchical modeling is to specify a hierarchical modeling structure for the subnational regions, such as the one from Figure 3.
+Just as for national-level hierarchical modeling, the starting point for subnational hierarchical modeling is to specify a hierarchical modeling structure for the subnational regions.
 When following the same principles as for national-level hierarchical modeling, a global model can be fitted to a global data base of subnational data, followed by local models for subnational regions.
@@
 
-The approach followed in NISE is illustrated in Figure 4, for the hierarchical structure from Figure 2.
+The approach followed in NISE is illustrated in Figure \ref{fig-nise}.
 In the NISE approach, the first step is to fit a model to a global data base with national-level data, to obtain estimates of model parameters up to and including national level parameters.
-This national-level model fit is used in a global subnational fit to a global data base of subnational data.
+This national-level model fit isthen used in a global subnational fit to a global data base of subnational data.
 Specifically, in the global subnational fitting step, national level parameters are fixed at point estimates from the global national fit, i.e., $\hat{\sigma}_{ l}$ and $\mu_{\cdot}^{(l)}$ for level $l$ up to the country level.
-Finally, in a 3rd step, local models can be fitted, using fixed values from the global subnational fit.
+Finally, in a third step, local models can be fitted, using fixed values from the global subnational fit.
 In the remainder of this section, we illustrate this approach using simulated data, and present an application of this approach for family planning estimation.
@@
 \end{adjustbox}
-\caption{Workflow for nationally-informed global and local subnational modeling. }
+\caption{Workflow for nationally-informed global and local subnational modeling. }\label{fig-nise}
 \end{figure}
@@
 
-The simulation setup for subnational parameters and data is summarized in Figure 8, Panels A-B-C.
-As explained above for national estimation, Panel A indicates the hierarchical model considered.
+The simulation setup for subnational parameters and data is summarized in Figure \ref{fig-simusetup} in three panels.
+As explained above for national estimation, the top panel indicates the hierarchical model considered.
 For subnational estimation, an additional level is added, with $\mu_r^{(\subnat)} = \mu_{c[r]}^{(\country)} + \eta_{r}^{(\subnat)}$, $\eta_{r}^{(\subnat)}|\sigma_{ \subnat} \sim N(0, \sigma_{ \subnat})$.
 
-As explained earlier as well, we chose simple normal densities as data generating mechanisms in the simulation, to focus the modeling on the hierarchical model and the `localhierarchy` functionality as opposed to the data model aspects of modeling.
+As previously explained, we chose simple normal densities as data generating mechanisms in the simulation, to focus the modeling on the hierarchical model and the `localhierarchy` functionality as opposed to the data model aspects of modeling.
 Subnational data follows a normal distribution with constant variance: $$y^{(subnat)}_i|\mu^{(subnat)}_{r[i]}, \sigma_y^2 \sim N(\mu^{(subnat)}_{r[i]}, \sigma_y^2),$$ where $r[i]$ refers to the subnational region of subnational observation $i$.
@@
 
-Figure @fig-subnatfits show illustrations of regional fits for Benin.
+@fig-subnatfits show illustrations of regional fits for Benin.
 The subnational model is fitted to all regional data from the country.
@@
 Availability of a local modeling tool has enabled users to do additional analyses and check sensitivity of estimates to specific data sources or data quality assumptions.
-In addition to fitting the model to survey data, users of the local family planning model can also incorporate additional data from routine information systems (e.g., the number of contraceptive commodities supplied, see [@mooney_emu]).
+In addition to fitting the model to survey data, users of the local family planning model can also incorporate additional data from routine information systems (e.g., the number of contraceptive commodities supplied at family planning facilities, see [@mooney_emu]).
 These use cases illustrate the value of local models to enable local analysis.
@@
 
+## Appendix II: Documentation of the `fit_model_localhierarchy` function {#sec-helpfilefitmodel}
+
+| **Argument** | **Description** |
+|----|----|
+| **`survey_df`** | Tibble with survey data |
+| **`y`** | Column name of outcome, defaults to **`y`**. |
+| **`area`** | Column name of the area of each observation (used as geounit; iso or subnational region) |
+| **`area_select`** | Area name to use for local run (e.g., iso code or subnational region name) |
+| **`runstep`** | Type of run, defines which model fitting step to perform (see Details for options). |
+| **`global_fit`** | Optional global fit object, used to obtain fixed values for some parameters in the current fit (see Details). |
+| **`hierarchical_level`** | Vector specifying hierarchical structure used for mu. Should list elements in hierarchical order from `intercept` downward (see Details for examples). |
+| **`add_subnational_hierarchy`** | Level that's added to the hierarchy for subnational analysis, defaults to `subnat`. |
+| **`use_globalsubnat_fromnat`** | Logical; for local subnational runs, whether to use the global fit derived from national data (requires **`fit_globalsubnat_fromnat`** in **`global_fit`** if TRUE).
+|
+| **`mu_isvector`** | Logical, TRUE if mu is a vector, defaults to FALSE |
+| **`chains`** | Number of chains to run (sampling parameter) |
+| **`iter_sampling`** | Number of posterior samples to draw (sampling parameter) |
+| **`iter_warmup`** | Number of warmup iterations (sampling parameter) |
+| **`compile_model`** | Boolean indicator of whether to compile the Stan model |
+| **`force_recompile`** | Boolean indicator of whether to force recompilation of the Stan model |
+| **`seed`** | Random seed |
+| **`refresh`** | Number of iterations between progress updates |
+| **`adapt_delta`** | Target acceptance rate for the No-U-Turn Sampler |
+| **`max_treedepth`** | Maximum tree depth for the No-U-Turn Sampler |
+
+### Details
+
+The `fit_model_localhierarchy` function fits the toy example for hierarchical models/seq fitting. The argument `runstep` determines the type of run to perform. The following run steps are supported:
+
+-   "global_national": Fit the global model.
+
+-   "local_national": Fit the model to data from a single country, using a global_national fit.
+
+-   "global_subnational": Fit the model to global database with subnational data, using a global_national fit.
+
+-   "local_subnational": Fit the model to subnational data from a single country or region, using a global_subnational fit. This is also explained in the documentation folder.
+
+Details on hierarchical set ups used The package allows the structure of the hierarchical prior to be configured by the user through the `hierarchical_level` argument.
+These arguments expect a character vector that specifies a nesting hierarchical structure.
+Each element of the vector must be either "intercept" or a column name in the survey dataset, where "intercept" will add a global intercept for the parameter.
+The vector must be in descending order in terms of the hierarchy: that is, it starts with "intercept" and proceeds down the hierarchy.
+
+For example, suppose we are fitting country-level data, where the dataset has columns "name_country", "name_sub_region", and "name_region" containing the name of the country, sub-region, and region that each observation belongs to.
+To specify that parameter mu should be fitted with a hierarchical model in which countries are nested within sub-regions within regions within world, we would use the argument `hierarchical_level = c("intercept", "name_region", "name_sub_region", "name_country")`.
+
+Optionally, model parameters can be fixed to values from a previous model fit provided via the `global_fit` argument.
+In a typical use case, the `global_fit` will have been fit to data from many geographic units (e.g., all countries), while the current fit uses data from a smaller number of locations.
+When using a global_fit to fix parameter values, what exactly is fixed is determined by the runstep and global_fit combi.
+see options in function
+
+### Value
+
+List that contains samples, stan_data, other information relevant to model fit (arguments), and for global fits, point estimates of relevant parameters (post_summ).
+For subnational global fits, the list includes fit_globalsubnat_fromnat, which is the global fit with additional subnational sigmas added to the postsum object.
+
 # References

### 2025-10-07 R/fit_model.R

@@ -11,11 +11,11 @@
 #' @param area column name of the area of each observation (used as geounit; iso or subnational region)
 #' @param area_select area name to use for local run (eg iso code or subnat region name)
 #'
-#' @param runstep type of run, currently one of XXX (see Details).
+#' @param runstep Type of run, defines which model fitting step to perform (see Details for options).
 #' @param global_fit optional global fit object, used to obtain fixed
 #'   values to use for some parameters in the current fit (see Details).
 #' @param hierarchical_level vector specifying hierarchical structure used for mu
-#' @param add_subnational_hierarchy level that's added to the hierarchy for subnational, defaults to 'subnat'
+#' @param add_subnational_hierarchy level that's added to the hierarchy for subnational, defaults to `subnat`
 #' @param use_globalsubnat_fromnat Logical, whether in a local subnational run,
 #' to use the global fit derived from national data
 #' if TRUE and local subnat run, global_fit needs to contain object `fit_globalsubnat_fromnat`

### 2025-10-08 paper/paper.qmd
clean up export and naming of functions

@@
   This article presents a practical introduction to the package for the applied user and illustrates the package’s functionality through examples for national and subnational estimation.
+
+keywords: Bayesian hierarchical model; R; Stan; Local models 
 bibliography: bibliography.bib
@@
 library(patchwork)
-devtools::load_all(here::here())
+library(localhierarchy)
 ```
@@
 
-![Schematic illustration of transition model used in the modeling of family planning indicators, modified from @susmann_flexible_2025.
-The transition model describes changes in an indicator (here $\eta_{c,t}$ in terms of three principal parameters: the rate of change, the level in a reference year (here labeled with $\Omega_c$), and the asymptotic level (the maximum attainable level for an indicator (here labeled with $\lambda_c$).](fig/transition.png){#fig-transition width="476"}
+![Schematic illustration of transition model used in the modeling of family planning indicators.
+The transition model describes changes in an indicator (here $\eta_{c,t}$ in terms of three principal parameters: the rate of change, the level in a reference year (here labeled with $\Omega_c$), and the asymptotic level (the maximum attainable level for an indicator (here labeled with $\lambda_c$).
+This figure has been reproduced with permission from @susmann_flexible_2025.](fig/transition.png){#fig-transition width="476"}
 
@@
 ```{r}
-plot_prior_post_sigmas(fit = global_fit, parname = "mu")
+plot_prior_post_sigmas_localhierarchy(fit = global_fit, parname = "mu")
 ```
@@
 
+### Data availability statement {.unnumbered}
+
+No data are associated with the software tool described in this article.
+
+### Software availability {.unnumbered}
+
+The R package is available at [github.com/AlkemaLab/localhierarchy](https://github.com/AlkemaLab/localhierarchy) and Zenodo \[to do: add reference\]. The code is licensed XXX.
+
+### Grant information {.unnumbered}
+
+This work was supported, in whole or in part, by the Gates Foundation (INV-00844 and OPP1192802).
+Under the grant conditions of the Foundation, a Creative Commons Attribution 4.0 Generic License has already been assigned to the Author Accepted Manuscript version that might arise from this submission.
+
 # Appendix

### 2025-10-08 R/fit_model.R

@@ -37,6 +37,8 @@
 #' For subnational global fits, the list includes fit_globalsubnat_fromnat, which is the global fit with
 #' additional subnational sigmas added to the postsum object.
 #'
+#' @export
+#'
 #' @details
 #' The `fit_model_localhierarchy` function fits the toy example for hierarchical models/seq fitting.
 #' The argument \code{runstep} determines the type of run to perform. The
@@ -45,7 +47,7 @@
 #' * "local_national": Fit the model to data from a single country, using a global_national fit.
 #' * "global_subnational": Fit the model to global database with subnational data, using a global_national fit.
 #' * "local_subnational": Fit the model to subnational data from a single country or region, using a global_subnational fit.
-#' This is also explained in the documentation folder.
+#' This is also explained in the article with the package.
 #'
 #' Details on hierarchical set ups used
 #' The package allows the structure of the hierarchical prior to be configured by the user
@@ -69,7 +71,6 @@
 #' (e.g., all countries), while the current fit uses data from a smaller number
 #' of locations.
 #' When using a global_fit to fix parameter values, what exactly is fixed is determined by the runstep and global_fit combi.
-#' see options in function
 #'
 #' @importFrom cmdstanr cmdstan_model write_stan_file
 #' @importFrom tibble tibble
@@ -323,7 +324,7 @@ fit_model_localhierarchy <- function(
     result <- c(result,
                 samples = fit)
     if (runstep %in% c("global_national", "global_subnational")){
-      post_summ <- get_posterior_summaries(result)
+      post_summ <- get_posterior_summaries_localhierarchy(result)
       result <- c(result,
           list(post_summ = post_summ))
       if (runstep %in% c("global_subnational")){

### 2025-10-08 R/get_posterior_summaries.R

@@ -12,7 +12,7 @@
 #' @returns tibble with variable (parname with index used in model) and postmean (posterior mean)
 #' @export
 #'
-get_posterior_summaries <- function(
+get_posterior_summaries_localhierarchy <- function(
     fit,
     params = c("mu_raw", "mu_sigma", "nonse")
 ) {
@@ -54,7 +54,6 @@ get_posterior_summaries <- function(
 #' it is just the stacking
 
 #' @returns tibble with variable (parname with index used in model) and postmean (posterior mean)
-#' @export
 #'
 get_posterior_summary_one_param <- function(
     fit,

### 2025-10-08 R/hier_data_helpers.R

@@ -1,5 +1,11 @@
 
 
+#' Is there a comma?
+#'
+#' @param variable character string to be checked for comma
+#'
+#' @returns Boolean
+#'
 is_there_a_comma <- function(variable){
   grepl(
     pattern = ",",
@@ -35,7 +41,6 @@ is_there_a_comma <- function(variable){
 #' variable examples are `a[1,1]`, `a[2,1]`, etc.
 #'
 #' @returns matrix with rows and columns corresponding to the indices in the variable
-#' @export
 #'
 create_a_matrix <- function(post_summ){
   res <- post_summ %>%
@@ -94,7 +99,6 @@ variable_exists_in_draws <- function(fit, var_name) {
 #'
 #' @returns data frame with the same columns as `estimate_df` plus columns containing
 #' just the variable name and the indices.
-#' @export
 #'
 #' @examples
 #' split_var_name_and_index(

### 2025-10-08 R/mu_raw.R

@@ -9,7 +9,6 @@
 #' @param morethan1param does parname refer to more than 1 parameter (a vector)
 #'
 #' @returns tibble with samples of parname_raw_estimate, including info on hierarchical level
-#' @export
 #'
 get_mu_raw_labeled <- function(fit, parname, morethan1param = FALSE){
   # mu_raw_estimate is saved directly so getting those is not difficult

### 2025-10-08 R/plot_posterior_summaries.R

@@ -1,12 +1,12 @@
 
 #' Plot summaries of hierarchical parameters
 #'
-#' Display outputs from posterior_summary_hierparam
+#' Display outputs from posterior_summary_hierparam_localhierarchy
 #'
-#' @param res output from posterior_summary_hierparam
+#' @param res output from posterior_summary_hierparam_localhierarchy
 #' @param hierarchy_select optional, what hierarchical level to show? if NULL, all levels are shown
 #' @param areas_select optional: specific areas in a level to filter by, allowed only if one hierarchical level is selected
-#' @param res2 optional, output from posterior_summary_hierparam for comparison
+#' @param res2 optional, output from posterior_summary_hierparam_localhierarchy for comparison
 #' @param modelname1 label for res
 #' @param modelname2 label for res2
 #' @param k_select optional, if res contains k, which k values to show? if NULL, all k values are shown
@@ -14,7 +14,7 @@
 #'
 #' @returns ggplot object
 #'
-plot_posterior_summaries <- function(
+plot_posterior_summaries_localhierarchy <- function(
     res,
     hierarchy_select = NULL,
     areas_select = NULL,

### 2025-10-08 R/posterior_summary_hierparam.R

@@ -1,6 +1,6 @@
 
 
-#' posterior_summary_hierparam
+#' posterior_summary_hierparam_localhierarchy
 #'
 #' Calculate posterior summaries for hierarchical parameters
 #'
@@ -14,7 +14,7 @@
 #' for morethan1param, each level has a list where k refers to the index
 #' @export
 #'
-posterior_summary_hierparam <- function(fit, parname, morethan1param = FALSE,
+posterior_summary_hierparam_localhierarchy <- function(fit, parname, morethan1param = FALSE,
                                         hierarchical_levels = fit$hierarchical_level){
   print("Calculating posterior summary for hierarchical parameters")
   print("This can take a little while")
@@ -76,7 +76,6 @@ posterior_summary_hierparam <- function(fit, parname, morethan1param = FALSE,
 #' these mus are obtained by summing up all relevant etas
 #' variables are `name`, `value`, `.draw/iteration/chain'
 #' and `k` referring to parameter index if morethan1param = TRUE
-#' @export
 #'
 extract_parameter_subhierarchical <- function(
     hierarchical_data,

### 2025-10-08 R/prior_post.R

@@ -13,9 +13,10 @@
 #' summary plots gives summary CIs, nresultsperpage at a time.
 #' plots_allmuraw gives all the individual plots of each mu_raw, with density per chain and prior added
 #' If morethan1param = TRUE, then each list contains a list per parameter k
-#' #' @export
 #'
-plot_mu_raw <- function(fit, parname, morethan1param = FALSE,
+#' @export
+#'
+plot_muraw_localhierarchy <- function(fit, parname, morethan1param = FALSE,
                         nresultsperpage = 30){
   helper_get_plots_allmuraw <- function(samples_tibble){
     parnames <- names(samples_tibble)[-seq(1,3)] # %>% select(-.chain, -.iteration, -.draw))
@@ -88,7 +89,7 @@ plot_mu_raw <- function(fit, parname, morethan1param = FALSE,
 #' @returns Plot with density of sigma_estimate and prior added
 #' @export
 #'
-plot_prior_post_sigmas <- function(fit, parname){
+plot_prior_post_sigmas_localhierarchy <- function(fit, parname){
   # sigmas, fine to look at all k's combined
   samp <- fit$samples$draws(paste0(parname, "_sigma_estimate"))
   parnames <- dimnames(samp)[[3]]

### 2025-10-08 paper/paper.qmd
clean up name space dependencies blabla

@@
   cache: true
-
 abstract: | 
@@
 #| include: false
-library(here)
-library(tidyverse)
-library(ggplot2)
-library(haven)
-library(stringr)
-library(cmdstanr)
-library(truncnorm) # for inits
-library(bayesplot) # diagnostic plots
-library(posterior) # else error for as_draws_list
-library(tidybayes)
-library(patchwork)
 library(localhierarchy)
@@
 ```{r}
-dat_nat |> select(level1,  mu_global, level1_eta, y)
+dat_nat |> dplyr::select(level1,  mu_global, level1_eta, y)
 ```
@@
 global_fit$post_summ |> 
-  filter(variable_no_index == "mu_sigma")
+  dplyr::filter(variable_no_index == "mu_sigma")
 ```
@@
 ```{r, message=FALSE, warning=FALSE}
-res_global <- posterior_summary_hierparam(fit = global_fit, parname = "mu")
-res_local <- posterior_summary_hierparam(fit = fit_local, parname = "mu")
+res_global <- posterior_summary_hierparam_localhierarchy(
+  fit = global_fit, parname = "mu")
+res_local <- posterior_summary_hierparam_localhierarchy(
+  fit = fit_local, parname = "mu")
 ```
@@
 
-The column `name` includes the country names, given by simple indicares in this simulation.
+The column `name` includes the country names, given by simple indices in this simulation.
 The column `quant` above refers to posterior means and upper and lower bounds of 95% credible intervals.
@@
 ```{r}
-plots <- plot_posterior_summaries(
+plots <- plot_posterior_summaries_localhierarchy(
   res = res_global, 
@@
 ```{r}
-plots <- plot_posterior_summaries(
+plots <- plot_posterior_summaries_localhierarchy(
   res = res_local, 
@@
 ```{r, message=FALSE, warning=FALSE}
-plots <- plot_mu_raw(fit = global_fit, parname = "mu")
+plots <- plot_muraw_localhierarchy(fit = global_fit, parname = "mu")
 ```
@@
 dat_subnat <- dat_all$data
-dat_subnat |> select(level1, level2, mu_global, level1_eta, level2_eta, y)
+dat_subnat |> dplyr::select(level1, level2, mu_global, level1_eta, level2_eta, y)
 ```
@@
 global_fit$post_summ |>
-  filter(str_detect(variable_no_index, "sigma"))
+  dplyr::filter(stringr::str_detect(variable_no_index, "sigma"))
 global_fitsubnat_seq$post_summ |>
-  filter(str_detect(variable_no_index, "sigma"))
+  dplyr::filter(stringr::str_detect(variable_no_index, "sigma"))
 ```
@@
 ```{r, message=FALSE, warning=FALSE}
-res_globalsubnat_seq <- posterior_summary_hierparam(
+res_globalsubnat_seq <- posterior_summary_hierparam_localhierarchy(
   fit = global_fitsubnat_seq,
@@
 
-res_local_seq <- posterior_summary_hierparam(
+res_local_seq <- posterior_summary_hierparam_localhierarchy(
   fit = fit_local_seq, 
@@
 ```{r}
-plots <- plot_posterior_summaries(
+plots <- plot_posterior_summaries_localhierarchy(
   res = res_local_seq, 
@@
 | **Argument** | **Description** |
-|----|----|
+|------------------------------------|------------------------------------|
 | **`survey_df`** | Tibble with survey data |

### 2025-10-08 R/LocalHierarchy-package.R

@@ -2,9 +2,23 @@
 "_PACKAGE"
 
 ## usethis namespace: start
+#' @import dplyr
+#' @import ggplot2
+#' @importFrom bayesplot mcmc_dens_overlay
+#' @importFrom bayesplot mcmc_intervals
+#' @importFrom cmdstanr cmdstan_model
+#' @importFrom cmdstanr read_cmdstan_csv
+#' @importFrom posterior as_draws_list
 #' @importFrom purrr list_flatten
 #' @importFrom purrr map
 #' @importFrom purrr map_int
-#' @importFrom purrr map2_int
+#' @importFrom purrr map2_int map_dbl
+#' @importFrom stats dnorm quantile rnorm
+#' @importFrom stringr fixed
+#' @importFrom tibble tibble
+#' @importFrom tidyr pivot_longer
+#' @importFrom tidyr pivot_wider
+#' @importFrom truncnorm rtruncnorm
+#' @importFrom utils data
 ## usethis namespace: end
 NULL

### 2025-10-08 R/fit_model.R

@@ -72,10 +72,9 @@
 #' of locations.
 #' When using a global_fit to fix parameter values, what exactly is fixed is determined by the runstep and global_fit combi.
 #'
-#' @importFrom cmdstanr cmdstan_model write_stan_file
+#' @importFrom cmdstanr cmdstan_model
 #' @importFrom tibble tibble
 #' @import dplyr
-#' @importFrom stringr str_replace_all
 #'
 #'
 #'

### 2025-10-08 R/get_geo_unit_index_data.R

@@ -11,8 +11,9 @@
 #' @returns tibble with unique hierarchical levels (one column for each level) and index "c" assigned based on area
 #' @export
 #'
+#'
 #' @examples
-#' survey_dat <- tibble(subcluster = c("A", "A", "B", "B"),
+#' survey_dat <- tibble::tibble(subcluster = c("A", "A", "B", "B"),
 #'     iso = c("iso1", "iso2", "iso3", "iso4"))
 #' get_geo_unit_index_data(survey_dat,
 #'         hierarchical_levels = c("intercept", "subcluster", "iso"),

### 2025-10-08 R/get_posterior_summaries.R

@@ -31,7 +31,8 @@ get_posterior_summaries_localhierarchy <- function(
     purrr::list_rbind()%>%
     # add column wtith parname w/o []
     dplyr::mutate(
-      variable_no_index = stringr::str_split_i(string = variable, pattern = fixed("["), i = 1)
+      variable_no_index = stringr::str_split_i(string = variable,
+                                               pattern = stringr::fixed("["), i = 1)
     )
 
   return(combined_summary)

### 2025-10-08 R/hier_data_helpers.R

@@ -42,6 +42,8 @@ is_there_a_comma <- function(variable){
 #'
 #' @returns matrix with rows and columns corresponding to the indices in the variable
 #'
+#' @importFrom tidyr pivot_wider pivot_longer
+#'
 create_a_matrix <- function(post_summ){
   res <- post_summ %>%
     mutate(
@@ -82,7 +84,7 @@ create_a_matrix <- function(post_summ){
 variable_exists_in_draws <- function(fit, var_name) {
   all_variables <- posterior::variables(fit$samples$draws())
   all_variables_no_index <- stringr::str_split_i(string = all_variables,
-                                                 pattern = fixed("["),
+                                                 pattern = stringr::fixed("["),
                                                  i = 1)
   return(any(all_variables_no_index == var_name))
 }
@@ -100,12 +102,7 @@ variable_exists_in_draws <- function(fit, var_name) {
 #' @returns data frame with the same columns as `estimate_df` plus columns containing
 #' just the variable name and the indices.
 #'
-#' @examples
-#' split_var_name_and_index(
-#' estimate_df = tibble(variable = c("a[1]", "a[2]")), num_inds = 1)
-#' split_var_name_and_index(
-#' estimate_df = tibble(variable = c("a[1,1]", "a[2,1]")), num_inds = 2)
-#'
+##'
 split_var_name_and_index <- function(estimate_df, num_inds) {
   # column names for indices, starting from "i"
   ind_names <- letters[seq(from = 9, length = num_inds)]
@@ -135,6 +132,11 @@ split_var_name_and_index <- function(estimate_df, num_inds) {
 # split_var_name_and_index(
 #   estimate_df = tibble(variable = c("a[1]", "a[2]")),
 #   num_inds = 2)
+#
+# split_var_name_and_index(
+# estimate_df = tibble::tibble(variable = c("a[1]", "a[2]")), num_inds = 1)
+# split_var_name_and_index(
+# estimate_df = tibble::tibble(variable = c("a[1,1]", "a[2,1]")), num_inds = 2)
 
 
 #' Starting from an array of parameter values (of dimension 1 or 2),

### 2025-10-08 R/inits.R

@@ -8,6 +8,8 @@
 #'
 #' @returns a list of initial values for the model parameters
 #'
+#' @importFrom truncnorm rtruncnorm
+#'
 init_fun_localhierarchy <- function(chain_id, stan_data){
   set.seed(chain_id)
   inits <- list()

### 2025-10-08 R/plot_posterior_summaries.R

@@ -12,6 +12,7 @@
 #' @param k_select optional, if res contains k, which k values to show? if NULL, all k values are shown
 #' @param dodge used for offsetting plots, default is 0.5
 #'
+#' @export
 #' @returns ggplot object
 #'
 plot_posterior_summaries_localhierarchy <- function(

### 2025-10-08 R/posterior_summary_hierparam.R

@@ -13,6 +13,7 @@
 #'  these mus are obtained by summing up all relevant etas
 #' for morethan1param, each level has a list where k refers to the index
 #' @export
+#' @importFrom tidybayes spread_draws
 #'
 posterior_summary_hierparam_localhierarchy <- function(fit, parname, morethan1param = FALSE,
                                         hierarchical_levels = fit$hierarchical_level){
@@ -55,10 +56,12 @@ posterior_summary_hierparam_localhierarchy <- function(fit, parname, morethan1pa
   }
   if (!morethan1param){
     res <- map(mu, function(tibble_samples)
-      tibble_samples %>% select(name, value)  %>% reframe(summary_df(value), .by = name))
+      tibble_samples %>% dplyr::select(name, value)  %>%
+        dplyr::reframe(summary_df(value), .by = name))
   } else {
     res <- map(mu, function(tibble_samples)
-      tibble_samples %>% select(name, value, k)  %>% reframe(summary_df(value), .by = c(name, k)))
+      tibble_samples %>% dplyr::select(name, value, k)  %>%
+        dplyr::reframe(summary_df(value), .by = c(name, k)))
   }
   return(res)
 }

### 2025-10-08 R/prior_post.R

@@ -16,14 +16,17 @@
 #'
 #' @export
 #'
+#' @importFrom posterior as_draws_list
+#' @importFrom bayesplot mcmc_dens_overlay
+#'
 plot_muraw_localhierarchy <- function(fit, parname, morethan1param = FALSE,
                         nresultsperpage = 30){
   helper_get_plots_allmuraw <- function(samples_tibble){
     parnames <- names(samples_tibble)[-seq(1,3)] # %>% select(-.chain, -.iteration, -.draw))
-    samp <- as_draws_list(samples_tibble)
+    samp <- posterior::as_draws_list(samples_tibble)
     plots_allmuraw <- list()
     for (parname in parnames){
-      plots_allmuraw[[parname]] <- mcmc_dens_overlay(samp, pars = parname) +
+      plots_allmuraw[[parname]] <- bayesplot::mcmc_dens_overlay(samp, pars = parname) +
         # warning when adding the dnorm, not sure why
         stat_function(fun = dnorm, color = "red")
     }
@@ -88,6 +91,8 @@ plot_muraw_localhierarchy <- function(fit, parname, morethan1param = FALSE,
 #'
 #' @returns Plot with density of sigma_estimate and prior added
 #' @export
+#'
+#'@importFrom truncnorm dtruncnorm
 #'
 plot_prior_post_sigmas_localhierarchy <- function(fit, parname){
   # sigmas, fine to look at all k's combined

### 2025-10-08 paper/paper.qmd
paper updates

@@
 editor: visual
-execute:
-  cache: true
 abstract: | 
@@
 ![Data, estimates, and forecasts from 1970 until 2030 for FP indicators for married women in Benin (BEN), Gambia (GMB), and Ghana (GHA).
-Graphs show mCPR, unmet need for modern methods, demand, and demand satisfied with modern methods.
+Graphs show modern contraceptive use, unmet need for modern methods, demand, and demand satisfied with modern methods.
 Survey data are shown in colored dots with 95% confidence intervals that reflect total error variance.
 FPET estimates are shown in black, with grey shaded areas representing 90% credible intervals.](fig/fp_local.jpg){#fig-localfits}
-
-<!-- ```{r} -->
-
-<!-- #| fig-height: 9 -->
-
-<!-- #| fig-width: 7 -->
-
-<!-- #| echo: false -->
-
-<!-- #| label: fig-localfits -->
-
-<!-- #| fig-cap: "Data, estimates, and forecasts from 1970 until 2030 for FP indicators for married women in Benin (BEN), Gambia (GMB), and Ghana (GHA).
-Graphs show mCPR, unmet need for modern methods, demand, and demand satisfied with modern methods.
-Survey data are shown in colored dots with 95% confidence intervals that reflect total error variance.
-FPET estimates are shown in black, with grey shaded areas representing 90% credible intervals." -->
-
-<!-- plots <- readRDS("outputfpet_localfits.rds") -->
-
-<!-- plots[[1]] /  plots[[2]]/ plots[[3]]  -->
-
-<!-- ``` -->
 
@@
 
-Given limited country-specific information on some parameters, Bayesian hierarchical models are used to share information between populations, using hierarchical structures such as the one illustrated in Figure \ref{fig-hierstruc}.
+Given limited country-specific information on some parameters, Bayesian hierarchical models are used to share information between populations, using hierarchical structures such as the one illustrated in \@fig-hierstruc.
 Specifically for family planning estimation and forecasting, a hierarchical model is used to estimate the level of demand in a reference year and the asymptote of the transition of demand from low to high values.
@@
 
-```{=latex}
-\begin{figure}[h]
-\begin{adjustbox}{max size={\textwidth}{\textheight},center}
-%\resizebox{\textwidth}{\textwidth}{
-\begin{tikzpicture}[%scale=0.8,
-  box/.style={rectangle, draw, rounded corners, minimum width=2.6cm, minimum height=1cm, align=center},
-  level0/.style={box, fill=world!40},
-  level1/.style={box, fill=blue!20},
-  level2/.style={box, fill=purple!20},
-  level3/.style={box, fill=country!40, minimum width=2.2cm},
-    level4/.style={box, fill=data!20, minimum width=2.2cm},
-  arrow/.style={-{Latex[length=3mm]}, thick},
-  faintarrow/.style={-{Latex[length=2mm]}, thick, opacity=0.4},
-  background/.style 2 args={rectangle, rounded corners, draw=none, fill=#1, fill opacity=0.15, inner sep=0.5cm, fit=#2}
-]
-
-% --- Level 0 ---
-\node[level0] (mu0) {World};
-
-% --- Level 1 ---
-\node[level1, below left=1cm and 3cm of mu0] (mu11) {Sub-Saharan Africa};
-\node[level1, right=of mu11] (mu12) {South-East Asia};
-\node[level1, right=of mu12] (dots1) {$\dots$};
-% --- Level 2 ---
-\node[level2, below left=1cm and 0cm of mu11] (mu21) {Western Africa};
-\node[level2, right=of mu21] (mu22) {Eastern Africa};
-\node[level2, right=of mu22] (dots2) {$\dots$};
-% % --- Level 3 ---
-\node[level3, below left=1cm and 0cm of mu21] (mu31) {Senegal};
-\node[level3, right=of mu31] (mu32) {Burkina Faso};
-\node[level3, right=of mu32] (dots3) {$\dots$};
-% % % --- Level 4 ---
-% \node[level4, below left=1cm and 0cm of mu31] (mu41) {Data point 1};
-% \node[level4, right=of mu41] (mu42) {Data point 2};
-% \node[level4, right=of mu42] (dots4) {$\dots$};
-% --- Background blocks ---
-\begin{pgfonlayer}{background}
-  \node[background={world!40}{(mu0)}] {};
-  \node[background={blue!20}{(mu11)(mu12)(dots1)}] (bg1) {};
-  \node[background={pink!60}{(mu21)(mu22)(dots2) }] (bg2) {};
- \node[background={country!40}{(mu31)(mu32)(dots3)}] (bg3) {};
- % \node[background={data!40}{(mu41)(mu42)(dots4)}] (bg4) {};
-\end{pgfonlayer}
-
-% Sigma labels inside background blocks
-\node[anchor=north west] at (bg1.east) {Major world regions};
-\node[anchor=north west] at (bg2.east) {Smaller world regions};
-\node[anchor=north west] at (bg3.east) {Countries};
-%\node[anchor=north west] at (bg4.east) %{Country data};
-% --- Arrows ---
-\draw[arrow] (mu0.south) -- (mu11.north);
-\draw[arrow] (mu0.south) -- (mu12.north);
-\draw[arrow] (mu0.south) -- (dots1.north);
-
-\draw[arrow] (mu11.south) -- (mu21.north);
-\draw[arrow] (mu11.south) -- (mu22.north);
-\draw[arrow] (mu11.south) -- (dots2.north);
-
-\draw[arrow] (mu21.south) -- (mu31.north);
-\draw[arrow] (mu21.south) -- (mu32.north);
-\draw[arrow] (mu21.south) -- (dots3.north);
-
-% \draw[arrow] (mu31.south) -- (mu41.north);
-% \draw[arrow] (mu31.south) -- (mu42.north);
-% \draw[arrow] (mu31.south) -- (dots4.north);
-\end{tikzpicture}%}
-\end{adjustbox}
-\caption{Illustration of hierarchical grouping of countries in world regions.}\label{fig-hierstruc}
-\end{figure}
-```
-
-<!-- ```{r, echo = FALSE} -->
-
-<!-- plots <- readRDS("outputfpet_params.rds") -->
+<!-- ```{=latex} -->
+
+<!-- \begin{figure}[h] -->
+
+<!-- \begin{adjustbox}{max size={\textwidth}{\textheight},center} -->
+
+<!-- %\resizebox{\textwidth}{\textwidth}{ -->
+
+<!-- \begin{tikzpicture}[%scale=0.8, -->
+
+<!--   box/.style={rectangle, draw, rounded corners, minimum width=2.6cm, minimum height=1cm, align=center}, -->
+
+<!--   level0/.style={box, fill=world!40}, -->
+
+<!--   level1/.style={box, fill=blue!20}, -->
+
+<!--   level2/.style={box, fill=purple!20}, -->
+
+<!--   level3/.style={box, fill=country!40, minimum width=2.2cm}, -->
+
+<!--     level4/.style={box, fill=data!20, minimum width=2.2cm}, -->
+
+<!--   arrow/.style={-{Latex[length=3mm]}, thick}, -->
+
+<!--   faintarrow/.style={-{Latex[length=2mm]}, thick, opacity=0.4}, -->
+
+<!--   background/.style 2 args={rectangle, rounded corners, draw=none, fill=#1, fill opacity=0.15, inner sep=0.5cm, fit=#2} -->
+
+<!-- ] -->
+
+<!-- % --- Level 0 --- -->
+
+<!-- \node[level0] (mu0) {World}; -->
+
+<!-- % --- Level 1 --- -->
+
+<!-- \node[level1, below left=1cm and 3cm of mu0] (mu11) {Sub-Saharan Africa}; -->
+
+<!-- \node[level1, right=of mu11] (mu12) {South-East Asia}; -->
+
+<!-- \node[level1, right=of mu12] (dots1) {$\dots$}; -->
+
+<!-- % --- Level 2 --- -->
+
+<!-- \node[level2, below left=1cm and 0cm of mu11] (mu21) {Western Africa}; -->
+
+<!-- \node[level2, right=of mu21] (mu22) {Eastern Africa}; -->
+
+<!-- \node[level2, right=of mu22] (dots2) {$\dots$}; -->
+
+<!-- % % --- Level 3 --- -->
+
+<!-- \node[level3, below left=1cm and 0cm of mu21] (mu31) {Senegal}; -->
+
+<!-- \node[level3, right=of mu31] (mu32) {Burkina Faso}; -->
+
+<!-- \node[level3, right=of mu32] (dots3) {$\dots$}; -->
+
+<!-- % % % --- Level 4 --- -->
+
+<!-- % \node[level4, below left=1cm and 0cm of mu31] (mu41) {Data point 1}; -->
+
+<!-- % \node[level4, right=of mu41] (mu42) {Data point 2}; -->
+
+<!-- % \node[level4, right=of mu42] (dots4) {$\dots$}; -->
+
+<!-- % --- Background blocks --- -->
+
+<!-- \begin{pgfonlayer}{background} -->
+
+<!--   \node[background={world!40}{(mu0)}] {}; -->
+
+<!--   \node[background={blue!20}{(mu11)(mu12)(dots1)}] (bg1) {}; -->
+
+<!--   \node[background={pink!60}{(mu21)(mu22)(dots2) }] (bg2) {}; -->
+
+<!--  \node[background={country!40}{(mu31)(mu32)(dots3)}] (bg3) {}; -->
+
+<!--  % \node[background={data!40}{(mu41)(mu42)(dots4)}] (bg4) {}; -->
+
+<!-- \end{pgfonlayer} -->
+
+<!-- % Sigma labels inside background blocks -->
+
+<!-- \node[anchor=north west] at (bg1.east) {Major world regions}; -->
+
+<!-- \node[anchor=north west] at (bg2.east) {Smaller world regions}; -->
+
+<!-- \node[anchor=north west] at (bg3.east) {Countries}; -->
+
+<!-- %\node[anchor=north west] at (bg4.east) %{Country data}; -->
+
+<!-- % --- Arrows --- -->
+
+<!-- \draw[arrow] (mu0.south) -- (mu11.north); -->
+
+<!-- \draw[arrow] (mu0.south) -- (mu12.north); -->
+
+<!-- \draw[arrow] (mu0.south) -- (dots1.north); -->
+
+<!-- \draw[arrow] (mu11.south) -- (mu21.north); -->
+
+<!-- \draw[arrow] (mu11.south) -- (mu22.north); -->
+
+<!-- \draw[arrow] (mu11.south) -- (dots2.north); -->
+
+<!-- \draw[arrow] (mu21.south) -- (mu31.north); -->
+
+<!-- \draw[arrow] (mu21.south) -- (mu32.north); -->
+
+<!-- \draw[arrow] (mu21.south) -- (dots3.north); -->
+
+<!-- % \draw[arrow] (mu31.south) -- (mu41.north); -->
+
+<!-- % \draw[arrow] (mu31.south) -- (mu42.north); -->
+
+<!-- % \draw[arrow] (mu31.south) -- (dots4.north); -->
+
+<!-- \end{tikzpicture}%} -->
+
+<!-- \end{adjustbox} -->
+
+<!-- \caption{Illustration of hierarchical grouping of countries in world regions.}\label{fig-hierstruc} -->
+
+<!-- \end{figure} -->
 
@@
 
-<!-- ```{r} -->
-
-<!-- #| fig-height: 9 -->
-
-<!-- #| fig-width: 7 -->
-
-<!-- #| echo: false -->
-
-<!-- #| label: fig-level -->
-
-<!-- #| fig-cap: "Estimates of the level of demand for countries in Western Africa and for higher levels used in the hierarchical model (smaller world regions and major world regions).
-Point estimates and 95% credible intervals are shown.
-" -->
-
-<!-- plots$plot_reg_omega[[2]]/ plots$plot_reg_omega[[3]]/  plots$plot_iso_omega -->
-
-<!-- ``` -->
-
-<!-- ```{r} -->
-
-<!-- #| fig-height: 9 -->
-
-<!-- #| fig-width: 7 -->
-
-<!-- #| echo: false -->
-
-<!-- #| label: fig-asympt -->
-
-<!-- #| fig-cap: "Estimates of the asymptote for demand for countries in Western Africa and for higher levels used in the hierarchical model (major world regions).
-Point estimates and 95% credible intervals are shown.
-" -->
-
-<!-- plots$plot_reg_ptilde[[2]]  / plots$plot_iso_ptilde -->
-
-<!-- ``` -->
+![Illustration of hierarchical grouping of countries in world regions.](fig/worldregions.jpg){#fig-hierstruc width="643"}
 
@@
 For a given level $l$, index $r$ refers to the population in a particular region indexed by $r$, for example, a subnational region, a country, or a group of countries.
-We assume that $\mu$ is unconstrained (possibly after a transformation has been applied) and consider the following general hierarchical model setup, see Figure \ref{fig-hiermodel}:
+We assume that $\mu$ is unconstrained (possibly after a transformation has been applied) and consider the following general hierarchical model setup, see @fig-hiermodel:
 
@@
 
+![General specification of a hierarchical model.](fig/hiermodel.jpg){#fig-hiermodel width="664"}
+
 As a specific example, we consider the estimation of country-specific parameters, where countries are organized into subregions of countries (illustrated in Figure \ref{fig-hiermodel}).
@@
 We use the term local hierarchical models to refer to a hierarchical model in which a subset of parameters have been fixed at posterior mean estimates obtained from a fit of the model to a larger data set.
-A typical workflow for local estimation of national level outcomes is illustrated in Figure \ref{fig-workflow}.
+A typical workflow for local estimation of national level outcomes is illustrated in @fig-workflow.
 For national estimation using the model given by Eq.\ref{eq-countryhier}, the user starts by fitting a global model, using a global data base and estimating all parameters in a hierarchical model.
@@
 
-```{=latex}
-\begin{figure}[h]
-\begin{adjustbox}{max size={\textwidth}{\textheight},center}
-\begin{tikzpicture}[
-  box/.style={rectangle, draw, thick, rounded corners, minimum width=6cm, align=left, inner sep=6pt},
-  databox/.style={box, fill=data!20},
-  modelbox/.style={box, fill=param!20},
-  arrowdata/.style={-{Latex[length=3mm]}, thick, data},
-  arrowparam/.style={-{Latex[length=3mm]}, thick, param!80},
-  font=\sffamily
-]
-% --- Nodes ---
-\node[databox] (globaldata) {Global data base of country data};
-\node[modelbox, below=1cm of globaldata] (globalmodel) {%
-  Global national modeling:\\
-  Estimate all parameters in hierarchical model, e.g., \\
-\(
-  \begin{aligned}
-\mu_w^{(region)} &\sim N(\mu^{(global)}, \sigma_{ region}^2),\\
-\mu_s^{(subregion)} &\sim N(\mu_{w[s]}^{(region)}, \sigma_{ subregion}^2), \\
-\mu_c^{(country)} &\sim N(\mu_{s[c]}^{(subregion)}, \sigma_{ country}^2),  \end{aligned}
-\)
-};
-\node[modelbox, below=1cm of globalmodel] (localmodel) {%
-  Local national modeling:\\
-  Estimate national parameters only, \\
-  fix parameter estimates at higher levels of the hierarchy \\i.e., $\mu_c^{(country)}$ with $\mu_c^{(country)} \sim N(\hat{\mu}_{s[c]}^{(subregion)}, \hat{\sigma}_{ country}^2).$
-};
-
-% --- Data arrow (down then right) ---
-\draw[arrowdata] (globaldata.south) -- (globalmodel.north);
-\draw[arrowdata] (globaldata.east) ++(0,0) -- ++(4.5,0) |- (localmodel.east)
-node[midway,above] {Data for one country};
-
-% --- Parameter arrow ---
-\draw[arrowparam] (globalmodel.south) -- (localmodel.north);
-
-% --- Legend ---
-% \node[arrowparam, below=1.2cm of localmodel.west, anchor=west] (leg1) {}; 
-% \node[right=0.3cm of leg1] {\textcolor{param}{$\rightarrow$ Parameter estimates}};
-\node[below=1cm of localmodel](parameter) {\textcolor{param}{$\rightarrow$ Parameter estimates}};
-\node[right=1cm of parameter] {\textcolor{data}{$\rightarrow$ Data}};
-\end{tikzpicture}
-\end{adjustbox}
-\caption{Workflow for global and local national modeling. }\label{fig-workflow}
-\end{figure}
-```
+<!-- ```{=latex} -->
+
+<!-- \begin{figure}[h] -->
+
+<!-- \begin{adjustbox}{max size={\textwidth}{\textheight},center} -->
+
+<!-- \begin{tikzpicture}[ -->
+
+<!--   box/.style={rectangle, draw, thick, rounded corners, minimum width=6cm, align=left, inner sep=6pt}, -->
+
+<!--   databox/.style={box, fill=data!20}, -->
+
+<!--   modelbox/.style={box, fill=param!20}, -->
+
+<!--   arrowdata/.style={-{Latex[length=3mm]}, thick, data}, -->
+
+<!--   arrowparam/.style={-{Latex[length=3mm]}, thick, param!80}, -->
+
+<!--   font=\sffamily -->
+
+<!-- ] -->
+
+<!-- % --- Nodes --- -->
+
+<!-- \node[databox] (globaldata) {Global data base of country data}; -->
+
+<!-- \node[modelbox, below=1cm of globaldata] (globalmodel) {% -->
+
+<!--   Global national modeling:\\ -->
+
+<!--   Estimate all parameters in hierarchical model, e.g., \\ -->
+
+<!-- \( -->
+
+<!--   \begin{aligned} -->
+
+<!-- \mu_w^{(region)} &\sim N(\mu^{(global)}, \sigma_{ region}^2),\\ -->
+
+<!-- \mu_s^{(subregion)} &\sim N(\mu_{w[s]}^{(region)}, \sigma_{ subregion}^2), \\ -->
+
+<!-- \mu_c^{(country)} &\sim N(\mu_{s[c]}^{(subregion)}, \sigma_{ country}^2),  \end{aligned} -->
+
+<!-- \) -->
+
+<!-- }; -->
+
+<!-- \node[modelbox, below=1cm of globalmodel] (localmodel) {% -->
+
+<!--   Local national modeling:\\ -->
+
+<!--   Estimate national parameters only, \\ -->
+
+<!--   fix parameter estimates at higher levels of the hierarchy \\i.e., $\mu_c^{(country)}$ with $\mu_c^{(country)} \sim N(\hat{\mu}_{s[c]}^{(subregion)}, \hat{\sigma}_{ country}^2).$ -->
+
+<!-- }; -->
+
+<!-- % --- Data arrow (down then right) --- -->
+
+<!-- \draw[arrowdata] (globaldata.south) -- (globalmodel.north); -->
+
+<!-- \draw[arrowdata] (globaldata.east) ++(0,0) -- ++(4.5,0) |- (localmodel.east) -->
+
+<!-- node[midway,above] {Data for one country}; -->
+
+<!-- % --- Parameter arrow --- -->
+
+<!-- \draw[arrowparam] (globalmodel.south) -- (localmodel.north); -->
+
+<!-- % --- Legend --- -->
+
+<!-- % \node[arrowparam, below=1.2cm of localmodel.west, anchor=west] (leg1) {};  -->
+
+<!-- % \node[right=0.3cm of leg1] {\textcolor{param}{$\rightarrow$ Parameter estimates}}; -->
+
+<!-- \node[below=1cm of localmodel](parameter) {\textcolor{param}{$\rightarrow$ Parameter estimates}}; -->
+
+<!-- \node[right=1cm of parameter] {\textcolor{data}{$\rightarrow$ Data}}; -->
+
+<!-- \end{tikzpicture} -->
+
+<!-- \end{adjustbox} -->
+
+<!-- \caption{Workflow for global and local national modeling. }\label{fig-workflow} -->
+
+<!-- \end{figure} -->
+
+<!-- ``` -->
+
+![Workflow for global and local national modeling.](fig/workflow.jpg){#fig-workflow width="677"}
 
@@
 
-The simulation setup is summarized in Figure \ref{fig-simusetup}, top and center panels. Simulation of subnational data is discussed in a later section.
-
-```{=latex}
-\begin{figure}[h]
-\fbox{\begin{adjustbox}{max size={\textwidth}{\textheight},center}
-\begin{tikzpicture}[
-  box/.style={rectangle, draw, rounded corners, minimum width=2cm, minimum height=1cm, align=center},
-  level0/.style={box, fill=yellow!40},
-  level1/.style={box, fill=orange!40},
-  level2/.style={box, fill=green!20},
-  arrow/.style={-{Latex[length=3mm]}, thick},
-  faintarrow/.style={-{Latex[length=2mm]}, thick, opacity=0.4},
-  background/.style 2 args={rectangle, rounded corners, draw=none, fill=#1, fill opacity=0.15, inner sep=0.5cm, fit=#2}
-]
-
-% --- Level 0 ---
-\node[level0] (mu0) {$\mu^{(\glob)} \sim \mathcal{N}(0,s_\glob^2)$};
-
-% --- Level 1 ---
-\node[level1, below left=1.5cm and 3cm of mu0] (mu11) {$\mu^{(\country)}_1 \sim \mathcal{N}(\mu^{(\glob)},\sigma_{\country}^2)$};
-\node[level1, right=of mu11] (mu12) {$\mu^{(\country)}_2$};
-\node[level1, right=of mu12] (dots1) {$\dots$};
-% --- Level 2 ---
-\node[level2, below left=1.5cm and 0cm of mu11] (mu21) {$\mu^{(\subnat)}_1 \sim \mathcal{N}(\mu^{(\country)}_1,\sigma_{\subnat}^2)$};
-\node[level2, right=of mu21] (mu22) {$\mu^{(2)}_2$};
-\node[level2, right=of mu22] (dots2) {$\dots$};
-% --- Background blocks ---
-\begin{pgfonlayer}{background}
-  \node[background={yellow!40}{(mu0)}] {};
-  \node[background={orange!40}{(mu11)(mu12)(dots1)}] (bg1) {};
-  \node[background={green!40}{(mu21)(mu22)(dots2) }] (bg2) {};
-\end{pgfonlayer}
-
-% Sigma labels inside background blocks
-\node[anchor=north west] at (bg1.east) {$\sigma_{\country}$, country};
-\node[anchor=north west] at (bg2.east) {$\sigma_{\subnat}$, subnational region};
-% --- Arrows ---
-\draw[arrow] (mu0.south) -- (mu11.north);
-\draw[arrow] (mu0.south) -- (mu12.north);
-\draw[arrow] (mu0.south) -- (dots1.north);
-
-\draw[arrow] (mu11.south) -- (mu21.north);
-\draw[arrow] (mu11.south) -- (mu22.north);
-\draw[arrow] (mu11.south) -- (dots2.north);
-\end{tikzpicture}
-\end{adjustbox}}
-
-% local model nat
-\fbox{\begin{adjustbox}{max size={\textwidth}{\textheight},center}
-\begin{tikzpicture}[
-  box/.style={rectangle, draw, rounded corners, minimum width=2cm, minimum height=1cm, align=center},
-  level0/.style={box, fill=yellow!40},
-  level1/.style={box, fill=orange!40},
-  level2/.style={box, fill=green!20},
-    data/.style={box, fill=gray!20},
-  arrow/.style={-{Latex[length=3mm]}, thick},
-  faintarrow/.style={-{Latex[length=2mm]}, thick, opacity=0.4},
-  background/.style 2 args={rectangle, rounded corners, draw=none, fill=#1, fill opacity=0.15, inner sep=0.5cm, fit=#2}
-]
-
-\node[level1] (mu0) {$\mu^{(\country)}_1$};
-
-% --- Level 1 ---
-\node[data, below left=1.5cm and 3cm of mu0] (mu11) {$y^{(country)}_1 \sim \mathcal{N}(\mu^{(\country)}_1,\sigma_{y}^2)$};
-\node[data, right=of mu11] (mu12) {$y^{(country)}_2$};
-\node[data, right=of mu12] (dots1) {$\dots$};
-% --- Background blocks ---
-\begin{pgfonlayer}{background}
-  \node[background={gray!40}{(mu11)(mu12)(dots1)}] (bg1) {};
-\end{pgfonlayer}
-
-% Sigma labels inside background blocks
-\node[anchor=north west] at (bg1.east) {$\sigma_{y}$, country data for first country};
-
-% --- Arrows ---
-\draw[arrow] (mu0.south) -- (mu11.north);
-\draw[arrow] (mu0.south) -- (mu12.north);
-\draw[arrow] (mu0.south) -- (dots1.north);
-\end{tikzpicture}
-\end{adjustbox}}
-
-% local model subnat
-\fbox{
-\begin{adjustbox}{max size={\textwidth}{\textheight},center}
-\begin{tikzpicture}[
-  box/.style={rectangle, draw, rounded corners, minimum width=2cm, minimum height=1cm, align=center},
-  level0/.style={box, fill=yellow!40},
-  level1/.style={box, fill=orange!40},
-  level2/.style={box, fill=green!20},
-    data/.style={box, fill=gray!20},
-  arrow/.style={-{Latex[length=3mm]}, thick},
-  faintarrow/.style={-{Latex[length=2mm]}, thick, opacity=0.4},
-  background/.style 2 args={rectangle, rounded corners, draw=none, fill=#1, fill opacity=0.15, inner sep=0.5cm, fit=#2}
-]
-
-% keep same colors so now color of level 1 at top 
-\node[level2] (mu0) {$\mu^{(\subnat)}_1$};
-
-% --- Level 1 ---
-\node[data, below left=1.5cm and 3cm of mu0] (mu11) {$y^{(subnat)}_1 \sim \mathcal{N}(\mu^{(\subnat)}_1,\sigma_{y}^2)$};
-\node[data, right=of mu11] (mu12) {$y^{(subnat)}_2$};
-\node[data, right=of mu12] (dots1) {$\dots$};
-% --- Background blocks ---
-\begin{pgfonlayer}{background}
-  \node[background={gray!40}{(mu11)(mu12)(dots1)}] (bg1) {};
-\end{pgfonlayer}
-
-% Sigma labels inside background blocks
-\node[anchor=north west] at (bg1.east) {$\sigma_{y}$, subnational data for first region};
-
-% --- Arrows ---
-\draw[arrow] (mu0.south) -- (mu11.north);
-\draw[arrow] (mu0.south) -- (mu12.north);
-\draw[arrow] (mu0.south) -- (dots1.north);
-\end{tikzpicture}
-\end{adjustbox}}
-\caption{Diagram of simulation settings.
+The simulation setup is summarized in @fig-simusetup, top and center panels. Simulation of subnational data is discussed in a later section.
+
+<!-- ```{=latex} -->
+
+<!-- \begin{figure}[h] -->
+
+<!-- \fbox{\begin{adjustbox}{max size={\textwidth}{\textheight},center} -->
+
+<!-- \begin{tikzpicture}[ -->
+
+<!--   box/.style={rectangle, draw, rounded corners, minimum width=2cm, minimum height=1cm, align=center}, -->
+
+<!--   level0/.style={box, fill=yellow!40}, -->
+
+<!--   level1/.style={box, fill=orange!40}, -->
+
+<!--   level2/.style={box, fill=green!20}, -->
+
+<!--   arrow/.style={-{Latex[length=3mm]}, thick}, -->
+
+<!--   faintarrow/.style={-{Latex[length=2mm]}, thick, opacity=0.4}, -->
+
+<!--   background/.style 2 args={rectangle, rounded corners, draw=none, fill=#1, fill opacity=0.15, inner sep=0.5cm, fit=#2} -->
+
+<!-- ] -->
+
+<!-- % --- Level 0 --- -->
+
+<!-- \node[level0] (mu0) {$\mu^{(\glob)} \sim \mathcal{N}(0,s_\glob^2)$}; -->
+
+<!-- % --- Level 1 --- -->
+
+<!-- \node[level1, below left=1.5cm and 3cm of mu0] (mu11) {$\mu^{(\country)}_1 \sim \mathcal{N}(\mu^{(\glob)},\sigma_{\country}^2)$}; -->
+
+<!-- \node[level1, right=of mu11] (mu12) {$\mu^{(\country)}_2$}; -->
+
+<!-- \node[level1, right=of mu12] (dots1) {$\dots$}; -->
+
+<!-- % --- Level 2 --- -->
+
+<!-- \node[level2, below left=1.5cm and 0cm of mu11] (mu21) {$\mu^{(\subnat)}_1 \sim \mathcal{N}(\mu^{(\country)}_1,\sigma_{\subnat}^2)$}; -->
+
+<!-- \node[level2, right=of mu21] (mu22) {$\mu^{(2)}_2$}; -->
+
+<!-- \node[level2, right=of mu22] (dots2) {$\dots$}; -->
+
+<!-- % --- Background blocks --- -->
+
+<!-- \begin{pgfonlayer}{background} -->
+
+<!--   \node[background={yellow!40}{(mu0)}] {}; -->
+
+<!--   \node[background={orange!40}{(mu11)(mu12)(dots1)}] (bg1) {}; -->
+
+<!--   \node[background={green!40}{(mu21)(mu22)(dots2) }] (bg2) {}; -->
+
+<!-- \end{pgfonlayer} -->
+
+<!-- % Sigma labels inside background blocks -->
+
+<!-- \node[anchor=north west] at (bg1.east) {$\sigma_{\country}$, country}; -->
+
+<!-- \node[anchor=north west] at (bg2.east) {$\sigma_{\subnat}$, subnational region}; -->
+
+<!-- % --- Arrows --- -->
+
+<!-- \draw[arrow] (mu0.south) -- (mu11.north); -->
+
+<!-- \draw[arrow] (mu0.south) -- (mu12.north); -->
+
+<!-- \draw[arrow] (mu0.south) -- (dots1.north); -->
+
+<!-- \draw[arrow] (mu11.south) -- (mu21.north); -->
+
+<!-- \draw[arrow] (mu11.south) -- (mu22.north); -->
+
+<!-- \draw[arrow] (mu11.south) -- (dots2.north); -->
+
+<!-- \end{tikzpicture} -->
+
+<!-- \end{adjustbox}} -->
+
+<!-- % local model nat -->
+
+<!-- \fbox{\begin{adjustbox}{max size={\textwidth}{\textheight},center} -->
+
+<!-- \begin{tikzpicture}[ -->
+
+<!--   box/.style={rectangle, draw, rounded corners, minimum width=2cm, minimum height=1cm, align=center}, -->
+
+<!--   level0/.style={box, fill=yellow!40}, -->
+
+<!--   level1/.style={box, fill=orange!40}, -->
+
+<!--   level2/.style={box, fill=green!20}, -->
+
+<!--     data/.style={box, fill=gray!20}, -->
+
+<!--   arrow/.style={-{Latex[length=3mm]}, thick}, -->
+
+<!--   faintarrow/.style={-{Latex[length=2mm]}, thick, opacity=0.4}, -->
+
+<!--   background/.style 2 args={rectangle, rounded corners, draw=none, fill=#1, fill opacity=0.15, inner sep=0.5cm, fit=#2} -->
+
+<!-- ] -->
+
+<!-- \node[level1] (mu0) {$\mu^{(\country)}_1$}; -->
+
+<!-- % --- Level 1 --- -->
+
+<!-- \node[data, below left=1.5cm and 3cm of mu0] (mu11) {$y^{(country)}_1 \sim \mathcal{N}(\mu^{(\country)}_1,\sigma_{y}^2)$}; -->
+
+<!-- \node[data, right=of mu11] (mu12) {$y^{(country)}_2$}; -->
+
+<!-- \node[data, right=of mu12] (dots1) {$\dots$}; -->
+
+<!-- % --- Background blocks --- -->
+
+<!-- \begin{pgfonlayer}{background} -->
+
+<!--   \node[background={gray!40}{(mu11)(mu12)(dots1)}] (bg1) {}; -->
+
+<!-- \end{pgfonlayer} -->
+
+<!-- % Sigma labels inside background blocks -->
+
+<!-- \node[anchor=north west] at (bg1.east) {$\sigma_{y}$, country data for first country}; -->
+
+<!-- % --- Arrows --- -->
+
+<!-- \draw[arrow] (mu0.south) -- (mu11.north); -->
+
+<!-- \draw[arrow] (mu0.south) -- (mu12.north); -->
+
+<!-- \draw[arrow] (mu0.south) -- (dots1.north); -->
+
+<!-- \end{tikzpicture} -->
+
+<!-- \end{adjustbox}} -->
+
+<!-- % local model subnat -->
+
+<!-- \fbox{ -->
+
+<!-- \begin{adjustbox}{max size={\textwidth}{\textheight},center} -->
+
+<!-- \begin{tikzpicture}[ -->
+
+<!--   box/.style={rectangle, draw, rounded corners, minimum width=2cm, minimum height=1cm, align=center}, -->
+
+<!--   level0/.style={box, fill=yellow!40}, -->
+
+<!--   level1/.style={box, fill=orange!40}, -->
+
+<!--   level2/.style={box, fill=green!20}, -->
+
+<!--     data/.style={box, fill=gray!20}, -->
+
+<!--   arrow/.style={-{Latex[length=3mm]}, thick}, -->
+
+<!--   faintarrow/.style={-{Latex[length=2mm]}, thick, opacity=0.4}, -->
+
+<!--   background/.style 2 args={rectangle, rounded corners, draw=none, fill=#1, fill opacity=0.15, inner sep=0.5cm, fit=#2} -->
+
+<!-- ] -->
+
+<!-- % keep same colors so now color of level 1 at top  -->
+
+<!-- \node[level2] (mu0) {$\mu^{(\subnat)}_1$}; -->
+
+<!-- % --- Level 1 --- -->
+
+<!-- \node[data, below left=1.5cm and 3cm of mu0] (mu11) {$y^{(subnat)}_1 \sim \mathcal{N}(\mu^{(\subnat)}_1,\sigma_{y}^2)$}; -->
+
+<!-- \node[data, right=of mu11] (mu12) {$y^{(subnat)}_2$}; -->
+
+<!-- \node[data, right=of mu12] (dots1) {$\dots$}; -->
+
+<!-- % --- Background blocks --- -->
+
+<!-- \begin{pgfonlayer}{background} -->
+
+<!--   \node[background={gray!40}{(mu11)(mu12)(dots1)}] (bg1) {}; -->
+
+<!-- \end{pgfonlayer} -->
+
+<!-- % Sigma labels inside background blocks -->
+
+<!-- \node[anchor=north west] at (bg1.east) {$\sigma_{y}$, subnational data for first region}; -->
+
+<!-- % --- Arrows --- -->
+
+<!-- \draw[arrow] (mu0.south) -- (mu11.north); -->
+
+<!-- \draw[arrow] (mu0.south) -- (mu12.north); -->
+
+<!-- \draw[arrow] (mu0.south) -- (dots1.north); -->
+
+<!-- \end{tikzpicture} -->
+
+<!-- \end{adjustbox}} -->
+
+<!-- \caption{Diagram of simulation settings.
 Top: Hierarchical model for generating country and subnational parameters.
 Center: Data generating mechanism for national data, illustrated for the first country.
-Bottom: Data generating mechanism for subnational data, illustrated for the first subnational region (used only in Section 5).}\label{fig-simusetup}
-\end{figure}
-\clearpage
-```
+Bottom: Data generating mechanism for subnational data, illustrated for the first subnational region (used only in Section 5).}\label{fig-simusetup} -->
+
+<!-- \end{figure} -->
+
+<!-- \clearpage -->
+
+<!-- ``` -->
+
+![Diagram of simulation settings.
+Top: Hierarchical model for generating country and subnational parameters.
+Center: Data generating mechanism for national data, illustrated for the first country.
+Bottom: Data generating mechanism for subnational data, illustrated for the first subnational region (used only in Section 5).](fig/simusetup.jpg){#fig-simusetup width="682"}
 
@@
 
-The approach followed in NISE is illustrated in Figure \ref{fig-nise}.
+The approach followed in NISE is illustrated in @fig-nise.
 In the NISE approach, the first step is to fit a model to a global data base with national-level data, to obtain estimates of model parameters up to and including national level parameters.
@@
 
-```{=latex}
-\begin{figure}[h]
-\begin{adjustbox}{max size={\textwidth}{\textheight},center}
-\begin{tikzpicture}[
-  box/.style={rectangle, draw, thick, rounded corners, minimum width=2.5cm, align=left, inner sep=6pt},
-  eqbox/.style={box},
-  databox/.style={box, fill=data!20},
-  modelbox/.style={box, fill=param!20},
-  arrowdata/.style={-{Latex[length=3mm]}, thick, data},
-  arrowparam/.style={-{Latex[length=3mm]}, thick, param!80},
-  font=\sffamily
-]
-% --- Nodes ---
-\node[eqbox](model){
-Example subnat. hierarchical model:\\
-\(
-  \begin{aligned}
-\mu_w^{(\region)} &\sim N(\mu^{(\glob)}, \sigma_{ \region}^2),\\
-\mu_s^{(\subregion)} &\sim N(\mu_{w[s]}^{(\region)}, \sigma_{ \subregion}^2), \\
-\mu_c^{(\country)} &\sim N(\mu_{s[c]}^{(\subregion)}, \sigma_{ \country}^2),  \\
-\mu_r^{(\subnat)} &\sim N(\mu_{c[r]}^{(\country)}, \sigma_{ \subnat}^2),  \\
-\end{aligned}
-\)
-};
-
-\node[databox, right=5cm of model.north] (globaldata) {Global data base of country data};
-\node[modelbox, below=1cm of globaldata] (globalmodel) {%
-  Global national modeling:\\
-  Estimate all parameters in hierarchical \\ model 
-  $\leq$ country level 
-};
-\node[modelbox, below=1cm of globalmodel] (subglobalmodel) {%
-  Global subnational modeling:\\
-  Estimate subset of parameters in \\ hierarchical model, 
-  i.e., $\sigma_{ subnat}$
-};
-\node[databox, left=1cm of subglobalmodel] (globalsubdata) {Global data base\\ of subnational data};
-\node[modelbox, below=1cm of subglobalmodel] (localmodel) {%
-  Local subnational modeling:\\
-  Estimate subnational parameters only, \\i.e., $\mu_r^{(subnat)}$ in $\mu_r^{(subnat)} \sim N(\hat{\mu}_{c[r]}^{(country)}, \hat{\sigma}_{ subnat}^2).$
-};
-
-% --- Data arrow  ---
-\draw[arrowdata] (globaldata.south) -- (globalmodel.north);
-\draw[arrowdata] (globaldata.east) ++(0,0) -- ++(1.5,0) |- (localmodel.east)
-node[midway,above] {};
-\draw[arrowdata] (globaldata.east) ++(0,0) -- ++(1.5,0) |- (subglobalmodel.east)
-node[midway,above] {};
-\draw[arrowdata] (globalsubdata.east) -- (subglobalmodel.west);
-\draw[arrowdata] (globalsubdata.south) -- (localmodel.west)
-node[midway,above] {};
-
-% --- Parameter arrow ---
-\draw[arrowparam] (globalmodel.south) -- (subglobalmodel.north);
-\draw[arrowparam] (subglobalmodel.south) -- (localmodel.north);
-
-% --- Legend ---
-% \node[arrowparam, below=1.2cm of localmodel.west, anchor=west] (leg1) {}; 
-% \node[right=0.3cm of leg1] {\textcolor{param}{$\rightarrow$ Parameter estimates}};
-\node[below=1.2cm of localmodel.west](parameter) {\textcolor{param}{$\rightarrow$ Parameter estimates}};
-\node[right=1cm of parameter] {\textcolor{data}{$\rightarrow$ Data}};
-\end{tikzpicture}
-\end{adjustbox}
-\caption{Workflow for nationally-informed global and local subnational modeling. }\label{fig-nise}
-\end{figure}
-\clearpage
-```
+<!-- ```{=latex} -->
+
+<!-- \begin{figure}[h] -->
+
+<!-- \begin{adjustbox}{max size={\textwidth}{\textheight},center} -->
+
+<!-- \begin{tikzpicture}[ -->
+
+<!--   box/.style={rectangle, draw, thick, rounded corners, minimum width=2.5cm, align=left, inner sep=6pt}, -->
+
+<!--   eqbox/.style={box}, -->
+
+<!--   databox/.style={box, fill=data!20}, -->
+
+<!--   modelbox/.style={box, fill=param!20}, -->
+
+<!--   arrowdata/.style={-{Latex[length=3mm]}, thick, data}, -->
+
+<!--   arrowparam/.style={-{Latex[length=3mm]}, thick, param!80}, -->
+
+<!--   font=\sffamily -->
+
+<!-- ] -->
+
+<!-- % --- Nodes --- -->
+
+<!-- \node[eqbox](model){ -->
+
+<!-- Example subnat. hierarchical model:\\ -->
+
+<!-- \( -->
+
+<!--   \begin{aligned} -->
+
+<!-- \mu_w^{(\region)} &\sim N(\mu^{(\glob)}, \sigma_{ \region}^2),\\ -->
+
+<!-- \mu_s^{(\subregion)} &\sim N(\mu_{w[s]}^{(\region)}, \sigma_{ \subregion}^2), \\ -->
+
+<!-- \mu_c^{(\country)} &\sim N(\mu_{s[c]}^{(\subregion)}, \sigma_{ \country}^2),  \\ -->
+
+<!-- \mu_r^{(\subnat)} &\sim N(\mu_{c[r]}^{(\country)}, \sigma_{ \subnat}^2),  \\ -->
+
+<!-- \end{aligned} -->
+
+<!-- \) -->
+
+<!-- }; -->
+
+<!-- \node[databox, right=5cm of model.north] (globaldata) {Global data base of country data}; -->
+
+<!-- \node[modelbox, below=1cm of globaldata] (globalmodel) {% -->
+
+<!--   Global national modeling:\\ -->
+
+<!--   Estimate all parameters in hierarchical \\ model  -->
+
+<!--   $\leq$ country level  -->
+
+<!-- }; -->
+
+<!-- \node[modelbox, below=1cm of globalmodel] (subglobalmodel) {% -->
+
+<!--   Global subnational modeling:\\ -->
+
+<!--   Estimate subset of parameters in \\ hierarchical model,  -->
+
+<!--   i.e., $\sigma_{ subnat}$ -->
+
+<!-- }; -->
+
+<!-- \node[databox, left=1cm of subglobalmodel] (globalsubdata) {Global data base\\ of subnational data}; -->
+
+<!-- \node[modelbox, below=1cm of subglobalmodel] (localmodel) {% -->
+
+<!--   Local subnational modeling:\\ -->
+
+<!--   Estimate subnational parameters only, \\i.e., $\mu_r^{(subnat)}$ in $\mu_r^{(subnat)} \sim N(\hat{\mu}_{c[r]}^{(country)}, \hat{\sigma}_{ subnat}^2).$ -->
+
+<!-- }; -->
+
+<!-- % --- Data arrow  --- -->
+
+<!-- \draw[arrowdata] (globaldata.south) -- (globalmodel.north); -->
+
+<!-- \draw[arrowdata] (globaldata.east) ++(0,0) -- ++(1.5,0) |- (localmodel.east) -->
+
+<!-- node[midway,above] {}; -->
+
+<!-- \draw[arrowdata] (globaldata.east) ++(0,0) -- ++(1.5,0) |- (subglobalmodel.east) -->
+
+<!-- node[midway,above] {}; -->
+
+<!-- \draw[arrowdata] (globalsubdata.east) -- (subglobalmodel.west); -->
+
+<!-- \draw[arrowdata] (globalsubdata.south) -- (localmodel.west) -->
+
+<!-- node[midway,above] {}; -->
+
+<!-- % --- Parameter arrow --- -->
+
+<!-- \draw[arrowparam] (globalmodel.south) -- (subglobalmodel.north); -->
+
+<!-- \draw[arrowparam] (subglobalmodel.south) -- (localmodel.north); -->
+
+<!-- % --- Legend --- -->
+
+<!-- % \node[arrowparam, below=1.2cm of localmodel.west, anchor=west] (leg1) {};  -->
+
+<!-- % \node[right=0.3cm of leg1] {\textcolor{param}{$\rightarrow$ Parameter estimates}}; -->
+
+<!-- \node[below=1.2cm of localmodel.west](parameter) {\textcolor{param}{$\rightarrow$ Parameter estimates}}; -->
+
+<!-- \node[right=1cm of parameter] {\textcolor{data}{$\rightarrow$ Data}}; -->
+
+<!-- \end{tikzpicture} -->
+
+<!-- \end{adjustbox} -->
+
+<!-- \caption{Workflow for nationally-informed global and local subnational modeling. }\label{fig-nise} -->
+
+<!-- \end{figure} -->
+
+<!-- \clearpage -->
+
+<!-- ``` -->
+
+![Workflow for nationally-informed global and local subnational modeling.](fig/nise.jpg){#fig-nise width="630"}
 
@@
 ![Data, estimates, and forecasts from 1970 until 2030 for FP indicators for married women in subnational regions in Benin.
-Graphs show mCPR, unmet need for modern methods, demand, and demand satisfied with modern methods.
+Graphs show modern contraceptive use, unmet need for modern methods, demand, and demand satisfied with modern methods.
 Survey data are shown in colored dots with 95% confidence intervals that reflect total error variance.
@@
 <!-- #| fig-cap: "Data, estimates, and forecasts from 1970 until 2030 for FP indicators for married women in subnational regions in Benin.
-Graphs show mCPR, unmet need for modern methods, demand, and demand satisfied with modern methods.
+Graphs show modern contraceptive use, unmet need for modern methods, demand, and demand satisfied with modern methods.
 Survey data are shown in colored dots with 95% confidence intervals that reflect total error variance.

### 2025-10-15 README.Rmd
fix stan file paths; fix article; add vignette

@@
 ---
+title: "localHierarchy"
 output: github_document
@@
 
-# localHierarchy
+# Background
+
+Bayesian hierarchical models are widely used in global health estimation, where data availability may vary across national or subnational populations.
+Such models are typically fitted to a global database, e.g., to produce national-level estimates for all countries in the world or in some region.
+To facilitate analysis at a local level, models are often desirable that are informed by global models but can be fitted to just a subset of the data, such as data for one country.
+We refer to such models as local models: models that are derived from global hierarchical models but adapted such that they can be fitted to the data from one population alone.
 
 The `localHierarchy` R package provides functionality for fitting Bayesian hierarchical models in settings where both global and local estimation is required.
-The package provides R functions and Stan model components to support global modeling, in which all parameters in hierarchical models are estimated, and local modeling, where parameters are estimated for only one or a small number of populations, using fixed values from a global model fit.
+The package provides R functions and Stan model components to support global modeling, in which all parameters in hierarchical models are estimated, and local modeling, in which parameters are estimated for only one or a small number of populations, using fixed values from a global model fit.
 
-## Installation
+The article on this website presents a practical introduction to the package for the applied user and illustrates the package’s functionality through examples for national and subnational estimation.
 
--   Dependency on `cmdstanr:` Instructions for installing `cmdstanr` are available in their [Getting started](https://mc-stan.org/cmdstanr/articles/cmdstanr.html) guide.
+# Installation
+
+-   `localHierarchy` depends on `cmdstanr:` Instructions for installing `cmdstanr` are available in their [Getting started](https://mc-stan.org/cmdstanr/articles/cmdstanr.html) guide.
 
@@
 
-## Examples
+# Examples
 
-to do: add from paper
+See vignette and article.

### 2025-10-15 R/LocalHierarchy-package.R

@@ -13,6 +13,7 @@
 #' @importFrom purrr map
 #' @importFrom purrr map_int
 #' @importFrom purrr map2_int map_dbl
+#' @importFrom rlang sym
 #' @importFrom stats dnorm quantile rnorm
 #' @importFrom stringr fixed
 #' @importFrom tibble tibble

### 2025-10-15 R/fit_model.R

@@ -222,11 +222,19 @@ fit_model_localhierarchy <- function(
   ##### Set up hierarchical structures ######
   stan_data[["mu_isvector"]] <- mu_isvector
   parname <- "mu"
+
+  ### get the model file
+
+
   if (!mu_isvector){
-    stan_file_path = file.path(here::here("inst/stan/", "hierfunctions_seq_muscalar.stan"))
+    #stan_file_path = file.path(here::here("inst/stan/", "hierfunctions_seq_muscalar.stan"))
+    ### for pkgdown to be able to find the file, need to use the install
+    stan_file_path <- system.file("stan/hierfunctions_seq_muscalar.stan", package = "localhierarchy")
   } else {
     stan_data[["mu_k_terms"]] <- 3 # 2 or higher
-    stan_file_path = file.path(here::here("inst/stan/", "hierfunctions_seq_muvector.stan"))
+    #stan_file_path = file.path(here::here("inst/stan/", "hierfunctions_seq_muvector.stan"))
+    ### for pkgdown to be able to find the file, need to use the install
+    stan_file_path <- system.file("stan/hierfunctions_seq_muvector.stan", package = "localhierarchy")
   }
   stan_data[[paste0(parname, "_scalarprior_sd")]] <- 1
   stan_data[[paste0(parname, "_scalarprior_mean")]] <- 0

### 2025-10-16 README.Rmd
package related fixes

@@
 ---
-title: "localHierarchy"
+title: "localhierarchy"
 output: github_document
@@
 
-# Background
+## Background
 
@@
 
-The `localHierarchy` R package provides functionality for fitting Bayesian hierarchical models in settings where both global and local estimation is required.
+The `localhierarchy` R package provides functionality for fitting Bayesian hierarchical models in settings where both global and local estimation is required.
 The package provides R functions and Stan model components to support global modeling, in which all parameters in hierarchical models are estimated, and local modeling, in which parameters are estimated for only one or a small number of populations, using fixed values from a global model fit.
@@
 
-# Installation
+## Installation
 
--   `localHierarchy` depends on `cmdstanr:` Instructions for installing `cmdstanr` are available in their [Getting started](https://mc-stan.org/cmdstanr/articles/cmdstanr.html) guide.
+-   `localhierarchy` depends on `cmdstanr:` Instructions for installing `cmdstanr` are available in their [Getting started](https://mc-stan.org/cmdstanr/articles/cmdstanr.html) guide.
 
--   Install `localHierarchy` from Github: `remotes::install_github("AlkemaLab/localhierarchy")`
+-   Install `localhierarchy` from Github: `remotes::install_github("AlkemaLab/localhierarchy")`
 
-# Examples
+## Examples
 

### 2025-10-16 R/LocalHierarchy-package.R

@@ -21,5 +21,7 @@
 #' @importFrom tidyr pivot_wider
 #' @importFrom truncnorm rtruncnorm
 #' @importFrom utils data
+#' @importFrom glue glue
+#' @importFrom readr parse_number
 ## usethis namespace: end
 NULL

### 2025-10-16 R/check_nas.R

@@ -8,6 +8,7 @@
 #'
 #' @returns stops if there are NAs in the column or if the column does not exist
 #'
+#' @keywords internal
 check_nas <- function(data, column) {
   if(!(column %in% names(data))) {
     stop(glue::glue("Could not find column {column}."))

### 2025-10-16 R/fit_model.R

@@ -2,9 +2,9 @@
 
 
 
-#' Fit model for localhierarchy model
+#' Fit localhierarchy example model
 #'
-#' This function fits the localhierarchy model
+#' This function fits the localhierarchy example model
 #'
 #' @param survey_df tibble with survey data
 #' @param y column name of outcome, defaults to `y`.

### 2025-10-16 R/get_geo_unit_index_data.R

@@ -1,6 +1,6 @@
 
 
-#' Get geo_unit_index_data
+#' Get unique hierarchical levels from survey data
 #'
 #' Get unique hierarchical levels from survey data and assign index "c" based on area.
 #'

### 2025-10-16 R/get_posterior_summaries.R

@@ -2,7 +2,7 @@
 
 
 
-#' Posterior summaries for use in local models
+#' Get posterior summaries for use in local models
 #'
 #' Get posterior summaries for a set of parameters for use in local model
 #'
@@ -56,6 +56,7 @@ get_posterior_summaries_localhierarchy <- function(
 
 #' @returns tibble with variable (parname with index used in model) and postmean (posterior mean)
 #'
+#' @keywords internal
 get_posterior_summary_one_param <- function(
     fit,
     param_name) {

### 2025-10-16 R/hier_data_helpers.R

@@ -6,6 +6,7 @@
 #'
 #' @returns Boolean
 #'
+#' @keywords internal
 is_there_a_comma <- function(variable){
   grepl(
     pattern = ",",
@@ -43,7 +44,7 @@ is_there_a_comma <- function(variable){
 #' @returns matrix with rows and columns corresponding to the indices in the variable
 #'
 #' @importFrom tidyr pivot_wider pivot_longer
-#'
+#' @keywords internal
 create_a_matrix <- function(post_summ){
   res <- post_summ %>%
     mutate(
@@ -81,6 +82,7 @@ create_a_matrix <- function(post_summ){
 #'
 #' @return TRUE or FALSE indicating whether or not the variable exists in the
 #' fit.
+#' @keywords internal
 variable_exists_in_draws <- function(fit, var_name) {
   all_variables <- posterior::variables(fit$samples$draws())
   all_variables_no_index <- stringr::str_split_i(string = all_variables,
@@ -91,7 +93,8 @@ variable_exists_in_draws <- function(fit, var_name) {
 
 
 
-#' split_var_name_and_index
+#' Split variable name and index
+#'
 #' For a data frame of estimates with a column that contains combined info
 #' about a variable name and row and/or column indices, add columns containing
 #' just the variable name and the indices.
@@ -101,8 +104,7 @@ variable_exists_in_draws <- function(fit, var_name) {
 #'
 #' @returns data frame with the same columns as `estimate_df` plus columns containing
 #' just the variable name and the indices.
-#'
-##'
+#' @keywords internal
 split_var_name_and_index <- function(estimate_df, num_inds) {
   # column names for indices, starting from "i"
   ind_names <- letters[seq(from = 9, length = num_inds)]
@@ -141,7 +143,16 @@ split_var_name_and_index <- function(estimate_df, num_inds) {
 
 #' Starting from an array of parameter values (of dimension 1 or 2),
 #' create a data frame with columns `variable` with the combined parameter name
-#' and indices, and `median` with the parameter values
+#' and indices, and `mean` with the parameter values
+#'
+#' @param param_values Parameter values
+#' @param num_inds Dimension (1 or 2)
+#' @param n Number of indices
+#' @param m For dimension 2, the number of indices in the 2nd dimension
+#' @param param_name Parameter name
+#' @returns Data frame with columns `variable` with the combined parameter name
+#' and indices, and `mean` with the parameter values
+#' @keywords internal
 param_array_to_indexed_df <- function(
     param_values,
     num_inds,

### 2025-10-16 R/hierarchical_data.R

@@ -1,5 +1,5 @@
 
-#' Hierarchical_data
+#' Create hierarchical data structure for a hierarchical model.
 #'
 #' Create hierarchical data structure for a hierarchical model.
 #'

### 2025-10-16 R/hierarchical_param_stan_data.R

@@ -10,7 +10,7 @@
 #' @param hierarchical_terms_fixed character vector specifying hierarchical
 #'   levels for which the terms should be fixed (subset of `hierarchical_levels`).
 #'   Used only if global fit is provided.
-#' @param hierarchical_terms_fixed character vector specifying hierarchical
+#' @param hierarchical_sigmas_fixed character vector specifying hierarchical
 #'   levels for which the terms should be fixed (subset of `hierarchical_levels`, intercept is always included)
 #'   Used only if global fit is provided.
 #'

### 2025-10-16 R/inits.R

@@ -10,6 +10,7 @@
 #'
 #' @importFrom truncnorm rtruncnorm
 #'
+#' @keywords internal
 init_fun_localhierarchy <- function(chain_id, stan_data){
   set.seed(chain_id)
   inits <- list()

### 2025-10-16 R/mu_raw.R

@@ -9,7 +9,7 @@
 #' @param morethan1param does parname refer to more than 1 parameter (a vector)
 #'
 #' @returns tibble with samples of parname_raw_estimate, including info on hierarchical level
-#'
+#' @keywords internal
 get_mu_raw_labeled <- function(fit, parname, morethan1param = FALSE){
   # mu_raw_estimate is saved directly so getting those is not difficult
   # the only challenge is to figure out what hier level they refer to

### 2025-10-16 R/posterior_summary_hierparam.R

@@ -1,7 +1,5 @@
 
 
-#' posterior_summary_hierparam_localhierarchy
-#'
 #' Calculate posterior summaries for hierarchical parameters
 #'
 #' @param fit needs to include parname_star
@@ -79,7 +77,7 @@ posterior_summary_hierparam_localhierarchy <- function(fit, parname, morethan1pa
 #' these mus are obtained by summing up all relevant etas
 #' variables are `name`, `value`, `.draw/iteration/chain'
 #' and `k` referring to parameter index if morethan1param = TRUE
-#'
+#' @keywords internal
 extract_parameter_subhierarchical <- function(
     hierarchical_data,
     subhierarchy,
@@ -139,7 +137,7 @@ extract_parameter_subhierarchical <- function(
 #' @param probs vector of quantiles
 #'
 #' @returns tibble with the requested quantiles
-#'
+#' @keywords internal
 quantile_df <- function(x, probs = c(0.025, 0.5, 0.975)) {
   tibble(
     val = quantile(x, probs, na.rm = TRUE),
@@ -155,7 +153,7 @@ quantile_df <- function(x, probs = c(0.025, 0.5, 0.975)) {
 #'
 #' @returns tibble with 2.5th quantile, posterior mean, and 97.5th quantile, and quant = c(0.025, 0.5, 0.975)
 #'
-#'
+#' @keywords internal
 summary_df <- function(x) {
   tibble(
     val = c(quantile(x, 0.025, na.rm = TRUE),

### 2025-10-16 R/prior_post.R

@@ -1,8 +1,8 @@
 
 
-#' Plot prior and posterior densities of mu_raw
+#' Plot prior and posterior densities of location parameters
 #'
-#' This function plots the posterior densities of mu_raw parameters, with prior added.
+#' This function plots the posterior densities of location (`mu_raw`) parameters, with priors added.
 #'
 #' @param fit List that includes "parname"_raw_estimate and stan_data
 #' @param parname Selected parameter name (example: mu)
@@ -84,7 +84,7 @@ plot_muraw_localhierarchy <- function(fit, parname, morethan1param = FALSE,
 
 #' Plot prior and posteriors of sigmas from hierarchical models
 #'
-#' This function plots the prior and posterior densities of sigma_estimate parameters,
+#' This function plots the prior and posterior densities of sigma_estimate parameters.
 #'
 #' @param fit List, needs to include parname_sigma_estimate and stan_data
 #' @param parname Selected parameter name (example: "mu")

### 2025-10-16 R/simulate.R

@@ -1,6 +1,6 @@
 #' Simulate multilevel data
 #'
-#' Simulates a nested multilevel data structure with etas at each level.
+#' Simulates a nested multilevel data structure.
 #' Lowest level can be used as observational error.
 #'
 #' @param n_levels Number of hierarchical levels (default: 3).

### 2025-10-16 R/check_nas.R
export check_nas; cleaning

@@ -7,8 +7,8 @@
 #' @param column column to check for NAs
 #'
 #' @returns stops if there are NAs in the column or if the column does not exist
+#' @export
 #'
-#' @keywords internal
 check_nas <- function(data, column) {
   if(!(column %in% names(data))) {
     stop(glue::glue("Could not find column {column}."))

### 2025-10-16 R/hierarchical_model_matrix.R

@@ -1,6 +1,6 @@
 
 
-#' hierarchical_model_matrix
+#' Hierarchical model matrix
 #'
 #' Create list with information including a model matrix, associated with the geo_unit_data and hierarchical_levels
 #'

### 2025-11-21 README.Rmd
clean up before release

@@
 
-The article on this website presents a practical introduction to the package for the applied user and illustrates the package’s functionality through examples for national and subnational estimation.
+The article on this website presents a practical introduction to the package for the applied user and illustrates the package’s functionality through examples for national estimation.
 
@@
 
-See vignette and article.
+See article.
+
+## Citation
+
+Please cite as follows:
+
+```{r}
+citation("localhierarchy")
+```

### 2025-11-21 R/fit_model.R

@@ -45,9 +45,10 @@
 #' following run steps are supported:
 #' * "global_national": Fit the global model.
 #' * "local_national": Fit the model to data from a single country, using a global_national fit.
-#' * "global_subnational": Fit the model to global database with subnational data, using a global_national fit.
+#' * "global_subnational": Fit the model to global database with subnational data, using a global_national fit (NISE modeling).
 #' * "local_subnational": Fit the model to subnational data from a single country or region, using a global_subnational fit.
-#' This is also explained in the article with the package.
+#'
+#' The options are also explained in the article with the package.
 #'
 #' Details on hierarchical set ups used
 #' The package allows the structure of the hierarchical prior to be configured by the user

### 2025-11-25 README.Rmd
update readme
fix URLs

@@
 
--   `localhierarchy` depends on `cmdstanr:` Instructions for installing `cmdstanr` are available in their [Getting started](https://mc-stan.org/cmdstanr/articles/cmdstanr.html) guide.
+-   `localhierarchy` depends on `cmdstanr`: Instructions for installing `cmdstanr` are available in their [Getting started](https://mc-stan.org/cmdstanr/articles/cmdstanr.html) guide.
 
@@
 
-See article.
+See [article](https://alkemalab.github.io/localhierarchy/articles/localhierarchy_article.html) on package website, at [https://alkemalab.github.io/localhierarchy](https://alkemalab.github.io/localhierarchy/index.html){.uri}.
 
@@
 ```
+
+## Funding
+This work was supported, in whole or in part, by the Bill & Melinda Gates Foundation (INV-00844).
